### 2018-09-06 tex/article.tex
add paper draft and docs

\documentclass[twocolumn]{svjour3}
\usepackage{style}
\usepackage{subfiles}
\usepackage{microtype}

% Title
\title{Multiset-trie data structure}
\author{Mikita Akulich \and Iztok Savnik \and Matja\v z Krnc \and Riste \v Skrekovski}
\institute{
		Mikita Akulich 
		\at 
		University of Primorska, FAMNIT, Koper, Slovenia\\
		\email{mikita.akulich@gmail.com}
		\and
		Iztok Savnik
		\at 
		University of Primorska, FAMNIT, Koper, Slovenia\\
		\email{iztok.savnik@famnit.upr.si}
		\and 
		Matja\v z Krnc 
		\at 
		University of Primorska, FAMNIT, Koper, Slovenia\\
		\email{matjaz.krnc@famnit.upr.si}
		\and 
		Riste \v Skrekovski
		\at
		University of Ljubljana, FMF, Ljubljana, Slovenia\\
		\email{riste.skrekovski@famnit.upr.si}
		}

\begin{document}
\maketitle

%Abstract
\begin{abstract}
[Will be rewritten]
In this paper we will present the multiset-trie, a new data structure that operates 
on objects represented as multisets. The multiset-trie is a search-tree-based 
data structure with the properties similar to those of a trie. It implements all standard search 
tree operations together with the special multiset containment operations. Multiset 
containment operations supported by the multiset-trie are submultiset and supermultiset. 
These operations are used for implementation of different queries that can be performed 
on multisets in a multiset-trie. One of the most important queries is the search of 
the nearest neighbor given an input object. The nearest neighbor search of a 
multiset-trie makes it a good alternative for the index data structures that are 
used in information retrieval systems. In particular, our research is focused on the 
application of the multiset-trie to full-text search systems. 
\end{abstract}
\keywords{multiset, bag-of-words, trie, multiset-trie, information retrieval, inverted index, full-text search}
%\subclass{68P05, 68P20, 68Q87}

% section 1
\input{tex/s_intro}
% section 2
\input{tex/s_data_structure}
% section 3
\input{tex/s_operations}
% section 4
\input{tex/s_math}
% section 5
\input{tex/s_experiments}
% section 6
\input{tex/s_relwork}
% section 7
\input{tex/s_conclusions}

\bibliographystyle{abbrv}
\bibliography{multiset-trie}

\end{document}

### 2018-09-06 tex/tex/head.tex
add paper draft and docs

\pagenumbering{Roman}
\pagestyle{empty}
\begin{center}
\noindent \large UNIVERZA NA PRIMORSKEM\\
\large FAKULTETA ZA MATEMATIKO, NARAVOSLOVJE IN\\
INFORMACIJSKE TEHNOLOGIJE

\normalsize
\vspace{5.5cm}
Zaklju\v cna naloga\\
(Final project paper)\\
\textbf{\large Predstavitev objektov z multimno\v zicami v sistemih za informacijsko povpra\v sevanje}\\
\normalsize
(Multiset representation of objects in information retrieval systems)\\
\end{center}

\begin{flushleft}
\vspace{5cm}
\noindent Ime in priimek: Mikita Akulich
% add your first name and last name in the line above
\\
\noindent \v Studijski program: Matematika
% add your study program in the line above
\\
\noindent Mentor: doc. dr. Iztok Savnik
% add the academic title, first name, and last name of your mentor in the line above
\\
\noindent Somentor: doc. dr. Matja\v z Krnc
% if you have a co-mentor, add his/her academic title, first name, and last name in the line above
% if you do not have a co-mentor, delete the line above and the line below
\\
\end{flushleft}

\vspace{4cm}
\begin{center}
\large \textbf{Koper, september 2017}
% add the month and year of submission of your final project paper
\end{center}
\newpage

\pagestyle{fancy}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Key words documentation (Slovene and English) %%%%%%%%%%%

\section*{Klju\v cna dokumentacijska informacija}

\medskip
\begin{center}
\fbox{\parbox{\linewidth}{
\vspace{0.2cm}
\noindent
Ime in PRIIMEK: Mikita AKULICH \vspace{0.5cm}\\
Naslov zaklju\v cne naloge: Predstavitev objektov z multimno\v zicami v sistemih za informacijsko povpra\v sevanje \vspace{0.5cm}\\
Kraj: Koper \vspace{0.5cm}\\
Leto: 2017 \vspace{0.5cm}\\
\v Stevilo listov: 40\hspace{2cm} \v Stevilo slik: 15\hspace{2.6cm} \\
\v Stevilo referenc: 14 \vspace{0.5cm}\\
Mentor: doc. dr. Iztok Savnik \vspace{0.5cm}\\
Somentor: doc. dr. Matja\v z Krnc \vspace{0.5cm}\\
Klju\v cne besede: multimno\v zica, vre\v ca-besed, trie, multiset-trie, informacijsko povpra\v sevanje, obrnjen indeks, full-text search \vspace{0.5cm} \\
Math.~Subj.~Class.~(2010): 68P05, 68P20, 68Q87 \vspace{0.5cm}\\
{\bf Izvle\v cek:}
\vspace{0.2cm} \\
V zaklju\v cni nalogi je predstavljena multiset-trie, nova podatkovna struktura, ki deluje na objektih predstavljenih 
z multimno\v zicami. 
Multiset-trie je podatkovna struktura, ki temelji na iskalnem drevesu in ima podobne lastnosti kot trie. 
Vklju\v cuje vse standardne operacije iskalnega drevesa skupaj z operacijami vsebovanja nad multimno\v zicami. 
Operacije vsebovanja, ki so definirane z multiset-trie, so izpeljane iz operacij pod-multimno\v zica in nad-multimno\v zica. 
Te operacije se uporabljajo za izvajanje razli\v cnih poizvedb, ki delujejo na multimno\v zicah v multiset-trie. 
Ena izmed najpomembnej\v sih poizvedb je iskanje najbli\v zjega soseda glede na vhodno multimno\v zico. 
Iskanje najbli\v zjega soseda v multiset-trie je dobra alternativa operacijam nad indeksnimi strukturami, ki se uporabljajo v sistemih 
za informacijsko povpra\v sevanje.
Na\v sa raziskava je osredoto\v cena na uporabo multiset-trie v sistemih za iskanje po celotnem besedilu.
}}
\end{center}

\newpage

\section*{Key words documentation}

\medskip

\begin{center}
\fbox{\parbox{\linewidth}{
\vspace{0.2cm}
\noindent
Name and SURNAME: Mikita AKULICH \vspace{0.5cm}\\
Title of final project paper: Multiset representation of objects in information retrieval systems \vspace{0.5cm}\\
Place: Koper \vspace{0.5cm}\\
Year: 2017 \vspace{0.5cm}\\
Number of pages: 40\hspace{1.6cm} Number of figures: 15\\
Number of references: 14 \vspace{0.5cm}\\
Mentor: Assist.~Prof.~Iztok~Savnik, PhD\vspace{0.5cm}\\
Co-Mentor: Assist.~Prof.~Matja\v z~Krnc, PhD\vspace{0.5cm}\\
Keywords: multiset, bag-of-words, trie, multiset-trie, information retrieval, inverted index, full-text search \vspace{0.5cm}\\
Math.~Subj.~Class.~(2010): 68P05, 68P20, 68Q87 \vspace{0.5cm}\\
{\bf Abstract:}
\vspace{0.2cm} \\
In this thesis we will present the multiset-trie, a new data structure that operates 
on objects represented as multisets. The multiset-trie is a search-tree-based 
data structure with the properties similar to those of a trie. It implements all standard search 
tree operations together with the special multiset containment operations. Multiset 
containment operations supported by the multiset-trie are submultiset and supermultiset. 
These operations are used for implementation of different queries that can be performed 
on multisets in a multiset-trie. One of the most important queries is the search of 
the nearest neighbor given an input object. The nearest neighbor search of a 
multiset-trie makes it a good alternative for the index data structures that are 
used in information retrieval systems. In particular, our research is focused on the 
application of the multiset-trie to full-text search systems. 
}}
\end{center}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Acknowledgement %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\section*{Acknowledgement}

I would like to thank my mentor assist. prof. Iztok Savnik and co-mentor 
assist. prof. Matja\v z Krnc for introducing me to this topic and their guidance 
for completing my final thesis. I would also like to thank assoc. prof. Riste 
\v Skrekovski for organizing a lecture at Faculty of Mathematics and Physics at 
University of Ljubljana where I had an opportunity to present our project and 
receive a practical feedback. Finally, I would like to thank Faculty of Mathematics, 
Natural Sciences and Information Technologies for given support trough supporting me 
with a scholarship during my stay, as well as the financial help for attending competitions 
and conferences.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Table of contents, list of figures, etc. %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage

\tableofcontents
\addtocontents{toc}{\protect\thispagestyle{fancy}}
% if there are no tables in your final project paper, delete the following three lines
%\newpage
%\listoftables
%\addtocontents{lot}{\protect\thispagestyle{fancy}}
% if there are no figures in your final project paper, delete the following three lines
\newpage
\listoffigures
\addtocontents{lof}{\protect\thispagestyle{fancy}}
\newpage
% Since the appendices are not numbered, we also do not want to show the dots to their (non-existing) page numbers.
%\renewcommand{\cftdot}{}
%\listofappendices
%\thispagestyle{fancy}
%\newpage

\normalsize

### 2018-09-06 tex/tex/s_conclusions.tex
add paper draft and docs

\section{Conclusions and future work} \label{c:conclusions}
[Will be rewritten]
%
% Experiments conclusions
%
One of the conclusions of studying the multiset-trie both theoretically and 
empirically is that our data structure is input sensitive. Input sensitivity implies 
a non consistent performance on different input data. However, our argument 
that the performance can be optimized by pre-processing the input data 
is confirmed in the Experiment~\hyperref[ss:exp3]{4}. 
Pre-processing determines the optimal encoding for input data and ensures the 
best performance of the multiset-trie on particular input data. In case of storing 
words in the multiset-trie, the search queries can be always optimized based on 
the frequencies of letters in a specific language. We also see from 
Experiments~\hyperref[s:exp1]{1} and~\hyperref[s:exp2]{2} that dependence of 
the multiset-trie performance on the density is not a linear function. Yet the function 
is continuous and the point of inflection is unique on the whole domain as it is shown 
in Experiment~\hyperref[s:exp3]{3}. This allows us to predict whether multiset-trie 
can be used for some particular application, serving a high performance. 
%

% Mathematical analysis conclusions
%
The mathematical analysis of the space complexity shows that multiset-trie requires 
only $O(|M|)$ space, which is the minimal possible space that is required by any 
data structure for storage of $|M|$ objects. As for the running time complexity of the 
algorithms the basic tree functions such as \textsc{insert}, \textsc{search} and 
\textsc{delete} all have a constant complexity once the multiset-trie is defined. 
The "getAll" multiset containment functions have worst case running time complexity of 
$O(|\mathcal{M}|),$ where $|\mathcal{M}|$ is the cardinality of the multiset-trie data structure. 
The "existence" multiset containment functions have the worst case running time 
complexity of $O(|\mathcal{M}| - |M|),$ where $|\mathcal{M}|$ is the cardinality of 
the multiset-trie and $|M|$ is the number of inserted multisets (nodes on leaf level).

It can also be concluded that the multiset-trie is an input sensitive data structure, 
because the size of multiset-trie $|\mathcal{M}|$ depends on the distribution of 
multisets in $M.$ Our mathematical model assumes that multisets $m$ in $M$ 
are distributed uniformly. However, in a real world models such an assumption 
is not true in a lot of the cases. Specifically, the probability $P(m\in M)$ may vary 
dramatically and can be even equal to $0.$ For example, if words are mapped to 
multisets, then the sample space contains very large multisets. Nonetheless, most of 
them will have zero probability to appear in $M$, because a word that would 
correspond to such a multiset simply does not exist.
%

% Future work notes
%
The above results have opened even more interesting questions for the future research. 
Further steps in our research will be to extend the functionality of the multiset-trie. 
We are interested in more flexible multiset containment queries, where the types 
of sub and supermultisets can be specified. As an example, the multiplicity of an 
element in a multiset can be bounded in operations getAllSubmsets and 
getAllSupermsets. Such functionality would allow more specific queries of multisets.
%
The second line of research is to investigate the multiset-trie as an index data 
structure in detail. It will be very interesting to study the comparison of the 
multiset-trie with other existing index data structures.

### 2018-09-06 tex/tex/s_data_structure.tex
add paper draft and docs

\section{Multiset-trie data structure} \label{c:description}
%
Let $\Sigma$ be a set of distinct symbols that define an alphabet and let 
$\sigma$ be the cardinality of $\Sigma.$ The \emph{multiset-trie} data structure 
stores multisets that are composed of symbols from the alphabet $\Sigma.$ It 
provides the basic tree data structure operations such as insert, delete and 
search together with multiset containment and membership operations such as 
submultiset and supermultiset that will be discussed in the next section 
in greater details.

Multiset ignores the ordering of its elements by definition, which allows us to 
define a bijective mapping $\phi:\Sigma \rightarrow I,$ where $I$ is the set of 
integers $\{ 1,2,3, \ldots, \sigma \}.$ In this way, we obtain an indexing of 
elements from the alphabet $\Sigma,$ so we can work directly with integers 
rather then with specific symbols from $\Sigma.$

The multiset-trie is an $n$-ary tree based data structure with the properties of 
trie. A node in multiset-trie always has degree $n,$ i.e. $n$ children. Some of 
the children may be \emph{Null} (non-existing), but the number of \emph{Null} 
children can be at most $n-1.$ All the children of a node, including the \emph{Null} 
children, are labeled from left to right with labels $c_j,$ where $j\in \{ 0, 1, \ldots, n-1 \}.$ 
Every two child nodes $u$ and $v$ that share the same parent node have different labels.

%The multiset-trie is an $n$-ary tree based structure. A node can have a degree 
%from $1$ to $n,$ depending on the number of existing children. The edges from 
%parent node to children have labels $c_j,$ where $j\in \{ 0, 1, \ldots, n-1 \}.$ 
%The edges are labeled from left to right with the condition that there are no 
%two edges with the same label coming from a parent node.

Nodes that have equal height in a multiset-trie form a level. The height of a 
multiset-trie is always $\sigma+1$ if at least one multiset is in structure. 
The height of the root node (the first level) is defined to be 1.
%
Levels in multiset-trie are enumerated by their height, i.e. a level $L_i$ has 
height $i.$ The connection between level height in a multiset-trie and symbols 
from alphabet $\Sigma$ is defined as follows. A level $L_i,$ where 
$i\in\{ 1,2,\ldots, \sigma \}$ represents a symbol $s\in\Sigma,$ such that 
$\phi^{-1}(i) = s.$ The last level $L_{\sigma+1}$ does not represent any symbol 
and is named \emph{leaf level} ($LL$ for short).

Since every level, except $LL$ represents a symbol from $\Sigma,$ we can define 
a transition between nodes that are located at different levels in a multiset-trie. 
%
Consider two nodes $u,v$ in a multiset-trie at levels $L_i, L_{i+1}$ respectively, 
where $i\in\{1,2,\ldots,\sigma\}.$ Let a node $u$ be a parent node of a node $v$ 
and consequently a node $v$ be a child node of a node $u.$ Suppose that a child 
node $v$ is not \emph{Null} and has a label $c_j,$ where $j\in\{ 0,1,\ldots, n-1 \}.$ 
%
Then the \emph{path} $u\rightarrow v$ represents a symbol $s\in\Sigma$ with 
multiplicity $j,$ such that $\phi^{-1}(i) = s.$ 
%
Such a transition $u\rightarrow v$ is called a \emph{path of length} $1$ and is 
allowed if and only if a node $v$ is not \emph{Null} and $u$ is a parent node of 
a node $v.$ If a node $v$ has label $c_0,$ then the path $u\rightarrow v$ 
represents a symbol with the multiplicity $0$ respectively i.e. an empty symbol.

We define a \emph{complete path} to be the path of length $\sigma$ in a 
multiset-trie with the end points at root node (the 1st level) and $LL$. Thus, 
a multiset $m$ is inserted into a multiset-trie if and only if there exists a 
complete path in a multiset-trie that corresponds to $m.$
%
Note that every complete path in a multiset-trie is unique. Therefore, the multisets 
that share a common prefix in a multiset-trie can have a common path of length at 
most $\sigma-1.$ The complete path that passes through nodes labeled by $c_0$ 
on all levels represents an empty multiset or an empty set.
%
Thus, any multiset $m$ that is composed of symbols from $\Sigma$ with maximum 
multiplicity not greater than $n-1$ can be represented by a complete path in a multiset-trie.

Let us have an example of a multiset-trie data structure. Let $\sigma = 2$ and 
$\Sigma = I = \{ 1,2 \}$ respectively, so the mapping $\phi$ is an identity mapping. 
Fix the degree of a node $n=3,$ so the maximal multiplicity of an element in 
a multiset is $n-1=2.$ The figure~\ref{fig:sketch} presents the multiset-trie that 
contains multisets $\emptyset, \{ 1,1,2 \},$ $\{ 1,2,2 \},$ $\{ 2 \},$ $\{ 1,2 \},$ $\{ 2,2 \}.$ 
The \emph{Null} children are omitted on the figure.

\begin{figure}[h!]
\centering
\begin{tikzpicture}[>=stealth',
level 1/.style={sibling distance = 2.5cm},
level 2/.style={sibling distance = 1cm},
level distance = 1.5cm]
\node (R) [node1] {\tiny Root}
	child { node [node1] {} 
		child { node [node1] {} edge from parent node [near end, left] {$c_0$}}
		child { node [node1] {} edge from parent node [near end, left] {$c_1$}}
		child { node [node1] {} edge from parent node [near end, right] {$c_2$}}
		edge from parent node [near end, left] {$c_0$}	
	}
	child { node [node1] {}
		child { node [node1] {} edge from parent node [near end, left] {$c_1$}}
		child { node [node1] {} edge from parent node [near end, right] {$c_2$}}
		edge from parent node [near end, left] {$c_1$}
	}
	child { node [node1] {} 
		child { node [node1] {} edge from parent node [near end, left] {$c_1$}}
		edge from parent node [near end, right] {$c_2$} 
};	
%----------------- Level labels --------------	
	\begin{scope}[every node/.style={right}]
		\path (R 				-| R-3-1) ++(5mm,0) node {$L_1$};
		\path (R-3 			-| R-3-1) ++(5mm,0) node {$L_2$};
		\path (R-3-1		-| R-3-1) ++(5mm,0) node {$LL$};
	\end{scope}			
\end{tikzpicture}
\caption{Example of multiset-trie structure.}
\label{fig:sketch}
\end{figure}

Let a pair $(L_i, c_j)$ represents a node with label $c_j$ at a level $L_i.$ 
The pair $(L_1,c_j)$ is equivalent to $(L_1,root),$ since the first level has 
the root node only. According to the figure~\ref{fig:sketch} we can extract 
inserted multisets as follows:

\begin{eqnarray*}
(L_1,root) \rightarrow (L_2,c_0) \rightarrow (LL,c_0) & \textrm{equivalent to} & \{ 1^0, 2^0 \} = \emptyset \\
(L_1,root) \rightarrow (L_2,c_0) \rightarrow (LL,c_1) & \textrm{equivalent to} & \{ 1^0, 2^1 \} = \{ 2 \} \\
(L_1,root) \rightarrow (L_2,c_0) \rightarrow (LL,c_2) & \textrm{equivalent to} & \{ 1^0, 2^2 \} = \{ 2,2 \} \\
(L_1,root) \rightarrow (L_2,c_1) \rightarrow (LL,c_1) & \textrm{equivalent to} & \{ 1^1, 2^1 \} = \{ 1,2 \} \\
(L_1,root) \rightarrow (L_2,c_1) \rightarrow (LL,c_2) & \textrm{equivalent to} & \{ 1^1, 2^2 \} = \{ 1,2,2 \} \\
(L_1,root) \rightarrow (L_2,c_2) \rightarrow (LL,c_1) & \textrm{equivalent to} & \{ 1^2, 2^1 \} = \{ 1,1,2 \} 
\end{eqnarray*}
where $e^k$ represents an element $e$ with multiplicity $k.$

### 2018-09-06 tex/tex/s_experiments.tex
add paper draft and docs

\section{Experiments} \label{c:experiments}
%
This section contains results of experiments that were performed on the multiset-trie 
data structure. The implementation of multiset-trie is done in the \CC 
{ programming} language. The current implementation uses only the standard 
library of \CC14 version of the standard and has a command line interface. The implementation 
of the program was optimized for testing and therefore, the program operates 
with files, in order to process queries. After processing all the queries 
the results are stored in files for further analysis.

In our experiments we will test the functions: \textsc{submsetExistence}, 
\textsc{supermsetExistence}, \textsc{getAllSubmsets} and 
\textsc{getAllSupermsets}. Performance of the functions will be measured by 
the number of visited nodes in multiset-trie by the particular function. In 
particular the performance is inversely proportional to the number of visited 
nodes.

Before we start, we will give a few definitions about the parameters 
that will be varied throughout the experiments and discuss the experimental data 
that was used.

Let $M$ be a set of multisets that are inserted to multiset-trie and let $n$ be 
the maximal node degree. Let $N$ be the power multiset of $\Sigma,$ where 
the multiplicity of each element is bounded from above by $n-1.$ We define the 
\emph{density} of a multiset-trie to be the ratio $\frac{|M|}{|N|},$ where 
$|\cdot|$ denotes cardinality.

The selected parameters of the data structure that will be varied in experiments 
are as follows:
%
\begin{itemize}
\item $\sigma$ - the cardinality of the alphabet $\Sigma;$
%
\item $n$ - the maximal degree of a node, which explicitly defines the maximal 
multiplicity of elements in a multiset;
%
\item $\phi$ - mapping of letters from $\Sigma$ into a set of consecutive 
integers;
%
\item $d$ - density of a multiset-trie.
%
\end{itemize}
The cardinality of a power multiset $N$ is equal to $n^\sigma,$ which means that 
density $d$ of a multiset-trie depends on parameters $|M|,$ $\sigma$ and $n.$ 
Because parameters $\sigma$ and $n$ are set when a multiset-trie is initialized, 
the parameter $|M|$ will be varied to change the density in experiments. As we 
mentioned in Chapter~\ref{c:description}, the mapping $\phi$ determines the 
correspondence of letters to levels in multiset-trie, i.e. it defines the ordering of 
levels in multiset-trie. It is also true, that $\phi$ defines the ordering in multisets.

% anouncement of the experiments
In the next sections we will present the behavior of the multiset-trie data 
structure depending on the selected parameters as well as the benchmark 
of the multiset-trie against trie and hash map data structures. 
% outline of the experiment section
We start with experiments that are performed on an artificially generated data in 
order to give a general picture of the multiset-trie performance. In the 
Experiment~\hyperref[s:exp1]{1} a special case of the multiset-trie is considered. 
Only sets are allowed to be stored in the data structure, i.e. the maximal allowed 
multiplicity is set to 1. The performance is measured with respect to the density 
of the multiset-trie.
%
The Experiment~\hyperref[s:exp2]{2} is an extension of the previous one. Here, 
we also measure the performance of the multiset-trie depending on its density. 
The difference is that the allowed multiplicity of an element is raised, i.e. 
multisets are allowed to be stored in the data structure. 
%
Summarizing the tests of performance depending on the density we present the 
Experiment~\hyperref[s:exp3]{3}. It shows a non linearity of the performance 
with respect to the density of the multiset-trie.
%
The next experiment on the multiset-trie uses the real world data. In 
Experiment~\hyperref[s:exp4]{4} the influence of the mapping $\phi$ is studied. 
The input data is obtained by mapping of the real words from English dictionary 
to the set of consecutive integers using the function $\phi.$ The experiment 
shows that the performance of the multiset-trie can be noticeably optimized using 
different mappings $\phi.$ It also shows the usability of the multiset-trie in terms 
of real data demonstrating the high performance of search queries.
%
After all the experiments we present an empirical comparison of multiset-trie 
data structure with inverted index. Inverted index is considered in two versions, 
when it uses trie and hash map for storing multisets. In the comparison we use 
two types of queries exact retrieval and searching for the first submultiset.

\paragraph*{Data generation}
We denote by \emph{input data} the data that is inserted into multiset-trie and 
by \emph{test data} the data which will be used for queries in order to test the 
performance of the functions.

The artificially generated input data is obtained by sampling $|M|$ multisets 
from $N.$ All the multisets in $N$ are constructed according to parameters 
$\sigma$ and $n$ and represent the power multiset of the alphabet $\Sigma.$ 
Every multiset in $M$ is chosen from $N$ with equal probability $p.$ Thus, the 
probability $p$ gives a collection $M$ of multisets that are sampled from $N$ 
with uniform distribution. Uniform distribution is chosen in order to simulate a 
random user input.

% test data explanation
The test data is generated artificially and constructed as follows. Given the 
parameters $\sigma$ and $n,$ the possible size of a multiset varies from~1 
to~$\sigma n.$ The number of randomly generated test multisets for every 
value of multiset size is 1500. In other words, we perform 1500 experiments 
in order to measure the number of visited nodes for the queries with test multiset 
of a distinct size. The final value of visited nodes is calculated by taking an 
arithmetic mean among all 1500 measurements.

\subsection{Experiment 1} \label{s:exp1}
This experiment shows the performance of multiset-trie being used for storing 
and retrieving \emph{sets} instead of \emph{multisets}. We restrict multiset-trie in order 
to make a closer comparison with the \emph{set-trie} data structure~\cite{savnik2013index}.
In this case we set the maximal node degree $n$ to be $2$ and $\sigma$ to be 25. 
The mapping $\phi$ does not have an influence in this particular experiment, 
because the input data is generated artificially with uniform distribution. On 
average the results will be the same for any $\phi,$ since all the multisets are 
equally likely to appear in $M.$ The parameter $|M|$ varies from 10000 sets up 
to 320000 sets. According to the parameters $n$ and $\sigma,$ the cardinality of 
$N$ is $33554432\approx \num{3.36e+7}.$ Thus, the calculated density of the 
multiset-trie with respect to $|M|$ varies from~$0.03\%$ to~$0.95\%.$

\begin{figure}
\center
\includegraphics[width=.4\textwidth]{exp1-m1.pdf}
\caption{Experiment 1, submsetExistence function.}
\label{fig:e1m1}
\end{figure}

\begin{figure}
\center
\includegraphics[width=.4\textwidth]{exp1-m2.pdf}
\caption{Experiment 1, supermsetExistence function.}
\label{fig:e1m2}
\end{figure}

\begin{figure}
\center
\includegraphics[width=.4\textwidth]{exp1-m3.pdf}
\caption{Experiment 1, getAllSubmsets function.}
\label{fig:e1m3}
\end{figure}

\begin{figure}
\center
\includegraphics[width=.4\textwidth]{exp1-m4.pdf}
\caption{Experiment 1, getAllSupermsets function.}
\label{fig:e1m4}
\end{figure}

The performance of the functions \textsc{submsetExistence} and 
\textsc{supermsetExistence} increases as the density increases (see figures~\ref{fig:e1m1}
and~\ref{fig:e1m2}). The results are as expected, because the increase of the 
density increases the probability of finding submultiset or supermultiset in 
multiset-trie, which leads to the lower number of visited nodes.

The maxima are located between 175 and 375 for \textsc{submsetExistence} and 
between 175 and 350 for \textsc{supermsetExistence}. According to those maxima 
we can deduce that at least 7-15 multisets were checked in order to find 
submultiset or supermultiset, which is from $0.002\%$ to $0.15\%$ of the 
multiset-trie and from $\num{1.9e-5}\%$ to $\num{4.5e-5}\%$ of the complete 
multiset-trie.

As the density increases the peaks shift from the center to the left, or to the right, 
for \textsc{submsetExistence} and \textsc{supermsetExistence} respectively. 
The shifts are the consequence of the uniform distribution of sets in $M.$ 
Since every set has the same probability to appear in $M,$ the distribution of set 
sizes in $M$ is normal. Consequently, with increase of the density of the 
multiset-trie the number of sets in $M$ with cardinality $\frac{1}{2}\sigma$ will be 
larger than the number of sets with cardinality $\frac{1}{2}\sigma\pm\epsilon,$ 
for $\frac{1}{2}\sigma > \epsilon > 0.$ So the function \textsc{submsetExistence} 
needs to visit less nodes for test sets of size $\frac{1}{2}\sigma$ than for test 
sets of size $\frac{1}{2}\sigma\pm\epsilon.$ The function decreases the 
multiplicity of some elements (in some cases skips them) in order to find the 
closest subset. Hence, the peak shifts to the left. Oppositely the function 
\textsc{supermsetExistence} increases the multiplicity of some elements 
(in this case adding new elements) in order to find the closest superset. 
Thus, the peak shifts to the right.

Note that despite the peak shifts both functions \textsc{submsetExistence} and 
\textsc{supermsetExistence} have approximately the same worst case performance.

The performance of the functions \textsc{getAllSubmsets} and \textsc{getAllSupermsets} 
decreases as the density increases (see figures~\ref{fig:e1m3} and~\ref{fig:e1m4}). 
This happens because the number of multisets in multiset-trie increases, which means 
that any multiset in the data structure will have more sub- and supermultisets. 
The maxima for both functions varies from $\num{8.0e4}$ to $\num{1.5e6}$ visited nodes. 
We can notice that local maxima for the functions \textsc{getAllSubmsets} and 
\textsc{getAllSupermsets} differs with respect to the length of input. The 
explanation is very simple. In order to find all submultisets of a small set the 
function has to traverse a small part of multist-trie. As the size of a set 
increases the part of a multiset-trie where all the submultisets of a given set 
are stored also increases. The opposite holds for the function 
\textsc{getAllSupermsets}.

Despite the fact that for a lookup of any set/multiset $\sigma$ nodes must be visited 
in multiset-trie on average case, the data structure has a very similar performance 
results in comparison to the \emph{set-trie} data structure.

\subsection{Experiment 2} \label{s:exp2}
In the Experiment~\hyperref[s:exp2]{2} we demonstrate the performance of 
the unrestricted multiset-trie allowing \emph{multisets} to be inserted into data structure. 
We set $n$ to be 6 and retain $\sigma = 25$ as it was in Experiment~\hyperref[s:exp1]{1}. 
The mapping $\phi$ does not have an influence on results, since the input 
data is generated artificially with uniform distribution. The cardinality of $M$ 
varies from 40000 to 640000 multisets. Thus, the calculated density $d$ varies 
from $\num{1.4e-13}\%$ to $\num{2.25e-12}\%.$ The density is much smaller than 
in Experiment~\hyperref[s:exp1]{1}, because now we allow multisets to be stored 
in the data structure and according to the parameters $n$ and $\sigma$ the 
cardinality of $N$ is $6^{25} = \num{2.84e19}.$

\begin{figure}
\center
\includegraphics[width=.4\textwidth]{exp2-m1.pdf}
\caption{Experiment 2, submsetExistence function.}
\label{fig:e2m1}
\end{figure}

\begin{figure}
\center
\includegraphics[width=.4\textwidth]{exp2-m2.pdf}
\caption{Experiment 2, supermsetExistence function.}
\label{fig:e2m2}
\end{figure}

\begin{figure}
\center
\includegraphics[width=.4\textwidth]{exp2-m3.pdf}
\caption{Experiment 2, getAllSubmsets function.}
\label{fig:e2m3}
\end{figure}

\begin{figure}
\center
\includegraphics[width=.4\textwidth]{exp2-m4.pdf}
\caption{Experiment 2, getAllSupermsets function.}
\label{fig:e2m4}
\end{figure}

As we can see from the graphs on figures~\ref{fig:e2m1} and~\ref{fig:e2m2}, 
the performance of the functions \textsc{submsetExistence} and 
\textsc{supermsetExistence} becomes worse as the density increases. 
In this case the number $|M|$ is slightly larger than in the 
Experiment~\hyperref[s:exp1]{1}, but the density is very small. Consequently 
multiset-trie become more sparse. Multisets in a sparse multiset-trie differs more, 
which leads to the larger number of visited nodes.

The maxima for both functions varies from 7500 to 25000 visited nodes. According 
to those maxima at least 300-1000 multisets were checked in order to find 
submultiset or supermultiset, which is from $0.15\%$ to $0.75\%$ of the entire 
multiset-trie and from $\num{1.1e-15}\%$ to $\num{3.4e-15}\%$ of the complete 
multiset-trie. The percentage of visited multisets with respect to $|M|$ is 
larger than in the Experiment~\hyperref[s:exp1]{1}. However, if one would compare 
the percentage of visited multiset with respect to complete multiset-trie, then 
in case of Experiment~\hyperref[s:exp2]{2} it is less by 10 orders than in the 
Experiment~\hyperref[s:exp1]{1}.

The peaks are shifted from the center to the left and right for 
\textsc{submsetExistence} and \textsc{supermsetExistence} respectively. Such a 
behavior was previously observed in the Experiment~\hyperref[s:exp1]{1}. The 
explanation is the same: the input data has uniform distribution, implying that 
the size of multisets in $M$ is normally distributed. Because of the normal 
distribution of size of multisets the shift of the peak occurs as the density increases.

It can be also observed that as in previous Experiment~\hyperref[s:exp1]{1} both 
functions \textsc{submsetExistence} and \textsc{supermsetExistence} have similar 
worst case performance.

The functions \textsc{getAllSubmsets} and \textsc{getAllSupermsets} decrease 
their performance as the density increases (see figures~\ref{fig:e2m3} 
and~\ref{fig:e2m4}). It happens, because the number of multisets increases as 
the density increases. So there are more nodes have to be visited in order to 
retrieve all sub- or supermultisets of some multiset. The maximum for both functions 
varies from $\num{0.9e5}$ to $\num{1.5e7}$ visited nodes. As it was observed in 
Experiment~\hyperref[s:exp1]{1} the maxima occur at the opposite points. For the 
function \textsc{getAllSubmsets} it will always be at the largest size of multiset, 
which is 125 in our case. Conversely the maximum for the \textsc{getAllSupermsets} 
is at the smallest size of multiset, which is 0 (an empty set).

The results of the Experiment~\hyperref[s:exp1]{1} show that the performance 
of functions \textsc{submsetExistence} and \textsc{supermsetExistence} increases 
as the density increases. However, we observe the opposite behavior in the 
Experiment~\hyperref[s:exp2]{2}. We explain the reason of such a contradiction 
in the next Experiment~\hyperref[s:exp3]{3}

%\subsection{Experiment 3}
%In this experiment we vary $\sigma$ from 10 up to 70. The maximal degree node 
%$n$ is set to 6 and the input data $|M|$ is set to 640000 multisets. This way 
%we extremly decrease density $d$ as we increase $\sigma.$ 
%The density varies from $1.0\%$ to $\num{2.1e-47}\%.$
%
%\begin{figure}
%\center
%\includegraphics[width=.9\textwidth]{exp3-m1.pdf}
%\caption{Experiment 3, submsetExistence function.}
%\end{figure}
%
%\begin{figure}
%\center
%\includegraphics[width=.9\textwidth]{exp3-m2.pdf}
%\caption{Experiment 3, supermsetExistence function.}
%\end{figure}
%
%\begin{figure}
%\center
%\includegraphics[width=.9\textwidth]{exp3-m3.pdf}
%\caption{Experiment 3, getAllSubmsets function.}
%\end{figure}
%
%\begin{figure}
%\center
%\includegraphics[width=.9\textwidth]{exp3-m4.pdf}
%\caption{Experiment 3, getAllSupermsets function.}
%\end{figure}

\subsection{Experiment 3} \label{s:exp3}
The results of the Experiment~\hyperref[s:exp1]{1} and Experiment~\hyperref[s:exp2]{2} 
have shown that as the density of a multiset-trie increases the performance of 
functions \textsc{submsetExistence} and \textsc{supermsetExistence} can both get 
better and worse. The reason of such a behavior is that the dependence of the 
number of visited nodes on density is not a linear function. It is obvious that the 
performance of the mentioned above functions is maximal when multiset-trie is 
complete. As multiset-trie becomes more sparse (the density is small) multisets 
differ more and the number of visited nodes increases. However, when the density 
is high, multisets differ less, so the number of visited nodes decreases. Since 
the dependence of the number of visited nodes on the density of multiset-trie 
is a continuous function on the interval $[0,1],$ there exists a global maximum. 
In other words there exists such a value of density where the number of visited 
nodes is maximal.

In this experiment, we empirically find the extremum of density for functions 
\textsc{submsetExistence} and \textsc{supermsetExistence}. The parameters 
$\sigma$ and $n$ are set to 12 and 5 respectively. The density varies from 
$\num{1.0e-4}\%$ to $1.0\%.$ The number of visited nodes was chosen to be 
maximal for each value of particular density.

\begin{figure}
\center
\includegraphics[width=.4\textwidth]{exp4-m1.pdf}
\caption{Experiment 3, submsetExistence function.}
\label{fig:e3m1}
\end{figure}

\begin{figure}
\center
\includegraphics[width=.4\textwidth]{exp4-m2.pdf}
\caption{Experiment 3, supermsetExistence function.}
\label{fig:e3m2}
\end{figure}

As we see on figures~\ref{fig:e3m1} and~\ref{fig:e3m2} both functions 
\textsc{submsetExistence} and \textsc{supermsetExistence} have the maximum 
around $d\approx \num{7.0e-3}\%.$ The maximum is less than $0.03\%$ and 
greater than $\num{1.4e-13}\%,$ which explains the behavior of multiset-trie in 
Experiment~\hyperref[s:exp1]{1} and Experiment~\hyperref[s:exp2]{2}. It is safe 
to say that the maximum may vary depending on parameters $n$ and $\sigma,$ but 
such a maximum always exists. Therefore, we omit the experiments with different 
parameters $n$ and $\sigma.$

\subsection{Experiment 4} \label{s:exp4}
In previous experiments the input was generated artificially with uniform 
distribution, so there was no influence of the mapping function $\phi$ on 
performance of tested functions. This experiment shows the influence of the 
mapping $\phi$ from alphabet $\Sigma$ to a set of consecutive integers. 
We obtain the influence by taking the real world data as an input data.

The data is taken from English dictionary which contains 235883 different words. 
Those words are mapped to multisets of integers according to the $\phi.$ In 
particular, we are interested in cases when $\phi(\Sigma)$ enumerates 
letters by their relative frequency in English language. We say that $\phi(\Sigma)$ 
maps letters in \emph{ascending order} if the most frequent letter is mapped to 
number $\sigma.$ Conversely, in \emph{descending order} this letter is mapped to 
number $1.$ The size of the alphabet $\sigma$ is set to the size of the English 
alphabet 26. The degree of a node $n$ is set to 10. On average the multiplicity 
of letters is of course less than 10. We choose such a large node degree allowing 
the multiplicity to be up to 10, because the dictionary contains such words.

\begin{figure}
\center
\includegraphics[width=.4\textwidth]{exp3-m1.pdf}
\caption{Experiment 4, submsetExistence function.}
\label{fig:e4m1}
\end{figure}

\begin{figure}
\center
\includegraphics[width=.4\textwidth]{exp3-m2.pdf}
\caption{Experiment 4, supermsetExistence function.}
\label{fig:e4m2}
\end{figure}

%\begin{figure}
%\includegraphics[width=\textwidth]{exp4-m3.pdf}
%\caption{Experiment 4, getAllSubmsets function.}
%\end{figure}
%
%\begin{figure}
%\includegraphics[width=\textwidth]{exp4-m4.pdf}
%\caption{Experiment 4, getAllSupermsets function.}
%\end{figure}

The results on figures~\ref{fig:e4m1} and~\ref{fig:e4m2} are more balanced when 
letters are ordered by frequency in ascending order. The maxima for the functions 
\textsc{submsetExistence} and \textsc{supermsetExistence} are at 250 visited nodes.

According to the design of the data structure multiset-trie, we can say 
something about multiset only if we try to reach it, i.e. to find the complete 
path that corresponds to a particular multiset. It means that in order to give 
an answer whether some multiset exists or not one have to check the leaf level in 
multiset-trie.

Letters that have the least frequencies are now located at the top of 
multiset-trie according to ascending order of letters by frequency. This means 
that the search becomes narrower, because a lot of invalid paths will be 
discarded on top most levels. Thus, multiset-trie can be traversed faster.

As you may have noticed the functions \textsc{getAllSubmsets} and 
\textsc{getAllSupermsets} were not tested in this experiment. Those functions 
are not affected by variations of the mapping $\phi,$ because for any multiset 
they retrieve all sub/supermultisets. This means that the number of visited 
nodes will not be changed as $\phi$ varies.

\subsection{Experiment 5} \label{s:exp5}
In this experiment we demonstrate the performance of multiset-trie data structure 
compared to inverted index. The data structures that were used for inverted index 
implementation are trie and hash map. Both indexes were implemented in~\CC{ to} 
avoid any optimization or degradation of performance caused by different programming 
language implementation.

During tests all three data structures used the same interface for benchmark that operates 
with files as its input and output. Each data structure was loaded with the same input data 
and was queried with the same test data. The input data consists of multisets that are 
generated at random with respect to $\sigma = 15$ and $n = 3.$ All together the input dataset 
has 300000 randomly generated multisets. The test dataset was constructed with the same 
$\sigma$ and $n$ having cardinality 100000 multisets.

First, we present the results of performing an exact search query on each data structure. 
The results on figure~\ref{fig:e5ex} show that hash map performs the best, which is not 
surprising, since its access complexity is $O(1).$ However, note that performance of 
multiset-trie does not depend on the multiset being searched. The trie version of the 
inverted index performs the worst in this experiment, because its performance is dependent 
on the multiset size.

\begin{figure}
\center
\includegraphics[width=.4\textwidth]{exact.pdf}
\caption{Experiment 5, exact search.}
\label{fig:e5ex}
\end{figure}

The results of the submultiset query are presented on figure~\ref{fig:e5subm}. Here, each searching 
query was terminated either when the first submultiset is found or all potential submultisets were 
checked for submultiset inclusion. Multiset-trie outperforms both trie and hash map inverted 
indexes by a major factor due to its design. Multiset-trie operates multiplicities on a node level 
while trie does it on a tree level. In the trie one should traverse the tree in order to see if the element 
of some multiplicity is inside, while in multiset-trie one just has to check if the node has appropriate 
child. In the case of hash map we can operate on as low as the whole multiset level, which is of course 
inconvenient when it comes to multiset containment queries.

Note that multiset-trie is sparse in this experiment, however, the multisets that do not satisfy the query 
are discarded much quicker than in trie or hash map.

\begin{figure}
\center
\includegraphics[width=.4\textwidth]{submset.pdf}
\caption{Experiment 5, first submultiset search.}
\label{fig:e5subm}
\end{figure}

### 2018-09-06 tex/tex/s_intro.tex
add paper draft and docs

\section{Introduction} \label{c:introduction}
[Will be rewritten]
%
% some intro
During recent years the popularity of digital data has increased. All sorts of 
information such as text, audio and video can now be accessed by searching 
\emph{information retrieval systems}. Information retrieval (IR) is the science of 
searching information units in a collection of data. Most commonly IR systems 
are used for searching a text-based content such as text documents in a database. 
Such IR systems are called \emph{full-text search systems}.

Full-text search techniques can be applied directly on the database. However, 
it is a very expensive procedure in terms of running time complexity, because it 
requires a frequent accesses to the database. In order to reduce this number, indexes 
were invented. Indexes narrow down the search using pre-generated meta data 
constructed from the data in the database. Furthermore the meta data can be 
organized in a data structure that would provide fast retrieval of data according 
to search queries.

In IR most systems use the concept of an inverted index to achieve full-text 
indexing of a database. Inverted index consists of two parts: \emph{postings} 
and \emph{dictionary}, a search structure that is used to locate a specific entries 
in a posting. A posting entry can be created on different levels depending on data 
that needs to be indexed. Most common for full-text indexes are document level 
entries. In this context a posting is defined to be a list of identifiers or keys that are 
further used to locate a specific document in the database \cite{zobel1992efficient}.

An index is a search structure that is used to process user queries. The query 
can be processed in different ways according to the \emph{retrieval model}. The 
retrieval model that will be discussed in the thesis is the \emph{boolean retrieval 
model} that views each document in a database as a set of words. The document 
itself is an information unit a retrieval system is built over. In our case an 
information unit is defined to be a textual document.

The boolean retrieval model is based on set and multiset theory together with 
boolean algebra. \emph{Set and multiset containment operations} are used to 
derive the similarities between objects, and consequently make decisions on their 
association. Thus, the multiset containment operations allow us to 
search objects not only with exact queries but also to retrieve the most relevant 
set of results that satisfy a given search query \cite{baeza1999modern,manning2008introduction,zobel2006inverted}.

% research topic
% walking through the new terms and so on
% brief overview of the literature (relations to other work etc.)
%

The dictionary (search structure) of the inverted index can be organized in different ways in 
order to meet the required types of queries and specification of data. For example, 
it can be organized as a search tree, hash map, array, heap, linked list, etc. In our 
research we will be focused on search trees. The most efficient search tree index 
nowadays is the Generalized Search Tree (GiST). Its flexibility stems from combining 
functionality of B+-trees, R-trees and RD-trees. GiST further extends their functionality 
providing support for a variety of data types together with the nearest-neighbor 
search \cite{broder2006indexing,hellerstein1995generalized,kornacker1999high}.

% description of innovations
%
The proposed data structure \emph{multiset-trie} can be used as an alternative 
implementation of the search structure in an inverted index. It is an extension of 
the \emph{set-trie} data structure proposed by Savnik \cite{savnik2013index}. 
Set-trie is a trie based data structure that is used for storing and fast retrieval of 
objects represented as sets. The set-trie provides the nearest-neighbor search by 
implementing methods that perform set-containment queries. Multiset-trie extends 
the abilities of set-trie and provides support for storing and retrieving objects 
that can be represented as both sets and multisets. It also implements 
multiset-containment methods together with the basic tree methods such as search, 
deletion and insertion.

The multiset-trie is an $n$-ary tree based data structure with properties similar to 
those of a trie. 
This particular combination allows us to associate multisets with a collection 
of nodes in a tree. Every node represents a symbol with particular multiplicity. 
%
Multiset-trie is a kind of search tree. Similarly to a trie, it uses common prefixes 
to narrow down the search. Unlike the compact prefix tree, Patricia, the multiset-trie 
does not provide the ability to compress a path. However, the absence of path 
compression makes the multiset-trie a perfectly height-balanced tree.

The multiset-trie is designed for efficient execution of the multiset containment 
operations. In particular, it supports the operations \textsc{submsetExistence}, 
\textsc{supermsetExistence}, \textsc{getAllSubmsets} and \textsc{getAllSupermsets}. 
The so-called "existence" queries implement the nearest-neighbor search queries. The 
functions \textsc{submsetExistence} and \textsc{supermsetExistence} search for 
the closest submultiset and supermultiset in the multiset-trie respectively and return 
an answer whether such a multiset exists in the data structure. The so-called 
"getAll" functions act in the same way as "existence" functions, but they do not 
terminate once they have reached the desired multiset. Alternatively, these functions 
store the results and continue until all the multisets that satisfy the query are retrieved.

% paper organization
%
Let us now present the organization of the thesis. 
In the following Chapter~\ref{c:description} we present the description of the 
multiset-trie data structure. The representation of multisets in multiset-trie is 
explained in detail. The organization of the data structure is also presented 
graphically.
%
In Chapter~\ref{c:operations} we present operations that multiset-trie currently 
supports. The multiset containment functions \textsc{submsetExistence}, 
\textsc{supermsetExistence}, \textsc{getAllSubmsets} and \textsc{getAllSupermsets} 
are presented together with the basic search tree functions such as \textsc{insert}, 
\textsc{delete} and \textsc{search}. The algorithms in pseudo code are presented 
as well. 
%
The description of multiset-trie functions and procedures is followed by the 
mathematical analysis of their complexity in Chapter~\ref{c:analysis}. 
In this analysis, we make an assumption that multisets are constructed uniformly at 
random and are parametrized by several parameters, such as multiplicity and the alphabet 
$\Sigma.$ By using probabilistic tools we describe time complexity of the algorithms 
and space complexity of the structure.
%
Further, in Chapter~\ref{c:experiments} we present an empirical study of the 
multiset-trie. Artificially generated as well as real-world data sets are used in experiments. 
The experiments are dedicated to testing the performance of the data structure while 
varying selected parameters. The experiments also show some methods for 
optimizing a multiset-trie.
%
The Chapter~\ref{c:relwork} presents related work. The connection to the set-trie 
data structure \cite{savnik2013index} is discussed more explicitly. We also relate 
the multiset-trie to the information retrieval systems. In particular, we refer to the 
inverted index data structure and discuss how the multiset-trie can be used as a 
database index.
%
Finally, in Chapter~\ref{c:conclusions} our conclusion about the multiset-trie 
data structure and discussion of future work are presented.

### 2018-09-06 tex/tex/s_math.tex
add paper draft and docs

\section{Mathematical analysis of the structure} \label{c:analysis}
%
In this chapter we present theoretical results of time and space complexity of
the multiset-trie data structure. In the following Section~\ref{s:timecomplexity} 
we discuss the running time complexity of the presented algorithms. First, in 
Section~\ref{ss:mathmodel}, we present our mathematical model that we use 
to describe the distribution of multisets in the multiset-trie and input data. 
Using probabilistic approach and tools from a Galton-Watson process we 
measure the expected cardinality of the multiset-trie in Theorem~\ref{thm:exp_nodes}. 
Further, we derive the expected cardinality of the searched subtree of the 
multiset-trie parametrized by an input multiset in Corollary~\ref{cor:exp_nodes_param}.

In Section~\ref{ss:getall} we discuss the running time complexity of the functions 
\textsc{getAllSubmsets} and \textsc{getAllSupermsets}. We observe that the 
complexity of functions is exponential. Moreover, the worst case running time 
complexity is the same for both functions and its upper bound is the cardinality of 
the multiset-trie.

The remaining "existence" functions are discussed in the Section~\ref{ss:exists}. 
We observe that out of scope of our mathematical model unlike in functions 
\textsc{getAllSubmsets} and \textsc{getAllSupermsets} the mapping $\phi$ has an 
impact on performance of the functions \textsc{submsetExistence} and 
\textsc{supermsetExistence}. In particular, the frequency analysis of the symbols 
from $\Sigma$ in input data determines such a $\phi$ that gives a boost in performance.

We find that the performance of the functions \textsc{submsetExistence} and 
\textsc{supermsetExistence} in the worst case scenario is also exponential and 
does not depend on the outcome of the functions. We give a quite precise upper 
bound for the worst case running time complexity, which appears to be the same 
for both functions. However, it must be stressed that for the positive outcome an 
exponential behavior holds only on specific cases, such as presence of the emptyset 
in the multiset-trie.

Finalizing the mathematical analysis, we present the study of space complexity 
of the multiset-trie in the Section~\ref{s:spacecomplexity}. We show that the space 
used for the storage is asymptotically equal to the size of the input data.

\subsection{Time complexity of the algorithms}\label{s:timecomplexity}
The performance of the functions will be measured by the number of 
visited nodes in a multiset-trie during execution of a particular query by the 
functions \textsc{search}, \textsc{delete}, \textsc{submsetExistence}, 
\textsc{supermsetExistence}, \textsc{getAllSubmsets}, \textsc{getAllSupermsets} 
and the procedure \textsc{insert}.
%

By the design of the multiset-trie, it is easy to see that the functions \textsc{search},
\textsc{delete} and the procedure \textsc{insert} have complexity of $O(\sigma).$
Because $\sigma$ is defined when the structure is initialized and does not depend
on the user input afterwards, the asymptotic complexity of the functions \textsc{search}, 
\textsc{delete} and the procedure \textsc{insert} is $O(1).$ Nonetheless, in the general 
case the complexity is $O(\sigma).$

In what follows, we focus on analysis of the more involved functions:
\textsc{submsetExistence}, \textsc{supermsetExistence}, \textsc{getAllSubmsets}
and \textsc{getAllSupermsets}.

\subsubsection{Mathematical model} \label{ss:mathmodel}
We start with the basics of our mathematical model. Let $\Sigma$ be an alphabet of
cardinality $\sigma,$ such that $\Sigma = \{ 1,2, \ldots, \sigma \}.$ Define $N$
to be the set of all possible multisets that can be inserted in
multiset-trie. Let $n$ be the maximal degree of a node in multiset-trie.
Then the maximal multiplicity of an element in a multiset is equal to $n-1.$
Thus, the number of multisets in a complete multiset-trie is $ |N| = n^{\sigma}.$
%
Let $M$ be a collection of multisets inserted into multiset-trie $\mathcal{M}.$
All the multisets in $M$ are constructed from the alphabet $\Sigma$ according
to the parameters $\sigma$ and $n$. Hence, any multiset $m\in M,$ 
has at most $\sigma$ distinct elements that are members of $\Sigma$ and 
every distinct element in $m$ has multiplicity strictly less than $n.$
%
Because a multiset does not distinguish different orderings, it is assumed, for
simplicity that all elements are ordered in an ascending order. A multiset $m$ is
represented as $\{1^{k_1},2^{k_2},\ldots, \sigma^{k_\sigma}\},$ where $e^{k_e}$
represents an element $e\in\Sigma$ with multiplicity $k_e.$
%

Denote the nodes of multiset-trie on all levels but on $\sigma + 1$ as \emph{internal}
and nodes on leaf level as \emph{leaf} nodes.
%
Observe that every internal non-root node has a degree at least~1. Indeed an
insertion of a multiset requires a construction of a path of length~$\sigma + 1,$
meaning that if an internal node exists in a multiset-trie it must have a degree
at least~1. It also follows that the height of a multiset-trie is always~$\sigma +1$
as soon as at least one multiset is inserted into the data structure.

Our model assumes that all the inserted multisets are chosen with the same probability,
meaning that for some $p\in (0,1)$ the following holds:
\[
P(m\in M) = p, \quad \forall~m\in N.
\]
%
Let $\xi_1, \xi_2, \ldots, \xi_{\sigma+1}$ be random variables such that $\xi_i$
represents the number of nodes in a multiset-trie on $i$-th level. For every node $j$ 
on $i$-th level we assign a random variable $\xi_{ij}$ to be the number of its children, 
such that $j\in[1,\xi_i].$ Then $\forall~i\in[1,\sigma]$ the following holds:
%
\begin{equation}\label{eq:sum_recursive}
\xi_{i+1} = \sum_{j=1}^{\xi_i} \xi_{ij},
\end{equation}
%
where $\xi_1 = 1.$
%
It is easy to see that the variable $\xi_{i+1}$ can have values in the interval
$[\xi_i,n^{i}]$ and the value of the variable $\xi_{ij}$ is within the interval $[1,n].$
Without conditioning on the existence of any node in multiset-trie, 
it is easy to describe the probability of existence of any individual node.

\begin{lemma}\label{l:prob-node-existence}

Any potential node on a fixed level $i,$ where $i\in \{ 1,2,\ldots, \sigma +1 \}$ exists, %independently of others,
with probability
\begin{equation}
p_i=1-(1-p)^{n^{\sigma + 1 -i}}.
\end{equation}
\end{lemma}
\begin{proof}
Let $v$ be an arbitrary node in a multiset-trie on an arbitrary level $i.$ Consider
the sub tree with the root $v$ and call it $v$-sub tree. Since the height of the
multiset-trie is $\sigma + 1$ we can calculate the height of the $v$-sub tree.
Taking in account that the root node has height 1, the height of the $v$-sub tree is
%
\[
h_v = \sigma + 1 - i.
\]
%
A node in a multiset-trie exists if at least one node exists on the leaf level of its
sub tree, i.e. a node on the level $\sigma + 1$ that belongs to $v$-sub tree. The possible
number of nodes on the leaf level of $v$-sub tree can be easily calculated knowing its height.
It is equal to
%
\[
n^{\sigma + 1 - i}
\]
%
A node at level $\sigma+1$ exists with probability $p,$ where $p = P(m\in M).$
Thus, the probability that there are no nodes on leaf level in $v$-sub tree is
%
\[
(1-p)^{n^{\sigma +1 - i}}.
\]
%
The claim follows by taking the complement probability of the above result.
%
\end{proof}

However, in order to determine the distribution of $\xi_{ij}$, one needs a lemma of a different type.

\begin{lemma}\label{l:prob-children}
Suppose that a node $v$ exists at level $1\leq i\leq\sigma$.
Then the number of its children $\xi_{iv}$ is modeled by a zero-truncated binomially
distributed random variable on parameters $n$ and $p_{i+1}$. In particular,
the probability of node $v$ having $k$ children equals to
\begin{equation}\label{eq:pdf}
P(\xi_{iv} = k) = \frac{\binom{n}{k} (1-p_{i+1})^{n-k}}{1-(1-p_{i+1})^n}
\end{equation}
and the corresponding probability generating function equals to
\begin{equation}\label{eq:generating_func}
G_i(z) = \frac{(1+p_{i+1}(z-1))^n - (1-p_{i+1})^n}{1-(1-p_{i+1})^n}.
\end{equation}
\end{lemma}
\begin{proof}
In order to prove the lemma, we have to show that
$\xi_{iv}\sim\mathcal{B}_0(n, p_{i+1}).$
Consider an arbitrary node $v$ on level $1\leq i\leq\sigma.$ According to the
definition of the multiset-trie a node exists at level $i$ if and only if
it has at least one child. Note that this is not true for the nodes on the leaf
level $\sigma + 1.$ Implies, a node on level $i$ can have
$k\in\{ 1,2,\ldots, n\}$ children. Let $X_0, X_1, \ldots, X_{n-1}$ be random
variables, they are defined as follows:
\[
X_k = \begin{cases}
0 & \textrm{child k of node $v$ does not exist} \\
1 & \textrm{child k of node $v$ exists} \\
\end{cases}
\]
As it was shown in previous Lemma~\ref{l:prob-children}, the distribution of 
$X_k$ is $X_k\sim Bernoulli(p_{i+1}).$ Since our model assumes that all the 
multisets in $M$ are chosen uniformly at random, the variables 
$X_k,X_l$ are independent for $k\neq l.$ But in our case the node $v$ can not 
have 0 children, so the sum $\sum_{k=1}^n X_k$ has a zero-truncated binomial 
distribution:
%
\[
\sum_{k=1}^n X_k \sim\mathcal{B}_0(n,p_{i+1})
\]
%
which completes the proof.
\end{proof}
%
Knowing the probability density and probability generating functions of $\xi_{ij}$ 
from Lemma~\ref{l:prob-children}, we now can estimate the number of nodes in 
a randomly generated multiset-trie as follows:
%
\begin{equation}\label{eq:num_nodes}
\mathbb{E}( | \mathcal{M} | ) = \mathbb{E}\left[ \sum_{i=1}^{\sigma+1} \xi_i \right].
\end{equation}
%

%Note that in a real world models multisets in $M$ are not chosen with the same 
%probability. Specifically, in some cases, the probability $P(m\in M)$ varies 
%dramatically and can be even equal to $0.$ For example, if words are mapped 
%to multisets, then the sample space $N$ contains very large multisets. However, 
%most of them will have probability $P(m\in M)=0,$ because a word that would correspond 
%to such a multiset simply does not exist. So, we can safely conclude that multiset-trie 
%is an input sensitive data structure, because the size of multiset-trie $|\mathcal{M}|$ 
%depends on the probability distribution function $P(m\in M).$
%

In order to evaluate~(\ref{eq:num_nodes}) we will use some of the tools from 
a Galton-Watson process, see Gardiner~\cite{gardiner1985stochastic} for 
an introduction.
Using the equations~(\ref{eq:sum_recursive}) and~(\ref{eq:generating_func}) we can 
derive the probability generating function for the random variable $\xi_{i+1}$ as 
\begin{equation}\label{eq:g_rec}
G_{\xi_{i+1}}(z) = G_{\xi_i}(G_i(z)).
\end{equation}
Since there is always precisely one node at the root-level, we have $P(\xi_1 = 1) = 1.$ 
Hence, the probability generating function for the random variable $\xi_1$ is 
\begin{equation}\label{eq:g_xi_one}
G_{\xi_1}(z) = z^1 = z
\end{equation}
which is the initial condition for the recursive equation~(\ref{eq:g_rec}).

\begin{proposition}\label{prop:exp_rec}
The expectation of the random variable $\xi_{i+1}$ can be expressed 
as follows.
\[
\mathbb{E}(\xi_{i+1}) = \mathbb{E}(\xi_i)\mathbb{E}(\mathcal{B}_0(n,p_{i+1}))
\]
for $1\leq i\leq\sigma.$
\end{proposition}
\begin{proof}
Using the following property of probability generating function
\begin{equation}\label{eq:prop_exp-pgf}
G_X'(1^-) = \mathbb{E}(X)
\end{equation}
the expectation for the random variable $\xi_{i+1}$ can be derived in terms of 
the equation~(\ref{eq:g_rec}).
\begin{eqnarray}\label{eq:exp}
\mathbb{E}(\xi_{i+1}) &=& G_{\xi_{i+1}}'(1^-) \nonumber \\
&=& G_{\xi_i}'(G_i(1^-))G_i'(1^-).
\end{eqnarray}
According to~(\ref{eq:pdf}) and~(\ref{eq:generating_func}) the value of $G_i(z)$ at 
$1$ is $1$ and the value of its derivative at $1$ is $\mathbb{E}(\mathcal{B}_0(n,p_{i+1})).$ 
Substituting the values of $G_i(1^-)$ and $G_i'(1^-),$ and applying the 
property~(\ref{eq:prop_exp-pgf}) we complete the proof.
\end{proof}
%
From the Proposition~\ref{prop:exp_rec} above and Lemma~\ref{l:prob-children} we 
can conclude that 
\begin{eqnarray}
\mathbb{E}(\xi_{i}) &=& \mathbb{E}(\xi_{i-1})\mathbb{E}\left( \mathcal{B}_0(n,p_{i}) \right) \nonumber \\
& = & \mathbb{E}(\xi_{i-1})\frac{np_{i}}{1-(1-p_{i})^n}.
\end{eqnarray}

\begin{theorem}\label{thm:exp_level}
Let $\mathcal{M}$ be a multiset-trie defined with parameters $n,$ $\sigma,$ and denote the number 
of nodes on every level $i$ by a random variable $\xi_i.$ Furthermore, let all multisets appear in 
$\mathcal{M}$ with equal probability $p\in (0,1).$ Then the expected number 
of nodes on every level of $\mathcal{M},$ i.e. $\mathbb{E}(\xi_i)$ is defined as 
\begin{equation}\label{eq:nodes_level}
\mathbb{E}(\xi_{i}) = n^{i-1} \frac{1-(1-p)^{n^{\sigma +1 -i}}}{1-(1-p)^{n^{\sigma}}}.
\end{equation}
\end{theorem}
\begin{proof}
According to~(\ref{eq:g_xi_one}) the expected number of nodes on the first level is 1. \\
Using $\mathbb{E}(\xi_1) = 1$ and the result from Proposition~\ref{prop:exp_rec}
we get
\begin{eqnarray*}
\mathbb{E}(\xi_{i}) &=& \prod_{j=2}^{i} \frac{n p_j}{1-(1-p_j)^n}
= \prod_{j=2}^{i} n \frac{1-(1-p)^{n^{\sigma +1-j}}}{1-(1-p)^{n^{\sigma + 2 -j}}} \\
&=& n^{i-1} \frac{1-(1-p)^{n^{\sigma +1 -i}}}{1-(1-p)^{n^{\sigma}}}
\end{eqnarray*}
\end{proof}

Having derived the expected number of nodes on every level of multiset-trie, the
expected value of the total number of nodes in a multiset-trie can be calculated 
with respect to the parameters $n,$ $\sigma$ and $p.$ This result is obtained in 
the next theorem.

\begin{theorem}\label{thm:exp_nodes}
The expected cardinality of a multiset-trie defined on parameters $n,$ $\sigma$ 
and $p$ can be computed as 
\begin{equation}
\mathbb{E}( | \mathcal{M} | ) = \sum_{i=1}^{\sigma+1} n^{i-1} \frac{1-(1-p)^{n^{\sigma +1 -i}}}{1-(1-p)^{n^{\sigma}}},
\end{equation}
where $r = (1-p)^n,$ so $r\in(0,1).$
\end{theorem}
\begin{proof}
Using the results obtained from Theorem~\ref{thm:exp_level} we compute 
\begin{eqnarray*}
\mathbb{E}( | \mathcal{M} | ) &=& \mathbb{E}\left[ \sum_{i=1}^{\sigma+1} \xi_i \right] \\
&=& \sum_{i=1}^{\sigma + 1} n^{i-1} \frac{1-(1-p)^{n^{\sigma + 1 -i}}}{1-(1-p)^{n^{\sigma}}} \\
\end{eqnarray*}
\end{proof}

With the expected number of nodes in a multiset-trie $\mathcal{M}$ obtained from 
Theorem~\ref{thm:exp_nodes}, we can now generalize the result for a subtree 
in $\mathcal{M}$ parametrized by an input multiset $m.$ The subtrees that we 
are interested in are the ones that contain all the submultisets or all the 
supermultisets of $m.$ In order to calculate the expected cardinality of such 
subtrees we need the following definition.

\begin{definition}\label{def:params}
Let $m = \{ 1^{k_1}, 2^{k_2}, \ldots, \sigma^{k_\sigma} \},$ where 
$e^{k_e}$ is an element $e$ with multiplicity $k_e.$ Let $M_1, M_2$ be 
the subsets of the set $M,$ such that $M_1 = \{ x\in M : x\subseteq m \}$ and 
$M_2 = \{ x\in M : x\supseteq m \}.$
Define $\alpha_i$ and $\beta_i$ as follows 
\begin{equation*}
\alpha_i = \begin{cases}
1, & i=0 \\
\prod_{j=1}^i (k_j+1), & 1\leq i\leq\sigma \\
\end{cases}
\end{equation*} 
and 
\begin{equation*}
\beta_i = \begin{cases}
1, & i=0 \\
\prod_{j=1}^i (n-k_j-1), & 1\leq i\leq\sigma \\
\end{cases}.
\end{equation*}
\end{definition}

The expected cardinality of the subtrees containing the multisets from $M_1$ or $M_2$ 
is defined in the following corollary.

\begin{corollary}\label{cor:exp_nodes_param}
Let $M_1, M_2, \alpha_i$ and $\beta_i$ be defined as in previous Definition~\ref{def:params}, 
then the expected cardinality of a multiset-trie subtree $\mathcal{M}_{M_1}$ that contains all the multisets 
from the set $M_1$ is equal to 
\begin{equation}\label{eq:nodes_sub_param}
\mathbb{E}( |\mathcal{M}_{M_1}| ) = \sum_{i=1}^{\sigma + 1} \alpha_{i-1} \frac{1-(1-p)^{\alpha_{i-1}}}{1-(1-p)^{\alpha_{\sigma}}}.
\end{equation}
The expected cardinality of a multiset-trie subtree $\mathcal{M}_{M_2}$ that contains all the multisets 
from the set $M_2$ is equal to 
\begin{equation}\label{eq:nodes_super_param}
\mathbb{E}( |\mathcal{M}_{M_2}| ) = \sum_{i=1}^{\sigma + 1} \beta_{i-1} \frac{1-(1-p)^{\beta_{i-1}}}{1-(1-p)^{\beta_{\sigma}}}.
\end{equation}
\end{corollary}
%
\begin{proof}
Using the results from Theorem~\ref{thm:exp_level} and Theorem~\ref{thm:exp_nodes} 
we derive the formulas~(\ref{eq:nodes_sub_param}) and~(\ref{eq:nodes_super_param}) 
by specifying the possible number of nodes on every level in the multiset-trie according 
to the multiset $m.$ Note that the formula~(\ref{eq:nodes_level}) assumes that on every 
level but the first one there are $n$ possible nodes. Given submultiset or supermultiset 
query and an input multiset $m$ the number of nodes that will be traversed on level $i$ is 
defined by the number $k_{i-1}+1$ or $n-k_{i-1}-1$ for $i\geq 2.$ On level $i=1$ 
there is only one root node in any multiset-trie $\mathcal{M},$ which always exists 
if $M\neq\emptyset$ and is traversed for any type of query (submultiset and 
supermultiset).
\end{proof}

\subsubsection{GetAllSubmsets and GetAllSupermsets}\label{ss:getall}
In this subsection we discuss the running time complexity of the functions 
\textsc{getAllSubmsets} and \textsc{getAllSupermsets}. It is obvious that any 
other algorithm for retrieving all the submultisets or supermultisets has worst 
case running time complexity at least $O(|M|).$ Hence, the functions 
\textsc{getAllSubmsets} and \textsc{getAllSupermsets} have the worst case 
running time complexity $O(|\mathcal{M}|).$ Indeed, the case when the algorithms 
retrieve all the multisets stored in a multiset-trie by traversing the whole structure 
can be easily constructed.

Consider the function \textsc{getAllSubmsets}. The function takes some multiset 
$m$ as an input argument. Then it returns a set of multisets $\{x\in M : x\subseteq m \}$ 
from the multiset-trie $\mathcal{M}$. Having a multiset $m$ set to the largest 
possible multiset in $N$ (it can also be larger) 
\[
m = \{ 1^{n-1},2^{n-1},\ldots, \sigma^{n-1} \}
\]
the whole multiset-trie is traversed during the \textsc{getAllSubmsets} query.

Now let us consider the function \textsc{getAllSupermsets}. Similarly, the function 
takes a multiset $m$ as an input argument. However, in this case it returns the 
set of multisets $\{ x\in M : x\supseteq m \}$ from the multiset-trie 
$\mathcal{M}.$ In order to obtain a traversing of all the multiset-trie one must set 
$m$ to the smallest possible multiset, i.e. an empty multiset
\begin{equation*}
m = \{ \emptyset \} = \{ 1^0,2^0, \ldots, \sigma^0 \}.
\end{equation*}

Thus, we can conclude that the worst case running time complexity of the 
functions \textsc{getAllSubmsets} and \textsc{getAllSupermsets} is $O(\mathbb{E}( |\mathcal{M}| )).$ 
According to the Theorem~\ref{thm:exp_nodes} the expected number of visited 
nodes in the worst case is 
\begin{equation*}
O(\sum_{i=1}^{\sigma+1} n^{i-1} \frac{1-(1-p)^{n^{\sigma +1 -i}}}{1-(1-p)^{n^{\sigma}}}).
\end{equation*}
According to the Theorem~\ref{cor:exp_nodes_param} the worst case running 
time complexity given an input multiset $m$ for the function \textsc{getAllSubmsets} is 
\begin{equation*}
O(\sum_{i=1}^{\sigma + 1} \alpha_{i-1} \frac{1-(1-p)^{\alpha_{i-1}}}{1-(1-p)^{\alpha_{\sigma}}})
\end{equation*}
and for the function \textsc{getAllSupermsets} is 
\begin{equation*}
O(\sum_{i=1}^{\sigma + 1} \beta_{i-1} \frac{1-(1-p)^{\beta_{i-1}}}{1-(1-p)^{\beta_{\sigma}}}).
\end{equation*}

%Taking into account that $\sigma |M| \geq |\mathcal{M}|,$ we can say that the 
%running time complexity is $O(\sigma |M|).$ However, this is a very rough upper 
%bound, which overestimates the complexity a lot.

\subsubsection{SubmsetExistence and SupermsetExistence}\label{ss:exists}
We start the analysis of the functions \textsc{submsetExistence} and 
\textsc{supermsetExistence} with an observation. Our theoretical model assumes 
that all the multisets are inserted into multiset-trie at random. It was already concluded 
that the probability distribution function $P(m\in M)$ has an impact on the size of 
mulitset-trie $\mathcal{M}.$ Moreover, this distribution influences on the performance 
of the functions \textsc{submsetExistence} and \textsc{supermsetExistence} even more.

For a real world model, such that $P(m\in M)\neq const$ the performance of the search 
algorithms directly depends on the number of nodes on every level $\xi_i.$ When 
the search functions check if a multiset is in multiset-trie the complete path that 
corresponds to that multiset is checked. Knowing that fact the search can be optimized 
during the construction of a multiset-trie.

Recall that a multiset-trie is defined on parameters $n,$ $\Sigma,$ $\sigma = |\Sigma|$ 
and $\phi.$ Let the frequency of an element $e$ in a multiset $m$ be the multiplicity of 
$e$ in $m,$ denoted by $\emph{mult}_m(e).$ Then the frequency of an element $e$ 
can be defined as a sum $\sum_{m\in M} \emph{mult}_m(e).$ According to the frequencies 
of elements in $\Sigma,$ the performance of the multiset-trie can be optimized by 
the mapping $\phi : \Sigma \rightarrow I.$ Indeed the ordering of elements by their 
frequencies has an influence on the performance.
%
The frequency of an element $e\in\Sigma$ affects the distribution of $\xi_{\phi(e)}$ 
as follows. The larger the frequency of $e$ the larger the number of nodes on 
$\phi(e)$ level. 
So, if the number of nodes on lower levels is greater than on higher levels, then 
the search functions will discard complete paths that do not satisfy the query 
faster. Hence, the closest match will be found faster.

%
%\begin{lemma}\label{l:frequency}
%Let $M$ be a collection of multisets. Let the frequency of an element $e$ in a 
%multiset $m$ be the multiplicity of $e$ in $m,$ denoted by $\emph{mult}_m(e).$ 
%Then the frequency of an element $e$ is the sum 
%\[
%\sum_{m\in M} \emph{mult}_m(e).
%\] 
%According to the frequencies of elements in $\Sigma,$ the performance of the 
%multiset-trie can be optimized. In particular, the mapping $\phi : \Sigma \rightarrow I$ 
%must order elements by their frequencies as follows. The smaller the frequency of 
%an element the smaller its index.
%\end{lemma}
%%
%\begin{proof}
%The frequency of an element $e\in\Sigma$ affects the distribution of $\xi_{\phi(e)}$ 
%as follows. The larger the frequency of $e$ the larger the number of nodes on 
%$\phi(e)$ level. 
%So, if the number of nodes on lower levels is greater than on higher levels, then 
%the search functions will discard complete paths that do not satisfy the query 
%faster. Hence, the closest match will be found faster.
%\end{proof}
%%

Let us now switch back to our mathematical model and note that the influence 
of the mapping function $\phi$ in our model has inessential impact on performance, 
because all the multisets are equally likely and the whole domain $N$ is used for 
sampling multisets.

Consider both functions \textsc{submsetExistence} and \textsc{supermsetExistence}. 
Whenever the result is \emph{false}, i.e. no multiset in $M$ is 
a submultiset or supermultiset of an input multiset $m,$ both functions in the 
worst case visit all the nodes in $\mathcal{M}$ but the nodes on leaf level. 
Of course such a case would be very rare assuming a random input model, but 
it can be constructed as follows.

Consider the function \textsc{submsetExistence}. Then given an input multiset 
$m = \{ 1^{k_1}, 2^{k_2}, \ldots, \sigma^{k_\sigma} \},$ the collection of inserted 
multisets $M$ must be equal to $M = \{ x\in M : k_{x,\sigma} > k_{m,\sigma} \}.$ 
Analogically for the function \textsc{supermsetExistence} with an input multiset 
$m = \{ 1^{k_1}, 2^{k_2}, \ldots, \sigma^{k_\sigma} \},$ the collection of inserted 
multisets $M$ must be equal to $M = \{ x\in M : k_{x,\sigma} < k_{m,\sigma} \}.$

Thus, the worst case running time complexity of the functions 
\textsc{submsetExistence} and \textsc{supermsetExistence} is $O(|\mathcal{M}| - |M|).$ 
According to Theorem~\ref{thm:exp_nodes}, this value is 
\begin{equation*}
O(\sum_{i=1}^{\sigma} n^{i-1} \frac{1-(1-p)^{n^{\sigma +1 -i}}}{1-(1-p)^{n^{\sigma}}}).
\end{equation*}
According to Theorem~\ref{cor:exp_nodes_param} the worst case running 
time given an input multiset $m$ for the function \textsc{submsetExistence} is 
\begin{equation*}
O(\sum_{i=1}^{\sigma} \alpha_{i-1} \frac{1-(1-p)^{\alpha_{i-1}}}{1-(1-p)^{\alpha_{\sigma}}})
\end{equation*}
and for the function \textsc{supermsetExistence} is 
\begin{equation*}
O(\sum_{i=1}^{\sigma} \beta_{i-1} \frac{1-(1-p)^{\beta_{i-1}}}{1-(1-p)^{\beta_{\sigma}}}).
\end{equation*}
Note that the summation goes only up to $\sigma$ and not up to $\sigma + 1$ as 
in the Theorem~\ref{thm:exp_nodes} or in the Theorem~\ref{cor:exp_nodes_param}.

As for the case when the outcome of the functions \textsc{submsetExistence} and 
\textsc{supermsetExistence} is \emph{true} one has to guarantee the termination 
of the algorithm at some node on the leaf level. The worst case scenario can be 
constructed in the same way as for the \emph{false} outcome but with two more 
multisets in $M.$ The first multiset is the empty multiset. With the empty multiset 
the function \textsc{submsetExistence} will visit the same amount of nodes as for 
the \emph{false} case plus one more for the empty multiset. The second multiset 
is the maximal possible multiset from $N.$ In this case the function \textsc{supermsetExistence} 
will also visit the same amount of nodes as for the \emph{false} case plus one more 
for the maximal multiset. Hence, the worst case running time complexity for both 
outcomes (\emph{true} and \emph{false}) is the same.

\subsection{Space complexity}\label{s:spacecomplexity}
As in any efficient algorithm there is always some trade-off between space and 
time complexity. While offering efficient sub- and supermultiset queries an 
additional space must be provided for multisets storage. Clearly, the cardinality 
of the set $M$ is smaller than the size of $\mathcal{M},$ because the number 
of multisets in $\mathcal{M}$ is equal to the number of nodes only on the leaf 
level. The figure~\ref{f:exp-nodes} demonstrates the relation between the number 
of multisets stored and the number of nodes needed for storage, where parameters 
$\sigma$ and $n$ are $26$ and $10,$ respectively.

\begin{figure}[h!]
\center
\includegraphics[width=.4\textwidth, keepaspectratio]{exp-nodes-on-probab.pdf}
\caption{$\mathbb{E}(|\mathcal{M}|)$ and $\mathbb{E}(|M|)$ on probability.}
\label{f:exp-nodes}
\end{figure}

As we see on the figure~\ref{f:exp-nodes} the value of $|\mathcal{M}|$ is slightly 
shifted with respect to the value of $|M|.$

Now we demonstrate a more descriptive comparison between $|\mathcal{M}|$ and 
$|M|.$ Figure~\ref{f:ratio-exp-msets} shows the ratio between the expected cardinality 
of a multiset-trie $|\mathcal{M}|$ and the actual number of multisets stored $|M|$ for 
parameters $n$ and $\sigma$ being $10$ and $26$ respectively.

\begin{figure}[h!]
\center
\includegraphics[width=.4\textwidth, keepaspectratio]{ratio-exp-nodes-and-msets-on-prob.pdf}
\caption{Ratio $\mathbb{E}(\frac{|\mathcal{M}|}{|M|})$ on $p.$}
\label{f:ratio-exp-msets}
\end{figure}

Note that analyzing the graph on figure~\ref{f:ratio-exp-msets} we can safely 
say that the upper bound for the ratio is $\sigma + 1.$ The argument holds, 
because of the limit 
\begin{equation}
\lim_{p\rightarrow 0^+} \mathbb{E}(\xi_i) = 1,
\end{equation}
where $\xi_i$ is the number of nodes on $i$-th level and $1\leq i \leq \sigma + 1.$

However, the ratio $\sigma + 1$ can be obtained only with a very small cardinality 
of the set $M,$ in particular $|M| = 1.$ In order to obtain such a case the 
probability $p$ must be at most $\frac{1}{n^\sigma}.$

The lower bound for the ratio is obviously at $p=1$ and is equal to 1
\begin{equation}
\lim_{n,\sigma \rightarrow\infty} \frac{n^{\sigma + 1} - 1}{n^\sigma (n-1)} = 1.
\end{equation}

Since the ratio $\sigma + 1$ can be obtained for a very specific case only and 
with a small increase of probability the ratio drops rapidly it can be concluded 
that the space complexity of the multiset-trie is $O(|M|).$

### 2018-09-06 tex/tex/s_operations.tex
add paper draft and docs

\section{Multiset-trie operations} \label{c:operations}
%
Let $\mathcal{M}$ be a multiset-trie and let $M$ be a set of multisets that are 
inserted into the multiset-trie $\mathcal{M}.$ We define a type \emph{Multiset} in 
order to use it as a representation of a multiset. The type \emph{Multiset} is 
an array $m$ of constant length $\sigma,$ where $i-th$ cell represents the element 
$\phi^{-1}(i)$ from $\Sigma$ with multiplicity $m[i].$ From now on we 
agree that the first cell of an array has index 1. Let us have an example of a 
\emph{Multiset} instance with $\sigma = 2:$
%
\begin{center}
\begin{tabular}{ccc}
Multiset & & Instance of type Multiset \\
$\{ 1,1,2 \}$ & $\cong $ & 
\begin{tabular}{|c|c|}
\hline 
2 & 1 \\
\hline 
\multicolumn{1}{c}{\tiny 1} & \multicolumn{1}{c}{\tiny 2} \\
\end{tabular}
\end{tabular}
\end{center}
%
The operations supported by the multiset-trie data structure are as follows.
%
\begin{enumerate}
\item \textsc{insert}($\mathcal{M}$, $m$): inserts a multiset $m$ into 
$\mathcal{M}$ if $m\not\in M;$
%
\item \textsc{search}($\mathcal{M}$, $m$): returns true if a multiset $m\in M$ 
for a given $\mathcal{M},$ and returns false otherwise;
%
\item \textsc{delete}($\mathcal{M}$, $m$): returns true if a multiset $m$ was 
successfully deleted from $\mathcal{M},$ and returns false otherwise (in case 
$m\not\in M$);
%
\item \textsc{submsetExistence}($\mathcal{M}$, $m$): returns true if 
there exists a $x\in M$ for a given $\mathcal{M}$ such that $x\subseteq m,$ 
and returns false otherwise; 
%
\item \textsc{supermsetExistence}($\mathcal{M}$, $m$): returns true if 
there exists a $x\in M$ for a given $\mathcal{M}$ such that $x\supseteq m,$ and returns 
false otherwise; 
%
\item \textsc{getAllSubmsets}($\mathcal{M}$, $m$): returns the set of multisets 
$\{ x \in M : x\subseteq m \}$ for a given $\mathcal{M};$
%
\item \textsc{getAllSupermsets}($\mathcal{M}$, $m$): returns the set of multisets 
$\{ x\in M : x\supseteq m \}$ for a given $\mathcal{M}.$
%
\end{enumerate}

In the following subsections we will present each operation of the multiset-trie 
data structure separately.

Firstly we would like to describe some notations that will be used. The 
multiset-trie data structure is a recursive data structure. Hence, any sub 
tree of a multiset-trie $\mathcal{M}$ is again a multiset-trie. This fact allows 
us to use the root node of a multiset-trie as its representative. Thus, the notation 
$\mathcal{M}$ will be used instead of $\mathcal{M}.root$ to refer to the root 
node of $\mathcal{M}.$ Non-existing or \emph{Null} nodes in multiset-trie will 
be marked as \emph{Null} and existing nodes at the level $LL$ will be marked 
as \emph{accepting} nodes. The array slicing operation will be used as follows. 
For a given array $a,$ $a[i:]$ represents the array obtained from $a$ by taking 
only the cells from index $i$ until the last cell.

\subsection{Insert} \label{s:insert}
The procedure \textsc{insert}($\mathcal{M}$, $m$) inserts a new instance $m$ of 
type Multiset into multiset-trie $\mathcal{M}$. If the complete path already 
exists, then procedure leaves the structure unchanged. Otherwise it extends 
partially existing or creates a new complete path. The procedure does not return 
any result. The pseudocode for procedure \textsc{insert} is presented in 
Algorithm~\ref{alg:insert}.

\begin{algorithm}[h!]
\caption{Procedure \textsc{insert}}
\label{alg:insert}
\begin{algorithmic}[1]
\Procedure{\textsc{insert}}{$\mathcal{M}$, $m$}
\State $currentNode \gets \mathcal{M}$
\For{$i=1$ to $\sigma$}
\If {child $c_{m[i]}$ of $currentNode$ is \emph{Null}}
\State create new child $c_{m[i]}$ of $currentNode$
\EndIf
\State $currentNode \gets c_{m[i]}$
\EndFor
\State mark $currentNode$ as \emph{accepting}
\EndProcedure
\end{algorithmic}
\end{algorithm}

\subsection{Search}\label{s:search}
The function \textsc{search}($\mathcal{M}$, $m$) checks if the complete path corresponding to 
a given multiset $m$ exists in the structure $\mathcal{M}.$ The function returns 
true if the multiset $m$ exists in $\mathcal{M}$, and returns false otherwise. The 
function \textsc{search} is presented in Algorithm~\ref{alg:search}.

\begin{algorithm}[h!]
\caption{Function \textsc{search}}
\label{alg:search}
\begin{algorithmic}[1]
\Function{\textsc{search}}{$\mathcal{M}$, $m$}
\State $currentNode \gets \mathcal{M}$
\For{$i=1$ to $\sigma$}
\If {child $c_{m[i]}$ of $currentNode$ is \emph{Null}}
\State \Return False
\EndIf
\State $currentNode \gets c_{m[i]}$
\EndFor
\State \Return True
\EndFunction
\end{algorithmic}
\end{algorithm}

\subsection{Delete} \label{s:delete}
The function \textsc{delete}($\mathcal{M},$ $m$) searches for the complete path 
that corresponds to $m$ in order to remove it. If the path can not be found, the 
function immediately returns false. During search, the function keeps track of the 
number of children for every node. It marks the nodes that have more than one child 
as \emph{parent nodes} and remembers the label of the child which is a potential node 
where the sub-tree will be cut to remove the multiset. The parent node is needed to 
perform a removal, because the multiset-tire is an explicit data structure. When search 
is completed, the function removes the sub-tree of the last found parent node, and 
returns true. In such a way after deletion all the prefixes for other multisets are 
preserved in $\mathcal{M}$ and $m$ is removed. The function \textsc{delete} is 
presented in Algorithm~\ref{alg:delete}.

\begin{algorithm}[h!]
\caption{Function \textsc{delete}}
\label{alg:delete}
\begin{algorithmic}[1]
\Function{\textsc{delete}}{$\mathcal{M},$ $m$}
\State $currentNode \gets \mathcal{M}$
\State $parent \gets currentNode$ 
\State $position \gets 1$
\For {$i=1$ to $\sigma$}
\If {child $c_{m[i]}$ of $currentNode$ is \emph{Null}}
\State \Return False
\EndIf
\State $numChildren \gets 0$
\For {$j=0$ to $n-1$}
\If {child $c_j$ of $currentNode$ is not \emph{Null}}
\State $numChildren\gets numChildren+1$
\EndIf
\EndFor
\If {$numChildren$ is not $1$}
\State $parent\gets currentNode$
\State $position \gets i$
\EndIf
\State $currentNode \gets c_{m[i]}$
\EndFor
\State child $c_{m[position]}$ of $parent\gets$ \emph{Null}
\State \Return True
\EndFunction
\end{algorithmic}
\end{algorithm}

\subsection{Sub-multiset existence} \label{s:subexists}
The function \textsc{submsetExistence}($\mathcal{M}$, $m$) checks if there exists 
a multiset $x$ in $\mathcal{M},$ that satisfies the condition $x\subseteq m.$ 
The function starts with searching for an exact match $x=m$ in $\mathcal{M},$ 
since $m\subseteq m$ by definition of submultiset inclusion. If an exact match is 
not found in $\mathcal{M},$ the function uses multiset-trie to find the closest 
(the largest) submultiset of $m$ in $\mathcal{M}$ by decreasing multiplicity of 
elements in $m.$ At every level the function tries to proceed with the largest 
possible multiplicity of an element that is provided by $m.$ However, when the 
function reaches some level where it meets a \emph{Null} node and can not go 
further using path provided by $m,$ it decreases the multiplicity of an element 
that corresponds to a current level. Thus, the function can decrease multiplicity 
of an element or eventually skip it in order to find the closest $x\subseteq m.$ 
The function \textsc{submsetExistence} is presented in Algorithm~\ref{alg:subexists}.

\begin{algorithm}[h!]
\caption{Function \textsc{submsetExistence}}
\label{alg:subexists}
\begin{algorithmic}[1]
\Function{\textsc{submsetExistence}}{$\mathcal{M}$, $m$}
\State $currentNode \gets \mathcal{M}$
\If {$currentNode$ is \emph{accepting}}
\State \Return True
\EndIf
\For {$i=m[1]$ down to $0$}
\If {child $c_i$ of $currentNode$ is not \emph{Null}}
\If {\textsc{submsetExistence}($c_i,$ $m[2:]$)}
\State \Return True
\EndIf
\EndIf
\EndFor
\State \Return False
\EndFunction
\end{algorithmic}
\end{algorithm}

\subsection{Super-multiset existence} \label{s:superexists}
The function \textsc{supermsetExistence}($\mathcal{M},$ $m$) checks if there 
exists supermultiset $x$ of a given multiset $m$ in $\mathcal{M}.$ By analogy 
to the function \textsc{submsetExistence}, the function \textsc{supermsetExistence} 
starts by searching for an exact match $x=m$ in $\mathcal{M}.$ If an exact 
match is not found in $\mathcal{M},$ the function searches for the closest (the 
smallest) supermultiset $x$ of $m$ in $\mathcal{M}$ by increasing multiplicity 
of elements in $m.$ At every level the function tries to proceed with the smallest 
possible multiplicity of an element that is provided by $m.$ However, when function 
reaches some level where it meets a \emph{Null} node and can not go further 
using path provided by $m,$ it increases the multiplicity of an element that 
corresponds to a current level. Thus, the function \textsc{supermsetExistence} 
can increase multiplicity of an element up to $n-1,$ where $n$ is the degree of 
a node in $\mathcal{M},$ to find the closest supermultiset $x\supseteq m$ in 
$\mathcal{M}.$ The function \textsc{supermsetExistence} is presented in 
Algorithm~\ref{alg:superxists}.

\begin{algorithm}[h!]
\caption{Function \textsc{supermsetExistence}}
\label{alg:superxists}
\begin{algorithmic}[1]
\Function{\textsc{supermsetExistence}}{$\mathcal{M}$, $m$}
\State $currentNode \gets \mathcal{M}$
\If {$currentNode$ is \emph{accepting} }
\State \Return True
\EndIf
\For {$i=m[1]$ to $n-1$}
\If {child $c_i$ of $currentNode$ is not \emph{Null}}
\If {\textsc{supermsetExistence}($c_i,$ $m[2:]$)}
\State \Return True
\EndIf
\EndIf
\EndFor
\State \Return False
\EndFunction
\end{algorithmic}
\end{algorithm}

\subsection{Get all sub-multisets and get all super-multisets} \label{s:getall}
The algorithms for functions \textsc{getAllSubmsets} and 
\textsc{getAllSupermsets} are based entirely on algorithms for 
\textsc{submsetExistence} and \textsc{supermsetExistence} functions that do not 
terminate on the first existing sub/supermultiset, but store the results and 
continue procedure until all existing sub/supermultisets in $\mathcal{M}$ are 
found and stored.

### 2018-09-06 tex/tex/s_relwork.tex
add paper draft and docs

\section{Related work} \label{c:relwork}
[Will be rewritten]
%
% Set theory, multiest theory, bag-of-words model
\subsection{Multiset}
Multiset is a widely used data structure in different areas of mathematics, physics 
and computer science \cite{singh2007overview}. The theory of multisets is based 
entirely on the theory of sets. However, classical mathematics does not deal with 
multisets directly. Instead, one can define a multiset to be a \emph{family} of sets 
or the functions on ordered pairs, where the members of a pair are an element and its 
multiplicity. This means that mathematically the concepts of set such as cardinality, 
set-containment operation, power set, equivalence classes and others are well defined 
for multisets in terms of sets \cite{blizard1988multiset}.

The concept of a multiset can also be referred to the \emph{bag-of-words model}. 
This model takes its origin from a linguistic context studied by Harris~\cite{harris1954distributional}. 
According to the bag-of-words model, text can be represented as a bag (multiset) 
of words, where an element is a word and the number of its occurrences in the 
text is multiplicity. A bag of words does not keep track of grammar and ordering 
of words.

% Information retrieval
\subsection{Information retrieval}
\emph{Information retrieval (IR)} refers to a problem of finding material of an 
unstructured nature that satisfies an information need~\cite{manning2008introduction}. 
Usually, one is searching for a specific documents in a significantly large text 
documents database. The size of a database makes the search a time consuming 
operation. In order to resolve the issue IR systems pre-process data and create 
indexes for future use in search operation.

The bag-of-words model is widely used in IR. In particular, such a representation of 
text documents is used in database indexes when a full text search of a database 
is required. The full-text search problem refers to indexing techniques for full-text 
databases. The most efficient index nowadays uses the concept of an inverted index 
\cite{zobel1992efficient}.

The proposed data structure multiset-trie can be used as an alternative implementation 
of the search structure of an inverted index. It represents words as multisets and 
stores them into data structure. The query processing is achieved using boolean 
retrieval model \cite{manning2008introduction} and multiset containment operations. 
Multiset containment operations of the multiset-trie implement the nearest neighbor 
search queries which retrieve not only exact but also the most relevant results 
to a user.

\subsection{Generalized search tree}
The properties and operations of the multiset-trie makes it a competitor to the 
most efficient implementation of a search tree the \emph{Generalized Search Tree 
(GiST)}~\cite{broder2006indexing,hellerstein1995generalized,kornacker1999high} 
that is used in inverted index. GiST is a very flexible data structure that can be 
customized in order to behave like B+-tree, R-tree or RD-tree. It also provides 
support for an extensible set of queries and data types that B+-tree, R-tree or RD-tree 
do not support originally. GiST supports all the basic search tree operations such as 
insert, delete and search, and in addition provides such extensions as the 
nearest-neighbor search and multiset containment operations. The extensions 
provided by GiST are native in the multiset-trie. The multiset-trie also has a fixed 
height while GiST is a self-balanced tree and has to use additional methods in 
order to preserve its balance.

% Set-trie
\subsection{Set-trie}
The multiset containment queries are well studied in the area of relational 
databases. The queries are well-defined in the context of relational 
algebra~\cite{Lamperti2000msetsinds}. This problem was previously studied 
for a restricted case of queries. In particular, the storage and fast retrieval of sets 
were previously accomplished in data structure \emph{set-trie} proposed by 
Savnik~\cite{savnik2013index}.

The set-trie data structure is based on a trie data structure. It supports set 
containment operations such as retrieval of the nearest sub- and supersets and 
retrieval of all sub- and supersets from the data structure.

The data structure multiset-trie adapts the properties of the set-trie implementing 
the functions \textsc{submsetExistence}, \textsc{supermsetExistence}, 
\textsc{getAllSubmsets} and \textsc{getAllSupermsets} together with the basic 
tree functions such as \textsc{insert}, \textsc{delete} and \textsc{search}. 
Moreover, multiset-trie extends the abilities of the set-trie allowing to store and 
retrieve multisets. The downside of such an extension is that multiset-trie no 
longer supports path compression that was obtained in set-trie. However, the 
design of multiset-trie provides a constant worst case time complexity of search 
function independently of user input.

### 2018-09-12 tex/tex/s_experiments.tex
algorithms adjusted and get-all algorithms added

@@
 \section{Experiments} \label{c:experiments}
+
 %
@@
 $N$ is $33554432\approx \num{3.36e+7}.$ Thus, the calculated density of the 
-multiset-trie with respect to $|M|$ varies from~$0.03\%$ to~$0.95\%.$
+multiset-trie with respect to $|M|$ varies from~$\num{0.3e-3}$ to~$\num{9.5e-3}.$
 
@@
 we can deduce that at least 7-15 multisets were checked in order to find 
-submultiset or supermultiset, which is from $0.002\%$ to $0.15\%$ of the 
-multiset-trie and from $\num{1.9e-5}\%$ to $\num{4.5e-5}\%$ of the complete 
+submultiset or supermultiset, which is from $\num{0.02e-3}$ to $\num{1.5e-3}$ of the 
+multiset-trie and from $\num{1.9e-7}$ to $\num{4.5e-7}$ of the complete 
 multiset-trie.
@@
 varies from 40000 to 640000 multisets. Thus, the calculated density $d$ varies 
-from $\num{1.4e-13}\%$ to $\num{2.25e-12}\%.$ The density is much smaller than 
+from $\num{1.4e-15}$ to $\num{2.25e-14}.$ The density is much smaller than 
 in Experiment~\hyperref[s:exp1]{1}, because now we allow multisets to be stored 
@@
 to those maxima at least 300-1000 multisets were checked in order to find 
-submultiset or supermultiset, which is from $0.15\%$ to $0.75\%$ of the entire 
-multiset-trie and from $\num{1.1e-15}\%$ to $\num{3.4e-15}\%$ of the complete 
+submultiset or supermultiset, which is from $\num{1.5e-3}$ to $\num{7.5e-3}$ of the entire 
+multiset-trie and from $\num{1.1e-17}$ to $\num{3.4e-17}$ of the complete 
 multiset-trie. The percentage of visited multisets with respect to $|M|$ is 
@@
 $\sigma$ and $n$ are set to 12 and 5 respectively. The density varies from 
-$\num{1.0e-4}\%$ to $1.0\%.$ The number of visited nodes was chosen to be 
+$\num{1.0e-6}$ to $\num{1.0e-2}.$ The number of visited nodes was chosen to be 
 maximal for each value of particular density.
@@
 \textsc{submsetExistence} and \textsc{supermsetExistence} have the maximum 
-around $d\approx \num{7.0e-3}\%.$ The maximum is less than $0.03\%$ and 
-greater than $\num{1.4e-13}\%,$ which explains the behavior of multiset-trie in 
+around $d\approx \num{7.0e-5}.$ The maximum is less than $\num{0.3e-3}$ and 
+greater than $\num{1.4e-15},$ which explains the behavior of multiset-trie in 
 Experiment~\hyperref[s:exp1]{1} and Experiment~\hyperref[s:exp2]{2}. It is safe 

### 2018-09-12 tex/tex/s_operations.tex
algorithms adjusted and get-all algorithms added

@@
 \section{Multiset-trie operations} \label{c:operations}
+
 %
@@
 %
-\item \textsc{submsetExistence}($\mathcal{M}$, $m$): returns true if 
-there exists a $x\in M$ for a given $\mathcal{M}$ such that $x\subseteq m,$ 
-and returns false otherwise; 
+\item \textsc{submsetExistence}($\mathcal{M}$, $m$, $dev$): returns true if 
+there exists a $x\in M$ for a given $\mathcal{M}$ such that $x\subseteq m$ 
+and $| x[i] - m[i] |\leq dev$ for $1\leq i\leq \sigma$, and returns false otherwise; 
 %
-\item \textsc{supermsetExistence}($\mathcal{M}$, $m$): returns true if 
-there exists a $x\in M$ for a given $\mathcal{M}$ such that $x\supseteq m,$ and returns 
-false otherwise; 
+\item \textsc{supermsetExistence}($\mathcal{M}$, $m$, $dev$): returns true if 
+there exists a $x\in M$ for a given $\mathcal{M}$ such that $x\supseteq m$ 
+and $| x[i] - m[i] |\leq dev$ for $1\leq i\leq \sigma$, and returns false otherwise; 
 %
-\item \textsc{getAllSubmsets}($\mathcal{M}$, $m$): returns the set of multisets 
-$\{ x \in M : x\subseteq m \}$ for a given $\mathcal{M};$
+\item \textsc{getAllSubmsets}($\mathcal{M}$, $m$, $dev$): returns the set of multisets 
+$\{ x \in M : x\subseteq m \wedge |x[i]-m[i]|\leq dev \}$ for a given 
+$\mathcal{M},$ where $1\leq i\leq \sigma;$
 %
-\item \textsc{getAllSupermsets}($\mathcal{M}$, $m$): returns the set of multisets 
-$\{ x\in M : x\supseteq m \}$ for a given $\mathcal{M}.$
+\item \textsc{getAllSupermsets}($\mathcal{M}$, $m$, $dev$): returns the set of multisets 
+$\{ x\in M : x\supseteq m \wedge |x[i]-m[i]|\leq dev \}$ for a given $\mathcal{M},$
+where $1\leq i\leq \sigma.$
 %
@@
 \subsection{Sub-multiset existence} \label{s:subexists}
-The function \textsc{submsetExistence}($\mathcal{M}$, $m$) checks if there exists 
-a multiset $x$ in $\mathcal{M},$ that satisfies the condition $x\subseteq m.$ 
+The function \textsc{submsetExistence}($\mathcal{M},m,dev$) checks if there exists 
+a multiset $x$ in $\mathcal{M},$ that satisfies the condition $x\subseteq m$ and 
+$| x[i] - m[i] | \leq dev,$ where $1\leq i \leq \sigma.$ 
 The function starts with searching for an exact match $x=m$ in $\mathcal{M},$ 
@@
 (the largest) submultiset of $m$ in $\mathcal{M}$ by decreasing multiplicity of 
-elements in $m.$ At every level the function tries to proceed with the largest 
-possible multiplicity of an element that is provided by $m.$ However, when the 
-function reaches some level where it meets a \emph{Null} node and can not go 
-further using path provided by $m,$ it decreases the multiplicity of an element 
-that corresponds to a current level. Thus, the function can decrease multiplicity 
-of an element or eventually skip it in order to find the closest $x\subseteq m.$ 
-The function \textsc{submsetExistence} is presented in Algorithm~\ref{alg:subexists}.
+elements in $m.$ The parameter $dev$ is used to limit a maximal deviation of 
+multiplicity for a particular element in $x$ with respect to $m.$ 
+At every level the function tries to proceed with the largest possible multiplicity of 
+an element that is provided by $m.$ However, when the function reaches some level 
+where it meets a \emph{Null} node and can not go further using path provided by 
+$m,$ it decreases the multiplicity of an element 
+that corresponds to a current level with respect to the specified maximal deviation. 
+Thus, the function can decrease multiplicity of an element or eventually skip it in 
+order to find the closest $x\subseteq m.$ The function \textsc{submsetExistence} 
+is presented in Algorithm~\ref{alg:subexists}.
 
@@
 \begin{algorithmic}[1]
-\Function{\textsc{submsetExistence}}{$\mathcal{M}$, $m$}
+\Function{\textsc{submsetExistence}}{$\mathcal{M}, m, dev$}
 \State $currentNode \gets \mathcal{M}$
@@
 \EndIf
-\For {$i=m[1]$ down to $0$}
+\For {$i=m[1]$ down to max($0, m[1]-dev$)}
 \If {child $c_i$ of $currentNode$ is not \emph{Null}}
-\If {\textsc{submsetExistence}($c_i,$ $m[2:]$)}
+\If {\textsc{submsetExistence}($c_i,m[2:], dev$)}
 \State \Return True
@@
 \subsection{Super-multiset existence} \label{s:superexists}
-The function \textsc{supermsetExistence}($\mathcal{M},$ $m$) checks if there 
-exists supermultiset $x$ of a given multiset $m$ in $\mathcal{M}.$ By analogy 
+The function \textsc{supermsetExistence}($\mathcal{M},m,dev$) checks if there 
+exists supermultiset $x$ of a given multiset $m$ in $\mathcal{M},$ such that 
+condition $| x[i] - m[i] |\leq dev$ is satisfied, where $1\leq i \leq \sigma.$ By analogy 
 to the function \textsc{submsetExistence}, the function \textsc{supermsetExistence} 
@@
 possible multiplicity of an element that is provided by $m.$ However, when function 
-reaches some level where it meets a \emph{Null} node and can not go further 
-using path provided by $m,$ it increases the multiplicity of an element that 
-corresponds to a current level. Thus, the function \textsc{supermsetExistence} 
+reaches some level where it meets a \emph{Null} node and can not go further using 
+path provided by $m,$ it increases the multiplicity of an element that corresponds to a 
+current level according to the deviation parameter $dev.$ Thus, the function \textsc{supermsetExistence} 
 can increase multiplicity of an element up to $n-1,$ where $n$ is the degree of 
@@
 \begin{algorithmic}[1]
-\Function{\textsc{supermsetExistence}}{$\mathcal{M}$, $m$}
+\Function{\textsc{supermsetExistence}}{$\mathcal{M},m,dev$}
 \State $currentNode \gets \mathcal{M}$
@@
 \EndIf
-\For {$i=m[1]$ to $n-1$}
+\For {$i=m[1]$ to min($n-1, m[1] + dev$)}
 \If {child $c_i$ of $currentNode$ is not \emph{Null}}
-\If {\textsc{supermsetExistence}($c_i,$ $m[2:]$)}
+\If {\textsc{supermsetExistence}($c_i,$ $m[2:], dev$)}
 \State \Return True
@@
 continue procedure until all existing sub/supermultisets in $\mathcal{M}$ are 
-found and stored.
+found and stored. The functions \textsc{getAllSubmsets} and \textsc{getAllSupermsets} 
+are presented in Algorithm~\ref{alg:getallsub} and Algorithm~\ref{alg:getallsup} respectively.
+
+In order to record a multiset during multiset-trie traversal we use the variable $x$ in the algorithms.
+It is an empty array of size $\sigma$ where we store multiplicities of elements at 
+each level as we traverse the tree. The variable $result$ is used as a container for storing 
+submultisets of $m$ found during traversal. Both variables $x$ and $result$ are presented 
+as global, however they could be passed to the recursive function as parameters.
+
+\begin{algorithm}[h!]
+\caption{Function \textsc{getAllSubmsets}}
+\label{alg:getallsub}
+\begin{algorithmic}[1]
+\State $result \gets$ empty container
+\State $x \gets$ empty array of size $\sigma$
+\Function{\textsc{getAllSubmsets}}{$\mathcal{M}, m, dev$}
+\State $currentNode \gets \mathcal{M}$
+\If {$currentNode$ is \emph{accepting}}
+\State add copy of $x$ to $result$ 
+\EndIf
+\For {$i=m[1]$ down to max($0, m[1]-dev$)}
+\If {child $c_i$ of $currentNode$ is not \emph{Null}}
+\State $x[1]\gets i$
+\State \textsc{getAllSubmsets}($c_i,m[2:], dev$)
+\EndIf
+\EndFor
+\EndFunction
+\end{algorithmic}
+\end{algorithm}
+
+\begin{algorithm}[h!]
+\caption{Function \textsc{getAllSupermsets}}
+\label{alg:getallsup}
+\begin{algorithmic}[1]
+\State $result \gets$ empty container
+\State $x \gets$ empty array of size $\sigma$
+\Function{\textsc{getAllSupermsets}}{$\mathcal{M},m,dev$}
+\State $currentNode \gets \mathcal{M}$
+\If {$currentNode$ is \emph{accepting} }
+\State add copy of $x$ to $result$ 
+\EndIf
+\For {$i=m[1]$ to min($n-1, m[1] + dev$)}
+\If {child $c_i$ of $currentNode$ is not \emph{Null}}
+\State $x[1]\gets i$
+\textsc{getAllSupermsets}($c_i,$ $m[2:], dev$)
+\EndIf
+\EndFor
+\EndFunction
+\end{algorithmic}
+\end{algorithm}

### 2018-11-13 tex/tex/s_experiments.tex
Paper updates (introduction, experiments)

@@
 This section contains results of experiments that were performed on the multiset-trie 
-data structure. The implementation of multiset-trie is done in the \CC 
+data structure. In particular, we will test the functions: \textsc{submsetExistence}, 
+\textsc{supermsetExistence}, \textsc{getAllSubmsets} and 
+\textsc{getAllSupermsets}.
+
+The implementation of multiset-trie is done in the \CC 
 { programming} language. The current implementation uses only the standard 
@@
 
-In our experiments we will test the functions: \textsc{submsetExistence}, 
-\textsc{supermsetExistence}, \textsc{getAllSubmsets} and 
-\textsc{getAllSupermsets}. Performance of the functions will be measured by 
-the number of visited nodes in multiset-trie by the particular function. In 
-particular the performance is inversely proportional to the number of visited 
-nodes.
+%Performance of the functions will be measured by 
+%the number of visited nodes in multiset-trie by the particular function. In 
+%particular the performance is inversely proportional to the number of visited 
+%nodes.
 
@@
 the parameter $|M|$ will be varied to change the density in experiments. As we 
-mentioned in Chapter~\ref{c:description}, the mapping $\phi$ determines the 
+mentioned in Section~\ref{c:description}, the mapping $\phi$ determines the 
 correspondence of letters to levels in multiset-trie, i.e. it defines the ordering of 
@@
 In the next sections we will present the behavior of the multiset-trie data 
-structure depending on the selected parameters as well as the benchmark 
-of the multiset-trie against trie and hash map data structures. 
+structure depending on the selected parameters as well as the comparative 
+benchmark of the multiset-trie against B+ tree implementation of inverted index. 
 % outline of the experiment section
@@
 The difference is that the allowed multiplicity of an element is raised, i.e. 
-multisets are allowed to be stored in the data structure. 
+the data structure is populated with multisets. 
 %
@@
 to the set of consecutive integers using the function $\phi.$ The experiment 
-shows that the performance of the multiset-trie can be noticeably optimized using 
+shows that the performance of the multiset-trie is noticeably influenced by 
 different mappings $\phi.$ It also shows the usability of the multiset-trie in terms 
@@
 After all the experiments we present an empirical comparison of multiset-trie 
-data structure with inverted index. Inverted index is considered in two versions, 
-when it uses trie and hash map for storing multisets. In the comparison we use 
-two types of queries exact retrieval and searching for the first submultiset.
+data structure with B+ tree based inverted index. We use inverted index 
+to store and retrieve multisets in the same way as it is described in the paper by 
+Helmer and Moerkotte~\cite{Helmer2003} for sets. In the comparison we use 
+three types of queries exact, submultiset and supermultiset retrieval.
 
-\paragraph*{Data generation}
-We denote by \emph{input data} the data that is inserted into multiset-trie and 
-by \emph{test data} the data which will be used for queries in order to test the 
+\subsection*{Data generation}
+We denote by \emph{input data} the data that is used to fill the structure prior 
+to testing and by \emph{test data} the set of queries that is used to test the 
 performance of the functions.
@@
 In this experiment we demonstrate the performance of multiset-trie data structure 
-compared to inverted index. The data structures that were used for inverted index 
-implementation are trie and hash map. Both indexes were implemented in~\CC{ to} 
-avoid any optimization or degradation of performance caused by different programming 
-language implementation.
+compared to B+ tree based inverted index. The data structure is also implemented 
+in~\CC{ to} avoid any optimization or degradation of performance caused by different 
+programming language implementation.
 
-During tests all three data structures used the same interface for benchmark that operates 
+During tests both data structures used the same interface for benchmark that operates 
 with files as its input and output. Each data structure was loaded with the same input data 
-and was queried with the same test data. The input data consists of multisets that are 
-generated at random with respect to $\sigma = 15$ and $n = 3.$ All together the input dataset 
-has 300000 randomly generated multisets. The test dataset was constructed with the same 
-$\sigma$ and $n$ having cardinality 100000 multisets.
+and was queried with the same test data. The input and test data were generated artificially 
+according to parameters $\sigma$ and $n.$ The combinations of parameters is as follows in the 
+Table~\ref{t:benchmark}.
 
-First, we present the results of performing an exact search query on each data structure. 
-The results on figure~\ref{fig:e5ex} show that hash map performs the best, which is not 
-surprising, since its access complexity is $O(1).$ However, note that performance of 
-multiset-trie does not depend on the multiset being searched. The trie version of the 
-inverted index performs the worst in this experiment, because its performance is dependent 
-on the multiset size.
+\begin{table}[h]
+\center
+\begin{tabular}{|c|c|}
+\hline
+$\sigma$ & $n$ \\
+\hline
+5		& 1\\
+\hline
+30	& 1 \\
+\hline
+5		& 3 \\
+\hline
+15	& 3 \\
+\hline
+30	& 3 \\
+\hline
+10	& 10 \\
+\hline
+\end{tabular}
+\caption{Configuration of $\sigma$ and $n$ in benchmark.}
+\label{t:benchmark}
+\end{table}
 
-\begin{figure}
+We tested all the three types of query on all of the configurations from the Table~\ref{t:benchmark} which 
+results in 18 experiments total, i.e. 6 experiments per query type. In each of the experiment we measured an 
+average time consumed by the data structure to process the query.
+
+The results of the exact search, submultiset search and supermultiset search experiments are shown in the tables 
+Table~\ref{t:res_ex}, Table~\ref{t:res_sub} and Table~\ref{t:res_sup} respectively .
+
+\begin{table}[h]
 \center
-\includegraphics[width=.4\textwidth]{exact.pdf}
-\caption{Experiment 5, exact search.}
-\label{fig:e5ex}
-\end{figure}
+\begin{tabular}{|c|c|c|c|}
+\hline
+$\sigma$ & $n$ & Multiset-trie ($\mu s$) & Inverted index ($\mu s$) \\
+\hline
+5		& 1 & 3.45 & 17782.35\\
+\hline
+30	& 1 & 4.18 & 24865.93\\
+\hline
+5		& 3 & 2.20 & 1508.81\\
+\hline
+15	& 3 & 4.39 & 2146.36\\
+\hline
+30	& 3 & 10.67 & 3639.97\\
+\hline
+10	& 10 & 6.93 & 384.05\\
+\hline
+\end{tabular}
+\caption{Exact search.}
+\label{t:res_ex}
+\end{table}
 
-The results of the submultiset query are presented on figure~\ref{fig:e5subm}. Here, each searching 
-query was terminated either when the first submultiset is found or all potential submultisets were 
-checked for submultiset inclusion. Multiset-trie outperforms both trie and hash map inverted 
-indexes by a major factor due to its design. Multiset-trie operates multiplicities on a node level 
-while trie does it on a tree level. In the trie one should traverse the tree in order to see if the element 
-of some multiplicity is inside, while in multiset-trie one just has to check if the node has appropriate 
-child. In the case of hash map we can operate on as low as the whole multiset level, which is of course 
-inconvenient when it comes to multiset containment queries.
+\begin{table}[h]
+\center
+\begin{tabular}{|c|c|c|c|}
+\hline
+$\sigma$ & $n$ & Multiset-trie ($\mu s$) & Inverted index ($\mu s$) \\
+\hline
+5		& 1			& 8.96 & 73500.84\\
+\hline
+30	& 1			& 17.33 & 547572.74\\
+\hline
+5		& 3			& 117.95 & 162360.43\\
+\hline
+15	& 3			& 20.74 & 443321.39\\
+\hline
+30	& 3 		& 23.75 & 947706.14\\
+\hline
+10	& 10		& 55.59 & 466022.68\\
+\hline
+\end{tabular}
+\caption{Submultiset search.}
+\label{t:res_sub}
+\end{table}
 
-Note that multiset-trie is sparse in this experiment, however, the multisets that do not satisfy the query 
-are discarded much quicker than in trie or hash map.
+\begin{table}[h]
+\center
+\begin{tabular}{|c|c|c|c|}
+\hline
+$\sigma$ & $n$ & Multiset-trie ($\mu s$) & Inverted index ($\mu s$)\\
+\hline
+5		& 1 		& 10.63 & 63073.86\\
+\hline
+30	& 1 		& 14.65 & 449251.68\\
+\hline
+5		& 3 		& 171.04 & 163256.77\\
+\hline
+15	& 3 		& 43.42 & 425733.80\\
+\hline
+30	& 3 		& 22.06 & 729831.34\\
+\hline
+10	& 10 	& 58.32 & 373784.81\\
+\hline
+\end{tabular}
+\caption{Supermultiset search.}
+\label{t:res_sup}
+\end{table}
 
-\begin{figure}
-\center
-\includegraphics[width=.4\textwidth]{submset.pdf}
-\caption{Experiment 5, first submultiset search.}
-\label{fig:e5subm}
-\end{figure}
+We can see that multiset-trie outperforms inverted index in all of the experiments by up to 4 orders of magnitude. 
+In exact search multiset-trie has to traverse only up to $\sigma+1$ nodes in order to get query result. It is seen 
+from results that with increase of $\sigma$ the processing time for multiset-tire also increases. Interestingly, the 
+multiplicity also affects the time of processing, however, this happens passively. Multiplicity or degree of a node $n$ 
+defines what multisets can be stored in multiset-trie. Thus, it affects the structure and density of the multiset-trie.
+
+As for the inverted index, it must fetch all the postings according to each element in a test multiset and then intersect 
+the results in order to answer the query. This is a more time consuming operation than a simple tree traversal, which we 
+can see from results. Postings are of course filtered on-the-fly to reduce the cost of intersection, however these are still 
+additional operations to be performed. Note, that inverted file will also access at most $\sigma$ postings during search, 
+but in addition to access some extra processing is required.
+
+The submultiset and supermultiset basically do not change to much in the case of inverted file, because the algorithm consists 
+of the same steps: the postings are fetched according to elements of the test multiset, then they are filtered on-the-fly 
+according to the query type, and at the end the postings either has to be united or intersected. So, besides a traversal of at most 
+$\sigma$ keys of the index additional processing is required. Note, that every key is mapped to a list of postings which 
+can be large as index grows.
+
+In the case of multiset-trie only a traversal of the tree is required which is much faster than processing of postings as we can 
+see from results. In the worst case the whole tree is traversed, but so is for inverted file.

### 2018-11-13 tex/tex/s_intro.tex
algorithms adjusted and get-all algorithms added
Paper updates (introduction, experiments)

@@
 \section{Introduction} \label{c:introduction}
-[Will be rewritten]
 %
 % some intro
-During recent years the popularity of digital data has increased. All sorts of 
-information such as text, audio and video can now be accessed by searching 
-\emph{information retrieval systems}. Information retrieval (IR) is the science of 
-searching information units in a collection of data. Most commonly IR systems 
-are used for searching a text-based content such as text documents in a database. 
-Such IR systems are called \emph{full-text search systems}.
+During recent years the popularity of digital data has increased. Huge amount of diverse data is collected in databases and 
+processed regularly. Data processing usually consists of operations such as filtering, sorting and aggregation. This requires 
+indexing for optimization of data processing especially when it is repetitive. Data often includes duplicates or multivalued 
+attributes that are complicated to process efficiently when sophisticated queries are required.
 
-Full-text search techniques can be applied directly on the database. However, 
-it is a very expensive procedure in terms of running time complexity, because it 
-requires a frequent accesses to the database. In order to reduce this number, indexes 
-were invented. Indexes narrow down the search using pre-generated meta data 
-constructed from the data in the database. Furthermore the meta data can be 
-organized in a data structure that would provide fast retrieval of data according 
-to search queries.
+Multisets or bags are not well supported by storage systems for processing. Usually multisets have to be stored in a 
+simpler format that does not allow a direct analysis of the data. Such storage results in a more complex processing, 
+since multisets have to be re-constructed or pre-processed first in order to run the query. It also becomes impossible
+to execute containment queries on multisets.
 
-In IR most systems use the concept of an inverted index to achieve full-text 
-indexing of a database. Inverted index consists of two parts: \emph{postings} 
-and \emph{dictionary}, a search structure that is used to locate a specific entries 
-in a posting. A posting entry can be created on different levels depending on data 
-that needs to be indexed. Most common for full-text indexes are document level 
-entries. In this context a posting is defined to be a list of identifiers or keys that are 
-further used to locate a specific document in the database \cite{zobel1992efficient}.
+Currently the most efficient ways for multisets storage and processing involve inverted files, signature trees and B+ tree.
+Such solutions provide key-value look-up only, which does not allow to perform containment queries on multisets directly and efficiently.
+Additional processing is required such as intersection or union of obtained partial results from key-value look-ups. The post 
+processing of partial results can be optimized, however, it is still not as efficient and usually requires additional encoded 
+signatures or other types of identifiers to speed-up the process.
 
-An index is a search structure that is used to process user queries. The query 
-can be processed in different ways according to the \emph{retrieval model}. The 
-retrieval model that will be discussed in the thesis is the \emph{boolean retrieval 
-model} that views each document in a database as a set of words. The document 
-itself is an information unit a retrieval system is built over. In our case an 
-information unit is defined to be a textual document.
-
-The boolean retrieval model is based on set and multiset theory together with 
-boolean algebra. \emph{Set and multiset containment operations} are used to 
-derive the similarities between objects, and consequently make decisions on their 
-association. Thus, the multiset containment operations allow us to 
-search objects not only with exact queries but also to retrieve the most relevant 
-set of results that satisfy a given search query \cite{baeza1999modern,manning2008introduction,zobel2006inverted}.
-
-% research topic
-% walking through the new terms and so on
-% brief overview of the literature (relations to other work etc.)
-%
-
-The dictionary (search structure) of the inverted index can be organized in different ways in 
-order to meet the required types of queries and specification of data. For example, 
-it can be organized as a search tree, hash map, array, heap, linked list, etc. In our 
-research we will be focused on search trees. The most efficient search tree index 
-nowadays is the Generalized Search Tree (GiST). Its flexibility stems from combining 
-functionality of B+-trees, R-trees and RD-trees. GiST further extends their functionality 
-providing support for a variety of data types together with the nearest-neighbor 
-search \cite{broder2006indexing,hellerstein1995generalized,kornacker1999high}.
+% --------------------------------------------------------------------------------------------------------------------------------------------------------------------
 
@@
 %
-The proposed data structure \emph{multiset-trie} can be used as an alternative 
-implementation of the search structure in an inverted index. It is an extension of 
-the \emph{set-trie} data structure proposed by Savnik \cite{savnik2013index}. 
+In this paper we propose a data structure \emph{multiset-trie} that is designed for storage and processing of multisets.
+It is an extension of the \emph{set-trie} data structure proposed by Savnik \cite{savnik2013index}. 
 Set-trie is a trie based data structure that is used for storing and fast retrieval of 
 objects represented as sets. The set-trie provides the nearest-neighbor search by 
-implementing methods that perform set-containment queries. Multiset-trie extends 
-the abilities of set-trie and provides support for storing and retrieving objects 
-that can be represented as both sets and multisets. It also implements 
-multiset-containment methods together with the basic tree methods such as search, 
-deletion and insertion.
+implementing methods that perform set-containment queries. Multiset-trie generalizes 
+set-trie and provides support for storing and retrieving objects that can be represented 
+as both sets and multisets.
 
-The multiset-trie is an $n$-ary tree based data structure with properties similar to 
-those of a trie. 
-This particular combination allows us to associate multisets with a collection 
-of nodes in a tree. Every node represents a symbol with particular multiplicity. 
+The multiset-trie is an $n$-ary tree based data structure with properties similar to those of a trie. 
+This particular combination allows us to associate multisets with a collection of nodes in a tree. 
+Every node represents an element of a multiset with particular multiplicity. 
 %
 Multiset-trie is a kind of search tree. Similarly to a trie, it uses common prefixes 
-to narrow down the search. Unlike the compact prefix tree, Patricia, the multiset-trie 
+to narrow down the search. It also implements the basic tree methods such as 
+search, deletion and insertion. Unlike the compact prefix tree, Patricia, the multiset-trie 
 does not provide the ability to compress a path. However, the absence of path 
@@
 The multiset-trie is designed for efficient execution of the multiset containment 
-operations. In particular, it supports the operations \textsc{submsetExistence}, 
-\textsc{supermsetExistence}, \textsc{getAllSubmsets} and \textsc{getAllSupermsets}. 
-The so-called "existence" queries implement the nearest-neighbor search queries. The 
-functions \textsc{submsetExistence} and \textsc{supermsetExistence} search for 
-the closest submultiset and supermultiset in the multiset-trie respectively and return 
-an answer whether such a multiset exists in the data structure. The so-called 
-"getAll" functions act in the same way as "existence" functions, but they do not 
-terminate once they have reached the desired multiset. Alternatively, these functions 
-store the results and continue until all the multisets that satisfy the query are retrieved.
+operations. It handles multisets directly by having access to each of the element 
+without the need to reconstruct them for processing, which allows fast retrieval and 
+containment queries. In particular, it supports submultiset and supermultiset queries.
+The operations allow to find and retrieve the closest submultisets or supermultisets 
+as well as to find and retrieve all of them.
+%
+Since multiset is a generalized version of sets that allows duplicates multiset-trie can be 
+used for storing and processing sets supporting the same set of operations and providing 
+the same efficiency.
+
+% --------------------------------------------------------------------------------------------------------------------------------------------------------------------
+
+% key results
+Empirical studies on real data show, that multiset-trie is sensitive to the context, which can 
+be further used to optimize data structure for particular data. Moreover, knowing the data 
+it is possible to estimate the overall performance of the multiset-trie both time and space 
+related using our mathematical theory that describes multiset-trie.
+%
+The comparative study shows how efficient multiset-trie is by outperforming inverted file in 
+both exact and containment queries by up to 4 orders of magnitude in the time consumed by 
+query processing.
+
+% --------------------------------------------------------------------------------------------------------------------------------------------------------------------
 
@@
 %
-Let us now present the organization of the thesis. 
-In the following Chapter~\ref{c:description} we present the description of the 
+In the following Section~\ref{c:description} we present the organization of 
 multiset-trie data structure. The representation of multisets in multiset-trie is 
-explained in detail. The organization of the data structure is also presented 
-graphically.
+explained in detail.
 %
-In Chapter~\ref{c:operations} we present operations that multiset-trie currently 
-supports. The multiset containment functions \textsc{submsetExistence}, 
-\textsc{supermsetExistence}, \textsc{getAllSubmsets} and \textsc{getAllSupermsets} 
-are presented together with the basic search tree functions such as \textsc{insert}, 
-\textsc{delete} and \textsc{search}. The algorithms in pseudo code are presented 
-as well. 
+In Section~\ref{c:operations} we present operations that multiset-trie currently 
+supports. Those are the multiset containment and the basic search tree functions. 
+The algorithms in pseudo code are presented as well. 
 %
 The description of multiset-trie functions and procedures is followed by the 
-mathematical analysis of their complexity in Chapter~\ref{c:analysis}. 
-In this analysis, we make an assumption that multisets are constructed uniformly at 
-random and are parametrized by several parameters, such as multiplicity and the alphabet 
-$\Sigma.$ By using probabilistic tools we describe time complexity of the algorithms 
-and space complexity of the structure.
+mathematical analysis of their complexity in Section~\ref{c:analysis}. 
+We assume that multisets are constructed uniformly at random and are parametrized 
+by several parameters, such as multiplicity and the alphabet. By using probabilistic 
+tools we describe time complexity of the algorithms and space complexity of the structure.
 %
-Further, in Chapter~\ref{c:experiments} we present an empirical study of the 
+Further, in Section~\ref{c:experiments} we present an empirical study of the 
 multiset-trie. Artificially generated as well as real-world data sets are used in experiments. 
 The experiments are dedicated to testing the performance of the data structure while 
-varying selected parameters. The experiments also show some methods for 
-optimizing a multiset-trie.
+varying selected parameters. The experiments also highlight the methods for optimizing a multiset-trie.
 %
-The Chapter~\ref{c:relwork} presents related work. The connection to the set-trie 
+The Section~\ref{c:relwork} presents related work. The connection to the set-trie 
 data structure \cite{savnik2013index} is discussed more explicitly. We also relate 
@@
 %
-Finally, in Chapter~\ref{c:conclusions} our conclusion about the multiset-trie 
+Finally, in Section~\ref{c:conclusions} our conclusion about the multiset-trie 
 data structure and discussion of future work are presented.

### 2019-02-18 tex/article.tex
article changes

@@
 \begin{abstract}
-[Will be rewritten]
-In this paper we will present the multiset-trie, a new data structure that operates 
-on objects represented as multisets. The multiset-trie is a search-tree-based 
-data structure with the properties similar to those of a trie. It implements all standard search 
-tree operations together with the special multiset containment operations. Multiset 
-containment operations supported by the multiset-trie are submultiset and supermultiset. 
-These operations are used for implementation of different queries that can be performed 
-on multisets in a multiset-trie. One of the most important queries is the search of 
-the nearest neighbor given an input object. The nearest neighbor search of a 
-multiset-trie makes it a good alternative for the index data structures that are 
-used in information retrieval systems. In particular, our research is focused on the 
-application of the multiset-trie to full-text search systems. 
+This work proposes a new data structure \emph{multiset-trie} that is designed for storage 
+and efficient processing of a set of multisets. Moreover, multiset-trie can operate on 
+a set of sets without efficiency loss.
+%
+The multiset-trie is a search-tree-based data structure with the properties similar to those of 
+a trie. It implements all standard search tree operations together with the multiset containment 
+operations such as submultiset and supermultiset. Containment operations allow efficient search and 
+retrieval of the individual multisets and also collections of multsets. In particular, it can find and retrieve 
+closest submultisets or supermultisets as well as all submultisets or supermultisets that are parameterized
+by user input.
+%
+Multiset-trie is compared to modern solutions that provide containment operations on multisets. 
+The study shows how efficient multiset-trie is by outperforming inverted file in both equality and 
+containment queries by up to 4 orders of magnitude in query execution time.
 \end{abstract}
-\keywords{multiset, bag-of-words, trie, multiset-trie, information retrieval, inverted index, full-text search}
+\keywords{multiset-trie, multiset, bag-of-words, trie, data structure, containment operations, inverted file}
 %\subclass{68P05, 68P20, 68Q87}

### 2019-02-18 tex/tex/s_data_structure.tex
article changes

@@
 \begin{eqnarray*}
-(L_1,root) \rightarrow (L_2,c_0) \rightarrow (LL,c_0) & \textrm{equivalent to} & \{ 1^0, 2^0 \} = \emptyset \\
-(L_1,root) \rightarrow (L_2,c_0) \rightarrow (LL,c_1) & \textrm{equivalent to} & \{ 1^0, 2^1 \} = \{ 2 \} \\
-(L_1,root) \rightarrow (L_2,c_0) \rightarrow (LL,c_2) & \textrm{equivalent to} & \{ 1^0, 2^2 \} = \{ 2,2 \} \\
-(L_1,root) \rightarrow (L_2,c_1) \rightarrow (LL,c_1) & \textrm{equivalent to} & \{ 1^1, 2^1 \} = \{ 1,2 \} \\
-(L_1,root) \rightarrow (L_2,c_1) \rightarrow (LL,c_2) & \textrm{equivalent to} & \{ 1^1, 2^2 \} = \{ 1,2,2 \} \\
-(L_1,root) \rightarrow (L_2,c_2) \rightarrow (LL,c_1) & \textrm{equivalent to} & \{ 1^2, 2^1 \} = \{ 1,1,2 \} 
+(L_1,root) \rightarrow (L_2,c_0) \rightarrow (LL,c_0) & = & \{ 1^0, 2^0 \} = \emptyset \\
+(L_1,root) \rightarrow (L_2,c_0) \rightarrow (LL,c_1) & = & \{ 1^0, 2^1 \} = \{ 2 \} \\
+(L_1,root) \rightarrow (L_2,c_0) \rightarrow (LL,c_2) & = & \{ 1^0, 2^2 \} = \{ 2,2 \} \\
+(L_1,root) \rightarrow (L_2,c_1) \rightarrow (LL,c_1) & = & \{ 1^1, 2^1 \} = \{ 1,2 \} \\
+(L_1,root) \rightarrow (L_2,c_1) \rightarrow (LL,c_2) & = & \{ 1^1, 2^2 \} = \{ 1,2,2 \} \\
+(L_1,root) \rightarrow (L_2,c_2) \rightarrow (LL,c_1) & = & \{ 1^2, 2^1 \} = \{ 1,1,2 \} 
 \end{eqnarray*}

### 2019-02-18 tex/tex/s_intro.tex
article changes

@@
 %
-% some intro
-During recent years the popularity of digital data has increased. Huge amount of diverse data is collected in databases and 
-processed regularly. Data processing usually consists of operations such as filtering, sorting and aggregation. This requires 
-indexing for optimization of data processing especially when it is repetitive. Data often includes duplicates or multivalued 
-attributes that are complicated to process efficiently when sophisticated queries are required.
+% general intro into problem
+%
+A multiset is a data structure that represents a collection of elements. It generalizes a set data structure 
+by allowing duplicate elements in a collection. 
+%
+Data with duplicate elements is encountered in a wide variety of domains and applications. For example, 
+in databases or data mining. In particular, processing of multisets has attracted significant attention in 
+recent years. 
+%
+There are discussions about storage and representation of multi-valued attributes in databases, organization 
+of data with shared content, containment joins, efficient algorithms for containment queries, compression 
+of multisets, etc~\cite{bouros2016set,gripon2012compressing,ross2004symmetric,steinruecken2015compressing}.
 
-Multisets or bags are not well supported by storage systems for processing. Usually multisets have to be stored in a 
-simpler format that does not allow a direct analysis of the data. Such storage results in a more complex processing, 
-since multisets have to be re-constructed or pre-processed first in order to run the query. It also becomes impossible
-to execute containment queries on multisets.
-
-Currently the most efficient ways for multisets storage and processing involve inverted files, signature trees and B+ tree.
-Such solutions provide key-value look-up only, which does not allow to perform containment queries on multisets directly and efficiently.
-Additional processing is required such as intersection or union of obtained partial results from key-value look-ups. The post 
-processing of partial results can be optimized, however, it is still not as efficient and usually requires additional encoded 
-signatures or other types of identifiers to speed-up the process.
+Multiset data structure is usually implemented as a search tree and its elements can be accessed by a key. 
+%
+This approach is good when individual elements of a multiset are of interest. However, it is not as optimal for operations 
+that use containment relation between multisets, also known as containment queries. 
+%
+Containment relation allows us to operate directly with the collections that are associated with a multiset such as a set 
+of submultiset and a set of supermultisets.
+%
+Currently, the solution to searching or retrieval of a set of multisets using containment queries involves inverted file, 
+signature tree or B+ tree~\cite{Helmer2003}. The solution provides a key-value look-up for elements of a multiset implying that 
+containment operations are still element-wise.
 
@@
 %
-In this paper we propose a data structure \emph{multiset-trie} that is designed for storage and processing of multisets.
-It is an extension of the \emph{set-trie} data structure proposed by Savnik \cite{savnik2013index}. 
-Set-trie is a trie based data structure that is used for storing and fast retrieval of 
-objects represented as sets. The set-trie provides the nearest-neighbor search by 
-implementing methods that perform set-containment queries. Multiset-trie generalizes 
-set-trie and provides support for storing and retrieving objects that can be represented 
-as both sets and multisets.
+In this paper we propose a data structure \emph{multiset-trie} that is designed for storage and 
+processing of a set of multisets. It extends the \emph{set-trie} data structure proposed by 
+Savnik~\cite{savnik2013index} that was designed for storage and processing of a set of sets. 
+%
+Set-trie is a trie based data structure that provides a fast search and retrieval of sets and implements 
+set-containment operations. 
+%
+Multiset-trie generalizes set-trie and can work with a set of sets (as set-trie) or with 
+a set of multisets providing efficient multiset-containment operations.
 
+% intro into mstrie
 The multiset-trie is an $n$-ary tree based data structure with properties similar to those of a trie. 
-This particular combination allows us to associate multisets with a collection of nodes in a tree. 
-Every node represents an element of a multiset with particular multiplicity. 
+Multisets are associated with a collection of nodes in a tree such that every node represents an element 
+of a multiset with particular multiplicity. 
 %
-Multiset-trie is a kind of search tree. Similarly to a trie, it uses common prefixes 
-to narrow down the search. It also implements the basic tree methods such as 
-search, deletion and insertion. Unlike the compact prefix tree, Patricia, the multiset-trie 
-does not provide the ability to compress a path. However, the absence of path 
+Multiset-trie is a kind of search tree. Similarly to a trie, it uses common prefixes for shared data representation.
+Unlike the compact prefix tree, the multiset-trie does not support path compression. However, the absence of path 
 compression makes the multiset-trie a perfectly height-balanced tree.
-
-The multiset-trie is designed for efficient execution of the multiset containment 
-operations. It handles multisets directly by having access to each of the element 
+%
+The multiset-trie handles multisets directly by having access to each of the element 
 without the need to reconstruct them for processing, which allows fast retrieval and 
-containment queries. In particular, it supports submultiset and supermultiset queries.
+containment operations. In particular, it supports submultiset and supermultiset queries.
 The operations allow to find and retrieve the closest submultisets or supermultisets 
 as well as to find and retrieve all of them.
-%
-Since multiset is a generalized version of sets that allows duplicates multiset-trie can be 
-used for storing and processing sets supporting the same set of operations and providing 
-the same efficiency.
 
@@
 The comparative study shows how efficient multiset-trie is by outperforming inverted file in 
-both exact and containment queries by up to 4 orders of magnitude in the time consumed by 
-query processing.
+both equality and containment queries by up to 4 orders of magnitude in the time consumed by 
+query.
 
@@
 In the following Section~\ref{c:description} we present the organization of 
-multiset-trie data structure. The representation of multisets in multiset-trie is 
-explained in detail.
+multiset-trie data structure in detail.
 %
-In Section~\ref{c:operations} we present operations that multiset-trie currently 
-supports. Those are the multiset containment and the basic search tree functions. 
+Next, in Section~\ref{c:operations} we present operations that multiset-trie currently 
+supports. This includes multiset containment and the basic search tree functions. 
 The algorithms in pseudo code are presented as well. 
@@
 mathematical analysis of their complexity in Section~\ref{c:analysis}. 
-We assume that multisets are constructed uniformly at random and are parametrized 
-by several parameters, such as multiplicity and the alphabet. By using probabilistic 
-tools we describe time complexity of the algorithms and space complexity of the structure.
+The main assumption is that multisets are constructed uniformly at random 
+with bounded cardinality. By using probabilistic tools we describe time complexity of 
+the algorithms and space complexity of the structure.
 %
 Further, in Section~\ref{c:experiments} we present an empirical study of the 
-multiset-trie. Artificially generated as well as real-world data sets are used in experiments. 
-The experiments are dedicated to testing the performance of the data structure while 
-varying selected parameters. The experiments also highlight the methods for optimizing a multiset-trie.
+multiset-trie. Synthetic and real-world data sets are used in experiments to test the performance 
+of the data structure. The experiments also highlight the methods for optimizing a multiset-trie.
 %
-The Section~\ref{c:relwork} presents related work. The connection to the set-trie 
-data structure \cite{savnik2013index} is discussed more explicitly. We also relate 
-the multiset-trie to the information retrieval systems. In particular, we refer to the 
-inverted index data structure and discuss how the multiset-trie can be used as a 
-database index.
+The related work is reviewed in Section~\ref{c:relwork}. 
 %
-Finally, in Section~\ref{c:conclusions} our conclusion about the multiset-trie 
-data structure and discussion of future work are presented.
+Finally, we conclude with the discussion of future work in Section~\ref{c:conclusions}.
+%

### 2019-02-18 tex/tex/s_math.tex
article changes

@@
 \begin{lemma}\label{l:prob-node-existence}
-
 Any potential node on a fixed level $i,$ where $i\in \{ 1,2,\ldots, \sigma +1 \}$ exists, %independently of others,
@@
 \[
-n^{\sigma + 1 - i}
+n^{\sigma + 1 - i}.
 \]
@@
 The claim follows by taking the complement probability of the above result. 
+\hspace*{\fill}\qed
 %
@@
 which completes the proof.
+\hspace*{\fill}\qed
 \end{proof}
@@
 property~(\ref{eq:prop_exp-pgf}) we complete the proof.
+\hspace*{\fill}\qed
 \end{proof}
@@
 \end{eqnarray*}
+\hspace*{\fill}\qed
 \end{proof}
@@
 supermultiset).
+\hspace*{\fill}\qed
 \end{proof}

### 2019-02-28 tex/tex/s_intro.tex
introduction changes

@@
 
-Multiset data structure is usually implemented as a search tree and its elements can be accessed by a key. 
 %
-This approach is good when individual elements of a multiset are of interest. However, it is not as optimal for operations 
-that use containment relation between multisets, also known as containment queries. 
+% problem
+%
+The problem we would like to address in this paper is efficient execution of containment queries on a set of multisets.
+The modern implementations of data structure for a set or a multiset is usually implemented as a search tree and 
+its elements can be accessed by a key. 
+%
+This approach is good when individual elements of a set or a multiset are of interest. However, it is not as optimal for operations 
+that use containment relation between multisets. 
 %
@@
 Currently, the solution to searching or retrieval of a set of multisets using containment queries involves inverted file, 
-signature tree or B+ tree~\cite{Helmer2003}. The solution provides a key-value look-up for elements of a multiset implying that 
-containment operations are still element-wise.
+signature tree or B+ tree~\cite{Helmer2003}. These solutions provide a key-value look-up for elements of a multiset implying that 
+containment operations are still element-wise. In order to improve the efficiency of operations we organize a set of 
+multisets in a different way described further below.
 
@@
 In this paper we propose a data structure \emph{multiset-trie} that is designed for storage and 
-processing of a set of multisets. It extends the \emph{set-trie} data structure proposed by 
-Savnik~\cite{savnik2013index} that was designed for storage and processing of a set of sets. 
+processing of a finite bounded set of multisets. It extends the \emph{set-trie} data structure proposed 
+by Savnik~\cite{savnik2013index} that was designed for storage and processing of a finite bounded (?or 
+unbounded?) set of sets. 
 %
@@
 Multisets are associated with a collection of nodes in a tree such that every node represents an element 
-of a multiset with particular multiplicity. 
+of a multiset with particular multiplicity bounded by a node degree $n.$
 %
@@
 Unlike the compact prefix tree, the multiset-trie does not support path compression. However, the absence of path 
-compression makes the multiset-trie a perfectly height-balanced tree.
+compression makes the multiset-trie a perfectly height-balanced tree. Moreover, when multiset-trie is full it forms 
+a complete $n$-ary tree.
 %

### 2019-03-04 tex/tex/s_experiments.tex
article changes
changed s_experiments.tex (experiment 5).

@@
 data structure. In particular, we will test the functions: \textsc{submsetExistence}, 
-\textsc{supermsetExistence}, \textsc{getAllSubmsets} and 
-\textsc{getAllSupermsets}.
+\textsc{supermsetExistence}, \textsc{getAllSubmsets} and \textsc{getAllSupermsets}.
 
-The implementation of multiset-trie is done in the \CC 
-{ programming} language. The current implementation uses only the standard 
-library of \CC14 version of the standard and has a command line interface. The implementation 
-of the program was optimized for testing and therefore, the program operates 
-with files, in order to process queries. After processing all the queries 
-the results are stored in files for further analysis.
+The implementation of multiset-trie is done in the \CC { programming} language. 
+The current implementation uses only the standard library of \CC14 version of the 
+standard and has a command line interface~\cite{akulich2019mstrie}. The implementation of the program was 
+optimized for testing and therefore, the program operates with files, in order to 
+process queries. After processing all the queries the results are stored in files for further analysis.
 
@@
 \subsection{Experiment 5} \label{s:exp5}
-In this experiment we demonstrate the performance of multiset-trie data structure 
-compared to B+ tree based inverted index. The data structure is also implemented 
-in~\CC{ to} avoid any optimization or degradation of performance caused by different 
-programming language implementation.
+In this experiment, we demonstrate the performance of multiset-trie data structure compared to the inverted index based on the B+ tree.
+Both data structures are implemented in the programming language \CC, providing in this way an experimental setup for a fair comparison  [ref.
+to impl].
 
-During tests both data structures used the same interface for benchmark that operates 
-with files as its input and output. Each data structure was loaded with the same input data 
-and was queried with the same test data. The input and test data were generated artificially 
-according to parameters $\sigma$ and $n.$ The combinations of parameters is as follows in the 
-Table~\ref{t:benchmark}.
+%% more details about implementation of the inverted files
+((Presentation of the inverted index implementation? What about the smallest postings first?))
+
+The experiment uses the input data for the construction of the given data structure and the test data for the execution of the operations on the given data structure.
+The input data comprises a set of randomly generated multisets used for the construction of a data structure.
+The test data includes the set of multisets together with the operations that are evaluated.
+The input and test data were generated with respect to parameters $\sigma$ and $n$ as presented in Table~\ref{t:benchmark}.
+
+((Is there anything special to be said about the artificially generated data?))
 
@@
 
-We tested all the three types of query on all of the configurations from the Table~\ref{t:benchmark} which 
-results in 18 experiments total, i.e. 6 experiments per query type. In each of the experiment we measured an 
-average time consumed by the data structure to process the query.
-
-The results of the exact search, submultiset search and supermultiset search experiments are shown in the tables 
-Table~\ref{t:res_ex}, Table~\ref{t:res_sub} and Table~\ref{t:res_sup} respectively .
+We tested all three types of a query on all of the configurations from Table 1 which results in 18 experiments total, i.e., 6 experiments per query type.
+In each of the experiment, we measured an average time consumed by the data structure to process the query.
+The results of the exact search, sub-multiset and super-multiset search experiments are presented in tables Table~\ref{t:res_ex}, Table~\ref{t:res_sub} and Table~\ref{t:res_sup}, respectively.
 
@@
 \center
-\begin{tabular}{|c|c|c|c|}
+\begin{tabular}{|r|r|r|r|}
 \hline
-$\sigma$ & $n$ & Multiset-trie ($\mu s$) & Inverted index ($\mu s$) \\
+\multicolumn{1}{|c|}{$\sigma$} & 
+\multicolumn{1}{c|}{$n$} & 
+\multicolumn{1}{c|}{Multiset-trie ($\mu s$)} & 
+\multicolumn{1}{c|}{Inverted index ($\mu s$)} \\
 \hline
@@
 \end{tabular}
-\caption{Submultiset search.}
+\caption{Sub-multiset search.}
 \label{t:res_sub}
@@
 \center
-\begin{tabular}{|c|c|c|c|}
+\begin{tabular}{|r|r|r|r|}
 \hline
-$\sigma$ & $n$ & Multiset-trie ($\mu s$) & Inverted index ($\mu s$)\\
+\multicolumn{1}{|c|}{$\sigma$} & 
+\multicolumn{1}{c|}{$n$} & 
+\multicolumn{1}{c|}{Multiset-trie ($\mu s$)} & 
+\multicolumn{1}{c|}{Inverted index ($\mu s$)} \\
 \hline
@@
 \end{tabular}
-\caption{Supermultiset search.}
+\caption{Super-multiset search.}
 \label{t:res_sup}
@@
 We can see that multiset-trie outperforms inverted index in all of the experiments by up to 4 orders of magnitude.
-In exact search multiset-trie has to traverse only up to $\sigma+1$ nodes in order to get query result. It is seen 
-from results that with increase of $\sigma$ the processing time for multiset-tire also increases. Interestingly, the 
-multiplicity also affects the time of processing, however, this happens passively. Multiplicity or degree of a node $n$ 
-defines what multisets can be stored in multiset-trie. Thus, it affects the structure and density of the multiset-trie.
+In an exact search, multiset-trie has to traverse only up to $\sigma+1$ nodes to get query result.
+It can be seen from results that with the increase of $\sigma$ the processing time for multiset-tire also increases.
+Multiplicity also affects the processing time; however, this happens passively.
+((equivalent to $\sigma+1$?)) Multiplicity, or degree of a node $n$, defines the shape of multisets that are stored in multiset-trie.
+Thus, it affects the structure and density of the multiset-trie.
 
-As for the inverted index, it must fetch all the postings according to each element in a test multiset and then intersect 
-the results in order to answer the query. This is a more time consuming operation than a simple tree traversal, which we 
-can see from results. Postings are of course filtered on-the-fly to reduce the cost of intersection, however these are still 
-additional operations to be performed. Note, that inverted file will also access at most $\sigma$ postings during search, 
-but in addition to access some extra processing is required.
+As for the inverted index, all three operations must first fetch all the postings for each particular element of a test multi-set.
+Afterward, the intersection of postings is computed to answer the query.
+The operations use more processing time than a simple tree traversal, which we can see from results.
+Postings are filtered on-the-fly to reduce the cost of the intersection.
+((I do not get this: "Note that inverted index will also access at most $\sigma$ postings during search, but in addition to access some extra processing is required."))
 
-The submultiset and supermultiset basically do not change to much in the case of inverted file, because the algorithm consists 
-of the same steps: the postings are fetched according to elements of the test multiset, then they are filtered on-the-fly 
-according to the query type, and at the end the postings either has to be united or intersected. So, besides a traversal of at most 
-$\sigma$ keys of the index additional processing is required. Note, that every key is mapped to a list of postings which 
-can be large as index grows.
+Implementations of the sub-multiset and super-multiset are similar in the case of the inverted file.
+The algorithm consists of the same steps.
+First, the postings are fetched for each element of the test multiset.
+Depending on the particular operation, postings are filtered on-the-fly.
+Finally, the union or the intersection of the filtered set of postings is computed.
+Note that the processing time increases with the size of the inverted index because of the increased sizes of postings.
 
-In the case of multiset-trie only a traversal of the tree is required which is much faster than processing of postings as we can 
-see from results. In the worst case the whole tree is traversed, but so is for inverted file.
+%% more details on this paragraph
+((Can you characterize the sub-tree (num of nodes visited) visited by an operation more precisely?
+Can we say something about this?
+What is the size of the sub-tree visited by an operation?
+Why it is not very large?
+I did not try to finalize the following text.))
+In the case of multiset-trie, only a traversal of the tree is required which is much faster than the processing of postings as we can see from results.
+In the worst case the whole tree is traversed, but so is for an inverted file.
+((Can we see how...?))

### 2019-03-06 tex/tex/s_experiments.tex
Inline answers to comments

@@
 %% more details about implementation of the inverted files
-((Presentation of the inverted index implementation? What about the smallest postings first?))
+((Presentation of the inverted index implementation? What about the smallest postings first?) -- (What do you mean by "smallest postings"?))
 
@@
 
-((Is there anything special to be said about the artificially generated data?))
+((Is there anything special to be said about the artificially generated data?) -- (Not really.
+We say that the data is generated with respect to ms-trie parameters and it is random.
+Other detail, such as how we generated the data (with Python) does not really matter.))
 
@@
 Multiplicity also affects the processing time; however, this happens passively.
-((equivalent to $\sigma+1$?)) Multiplicity, or degree of a node $n$, defines the shape of multisets that are stored in multiset-trie.
+((equivalent to $\sigma+1$?) -- (No, much smaller impact.
+In fact, if we consider the same data set and vary $\sigma$ and max multiplicity, the multiplicity will have almost no influence.
+The $\sigma$ will affect every single lookup of a multiset in ms-trie, when multiplicity will only influence supermultiset queries, where we will check one extra multiplicity.)) Multiplicity, or degree of a node $n$, defines the shape of multisets that are stored in multiset-trie.
 Thus, it affects the structure and density of the multiset-trie.
@@
 Postings are filtered on-the-fly to reduce the cost of the intersection.
-((I do not get this: "Note that inverted index will also access at most $\sigma$ postings during search, but in addition to access some extra processing is required."))
+((I do not get this: "Note that inverted index will also access at most $\sigma$ postings during search, but in addition to access some extra processing is required.") -- (Inverted index will search for multisets in all of the postings ($\sigma$ postings) in the worst case.
+Ms-trie needs to traverse $\sigma + 1$ nodes to lookup a multiset.
+So the access time would be approximately the same.
+However, inverted index requires additional processing of postings to search for a multiset when the ms-trie does not.))
 
@@
 Why it is not very large?
-I did not try to finalize the following text.))
+I did not try to finalize the following text.) -- (The sub-tree we visit is quite simple to picture.
+Take a multiset and draw the complete path defined by this multiset in the ms-trie.
+Everything that is on the left side of this path will be visited by submset query, everything on the right will be visited by supermset query.
+The size of this sub-tree we actually described in the math analysis.
+There is a formula that is parameterized by input multiset.
+I think I don't say that it is not large anywhere.
+It can be the whole tree.))
 In the case of multiset-trie, only a traversal of the tree is required which is much faster than the processing of postings as we can see from results.
 In the worst case the whole tree is traversed, but so is for an inverted file.
-((Can we see how...?))
+((Can we see how...?) -- (The same way as with ms-trie.
+Just search for a submsets of some large multiset that would result in retrieving the whole structure.))

### 2019-03-10 tex/tex/s_intro.tex
First part of the introduction is corrected. Not completed!

@@
 %
-A multiset is a data structure that represents a collection of elements. It generalizes a set data structure 
-by allowing duplicate elements in a collection. 
-%
-Data with duplicate elements is encountered in a wide variety of domains and applications. For example, 
-in databases or data mining. In particular, processing of multisets has attracted significant attention in 
-recent years. 
-%
-There are discussions about storage and representation of multi-valued attributes in databases, organization 
-of data with shared content, containment joins, efficient algorithms for containment queries, compression 
-of multisets, etc~\cite{bouros2016set,gripon2012compressing,ross2004symmetric,steinruecken2015compressing}.
-
+A multiset is a data structure that represents a collection of elements.
+It generalizes a set data structure by allowing duplicate elements in a collection.
+Multisets appear in a wide variety of domains and applications.
+The index structures for storing sets of multisets were studied in the area of object-relational database systems to efficiently store, compress and query multiset-valued attributes \cite{bouros2016set,gripon2012compressing,ross2004symmetric,steinruecken2015compressing}.
+Furthermore, the need to efficiently store and query the multisets appears also in the information retrieval \cite{}, the data mining \cite{}, and in the area of expert systems \cite{}.
 %
 % problem
+
+In this paper, we address the problems of storing, indexing and querying the sets of multisets.
+In particular, we deal with the design of an index data structure that provides an efficient implementation of the containment queries.
+Let $S$ be an index storing a set of multisets.
+For a given input multiset $m$, a \emph{containment query} searches for either sub-multisets or super-multisets of $m$ in $S$.
+
+Existent indexes for storing a set of multisets are rooted in the search trees.
+The elements of a search tree can be accessed through keys.
+This approach is efficient for checking the membership of individual multisets $m$ in $S$.
+However, it is not as efficient for operations that use containment relation between multisets.
+The search based on the containment relation requires access to the collections $C\subseteq\/S$ of multisets that are related to a multiset $m$ either by a sub-multiset or a super-multisets relationship.
+
+The existing solutions to the implementation of the containment queries include the inverted file \cite{}, signature tree \cite{} or B+ tree~\cite{Helmer2003}.
+These solutions provide a key-value look-up for elements of a multiset implying that containment operations are still element-wise.
+(?
+Is it possible to state more precisely the drawbacks of these solutions?
+?)
 %
-The problem we would like to address in this paper is efficient execution of containment queries on a set of multisets.
-The modern implementations of data structure for a set or a multiset is usually implemented as a search tree and 
-its elements can be accessed by a key. 
-%
-This approach is good when individual elements of a set or a multiset are of interest. However, it is not as optimal for operations 
-that use containment relation between multisets. 
-%
-Containment relation allows us to operate directly with the collections that are associated with a multiset such as a set 
-of submultiset and a set of supermultisets.
-%
-Currently, the solution to searching or retrieval of a set of multisets using containment queries involves inverted file, 
-signature tree or B+ tree~\cite{Helmer2003}. These solutions provide a key-value look-up for elements of a multiset implying that 
-containment operations are still element-wise. In order to improve the efficiency of operations we organize a set of 
-multisets in a different way described further below.
 
-% --------------------------------------------------------------------------------------------------------------------------------------------------------------------
+---------------------------------------------------------------------
 
@@
 %
-In this paper we propose a data structure \emph{multiset-trie} that is designed for storage and 
-processing of a finite bounded set of multisets. It extends the \emph{set-trie} data structure proposed 
-by Savnik~\cite{savnik2013index} that was designed for storage and processing of a finite bounded (?or 
-unbounded?) set of sets. 
+To improve the efficiency of containment operations, we propose a data structure \emph{multiset-trie} that is designed for storage and processing of a finite bounded set of multisets.
+It extends the \emph{set-trie} data structure proposed by Savnik~\cite{savnik2013index} that was designed for storage and processing of a finite bounded (?or unbounded?) set of sets.
 %
-Set-trie is a trie based data structure that provides a fast search and retrieval of sets and implements 
-set-containment operations. 
+Set-trie is a trie based data structure that provides a fast search and retrieval of sets and implements set-containment operations. 
 %

### 2019-03-11 tex/tex/s_intro.tex
minor updates.

@@
 The index structures for storing sets of multisets were studied in the area of object-relational database systems to efficiently store, compress and query multiset-valued attributes \cite{bouros2016set,gripon2012compressing,ross2004symmetric,steinruecken2015compressing}.
-Furthermore, the need to efficiently store and query the multisets appears also in the information retrieval \cite{}, the data mining \cite{}, and in the area of expert systems \cite{}.
+Furthermore, the need to efficiently store and query the multisets appears also in the information retrieval \cite{manningBook}, the data mining \cite{interestingSets}, and in the area of expert systems \cite{reteAlgorithm}.
 %
@@
 
-Existent indexes for storing a set of multisets are rooted in the search trees.
+Existent indexes for storing a set of multisets are rooted in the search trees \cite{rivestBook}.
 The elements of a search tree can be accessed through keys.
 This approach is efficient for checking the membership of individual multisets $m$ in $S$.
-However, it is not as efficient for operations that use containment relation between multisets.
+However, it is not as efficient for  containment queries.
 The search based on the containment relation requires access to the collections $C\subseteq\/S$ of multisets that are related to a multiset $m$ either by a sub-multiset or a super-multisets relationship.
 
-The existing solutions to the implementation of the containment queries include the inverted file \cite{}, signature tree \cite{} or B+ tree~\cite{Helmer2003}.
+The existing solutions to the implementation of the containment queries include the inverted file \cite{manningBook}, signature tree \cite{} or B+ tree~\cite{Helmer2003}.
 These solutions provide a key-value look-up for elements of a multiset implying that containment operations are still element-wise.

### 2019-03-11 tex/tex/s_intro.tex
first pass through the 2nd part of the paper (see text about parts).

@@
 %
-% problem
+% problem definition
 
@@
 
-% description of innovations
+% description of contributions
 %
-To improve the efficiency of containment operations, we propose a data structure \emph{multiset-trie} that is designed for storage and processing of a finite bounded set of multisets.
+To improve the efficiency of containment operations, we propose a data structure \emph{multiset-trie} that is designed for the storage and processing of a finite bounded set of multisets.
 It extends the \emph{set-trie} data structure proposed by Savnik~\cite{savnik2013index} that was designed for storage and processing of a finite bounded (?or unbounded?) set of sets.
 %
-Set-trie is a trie based data structure that provides a fast search and retrieval of sets and implements set-containment operations. 
+Set-trie is a tree data structure that provides a fast search and retrieval of sets and implements set-containment operations. 
 %
-Multiset-trie generalizes set-trie and can work with a set of sets (as set-trie) or with 
-a set of multisets providing efficient multiset-containment operations.
+Multiset-trie generalizes set-trie data structure.
+It provides a space efficient representation of a set of multisets and, it provides efficient multiset-containment operations.
+In addition, it can also represent and query sets of sets.
 
 % intro into mstrie
-The multiset-trie is an $n$-ary tree based data structure with properties similar to those of a trie. 
-Multisets are associated with a collection of nodes in a tree such that every node represents an element 
-of a multiset with particular multiplicity bounded by a node degree $n.$
+The multiset-trie is an $n$-ary tree data structure.
+((letters are levels of tree; each level states 0-n leters; I will correct and balance this further)) Multisets are represented with a collection of nodes in a tree such that every node represents an element of a multiset with particular multiplicity bounded by a node degree $n.$
 %
-Multiset-trie is a kind of search tree. Similarly to a trie, it uses common prefixes for shared data representation.
-Unlike the compact prefix tree, the multiset-trie does not support path compression. However, the absence of path 
-compression makes the multiset-trie a perfectly height-balanced tree. Moreover, when multiset-trie is full it forms 
-a complete $n$-ary tree.
+Multiset-trie is also a kind of search tree.
+Similarly to a trie, it uses common prefixes for a shared data representation.
+Unlike the compact prefix tree \cite{patriciaTree} ((radix tree?)), the multiset-trie does not support path compression.
+However, the absence of path compression makes the multiset-trie a perfectly height-balanced tree.
+Moreover, when multiset-trie is full it forms a complete $n$-ary tree.
 %
-The multiset-trie handles multisets directly by having access to each of the element 
-without the need to reconstruct them for processing, which allows fast retrieval and 
-containment operations. In particular, it supports submultiset and supermultiset queries.
-The operations allow to find and retrieve the closest submultisets or supermultisets 
-as well as to find and retrieve all of them.
+((this needs to be rewritten)) The multiset-trie handles multisets directly by having access to each of the element without the need to reconstruct them for processing, which allows fast retrieval and containment operations.
+In particular, it supports submultiset and super-multiset queries.
+The operations allow to find and retrieve the closest submultisets or super-multisets as well as to find and retrieve all of them.
 
-% --------------------------------------------------------------------------------------------------------------------------------------------------------------------
+% --------------------------------------------------------------------------------------------------------------------------------------------
 
@@
 
-% --------------------------------------------------------------------------------------------------------------------------------------------------------------------
+% --------------------------------------------------------------------------------------------------------------------------------------------
 

### 2019-03-12 tex/tex/s_intro.tex
2nd part updated.

@@
 %
-% general intro into problem
+% 1st part: general intro into problem
 %
@@
 ?)
+
+% ----------------------------------------------------------------------------------------------------------------------------------------
 %
-
----------------------------------------------------------------------
-
-% description of contributions
+% 2nd part: description of the proposed data structure
 %
 To improve the efficiency of containment operations, we propose a data structure \emph{multiset-trie} that is designed for the storage and processing of a finite bounded set of multisets.
-It extends the \emph{set-trie} data structure proposed by Savnik~\cite{savnik2013index} that was designed for storage and processing of a finite bounded (?or unbounded?) set of sets.
+It extends the \emph{set-trie} data structure proposed by Savnik~\cite{savnik2013index} that was designed for storage and processing of a finite bounded set of sets.
 %
 Set-trie is a tree data structure that provides a fast search and retrieval of sets and implements set-containment operations.
-%
 Multiset-trie generalizes set-trie data structure.
@@
 The multiset-trie is an $n$-ary tree data structure.
-((letters are levels of tree; each level states 0-n leters; I will correct and balance this further)) Multisets are represented with a collection of nodes in a tree such that every node represents an element of a multiset with particular multiplicity bounded by a node degree $n.$
+Each multiset in a multiset-trie is represented by a path from the root to a leaf node.
+Multiset elements are the symbols from an alpahabet.
+Each symbol from the alphabet is represented by a node at a certan level of the multiset-trie.
+The node stores the multiplicity of an element in a multiset.
 %
-Multiset-trie is also a kind of search tree.
+%Multisets are represented with a collection of nodes in a tree such that every node represents an element of a multiset with particular multiplicity bounded by a node degree $n.$
+%
+A multiset-trie is also a kind of search tree.
 Similarly to a trie, it uses common prefixes for a shared data representation.
-Unlike the compact prefix tree \cite{patriciaTree} ((radix tree?)), the multiset-trie does not support path compression.
+Unlike the compact prefix tree \cite{patriciaTree}, the multiset-trie does not support path compression.
 However, the absence of path compression makes the multiset-trie a perfectly height-balanced tree.
@@
 %
-((this needs to be rewritten)) The multiset-trie handles multisets directly by having access to each of the element without the need to reconstruct them for processing, which allows fast retrieval and containment operations.
-In particular, it supports submultiset and super-multiset queries.
+%The multiset-trie handles multisets directly by having access to each of the element without the need to reconstruct them for processing, which allows fast retrieval and containment operations.
+In particular, it supports sub-multiset and super-multiset queries.
 The operations allow to find and retrieve the closest submultisets or super-multisets as well as to find and retrieve all of them.
 
-% --------------------------------------------------------------------------------------------------------------------------------------------
+% 
+--------------------------------------------------------------------------------------------------------------------------------------------
+%
+% 3rd part: contributions
 
-% key results
 Empirical studies on real data show, that multiset-trie is sensitive to the context, which can 
@@
 % --------------------------------------------------------------------------------------------------------------------------------------------
+%
+% 4th part: paper organization
 
-% paper organization
-%
 In the following Section~\ref{c:description} we present the organization of 

### 2019-03-14 tex/tex/s_intro.tex
The first version of contributions paragraph (part 3)

@@
 
-The existing solutions to the implementation of the containment queries include the inverted file \cite{manningBook}, signature tree \cite{} or B+ tree~\cite{Helmer2003}.
+The existing solutions to the implementation of the containment queries include the inverted index \cite{manningBook}, signature tree \cite{} or B+ tree~\cite{Helmer2003}.
 These solutions provide a key-value look-up for elements of a multiset implying that containment operations are still element-wise.
@@
 % 
---------------------------------------------------------------------------------------------------------------------------------------------
+%--------------------------------------------------------------------------------------------------------------------------------------------
 %
 % 3rd part: contributions
-
-Empirical studies on real data show, that multiset-trie is sensitive to the context, which can 
-be further used to optimize data structure for particular data. Moreover, knowing the data 
-it is possible to estimate the overall performance of the multiset-trie both time and space 
-related using our mathematical theory that describes multiset-trie.
+The contributions of this paper are:
+\begin{itemize}
+\item A novel data structure multiset-trie which can store a set of multisets using common prefixes and provide fast containment queries. 
+\item A mathematical model of a general case multiset-trie that analyses its space complexity, an estimation of a sub-tree size traversed by a containment query and therefore gives insights on containment query time complexity.
+\item Empirical results of multiset-trie performance depending on a multiset space parameterization.
+\item A comparative study between an inverted index and multiset-trie where the later outperforms its competitor by up to 4 orders of magnitude in the time consumed by both equality and containment queries.
+\item Experimental result that shows an influence of ordering in multisets on their storage and processing.
+\end{itemize}
 %
-The comparative study shows how efficient multiset-trie is by outperforming inverted file in 
-both equality and containment queries by up to 4 orders of magnitude in the time consumed by 
-query.
 

### 2019-03-14 tex/tex/s_experiments.tex
updated experiment 5.

--- 2019-03-04 tex/tex/s_experiments.tex
@@
 In this experiment, we demonstrate the performance of multiset-trie data structure compared to the inverted index based on the B+ tree.
-Both data structures are implemented in the programming language \CC, providing in this way an experimental setup for a fair comparison  [ref.
-to impl].
+Both data structures are implemented in the programming language \CC, providing in this way an experimental setup for a fair comparison \cite{akulich2019mstrie}.
 
-%% more details about implementation of the inverted files
-((Presentation of the inverted index implementation? What about the smallest postings first?))
+((Brief description of the implementation of the inverted files?))
 
@@
 The input and test data were generated with respect to parameters $\sigma$ and $n$ as presented in Table~\ref{t:benchmark}.
-
-((Is there anything special to be said about the artificially generated data?))
 
@@
 We tested all three types of a query on all of the configurations from Table 1 which results in 18 experiments total, i.e., 6 experiments per query type.
-In each of the experiment, we measured an average time consumed by the data structure to process the query.
+In each experiment, we measured an average time consumed by the data structure to process the query.
 The results of the exact search, sub-multiset and super-multiset search experiments are presented in tables Table~\ref{t:res_ex}, Table~\ref{t:res_sub} and Table~\ref{t:res_sup}, respectively.
@@
 Multiplicity also affects the processing time; however, this happens passively.
-((equivalent to $\sigma+1$?)) Multiplicity, or degree of a node $n$, defines the shape of multisets that are stored in multiset-trie.
+Multiplicity, or degree of a node $n$, defines the shape of multisets that are stored in multiset-trie.
 Thus, it affects the structure and density of the multiset-trie.
 
-As for the inverted index, all three operations must first fetch all the postings for each particular element of a test multi-set.
+As for the inverted index, all three operations must first fetch all postings for each particular element of a test multi-set.
 Afterward, the intersection of postings is computed to answer the query.
@@
 Postings are filtered on-the-fly to reduce the cost of the intersection.
-((I do not get this: "Note that inverted index will also access at most $\sigma$ postings during search, but in addition to access some extra processing is required."))
 
@@
 
-%% more details on this paragraph
-((Can you characterize the sub-tree (num of nodes visited) visited by an operation more precisely?
-Can we say something about this?
-What is the size of the sub-tree visited by an operation?
-Why it is not very large?
-I did not try to finalize the following text.))
 In the case of multiset-trie, only a traversal of the tree is required which is much faster than the processing of postings as we can see from results.
 In the worst case the whole tree is traversed, but so is for an inverted file.
-((Can we see how...?))

### 2019-03-19 tex/tex/s_intro.tex
introduction updated. not finished.

@@
 The index structures for storing sets of multisets were studied in the area of object-relational database systems to efficiently store, compress and query multiset-valued attributes \cite{bouros2016set,gripon2012compressing,ross2004symmetric,steinruecken2015compressing}.
-Furthermore, the need to efficiently store and query the multisets appears also in the information retrieval \cite{manningBook}, the data mining \cite{interestingSets}, and in the area of expert systems \cite{reteAlgorithm}.
+Furthermore, the need to efficiently store and query the multisets appears also in the information retrieval \cite{manning2008introduction,baeza1999modern}, the data mining \cite{interestingSets}, and in the area of expert systems \cite{reteAlgorithm}.
 %
@@
 
-The existing solutions to the implementation of the containment queries include the inverted index \cite{manningBook}, signature tree \cite{} or B+ tree~\cite{Helmer2003}.
+The existing solutions to the implementation of the containment queries include the inverted index \cite{zobel1992efficient,zobel1998inverted,broder2006indexing}, the signature tree \cite{tousidou2002sigstruc,yangjun2005stree,zobel1998inverted} or, the B+ tree~\cite{Helmer2003,?}.
 These solutions provide a key-value look-up for elements of a multiset implying that containment operations are still element-wise.

### 2019-03-21 tex/tex/s_intro.tex
pharagraph on the contributions updated.

@@
 % 3rd part: contributions
-The contributions of this paper are:
-\begin{itemize}
-\item A novel data structure multiset-trie which can store a set of multisets using common prefixes and provide fast containment queries. 
-\item A mathematical model of a general case multiset-trie that analyses its space complexity, an estimation of a sub-tree size traversed by a containment query and therefore gives insights on containment query time complexity.
-\item Empirical results of multiset-trie performance depending on a multiset space parameterization.
-\item A comparative study between an inverted index and multiset-trie where the later outperforms its competitor by up to 4 orders of magnitude in the time consumed by both equality and containment queries.
-\item Experimental result that shows an influence of ordering in multisets on their storage and processing.
-\end{itemize}
+The main contribution of this paper is a novel data structure multiset-trie for storing and querying a set of multisets that provides efficient multiset containment operations.
+Furthermore, the contributions of this work are thorough mathematical and empirical analyses of the proposed data structure multiset-trie.
 %
-
+A mathematical model is developed to analyze the complexity of multiset containment operations.
+In particular, we estimate the size of a sub-tree traversed by a containment query and, therefore, give an insight into the time complexity of containment queries.
+In addition to the mathematical analysis, the size of a sub-tree visited by a multiset containment query is also the central focus of the empirical analysis.
+We carefully designed the experiments to unravel the main features of the time complexity space.
+%
+Finally, a comparative study between the inverted index and the multiset-trie shows that the multiset-trie outperforms the inverted index by 4 levels of magnitude in the time consumed by both equality and containment queries.
+%
+%Experimental result that shows an influence of ordering in multisets on their storage and processing.
+%
+%
 % --------------------------------------------------------------------------------------------------------------------------------------------
@@
 
-In the following Section~\ref{c:description} we present the organization of 
-multiset-trie data structure in detail.
+(? not finshed ... ?)
+The paper is organized as follows. In the following Section~\ref{c:description}, we present the multiset-trie data structure in detail.
 %
-Next, in Section~\ref{c:operations} we present operations that multiset-trie currently 
-supports. This includes multiset containment and the basic search tree functions. 
-The algorithms in pseudo code are presented as well. 
+Section~\ref{c:operations} presents the operations of the multiset-trie. This includes multiset containment and the basic search tree functions. The algorithms in pseudo code are presented as well. 
 %
-The description of multiset-trie functions and procedures is followed by the 
-mathematical analysis of their complexity in Section~\ref{c:analysis}. 
-The main assumption is that multisets are constructed uniformly at random 
-with bounded cardinality. By using probabilistic tools we describe time complexity of 
-the algorithms and space complexity of the structure.
+The description of multiset-trie functions and procedures is followed by the mathematical analysis of their complexity in Section~\ref{c:analysis}.
+The main assumption is that multisets are constructed uniformly at random with bounded cardinality.
+By using probabilistic tools we describe time complexity of the algorithms and space complexity of the structure.
 %
-Further, in Section~\ref{c:experiments} we present an empirical study of the 
-multiset-trie. Synthetic and real-world data sets are used in experiments to test the performance 
-of the data structure. The experiments also highlight the methods for optimizing a multiset-trie.
+Further, in Section~\ref{c:experiments} we present an empirical study of the multiset-trie.
+Synthetic and real-world data sets are used in experiments to test the performance of the data structure.
+The experiments also highlight the methods for optimizing a multiset-trie.
 %

### 2019-03-25 tex/tex/s_experiments.tex
Paragraph on inverted index implementation, added some references (.bib file is not completely reviewed)

@@
 \subsection{Experiment 5} \label{s:exp5}
-In this experiment, we demonstrate the performance of multiset-trie data structure compared to the inverted index based on the B+ tree.
+In this experiment, we demonstrate the performance of multiset-trie data structure compared to the inverted index based on the B+~tree.
 Both data structures are implemented in the programming language \CC, providing in this way an experimental setup for a fair comparison \cite{akulich2019mstrie}.
 
-((Brief description of the implementation of the inverted files?))
+%((Brief description of the implementation of the inverted files?))
+The inverted index is implemented using an idea from~\cite{Helmer2003}.
+An inverted index structure consist of two parts: a dictionary and postings.
+In our case a dictionary is implemented as an in-memory B+~tree where keys are all distinct values from a domain represented by a set $\Sigma.$ The postings are represented by lists of multisets that contain a particular element from $\Sigma.$ Each list item in postings contains a cardinality of a multiset, which speeds up the containment queries.
 

### 2019-03-25 tex/tex/s_intro.tex
Paragraph on inverted index implementation, added some references (.bib file is not completely reviewed)

@@
 The index structures for storing sets of multisets were studied in the area of object-relational database systems to efficiently store, compress and query multiset-valued attributes \cite{bouros2016set,gripon2012compressing,ross2004symmetric,steinruecken2015compressing}.
-Furthermore, the need to efficiently store and query the multisets appears also in the information retrieval \cite{manning2008introduction,baeza1999modern}, the data mining \cite{interestingSets}, and in the area of expert systems \cite{reteAlgorithm}.
+Furthermore, the need to efficiently store and query the multisets appears also in the information retrieval \cite{manning2008introduction,baeza1999modern}, the data mining \cite{interestingSets}, and in the area of expert systems \cite{reteAlgorithm,inbook}.
 %
@@
 
-The existing solutions to the implementation of the containment queries include the inverted index \cite{zobel1992efficient,zobel1998inverted,broder2006indexing}, the signature tree \cite{tousidou2002sigstruc,yangjun2005stree,zobel1998inverted} or, the B+ tree~\cite{Helmer2003,?}.
+The existing solutions to the implementation of the containment queries include the inverted index \cite{zobel1992efficient,zobel1998inverted,broder2006indexing}, the signature tree \cite{tousidou2002sigstruc,yangjun2005stree,zobel1998inverted} or, the B+ tree~\cite{Helmer2003}.
 These solutions provide a key-value look-up for elements of a multiset implying that containment operations are still element-wise.
@@
 Similarly to a trie, it uses common prefixes for a shared data representation.
-Unlike the compact prefix tree \cite{patriciaTree}, the multiset-trie does not support path compression.
+Unlike the compact prefix tree \cite{Sedgewick:2011:ALG:2011916}, the multiset-trie does not support path compression.
 However, the absence of path compression makes the multiset-trie a perfectly height-balanced tree.

### 2019-06-19 tex/tex/s_experiments.tex
Small addition to Introduction. Related work.

--- 2019-03-14 tex/tex/s_experiments.tex
@@
 structure depending on the selected parameters as well as the comparative 
-benchmark of the multiset-trie against B+ tree implementation of inverted index. 
+benchmark of the multiset-trie against B-tree implementation of inverted index. 
 % outline of the experiment section
@@
 After all the experiments we present an empirical comparison of multiset-trie 
-data structure with B+ tree based inverted index. We use inverted index 
+data structure with B-tree based inverted index. We use inverted index 
 to store and retrieve multisets in the same way as it is described in the paper by 
@@
 \subsection{Experiment 5} \label{s:exp5}
-In this experiment, we demonstrate the performance of multiset-trie data structure compared to the inverted index based on the B+ tree.
+In this experiment, we demonstrate the performance of multiset-trie data structure compared to the inverted index based on the B-tree.
 Both data structures are implemented in the programming language \CC, providing in this way an experimental setup for a fair comparison \cite{akulich2019mstrie}.
 
-((Brief description of the implementation of the inverted files?))
+%((Brief description of the implementation of the inverted files?))
+The inverted index is implemented using an idea from~\cite{Helmer2003}.
+An inverted index structure consist of two parts: a dictionary and postings.
+In our case a dictionary is implemented as an in-memory B-tree where keys are all distinct values from a domain represented by a set $\Sigma.$ The postings are represented by lists of multisets that contain a particular element from $\Sigma.$ Each list item in postings contains a cardinality of a multiset, which speeds up the containment queries.
 

### 2019-06-19 tex/tex/s_intro.tex
Small addition to Introduction. Related work.

@@
 The existing solutions to the implementation of the containment queries include the inverted index \cite{zobel1992efficient,zobel1998inverted,broder2006indexing}, the signature tree \cite{tousidou2002sigstruc,yangjun2005stree,zobel1998inverted} or, the B+ tree~\cite{Helmer2003}.
-These solutions provide a key-value look-up for elements of a multiset implying that containment operations are still element-wise.
-(?
-Is it possible to state more precisely the drawbacks of these solutions?
-?)
+These solutions provide a key-value look-up, where a key is an element of a multiset and a value is a corresponding multiset implying that containment operations require additional processing of partial results such as intersection or union.
 

### 2019-06-19 tex/tex/s_relwork.tex
algorithms adjusted and get-all algorithms added
Small addition to Introduction. Related work.

@@
 \section{Related work} \label{c:relwork}
-[Will be rewritten]
+
+% intro to related work
+
+Our work is related to storage of multisets and multiset containment queries. In this section we summarize previous work done 
+for storage and retrieval of sets and multisets as well as work on (multi)set containment queries. Also, we discuss the applications 
+of multiset-trie with respect to already existing data structures.
+
+% Construction of multiset-trie
+
+\subsection{Set-trie}
+% set-trie: 
+% - data structure that multiset-trie generalizes
+% - stores a set of sets
+% - supports set containment operations
+
+The multiset-trie is closely related to the set-trie data structure proposed by Savnik~\cite{savnik2013index}. 
+It is a trie-based data structure that is adapted to efficient storage and retrieval of sets. In particular, 
+it stores a set of sets and supports set containment operations such as retrieval of the nearest sub- 
+and supersets as well as retrieval of all sub- and supersets from the data structure.
+
+The data structure multiset-trie generalizes the set-trie allowing to store a set of multisets. When the multiset-trie 
+is restricted to store a set of sets the underlying data structure becomes a simple binary tree. Moreover, all the 
+operations of the set-trie are also supported by the multiset-trie. The generalization is, of course, comes with a small penalty 
+in performance if we compare the multiset-trie with the set-trie. The downside of such a generalization is that multiset-trie 
+no longer supporting path compression that was obtained in set-trie. However, the design of multiset-trie provides a storage 
+of multisets as well as constant worst-case time complexity of search operation independently of user input.
+
+\subsection{Inverted file}
+% inverted index:
+% - performance study of set containment queries (sequential signature files, signature trees, extendible signature hashing, inverted files)
+% - performance comparison of multiset-trie to inverted index
+% - containment query issues of inverted index
+
+The performance study~\cite{Helmer2003} shows that the inverted file is the most efficient data structure for set containment queries 
+among data structures: sequential signature files, signature trees, extendible signature hashing and inverted files. 
+In our work we replicate the experiment with inverted file and compare the performance of set containment queries between inverted file and multiset-trie data structures.
+We user the same methodology as in related paper in order to implement containment queries for inverted file.
+The results of this comparison can be found the Experiments section~\ref{s:exp5}.
+
+% Application areas of multiset-trie
+
+\subsection{Set containment joins}
+[WIP]
+% set containment joins:
+% - review the paper: Set containment join revisited\cite{bouros2016set}
+% - multiset containment join
+
+\subsection{Multisets in relational databases}
+[WIP]
+% multisets in RDBMS and ORDBMS:
+% - set-valued attributes
+% - multiset-valued attributes
+% - underlying datastructures for storage of set- and multiset-valued attributes
+
+\subsection{Application of multiset-trie}
+[WIP]
+% application of multiset-trie
+% - indexing of multisets
+% - containment queries
+% - multiset containment join
+% - information retrieval
+
 %
 % Set theory, multiest theory, bag-of-words model
-\subsection{Multiset}
-Multiset is a widely used data structure in different areas of mathematics, physics 
-and computer science \cite{singh2007overview}. The theory of multisets is based 
-entirely on the theory of sets. However, classical mathematics does not deal with 
-multisets directly. Instead, one can define a multiset to be a \emph{family} of sets 
-or the functions on ordered pairs, where the members of a pair are an element and its 
-multiplicity. This means that mathematically the concepts of set such as cardinality, 
-set-containment operation, power set, equivalence classes and others are well defined 
-for multisets in terms of sets \cite{blizard1988multiset}.
-
-The concept of a multiset can also be referred to the \emph{bag-of-words model}. 
-This model takes its origin from a linguistic context studied by Harris~\cite{harris1954distributional}. 
-According to the bag-of-words model, text can be represented as a bag (multiset) 
-of words, where an element is a word and the number of its occurrences in the 
-text is multiplicity. A bag of words does not keep track of grammar and ordering 
-of words.
+%\subsection{Multiset}
+%Multiset is a widely used data structure in different areas of mathematics, physics 
+%and computer science \cite{singh2007overview}. The theory of multisets is based 
+%entirely on the theory of sets. However, classical mathematics does not deal with 
+%multisets directly. Instead, one can define a multiset to be a \emph{family} of sets 
+%or the functions on ordered pairs, where the members of a pair are an element and its 
+%multiplicity. This means that mathematically the concepts of set such as cardinality, 
+%set-containment operation, power set, equivalence classes and others are well defined 
+%for multisets in terms of sets \cite{blizard1988multiset}. 
+%
+%The concept of a multiset can also be referred to the \emph{bag-of-words model}. 
+%This model takes its origin from a linguistic context studied by Harris~\cite{harris1954distributional}. 
+%According to the bag-of-words model, text can be represented as a bag (multiset) 
+%of words, where an element is a word and the number of its occurrences in the 
+%text is multiplicity. A bag of words does not keep track of grammar and ordering 
+%of words.
 
 % Information retrieval
-\subsection{Information retrieval}
-\emph{Information retrieval (IR)} refers to a problem of finding material of an 
-unstructured nature that satisfies an information need~\cite{manning2008introduction}. 
-Usually, one is searching for a specific documents in a significantly large text 
-documents database. The size of a database makes the search a time consuming 
-operation. In order to resolve the issue IR systems pre-process data and create 
-indexes for future use in search operation.
+%\subsection{Information retrieval}
+%\emph{Information retrieval (IR)} refers to a problem of finding material of an 
+%unstructured nature that satisfies an information need~\cite{manning2008introduction}. 
+%Usually, one is searching for a specific documents in a significantly large text 
+%documents database. The size of a database makes the search a time consuming 
+%operation. In order to resolve the issue IR systems pre-process data and create 
+%indexes for future use in search operation. 
+%
+%The bag-of-words model is widely used in IR. In particular, such a representation of 
+%text documents is used in database indexes when a full text search of a database 
+%is required. The full-text search problem refers to indexing techniques for full-text 
+%databases. The most efficient index nowadays uses the concept of an inverted index 
+%\cite{zobel1992efficient}. 
+%
+%The proposed data structure multiset-trie can be used as an alternative implementation 
+%of the search structure of an inverted index. It represents words as multisets and 
+%stores them into data structure. The query processing is achieved using boolean 
+%retrieval model \cite{manning2008introduction} and multiset containment operations. 
+%Multiset containment operations of the multiset-trie implement the nearest neighbor 
+%search queries which retrieve not only exact but also the most relevant results 
+%to a user.
 
-The bag-of-words model is widely used in IR. In particular, such a representation of 
-text documents is used in database indexes when a full text search of a database 
-is required. The full-text search problem refers to indexing techniques for full-text 
-databases. The most efficient index nowadays uses the concept of an inverted index 
-\cite{zobel1992efficient}.
-
-The proposed data structure multiset-trie can be used as an alternative implementation 
-of the search structure of an inverted index. It represents words as multisets and 
-stores them into data structure. The query processing is achieved using boolean 
-retrieval model \cite{manning2008introduction} and multiset containment operations. 
-Multiset containment operations of the multiset-trie implement the nearest neighbor 
-search queries which retrieve not only exact but also the most relevant results 
-to a user.
-
-\subsection{Generalized search tree}
-The properties and operations of the multiset-trie makes it a competitor to the 
-most efficient implementation of a search tree the \emph{Generalized Search Tree 
-(GiST)}~\cite{broder2006indexing,hellerstein1995generalized,kornacker1999high} 
-that is used in inverted index. GiST is a very flexible data structure that can be 
-customized in order to behave like B+-tree, R-tree or RD-tree. It also provides 
-support for an extensible set of queries and data types that B+-tree, R-tree or RD-tree 
-do not support originally. GiST supports all the basic search tree operations such as 
-insert, delete and search, and in addition provides such extensions as the 
-nearest-neighbor search and multiset containment operations. The extensions 
-provided by GiST are native in the multiset-trie. The multiset-trie also has a fixed 
-height while GiST is a self-balanced tree and has to use additional methods in 
-order to preserve its balance.
-
-% Set-trie
-\subsection{Set-trie}
-The multiset containment queries are well studied in the area of relational 
-databases. The queries are well-defined in the context of relational 
-algebra~\cite{Lamperti2000msetsinds}. This problem was previously studied 
-for a restricted case of queries. In particular, the storage and fast retrieval of sets 
-were previously accomplished in data structure \emph{set-trie} proposed by 
-Savnik~\cite{savnik2013index}.
-
-The set-trie data structure is based on a trie data structure. It supports set 
-containment operations such as retrieval of the nearest sub- and supersets and 
-retrieval of all sub- and supersets from the data structure.
-
-The data structure multiset-trie adapts the properties of the set-trie implementing 
-the functions \textsc{submsetExistence}, \textsc{supermsetExistence}, 
-\textsc{getAllSubmsets} and \textsc{getAllSupermsets} together with the basic 
-tree functions such as \textsc{insert}, \textsc{delete} and \textsc{search}. 
-Moreover, multiset-trie extends the abilities of the set-trie allowing to store and 
-retrieve multisets. The downside of such an extension is that multiset-trie no 
-longer supports path compression that was obtained in set-trie. However, the 
-design of multiset-trie provides a constant worst case time complexity of search 
-function independently of user input.
+%\subsection{Generalized search tree}
+%The properties and operations of the multiset-trie makes it a competitor to the 
+%most efficient implementation of a search tree the \emph{Generalized Search Tree 
+%(GiST)}~\cite{broder2006indexing,hellerstein1995generalized,kornacker1999high} 
+%that is used in inverted index. GiST is a very flexible data structure that can be 
+%customized in order to behave like B+-tree, R-tree or RD-tree. It also provides 
+%support for an extensible set of queries and data types that B+-tree, R-tree or RD-tree 
+%do not support originally. GiST supports all the basic search tree operations such as 
+%insert, delete and search, and in addition provides such extensions as the 
+%nearest-neighbor search and multiset containment operations. The extensions 
+%provided by GiST are native in the multiset-trie. The multiset-trie also has a fixed 
+%height while GiST is a self-balanced tree and has to use additional methods in 
+%order to preserve its balance.

### 2020-08-05 tex/article.tex
abstract and introduction updated.

@@
 \begin{abstract}
-This work proposes a new data structure \emph{multiset-trie} that is designed for storage 
-and efficient processing of a set of multisets. Moreover, multiset-trie can operate on 
-a set of sets without efficiency loss.
+This paper proposes a new data structure \emph{multiset-trie} that is designed for storing and efficiently processing of a set of multisets.
+Moreover, multiset-trie can operate on a set of sets without efficiency loss.
 %
-The multiset-trie is a search-tree-based data structure with the properties similar to those of 
-a trie. It implements all standard search tree operations together with the multiset containment 
-operations such as submultiset and supermultiset. Containment operations allow efficient search and 
-retrieval of the individual multisets and also collections of multsets. In particular, it can find and retrieve 
-closest submultisets or supermultisets as well as all submultisets or supermultisets that are parameterized
-by user input.
+The multiset-trie is a search tree with properties similar to those of a trie.
+It implements all standard search tree operations together with the multiset containment operations such as sub-multiset and super-multiset.
+Suppose we have a set of multisets $S$ and a multiset $X$.
+The multiset containment operations retrieve multisets from $S$ that are either sub-multisets or super-multisets of $X$.
+
+We present the mathematical analysis of a multiset-trie that gives the time complexity of the algorithms and the space complexity of the data structure.
+Further, the empirical analysis of the data structure is implemented in a series of experiments.
+The experiments illuminate the time complexity space of the multiset containment operations.
+Finally, the comparison of the multiset-trie with the inverted index shows that the multiset-trie outperforms the inverted index by up to 4 orders of magnitude.
 %
-Multiset-trie is compared to modern solutions that provide containment operations on multisets. 
-The study shows how efficient multiset-trie is by outperforming inverted file in both equality and 
-containment queries by up to 4 orders of magnitude in query execution time.
 \end{abstract}
-\keywords{multiset-trie, multiset, bag-of-words, trie, data structure, containment operations, inverted file}
+\keywords{multiset, bag-of-words, trie, data structure, containment operations, inverted file}
 %\subclass{68P05, 68P20, 68Q87}

### 2020-08-05 tex/tex/s_intro.tex
abstract and introduction updated.

@@
 The index structures for storing sets of multisets were studied in the area of object-relational database systems to efficiently store, compress and query multiset-valued attributes \cite{bouros2016set,gripon2012compressing,ross2004symmetric,steinruecken2015compressing}.
-Furthermore, the need to efficiently store and query the multisets appears also in the information retrieval \cite{manning2008introduction,baeza1999modern}, the data mining \cite{interestingSets}, and in the area of expert systems \cite{reteAlgorithm,inbook}.
+Furthermore, the need to efficiently store and query the multisets appears also in the information retrieval \cite{manning2008introduction,baeza1999modern}, the data mining \cite{mannila1997}, and in the area of expert systems \cite{forgy1982,inbook}.
 %
@@
 In this paper, we address the problems of storing, indexing and querying the sets of multisets.
-In particular, we deal with the design of an index data structure that provides an efficient implementation of the containment queries.
+In particular, we deal with the design of an index data structure that provides an efficient implementation of the multiset containment queries.
 Let $S$ be an index storing a set of multisets.
@@
 
-Existent indexes for storing a set of multisets are rooted in the search trees \cite{rivestBook}.
+Existent indexes for storing a set of multisets are rooted in the search trees \cite{corman2001}.
 The elements of a search tree can be accessed through keys.
@@
 
-The existing solutions to the implementation of the containment queries include the inverted index \cite{zobel1992efficient,zobel1998inverted,broder2006indexing}, the signature tree \cite{tousidou2002sigstruc,yangjun2005stree,zobel1998inverted} or, the B+ tree~\cite{Helmer2003}.
+The existing solutions to the implementation of the containment queries include the inverted index \cite{zobel1992efficient,zobel1998inverted,broder2006indexing}, the signature tree \cite{tousidou2002sigstruc,yangjun2005stree,zobel1998inverted} and, the B+ tree~\cite{Helmer2003}.
 These solutions provide a key-value look-up, where a key is an element of a multiset and a value is a corresponding multiset implying that containment operations require additional processing of partial results such as intersection or union.
@@
 %
-To improve the efficiency of containment operations, we propose a data structure \emph{multiset-trie} that is designed for the storage and processing of a finite bounded set of multisets.
-It extends the \emph{set-trie} data structure proposed by Savnik~\cite{savnik2013index} that was designed for storage and processing of a finite bounded set of sets.
+To improve the efficiency of containment operations, we propose a data structure \emph{multiset-trie} that is designed for storing and processing of a finite bounded set of multisets.
+It generalizes the \emph{set-trie} data structure proposed by Savnik~\cite{savnik2013index} that was designed for storage and processing of a finite bounded set of sets.
 %
-Set-trie is a tree data structure that provides a fast search and retrieval of sets and implements set-containment operations.
-Multiset-trie generalizes set-trie data structure.
+The set-trie is an extension of a trie data structure to provide fast search and retrieval of sets and, to implement efficiently the set containment operations.
+The set-trie is also a form of the binomial tree \cite{corman2001}.
+Multiset-trie generalizes the set-trie data structure.
 It provides a space efficient representation of a set of multisets and, it provides efficient multiset-containment operations.
@@
 % 3rd part: contributions
-The main contribution of this paper is a novel data structure multiset-trie for storing and querying a set of multisets that provides efficient multiset containment operations.
+\subsection{Contributions}
+
+The main contribution of this paper is a novel data structure \emph{multiset-trie} for storing and querying a set of multisets that provides efficient multiset containment operations.
 Furthermore, the contributions of this work are thorough mathematical and empirical analyses of the proposed data structure multiset-trie.
@@
 % 4th part: paper organization
+\subsection{Paper organization}
 
-(? not finshed ... ?)
 The paper is organized as follows. In the following Section~\ref{c:description}, we present the multiset-trie data structure in detail.
 %
-Section~\ref{c:operations} presents the operations of the multiset-trie. This includes multiset containment and the basic search tree functions. The algorithms in pseudo code are presented as well. 
+Section~\ref{c:operations} presents the operations of the multiset-trie.
+These include the basic operations of search trees and, the multiset containment operations.
+The algorithms are presented in detail using the pseudo code.
 %
-The description of multiset-trie functions and procedures is followed by the mathematical analysis of their complexity in Section~\ref{c:analysis}.
+The description of multiset-trie operations is followed by the mathematical analysis of their complexity in Section~\ref{c:analysis}.
 The main assumption is that multisets are constructed uniformly at random with bounded cardinality.
@@
 Further, in Section~\ref{c:experiments} we present an empirical study of the multiset-trie.
-Synthetic and real-world data sets are used in experiments to test the performance of the data structure.
-The experiments also highlight the methods for optimizing a multiset-trie.
+Synthetic and real-world data sets are used in experiments that are designed to study the performance of the multiset containment operations.
+The multiset-trie is empirically compared to the inverted index in a separate experiment.
+% The experiments highlight the methods for optimizing a multiset-trie.
 %

### 2020-08-07 tex/tex/s_conclusions.tex
section Conclusions & further work updated.

@@
 \section{Conclusions and future work} \label{c:conclusions}
-[Will be rewritten]
+
 %
@@
 %
-One of the conclusions of studying the multiset-trie both theoretically and 
-empirically is that our data structure is input sensitive. Input sensitivity implies 
-a non consistent performance on different input data. However, our argument 
-that the performance can be optimized by pre-processing the input data 
-is confirmed in the Experiment~\hyperref[ss:exp3]{4}. 
-Pre-processing determines the optimal encoding for input data and ensures the 
-best performance of the multiset-trie on particular input data. In case of storing 
-words in the multiset-trie, the search queries can be always optimized based on 
-the frequencies of letters in a specific language. We also see from 
-Experiments~\hyperref[s:exp1]{1} and~\hyperref[s:exp2]{2} that dependence of 
-the multiset-trie performance on the density is not a linear function. Yet the function 
-is continuous and the point of inflection is unique on the whole domain as it is shown 
-in Experiment~\hyperref[s:exp3]{3}. This allows us to predict whether multiset-trie 
-can be used for some particular application, serving a high performance. 
+One of the conclusions of studying the multiset-trie both theoretically and empirically is that our data structure is input sensitive.
+Input sensitivity implies a non-consistent performance on different input data.
+However, our argument that the performance can be optimized by pre-processing the input data is confirmed in the Experiment~\hyperref[ss:exp3]{4}.
+Pre-processing determines the optimal encoding for input data and ensures the best performance of the multiset-trie on particular input data.
+For example, in the case of storing words in the multiset-trie, the search queries can always be optimized based on the frequencies of letters in a specific language.
+We also see from Experiments~\hyperref[s:exp1]{1} and~\hyperref[s:exp2]{2} that dependence of the multiset-trie performance on the density is not a linear function.
+Yet the function is continuous, and the point of inflection is unique on the whole domain as it is shown in Experiment~\hyperref[s:exp3]{3}.
+This allows us to predict whether multiset-trie can be used for some particular application, serving a high performance.
+
 %
-
 % Mathematical analysis conclusions
 %
-The mathematical analysis of the space complexity shows that multiset-trie requires 
-only $O(|M|)$ space, which is the minimal possible space that is required by any 
-data structure for storage of $|M|$ objects. As for the running time complexity of the 
-algorithms the basic tree functions such as \textsc{insert}, \textsc{search} and 
-\textsc{delete} all have a constant complexity once the multiset-trie is defined. 
-The "getAll" multiset containment functions have worst case running time complexity of 
-$O(|\mathcal{M}|),$ where $|\mathcal{M}|$ is the cardinality of the multiset-trie data structure. 
-The "existence" multiset containment functions have the worst case running time 
-complexity of $O(|\mathcal{M}| - |M|),$ where $|\mathcal{M}|$ is the cardinality of 
-the multiset-trie and $|M|$ is the number of inserted multisets (nodes on leaf level).
+The mathematical analysis of the space complexity shows that multiset-trie requires only $O(|M|)$ space, which is the minimal possible space required by any data structure for storage of $|M|$ objects.
+As for the running time complexity of algorithms, the basic tree functions such as \textsc{insert}, \textsc{search}, and \textsc{delete} all have a constant complexity once the multiset-trie is defined.
+The "getAll" multiset containment functions have worst-case running time complexity of $O(|\mathcal{M}|),$ where $|\mathcal{M}|$ is the cardinality of the multiset-trie data structure.
+The "existence" multiset containment functions have the worst-case running time complexity of $O(|\mathcal{M}| - |M|),$ where $|\mathcal{M}|$ is the cardinality of the multiset-trie and $|M|$ is the number of inserted multisets (nodes on leaf level).
 
-It can also be concluded that the multiset-trie is an input sensitive data structure, 
-because the size of multiset-trie $|\mathcal{M}|$ depends on the distribution of 
-multisets in $M.$ Our mathematical model assumes that multisets $m$ in $M$ 
-are distributed uniformly. However, in a real world models such an assumption 
-is not true in a lot of the cases. Specifically, the probability $P(m\in M)$ may vary 
-dramatically and can be even equal to $0.$ For example, if words are mapped to 
-multisets, then the sample space contains very large multisets. Nonetheless, most of 
-them will have zero probability to appear in $M$, because a word that would 
-correspond to such a multiset simply does not exist.
-%
+The multiset-trie is an input sensitive data structure because the size of multiset-trie $|\mathcal{M}|$ depends on the distribution of multisets in $M.$ Our mathematical model assumes that multisets $m$ in $M$ are distributed uniformly.
+However, in real-world models, such an assumption is not true in a lot of cases.
+Specifically, the probability $P(m\in M)$ may vary dramatically and can be even equal to $0.$ For example, if words are mapped to multisets, then the sample space contains very large multisets.
+Nonetheless, most of the words will have zero probability to appear in $M$, because a word that would correspond to such a multiset does not exist.%
 
@@
 %
-The above results have opened even more interesting questions for the future research. 
+%The above results have opened even more interesting questions for future research. 
 Further steps in our research will be to extend the functionality of the multiset-trie.
-We are interested in more flexible multiset containment queries, where the types 
-of sub and supermultisets can be specified. As an example, the multiplicity of an 
-element in a multiset can be bounded in operations getAllSubmsets and 
-getAllSupermsets. Such functionality would allow more specific queries of multisets.
+We are interested in more flexible multiset containment queries where additional conditions constrain the sub and super-multisets.
+For example, the multiplicity of an element in a multiset can be bounded in operations getAllSubmsets and getAllSupermsets.
+Furthermore, the similarity search on multisets can be implemented by modifying the algorithms for searching the sub and super-multisets.
 %
-The second line of research is to investigate the multiset-trie as an index data 
-structure in detail. It will be very interesting to study the comparison of the 
-multiset-trie with other existing index data structures.
+The second line of research is to investigate the multiset-trie as a database index data structure.
+A disk-based index data structure allows storing and managing a huge amount of multisets.
+The mapping from a multiset-trie, i.e., a $n$-ary search tree, to a block-based index can be easily defined because of the regularity of multiset-trie.
+It will be interesting to compare the multiset-trie with other existing disk-based index data structures.

### 2020-09-29 tex/tex/s_relwork.tex
first draft of section on the related work.

@@
 
-Our work is related to storage of multisets and multiset containment queries. In this section we summarize previous work done 
-for storage and retrieval of sets and multisets as well as work on (multi)set containment queries. Also, we discuss the applications 
-of multiset-trie with respect to already existing data structures.
+The data structure multiset-trie is related to the data structures and indexes designed to store and manage sets and multisets.
+We mainly focus on the related data structures and indexes that efficiently support the set and multiset containment queries.
+Firstly, we summarize our previous work done on the storage and retrieval of sets in Section \ref{rel-strie}.
+Next, we present in Section \ref{rel-invfile} the related work on the inverted files, i.e., the index structure that serves as a central data structure in the area of Information retrieval but also for storing sets and multisets in database management systems.
+The alternative to the inverted file is the signature tree that is presented in Section \ref{rel-signature}.
+Finally, we describe the related work in the area of the database management systems in Section \ref{rel-dbms}.
+We review the novel index structures used for the containment queries and the proposed containment join algorithms.
 
@@
 
-\subsection{Set-trie}
+\subsection{Set-trie\label{rel-strie}}
 % set-trie: 
@@
 
-The multiset-trie is closely related to the set-trie data structure proposed by Savnik~\cite{savnik2013index}. 
-It is a trie-based data structure that is adapted to efficient storage and retrieval of sets. In particular, 
-it stores a set of sets and supports set containment operations such as retrieval of the nearest sub- 
-and supersets as well as retrieval of all sub- and supersets from the data structure.
+The multiset-trie is closely related to the set-trie data structure introduced by Savnik in~\cite{savnik2013index}.
+It is a trie-based data structure that is adapted for the efficient storage and retrieval of sets.
+The set-trie includes the set containment operations such as retrieval of the nearest sub- and super-sets as well as retrieval of all sub- and super-sets from the sets of sets.
 
-The data structure multiset-trie generalizes the set-trie allowing to store a set of multisets. When the multiset-trie 
-is restricted to store a set of sets the underlying data structure becomes a simple binary tree. Moreover, all the 
-operations of the set-trie are also supported by the multiset-trie. The generalization is, of course, comes with a small penalty 
-in performance if we compare the multiset-trie with the set-trie. The downside of such a generalization is that multiset-trie 
-no longer supporting path compression that was obtained in set-trie. However, the design of multiset-trie provides a storage 
-of multisets as well as constant worst-case time complexity of search operation independently of user input.
+The data structure multiset-trie generalizes the set-trie allowing to store a set of multisets.
+When the multiset-trie is restricted to store a set of sets, the underlying data structure becomes a simple binary tree.
+Moreover, all the operations of the set-trie are also supported by the multiset-trie.
+The generalization comes with a small penalty in performance if we compare the multiset-trie with the set-trie in the performance of the set containment operations.
+The downside of such a generalization is that multiset-trie no longer supporting path compression that was obtained in set-trie.
+However, the design of multiset-trie provides storage of multisets with constant worst-case time complexity of the set containment operations.
 
-\subsection{Inverted file}
+\subsection{Inverted file\label{rel-invfile}}
 % inverted index:
@@
 
-The performance study~\cite{Helmer2003} shows that the inverted file is the most efficient data structure for set containment queries 
-among data structures: sequential signature files, signature trees, extendible signature hashing and inverted files. 
-In our work we replicate the experiment with inverted file and compare the performance of set containment queries between inverted file and multiset-trie data structures.
-We user the same methodology as in related paper in order to implement containment queries for inverted file.
-The results of this comparison can be found the Experiments section~\ref{s:exp5}.
+%%%
+%
+%The information retrieval (abbr.
+IR) area deals with storing and querying huge collections of text documents \cite{manning2008introduction,zobel1992efficient,zobel1998inverted,zobel2006inverted}.
+The most common model used for the representation of texts is the bag-of-words model of documents.
+A document is represented by a multiset of words that are preprocessed using techniques such as stemming, stopping, lemmatization, and others.
+After the words of the documents are preprocessed, the inverted file is constructed to support a range of query types from simple word matchings to boolean queries typical for the early IR systems.
+The inverted file is the most important part of an IR  system; it is an all-in-one data structure that serves as the structural framework for the representation of the data, and for querying the collections of texts.
+%
+%The inverted file is composed of two main parts, the dictionary, and the postings.
+In the baseline inverted file, the dictionary is used to identify, for the given words, the postings that are composed of a list of document identifiers.
+In addition, postings usually include a pointer to the location of the word inside the document and the frequencies of words in documents.
+The data structures commonly used in inverted files for the representation of a dictionary are hash tables, B-trees, and tries \cite{Terrovitis2006}.
+The dictionary is typically stored in the main memory to speed up the access to postings.
+Because of the huge amount of texts stored by the IR system, the postings have to be stored on disks \cite{zobel2006inverted}.
+%
+%Inverted files support boolean queries.
+The conjunction of two or more one-word queries retrieves documents that include both words.
+The disjunction of one-word queries includes all documents listed for one-word queries.
+Finally, the negation of a one-word query includes all documents that do not include a given word.
+The boolean expressions can be arbitrarily combined by using operations conjunction, disjunction, and negation.
+The time complexity of query evaluation is $O(N)$, where $N$ is the number of documents in a collection.
+The efficiency of the query evaluation in practical IR systems depends on the careful implementation of the inverted file, and efficient methods for query evaluation.
+Let us present some optimizations of the baseline inverted file  \cite{zobel2006inverted}.
+%
+%Since postings include the integer numbers solely, the space used for the representation of integer numbers can be significantly reduced by using variable-length encodings.
+The benefits of compression are a more efficient representation of data and faster access to postings.
+The disadvantages of compression are the need to decode the postings before they can be used in calculations.
+Next, the processing of the postings can be sped up by introducing additional structure in postings.
+A technique called \emph{skipping} organizes postings into chunks that are used to guide the search during query evaluation.
+Further, it is beneficial that the postings are \emph{sorted}, either by frequencies or by the actual impacts of words to the similarity measure.
+Only the most influential hits are interesting for further processing.
+Finally, the \emph{ordering of postings} that are linked to the words from a query is very important for query processing.
+While the efficient order of processing the postings is an optimization problem, the greedy approach that selects the shortest lists first works well \cite{Manning2008}.
+%
+%%%
+
+The inverted file \cite{zobel1992efficient,zobel1998inverted,zobel2006inverted} is the most common data structure used to represent a collection of (multi)sets.
+It is composed of two parts: a dictionary and the postings.
+Originally, in the area of Information retrieval \cite{manning2008introduction}, the inverted files are used for searching documents that contain a given set of words.
+The dictionary maps each word to a list of document identifiers together with the locations of words in documents.
+The dictionary is most often implemented by a variant of a search tree such as, for example, a B+-tree.
+The postings are implemented as a list of positions that is stored on the disk because of the huge amount of documents usually indexed by the inverted file.
+Since we have a large number of postings for a given word, the postings are compressed.
+Furthermore, several possible optimizations exist in the representation and implementation of postings \cite{zobel2006inverted}, such as sorting of postings, a technique called skipping, and some others.
+
+The empirical analyses~\cite{Helmer2003,zobel1998inverted} show that the inverted file is the most efficient data structure for containment queries among the data structures: the sequential signature file, the signature tree, the extendible signature hashing, and the inverted file.
+In our work we replicate the experiment with the inverted file and compare the performance of the containment queries between the inverted file and the multiset-trie index.
+We designed the implementation of the inverted file and the containment operations as suggested in Helmer et al.
+in \cite{Helmer2003}.
+The results of this comparison can be found in Section~\ref{s:exp5}.
 
@@
 
-\subsection{Set containment joins}
-[WIP]
+\subsection{Signature trees\label{rel-signature}}
+%%%
+%
+%An alternative to the inverted file was proposed in the area of Information retrieval by Deppisch et al.\ in the form of a dynamically balanced signature tree \cite{Deppisch1986}.
+Signatures are hash-coded binary words of a fixed length that represent abstractions of objects.
+Each object attribute value is mapped to a sequence of bits that are carefully defined to represent an abstraction of the attribute value.
+The bits representing attribute values are then glued together to form a signature.
+The main advantage of signatures used as the access paths in database systems is the possibility to express partial matching, subset queries, substring matching, and fuzzy match queries.
+%
+%In the simplest instance, the \emph{signature file} stores the signatures of objects solely.
+The search is implemented by a sequential scan of the signature file retrieving the signatures having the bits in the query set to one \cite{Pfaltz1980}.
+The search can be improved by structuring the signatures hierarchically.
+The signatures are not just the keys; they have the additional information encoded into the signatures.
+We can decode from the signature the values of particular attributes of a stored object.
+Furthermore, given a set of signatures, they can be superimposed (computed using bitwise OR operation) to obtain a representative signature of a given set of signatures.
+The signatures are therefore used to guide the search.
+Given a superimposed signature and a signature that represents an object, we can determine just from the signatures if the superimposed signature can represent a given concrete signature.
+%
+%The dynamically balanced \emph{signature tree}, S-tree, is from many aspects similar to B+-tree \cite{Deppisch1986}.
+S-tree is comprised of index pages and leaf pages.
+The index pages store lists of pairs composed of signatures and pointers to the corresponding subtrees.
+The signatures at the higher levels are constructed by superimposing the signatures from the corresponding subtrees.
+The leaf pages store pairs composed of signatures and, either objects or pointers to objects stored in some other file.
+S-tree is a dynamically balanced tree; each leaf page of S-tree has equal height.
+%
+%The main operations of S-tree are \emph{retrieve}, \emph{insert}, \emph{update} and \emph{delete}.
+The operation \emph{retrieve} follows the paths from the root to a leaf node that is determined by relating the signature to the signatures from index pages.
+The operation \emph{insert} enters the new signature in a leaf page that is determined by the new signature.
+A leaf page is split into two pages if the number of signatures exceeds the maximal number of signatures.
+Splitting a page is implemented by grouping signatures by similarity into two pages.
+The operation \emph{delete}, on the other hand, deletes the signature from the leaf page and updates the affected signatures in index pages.
+When the number of signatures on the page is lower than the minimum, then two nodes are joined into a single node.
+The operation \emph{update} changes the value of a given signature and updates the affected signatures in index pages.
+%
+%%%
+
+A dynamically balanced signature tree \cite{Deppisch1986,Pfaltz1980}, or S-tree, is an alternative data structure for the representation of multisets.
+An S-tree stores objects on the basis of their attributes represented in the form of signatures.
+A signature of an object is formed by the discretization of object attributes.
+Each attribute is discretized by mapping the attribute values to a sequence of bits.
+The bit sequences are the abstractions of the values of object attributes.
+They are glued together to form a signature of an object.
+The mappings from attribute values to sequences of bits are defined in such a way that allows superimposing a set of signatures by a single signature.
+Such a signature is often formed by using the operation OR.
+This property of signatures allows for the construction of a hierarchy of signatures that is utilized for the efficient search.
+The mutisets can be effectively represented by means of signatures, and the superposition operation can be implemented by the operation OR.
+The use of the signature tree for the containment operations was studied by Tousidou et al.
+\cite{tousidou2002sigstruc}.
+They show that S-tree that uses linear hash partitioning can be used to implement the containment operations efficiently.
+
+%Another approach to the efficient implementation of the set containment queries is the use of signature-based structures.
+Tousidou et al.
+\cite{Tousidou2002} combine the advantages of two access paths: the linear hashing and the tree-structured methods.
+They show through the empirical analysis that S-tree that uses linear hash partitioning is an efficient data structure for the subset and superset queries.
+
+%The results of the experiments presented in \cite{Helmer2003} show that the signature tree is outperformed by the inverted index for storing sets with low cardinality.
+
+%\subsection{Set containment joins}
+%[WIP]
 % set containment joins:
@@
 
-\subsection{Multisets in relational databases}
-[WIP]
+\subsection{Multisets in relational databases\label{rel-dbms}}
 % multisets in RDBMS and ORDBMS:
@@
 
-\subsection{Application of multiset-trie}
-[WIP]
+%%%
+%
+%The sets are among the important data modeling constructs in the object-relational and object-oriented database systems.
+The \emph{set-valued} attributes are used for the representation of properties that range over the sets of atomic values or objects.
+The database community has shown significant interest in indexing structures that can be used as the access paths for querying set-valued attributes \cite{Terrovitis2006,Melnik2003,Helmer2003,Tousidou2002,Zhang2001}.
+\emph{Set containment queries} were studied in the frame of different index structures.
+%
+%Zhang et al.\ \cite{Zhang2001} investigated two alternatives for the implementation of the set containment queries: a) a separate  IR engine based on the inverted lists, and, b) the native tables of the relational database management system.
+They have shown that while RDBMS are poorly suited for the set containment queries, they can outperform the inverted list engine in some conditions.
+Furthermore, they have shown that with some modifications, RDBMS can support containment queries much more efficiently.
+%
+%Another approach to the efficient implementation of the set containment queries is the use of signature-based structures.
+Tousidou et al.
+\cite{Tousidou2002} combine the advantages of two access paths: the linear hashing and the tree-structured methods.
+They show through the empirical analysis that S-tree that uses linear hash partitioning is an efficient data structure for the subset and superset queries.
+%
+%Helmer and Moercotte investigated four index structures for querying set-valued attributes of low cardinality \cite{Helmer2003}.
+All four index structures are based on conventional techniques: signatures and inverted files.
+The index structures that they compare are the sequential signature files, the signature trees, the extendable signature hashing, and B-tree based implementation of inverted lists.
+The inverted file index showed the best performance over the other data structures in most operations.
+%
+%Terrovitis et al.\ \cite{Terrovitis2011} improve the performance of the inverted files by ordering the symbols (i.e., the set elements) from the vocabulary $\Sigma$.
+The sets are in an ordered inverted file (OIF) represented by using the ordered lists of elements.
+Therefore, the lexicographical ordering of sets can be defined.
+But then, the sets can be indexed in the same way as any other collection of key/value pairs by using B+ trees.
+The postings of an inverted file are stored in separate B+ trees that are merged in a single B+ tree index.
+The set containment operations are implemented efficiently by identifying the intervals of B+ index that are mapped to a range of disk blocks containing the candidate results.
+There are similarities between the OIF and the set-trie.
+The most evident similarity is in the ordering of the elements of sets.
+However, there are differences in the use of the ordering.
+A set-trie uses the ordering of sets for the space-efficient representation of the sets of sets.
+Further, the ordering of sets is in a set-trie employed for reducing the search space to a sub-tree of a set-trie.
+The search space is similarly reduced in the OIF.
+The difference is in the way a sub-tree of a given search tree (a set-trie, or an OIF B+ tree) is defined for a given query.
+%
+%A number of join algorithms based on set containment operations are proposed \cite{Ramasamy2000,Melnik2003,Jampani2005,Luo2015}.
+The partitioning set join \cite{Ramasamy2000} relies on the signature-based representation of the sets.
+The sets are abstracted by means of signatures to provide fast set comparison operations.
+The sets are converted to signatures by converting the set elements into the bits of the signatures using the modulo function.
+The partitioning set join \cite{Ramasamy2000} splits records of the input relations by using the values of the given set-valued attributes.
+In addition, partitioning eliminates unnecessary comparisons between signatures.
+The partitioning set join algorithm is further improved to handle efficiently large sets as well as to optimize the partitioning phase of the algorithm by the adaptive design of monotonic hash functions for particular relations \cite{Melnik2003}.
+%
+%Finally, Jampani and Pudi propose the use of a prefix tree together with an inverted index for the computation of the set-based joins of two relations \cite{Jampani2005}.
+The proposed set-based joins are referred to as PRETTI joins.
+Set containment join of the table $R$ with the table $S$ is computed by constructing a prefix tree for the sets from $R$ and an inverted index for the sets from $S$.
+Starting with the root of the prefix tree constructed for $R$, the algorithm for the set containment join traverses the prefix tree in depth-first order.
+In each node $n$ of the prefix tree, the list of rids of the records from $S$ that contain a given set represented by $n$ is computed.
+This can be done recursively by intersecting rids from $S$, computed previously for a common prefix, with the rids of records from $S$ containing $n$.
+The algorithm is efficient since a single intersection of two lists is required to be computed to enumerate matching tuple pairs from $S$ for a given set from $R$.
+Besides the algorithm for the set containment join, the paper proposes the algorithms for the set overlap join and the set equality join.
+The PRETTI set-based join algorithms proposed in \cite{Jampani2005} are further improved by Luo et al.
+\cite{Luo2015} by using the Patricia trie instead of the prefix tree.
+The empirical results show that their PRETTI+ algorithm outperforms the state-of-the-art set-based joins by an order of magnitude.
+%
+%%%
+
+The index structures for the efficient implementation of the (multi)set containment queries were studied in the frame of the relational DBMS as well as the object-relational DBMS where we can use multivalued attributes including sets, multisets (bags), and lists.
+Zhang et al.\ \cite{Zhang2001} compared the performance of the containment queries implemented in a standard relational DBMS (abbr.
+RDBMS) to an Information retrieval engine.
+The results show that, in general, the IR engine performs better than a RDBMS on containment queries.
+They have identified the problems that reason for the poor performance of the containment operations in a RDBMS and showed that with some modifications, a RDBMS could perform this class of queries more efficiently.
+
+The joins in an object-relational DBMS can be defined by means of the containment operations.
+A number of \emph{containment join} methods have been proposed \cite{Ramasamy2000,Melnik2003,Jampani2005,Luo2015}.
+Ramasamy et al.
+propose the use of the partitioning set join that relies on the representation of sets by using signatures \cite{Ramasamy2000}.
+The signature-based representation allows efficient implementation of the set comparison operations.
+The partitioning set join was further improved by Melnik at al.
+\cite{Melnik2003} to handle large sets and to speed-up the partitioning phase of the algorithm.
+Further, Jampani et al.
+introduce the PRETTI join algorithm that combines an inverted file with a prefix tree for the efficient implementation of the containment joins.
+The algorithm for $R\bowtie T$ recursively computes record identifiers from $T$ while traversing a prefix tree storing the sets from $R$.
+The algorithm uses a single intersection of two lists to enumerate the matching pairs of rid-s.
+The PRETTI algorithm was improved by Luo at al.
+\cite{Luo2015} by replacing the prefix tree with the Patricia tree.
+
+%\subsection{Application of multiset-trie}
+%[WIP]
 % application of multiset-trie

### 2021-04-25 tex/tex/s_math.tex
Riste remarks 1

@@
 on $i$-th level we assign a random variable $\xi_{ij}$ to be the number of its children, 
-such that $j\in[1,\xi_i].$ Then $\forall~i\in[1,\sigma]$ the following holds:
+such that $j\in[1,\xi_i].$ Then for every $i\in[1,\sigma]$ the following holds:
 %
@@
 Let $v$ be an arbitrary node in a multiset-trie on an arbitrary level $i.$ Consider
-the sub tree with the root $v$ and call it $v$-sub tree. Since the height of the
-multiset-trie is $\sigma + 1$ we can calculate the height of the $v$-sub tree.
-Taking in account that the root node has height 1, the height of the $v$-sub tree is
+the subtree with the root $v$ and call it $v$-subtree. Since the height of the
+multiset-trie is $\sigma + 1$ we can calculate the height of the $v$-subtree.
+Taking in account that the root node has height 1, the height of the $v$-subtree is
 %
@@
 A node in a multiset-trie exists if at least one node exists on the leaf level of its
-sub tree, i.e. a node on the level $\sigma + 1$ that belongs to $v$-sub tree. The possible
-number of nodes on the leaf level of $v$-sub tree can be easily calculated knowing its height.
+subtree, i.e. a node on the level $\sigma + 1$ that belongs to $v$-subtree. The possible
+number of nodes on the leaf level of $v$-subtree can be easily calculated knowing its height.
 It is equal to
@@
 A node at level $\sigma+1$ exists with probability $p,$ where $p = P(m\in M).$
-Thus, the probability that there are no nodes on leaf level in $v$-sub tree is
+Thus, the probability that there are no nodes on leaf level in $v$-subtree is
 %
@@
 
-However, in order to determine the distribution of $\xi_{ij}$, one needs a lemma of a different type.
+However, in order to determine the distribution of $\xi_{ij}$, one needs a lemma of a different kind.
 
@@
 X_k = \begin{cases}
-0 & \textrm{child k of node $v$ does not exist} \\
-1 & \textrm{child k of node $v$ exists} \\
+0 & \textrm{child $k$ of node $v$ does not exist} \\
+1 & \textrm{child $k$ of node $v$ exists} \\
 \end{cases}
@@
 
-In order to evaluate~(\ref{eq:num_nodes}) we will use some of the tools from 
+In order to evaluate~(\cref{eq:num_nodes}) we will use some of the tools from 
 a Galton-Watson process, see Gardiner~\cite{gardiner1985stochastic} for 
 an introduction.
-Using the equations~(\ref{eq:sum_recursive}) and~(\ref{eq:generating_func}) we can 
+Using the equations~(\cref{eq:sum_recursive}) and~(\cref{eq:generating_func}) we can 
 derive the probability generating function for the random variable $\xi_{i+1}$ as 
@@
 \end{equation}
-which is the initial condition for the recursive equation~(\ref{eq:g_rec}).
+which is the initial condition for the recursive equation~(\cref{eq:g_rec}).
 
@@
 the expectation for the random variable $\xi_{i+1}$ can be derived in terms of 
-the equation~(\ref{eq:g_rec}).
+the equation~(\cref{eq:g_rec}).
 \begin{eqnarray}\label{eq:exp}
@@
 \end{eqnarray}
-According to~(\ref{eq:pdf}) and~(\ref{eq:generating_func}) the value of $G_i(z)$ at 
+According to~(\cref{eq:pdf}) and~(\cref{eq:generating_func}) the value of $G_i(z)$ at 
 $1$ is $1$ and the value of its derivative at $1$ is $\mathbb{E}(\mathcal{B}_0(n,p_{i+1})).$ 
 Substituting the values of $G_i(1^-)$ and $G_i'(1^-),$ and applying the 
-property~(\ref{eq:prop_exp-pgf}) we complete the proof.
+property~(\cref{eq:prop_exp-pgf}) we complete the proof.
 \hspace*{\fill}\qed
@@
 \begin{proof}
-According to~(\ref{eq:g_xi_one}) the expected number of nodes on the first level is 1. \\
+According to~(\cref{eq:g_xi_one}) the expected number of nodes on the first level is 1. \\
 Using $\mathbb{E}(\xi_1) = 1$ and the result from Proposition~\ref{prop:exp_rec}
@@
 Using the results from Theorem~\ref{thm:exp_level} and Theorem~\ref{thm:exp_nodes} 
-we derive the formulas~(\ref{eq:nodes_sub_param}) and~(\ref{eq:nodes_super_param}) 
+we derive the formulas~(\cref{eq:nodes_sub_param}) and~(\cref{eq:nodes_super_param}) 
 by specifying the possible number of nodes on every level in the multiset-trie according 
-to the multiset $m.$ Note that the formula~(\ref{eq:nodes_level}) assumes that on every 
+to the multiset $m.$ Note that the formula~(\cref{eq:nodes_level}) assumes that on every 
 level but the first one there are $n$ possible nodes. Given submultiset or supermultiset 

### 2021-04-25 tex/tex/s_operations.tex
Riste remarks 1

@@
 order to use it as a representation of a multiset. The type \emph{Multiset} is 
-an array $m$ of constant length $\sigma,$ where $i-th$ cell represents the element 
+an array $m$ of constant length $\sigma,$ where $i$-th cell represents the element 
 $\phi^{-1}(i)$ from $\Sigma$ with multiplicity $m[i].$ From now on we 

### 2021-09-10 tex/article_els.tex
elsevier style

--- 2020-08-05 tex/article.tex
@@
-\documentclass[twocolumn]{svjour3}
+%\documentclass[twocolumn,numbook]{svjour3}
+\documentclass[preprint]{elsarticle}
+
+%\usepackage{times}
+%\normalfont
+%\usepackage{latexsym}
+%\usepackage[T1]{fontenc}
+%\usepackage[noend]{algpseudocode}
+%\usepackage{algorithm}
+%\usepackage{verbatim}
+%\usepackage{amsfonts}    % amssymb}
+%\usepackage{amsmath}
+%\usepackage{amsthm}
+%\usepackage{stmaryrd}
+%\usepackage{wrapfig}
+
 \usepackage{style}
@@
 \usepackage{microtype}
+\usepackage[capitalise]{cleveref}
 
-% Title
-\title{Multiset-trie data structure}
-\author{Mikita Akulich \and Iztok Savnik \and Matja\v z Krnc \and Riste \v Skrekovski}
-\institute{
-		Mikita Akulich 
-		\at 
-		University of Primorska, FAMNIT, Koper, Slovenia\\
-		\email{mikita.akulich@gmail.com}
-		\and
-		Iztok Savnik
-		\at 
-		University of Primorska, FAMNIT, Koper, Slovenia\\
-		\email{iztok.savnik@famnit.upr.si}
-		\and 
-		Matja\v z Krnc 
-		\at 
-		University of Primorska, FAMNIT, Koper, Slovenia\\
-		\email{matjaz.krnc@famnit.upr.si}
-		\and 
-		Riste \v Skrekovski
-		\at
-		University of Ljubljana, FMF, Ljubljana, Slovenia\\
-		\email{riste.skrekovski@famnit.upr.si}
-		}
+%\usepackage{graphics}
+%\usepackage{graphicx}
+
+% for math enviroments
+\newtheorem{theorem}{Theorem}
+\newtheorem{lemma}{Lemma}
+\newtheorem{proof}{Proof}
+\newtheorem{proposition}{Proposition}
+\newtheorem{definition}{Definition}
+\newtheorem{corollary}{Corollary}
+%\newtheorem{ex}[thm]{Exercise}
+%\newtheorem{obs}[thm]{Observation}
+%\newtheorem{example}[thm]{Example}
+
+% oštevilčenje. To be commented out in the end
+\pagestyle{plain}
 
 \begin{document}
-\maketitle
 
-%Abstract
+\begin{frontmatter}
+\title{Multiset-trie data structure}
+%
+%\titlerunning{}  % abbreviated title (for running head)
+%                                     also used for the TOC unless
+%                                     \toctitle is used
+%
+
+\author[FAMNIT]{Mikita Akulich}
+\ead{iztok.savnik@upr.si}
+
+\author[FAMNIT,FIS]{Iztok Savnik}
+\ead{iztok.savnik@upr.si}
+
+\author[FAMNIT]{Matja\v z Krnc\corref{cor1}} 
+\ead{matjaz.krnc@upr.si}
+
+\author[FMF,FIS]{Riste \v Skrekovski}
+\ead{skrekovski@gmail.com}
+
+%
+%\authorrunning{Iztok Savnik} % abbreviated author list (for running head)
+%
+%%%% list of authors for the TOC (use if author list has to be modified)
+%\tocauthor{Iztok Savnik}
+%
+\address[FAMNIT]{Faculty of mathematics, natural sciences and information technologies, University of Primorska, Slovenia}
+
+\address[FMF]{Department of Mathematics, University of Ljubljana, Slovenia}
+
+\address[FIS]{Faculty of Information Studies, Novo Mesto, Slovenia}
+
+\cortext[cor1]{Corresponding author}
+
+%\maketitle              % typeset the title of the contribution
+
 \begin{abstract}
@@
 Finally, the comparison of the multiset-trie with the inverted index shows that the multiset-trie outperforms the inverted index by up to 4 orders of magnitude.
+\end{abstract}
 % 
-\end{abstract}
-\keywords{multiset, bag-of-words, trie, data structure, containment operations, inverted file}
-%\subclass{68P05, 68P20, 68Q87}
+\begin{keyword}
+multiset, index data structure, containment operations, inverted file
+\end{keyword}
+
+\end{frontmatter}
 
@@
 
+\newpage
+\tableofcontents
+
 \end{document}

### 2021-11-23 tex/tex/abstract.tex
before plos submission

--- 2020-08-05 tex/article.tex
@@
-\documentclass[twocolumn]{svjour3}
-\usepackage{style}
-\usepackage{subfiles}
-\usepackage{microtype}
-
-% Title
-\title{Multiset-trie data structure}
-\author{Mikita Akulich \and Iztok Savnik \and Matja\v z Krnc \and Riste \v Skrekovski}
-\institute{
-		Mikita Akulich 
-		\at 
-		University of Primorska, FAMNIT, Koper, Slovenia\\
-		\email{mikita.akulich@gmail.com}
-		\and
-		Iztok Savnik
-		\at 
-		University of Primorska, FAMNIT, Koper, Slovenia\\
-		\email{iztok.savnik@famnit.upr.si}
-		\and 
-		Matja\v z Krnc 
-		\at 
-		University of Primorska, FAMNIT, Koper, Slovenia\\
-		\email{matjaz.krnc@famnit.upr.si}
-		\and 
-		Riste \v Skrekovski
-		\at
-		University of Ljubljana, FMF, Ljubljana, Slovenia\\
-		\email{riste.skrekovski@famnit.upr.si}
-		}
-
-\begin{document}
-\maketitle
-
-%Abstract
-\begin{abstract}
 This paper proposes a new data structure \emph{multiset-trie} that is designed for storing and efficiently processing of a set of multisets.
@@
 Finally, the comparison of the multiset-trie with the inverted index shows that the multiset-trie outperforms the inverted index by up to 4 orders of magnitude.
-%
-\end{abstract}
-\keywords{multiset, bag-of-words, trie, data structure, containment operations, inverted file}
-%\subclass{68P05, 68P20, 68Q87}
-
-% section 1
-\input{tex/s_intro}
-% section 2
-\input{tex/s_data_structure}
-% section 3
-\input{tex/s_operations}
-% section 4
-\input{tex/s_math}
-% section 5
-\input{tex/s_experiments}
-% section 6
-\input{tex/s_relwork}
-% section 7
-\input{tex/s_conclusions}
-
-\bibliographystyle{abbrv}
-\bibliography{multiset-trie}
-
-\end{document}

### 2022-07-24 cover.tex
minor changes

%% LyX 2.3.6.2 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english]{letter}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{babel}
\begin{document}
Dear Editor-In-Cheaf\medskip{}

we hereby submit our manuscript ``\emph{Multiset-trie data structure}''
with coauthors M. Akulich, I. Savnik, and R. ¦krekovski, for possible
publication in PLOS One.

\closing{Yours sincerely,\\
Matja¸ Krnc}
\end{document}

### 2022-07-24 tex/tex/s_data_structure.tex
Riste remarks 1
Riste's remarks
page 5: The figure 1  ----> Figure 1
page 6: sub tree ---> subtree
page 8: submultiset vs. sub-multiset
              (Podobno za Super....)

@@
 Fix the degree of a node $n=3,$ so the maximal multiplicity of an element in 
-a multiset is $n-1=2.$ The figure~\ref{fig:sketch} presents the multiset-trie that 
+a multiset is $n-1=2.$ The Figure~\ref{fig:sketch} presents the multiset-trie that 
 contains multisets $\emptyset, \{ 1,1,2 \},$ $\{ 1,2,2 \},$ $\{ 2 \},$ $\{ 1,2 \},$ $\{ 2,2 \}.$ 
@@
 The pair $(L_1,c_j)$ is equivalent to $(L_1,root),$ since the first level has 
-the root node only. According to the figure~\ref{fig:sketch} we can extract 
+the root node only. According to Figure~\ref{fig:sketch} we can extract 
 inserted multisets as follows:

### 2022-07-24 tex/tex/s_operations.tex
Riste's remarks
page 5: The figure 1  ----> Figure 1
page 6: sub tree ---> subtree
page 8: submultiset vs. sub-multiset
              (Podobno za Super....)

@@
 Firstly we would like to describe some notations that will be used. The 
-multiset-trie data structure is a recursive data structure. Hence, any sub 
-tree of a multiset-trie $\mathcal{M}$ is again a multiset-trie. This fact allows 
+multiset-trie data structure is a recursive data structure. Hence, any subtree 
+of a multiset-trie $\mathcal{M}$ is again a multiset-trie. This fact allows 
 us to use the root node of a multiset-trie as its representative. Thus, the notation 
@@
 The function starts with searching for an exact match $x=m$ in $\mathcal{M},$ 
-since $m\subseteq m$ by definition of submultiset inclusion. If an exact match is 
+since $m\subseteq m$ by definition of sub-multiset inclusion. If an exact match is 
 not found in $\mathcal{M},$ the function uses multiset-trie to find the closest 
-(the largest) submultiset of $m$ in $\mathcal{M}$ by decreasing multiplicity of 
+(the largest) sub-multiset of $m$ in $\mathcal{M}$ by decreasing multiplicity of 
 elements in $m.$ The parameter $dev$ is used to limit a maximal deviation of 
@@
 The function \textsc{supermsetExistence}($\mathcal{M},m,dev$) checks if there 
-exists supermultiset $x$ of a given multiset $m$ in $\mathcal{M},$ such that 
+exists super-multiset $x$ of a given multiset $m$ in $\mathcal{M},$ such that 
 condition $| x[i] - m[i] |\leq dev$ is satisfied, where $1\leq i \leq \sigma.$ By analogy 
@@
 match is not found in $\mathcal{M},$ the function searches for the closest (the 
-smallest) supermultiset $x$ of $m$ in $\mathcal{M}$ by increasing multiplicity 
+smallest) super-multiset $x$ of $m$ in $\mathcal{M}$ by increasing multiplicity 
 of elements in $m.$ At every level the function tries to proceed with the smallest 
@@
 can increase multiplicity of an element up to $n-1,$ where $n$ is the degree of 
-a node in $\mathcal{M},$ to find the closest supermultiset $x\supseteq m$ in 
+a node in $\mathcal{M},$ to find the closest super-multiset $x\supseteq m$ in 
 $\mathcal{M}.$ The function \textsc{supermsetExistence} is presented in 
@@
 \textsc{submsetExistence} and \textsc{supermsetExistence} functions that do not 
-terminate on the first existing sub/supermultiset, but store the results and 
-continue procedure until all existing sub/supermultisets in $\mathcal{M}$ are 
+terminate on the first existing sub/super-multiset, but store the results and 
+continue procedure until all existing sub/super-multisets in $\mathcal{M}$ are 
 found and stored. The functions \textsc{getAllSubmsets} and \textsc{getAllSupermsets} 
@@
 each level as we traverse the tree. The variable $result$ is used as a container for storing 
-submultisets of $m$ found during traversal. Both variables $x$ and $result$ are presented 
+sub-multisets of $m$ found during traversal. Both variables $x$ and $result$ are presented 
 as global, however they could be passed to the recursive function as parameters.

### 2022-07-25 tex/tex/s_intro.tex
s_intro.tex and s_relwork.tex updated.

@@
 \section{Introduction} \label{c:introduction}
-%
-% 1st part: general intro into problem
-%
-A multiset is a data structure that represents a collection of elements.
-It generalizes a set data structure by allowing duplicate elements in a collection.
-Multisets appear in a wide variety of domains and applications.
-The index structures for storing sets of multisets were studied in the area of object-relational database systems to efficiently store, compress and query multiset-valued attributes \cite{bouros2016set,gripon2012compressing,ross2004symmetric,steinruecken2015compressing}.
-Furthermore, the need to efficiently store and query the multisets appears also in the information retrieval \cite{manning2008introduction,baeza1999modern}, the data mining \cite{mannila1997}, and in the area of expert systems \cite{forgy1982,inbook}.
-%
+
+% multisets def; not ordered
+A multiset is a collection of elements where a particular element can have more than one instance.
+As in the case of ordinary sets, the ordering of the elements in multisets is not relevant.
+For example, multisets $\{1,1,2\}$ and $\{1,2,1\}$ represent the same multiset.
+
+% motivation for the design
+Multisets appear in a wide variety of domains and applications \cite{savnik2021plos}.
+The index structures for storing sets of multisets were studied in the area of object-relational database systems to store, compress and query multiset-valued attributes efficiently \cite{bouros2016set,gripon2012compressing,ross2004symmetric,steinruecken2015compressing}.
+The need to efficiently manage multisets also appears in the information retrieval \cite{zobel1992efficient,zobel2006inverted,manning2008introduction} where texts are represented as multisets.
+In data mining, sets are often used to represent and efficiently search hypotheses in the knowledge discovery process \cite{mannila1997,flach1999aicom}.
+In the area of expert systems, multisets are used for the representation and querying of the preconditions of rules \cite{forgy1982}.
+Finally, in recent internet applications, efficient representation, and search of multisets (such as user requests and object features) have become essential \cite{bayardo2007simlar,xiao2011tods,wang2017vldb} for data cleaning, information integration, community mining, and entity resolution.
+
 % problem definition
-
-In this paper, we address the problems of storing, indexing and querying the sets of multisets.
+In this paper, we address the problems of storing, indexing, and querying the sets of multisets.
 In particular, we deal with the design of an index data structure that provides an efficient implementation of the multiset containment queries.
@@
 The existing solutions to the implementation of the containment queries include the inverted index \cite{zobel1992efficient,zobel1998inverted,broder2006indexing}, the signature tree \cite{tousidou2002sigstruc,yangjun2005stree,zobel1998inverted} and, the B+ tree~\cite{Helmer2003}.
-These solutions provide a key-value look-up, where a key is an element of a multiset and a value is a corresponding multiset implying that containment operations require additional processing of partial results such as intersection or union.
+These solutions provide a key-value look-up, where a key is an element of a multiset and a value is a corresponding multiset.
+The containment operations require additional processing of partial results such as intersection or union of multisets.
 
-% ----------------------------------------------------------------------------------------------------------------------------------------
-%
-% 2nd part: description of the proposed data structure
-%
-To improve the efficiency of containment operations, we propose a data structure \emph{multiset-trie} that is designed for storing and processing of a finite bounded set of multisets.
-It generalizes the \emph{set-trie} data structure proposed by Savnik~\cite{savnik2013index} that was designed for storage and processing of a finite bounded set of sets.
-%
-The set-trie is an extension of a trie data structure to provide fast search and retrieval of sets and, to implement efficiently the set containment operations.
+% description of the proposed data structure
+To improve the efficiency of containment operations, we propose a data structure \emph{multiset-trie} that is designed for storing and processing a finite bounded set of multisets.
+It generalizes the \emph{set-trie} data structure proposed by Savnik~\cite{savnik2013index,savnik2021plos} that was designed for storage and processing of a finite bounded set of sets.
+The set-trie is an extension of a trie data structure to provide fast search and retrieval of sets and to implement the set containment operations efficiently.
 The set-trie is also a form of the binomial tree \cite{corman2001}.
+
 Multiset-trie generalizes the set-trie data structure.
-It provides a space efficient representation of a set of multisets and, it provides efficient multiset-containment operations.
-In addition, it can also represent and query sets of sets.
+It provides a space-efficient representation of a set of multisets and efficient multiset containment operations.
+As in the case of set-trie, the ordering of the elements from the alphabet is not important for the representation of multisets.
+The efficiency of the multiset containment operations is the consequence of selecting the specific ordering of multiset elements.
+The ordering can be exploited for the efficient search in a multiset-trie.
 
@@
 Each multiset in a multiset-trie is represented by a path from the root to a leaf node.
-Multiset elements are the symbols from an alpahabet.
-Each symbol from the alphabet is represented by a node at a certan level of the multiset-trie.
+Multiset elements are symbols from an alphabet.
+Each symbol from the alphabet is represented by a node at a certain level of the multiset-trie.
 The node stores the multiplicity of an element in a multiset.
-%
-%Multisets are represented with a collection of nodes in a tree such that every node represents an element of a multiset with particular multiplicity bounded by a node degree $n.$
-%
+
 A multiset-trie is also a kind of search tree.
-Similarly to a trie, it uses common prefixes for a shared data representation.
+Similar to a trie, it uses common prefixes for a shared data representation.
 Unlike the compact prefix tree \cite{Sedgewick:2011:ALG:2011916}, the multiset-trie does not support path compression.
 However, the absence of path compression makes the multiset-trie a perfectly height-balanced tree.
-Moreover, when multiset-trie is full it forms a complete $n$-ary tree.
-%
-%The multiset-trie handles multisets directly by having access to each of the element without the need to reconstruct them for processing, which allows fast retrieval and containment operations.
-In particular, it supports sub-multiset and super-multiset queries.
-The operations allow to find and retrieve the closest submultisets or super-multisets as well as to find and retrieve all of them.
+Moreover, when multiset-trie is full, it forms a complete $n$-ary tree.
 
-% 
-%--------------------------------------------------------------------------------------------------------------------------------------------
-%
-% 3rd part: contributions
 \subsection{Contributions}
@@
 Furthermore, the contributions of this work are thorough mathematical and empirical analyses of the proposed data structure multiset-trie.
-%
+
 A mathematical model is developed to analyze the complexity of multiset containment operations.
-In particular, we estimate the size of a sub-tree traversed by a containment query and, therefore, give an insight into the time complexity of containment queries.
+In particular, we estimate the size of a sub-tree traversed by a containment query and give an insight into the time complexity of containment queries.
 In addition to the mathematical analysis, the size of a sub-tree visited by a multiset containment query is also the central focus of the empirical analysis.
 We carefully designed the experiments to unravel the main features of the time complexity space.
-%
-Finally, a comparative study between the inverted index and the multiset-trie shows that the multiset-trie outperforms the inverted index by 4 levels of magnitude in the time consumed by both equality and containment queries.
-%
-%Experimental result that shows an influence of ordering in multisets on their storage and processing.
-%
-%
-% --------------------------------------------------------------------------------------------------------------------------------------------
-%
-% 4th part: paper organization
+
+Finally, the mathematical as well as empirical analyses show an influence of ordering the elements of multisets on the efficiency of storing and processing multisets.
+We show that the ordering which is based on the frequencies of the multiset elements can speed-up the multiset containment queries by orders of magnitude.
+
 \subsection{Paper organization}
@@
 The paper is organized as follows. In the following Section~\ref{c:description}, we present the multiset-trie data structure in detail.
-%
 Section~\ref{c:operations} presents the operations of the multiset-trie.
-These include the basic operations of search trees and, the multiset containment operations.
+These include the basic operations of search trees and multiset containment operations.
 The algorithms are presented in detail using the pseudo code.
-%
+
 The description of multiset-trie operations is followed by the mathematical analysis of their complexity in Section~\ref{c:analysis}.
@@
 By using probabilistic tools we describe time complexity of the algorithms and space complexity of the structure.
-%
-Further, in Section~\ref{c:experiments} we present an empirical study of the multiset-trie.
-Synthetic and real-world data sets are used in experiments that are designed to study the performance of the multiset containment operations.
-The multiset-trie is empirically compared to the inverted index in a separate experiment.
-% The experiments highlight the methods for optimizing a multiset-trie.
-%
-The related work is reviewed in Section~\ref{c:relwork}. 
-%
-Finally, we conclude with the discussion of future work in Section~\ref{c:conclusions}.
-%
+
+In Section~\ref{c:experiments} we present an empirical study of the multiset-trie.
+Synthetic data sets are used in experiments that are designed to study the performance of the multiset containment operations.
+The experiments highlight the methods for optimizing a multiset-trie.
+
+The related work is reviewed in Section~\ref{c:relwork}. We conclude with the discussion of future work in Section~\ref{c:conclusions}.

### 2022-07-25 tex/tex/s_intro.tex
update of multiset-trie.bib.

@@
 
-% problem definition
+%\subsubsection{Problem definition.}
 In this paper, we address the problems of storing, indexing, and querying the sets of multisets.
@@
 % description of the proposed data structure
+%\subsubsection{Proposed index data structure.}
 To improve the efficiency of containment operations, we propose a data structure \emph{multiset-trie} that is designed for storing and processing a finite bounded set of multisets.
@@
 
-\subsection{Contributions}
-
+%\subsubsection{Contributions.}
 The main contribution of this paper is a novel data structure \emph{multiset-trie} for storing and querying a set of multisets that provides efficient multiset containment operations.
@@
 
-\subsection{Paper organization}
-
+%\subsubsection{Paper organization.}
 The paper is organized as follows. In the following Section~\ref{c:description}, we present the multiset-trie data structure in detail.
@@
 
-The related work is reviewed in Section~\ref{c:relwork}. We conclude with the discussion of future work in Section~\ref{c:conclusions}.
+The related work is reviewed in Section~\ref{c:relwork}.
+We present the related approaches including a description of a set-trie data structure, the inverted indexes, the signature indexes, and the multisets in relational database systems.
+Finally, the concluding remarks and the future work are presented in Section~\ref{c:conclusions}.

### 2022-07-25 tex/tex/s_relwork.tex
update of multiset-trie.bib.

--- 2019-06-19 tex/tex/s_relwork.tex
@@
 % intro to related work
-
-Our work is related to storage of multisets and multiset containment queries. In this section we summarize previous work done 
-for storage and retrieval of sets and multisets as well as work on (multi)set containment queries. Also, we discuss the applications 
-of multiset-trie with respect to already existing data structures.
+The data structure multiset-trie is related to the data structures and indexes designed to store and manage sets and multisets.
+We mainly focus on the related data structures and indexes that efficiently support the set and multiset containment queries.
+Firstly, we summarize our previous work done on the storage and retrieval of sets in Section \ref{rel-strie}.
+Next, we present in Section \ref{rel-invfile} the related work on the inverted files, i.e., the index structure that serves as a central data structure in the area of Information retrieval but also for storing sets and multisets in database management systems.
+The alternative to the inverted file is the signature tree that is presented in Section \ref{rel-signature}.
+Finally, we describe the related work in the area of the database management systems in Section \ref{rel-dbms}.
+We review the novel index structures used for the containment queries and the proposed containment join algorithms.
 
@@
 
-The multiset-trie is closely related to the set-trie data structure proposed by Savnik~\cite{savnik2013index}. 
-It is a trie-based data structure that is adapted to efficient storage and retrieval of sets. In particular, 
-it stores a set of sets and supports set containment operations such as retrieval of the nearest sub- 
-and supersets as well as retrieval of all sub- and supersets from the data structure.
+The multiset-trie is closely related to the set-trie data structure introduced by Savnik in~\cite{savnik2013index,savnik2021plos}.
+A set-trie is a trie data structure that is adapted for the efficient storage and retrieval of sets instead of the sequences of symbols.
+The set-trie provides the set containment operations such as retrieval of the nearest sub- and super-sets as well as retrieval of all sub- and super-sets from the sets of sets.
 
-The data structure multiset-trie generalizes the set-trie allowing to store a set of multisets. When the multiset-trie 
-is restricted to store a set of sets the underlying data structure becomes a simple binary tree. Moreover, all the 
-operations of the set-trie are also supported by the multiset-trie. The generalization is, of course, comes with a small penalty 
-in performance if we compare the multiset-trie with the set-trie. The downside of such a generalization is that multiset-trie 
-no longer supporting path compression that was obtained in set-trie. However, the design of multiset-trie provides a storage 
-of multisets as well as constant worst-case time complexity of search operation independently of user input.
+Since we are storing sets where each element of the set can appear only once, and the ordering of elements is not important, the ordering of the elements from the alphabet can be used for guiding the search in set containment operations.
+Each set is represented in a set-trie by a path including the increasing elements of a set represented by set-trie nodes.
+Since all sets from a set-trie are ordered by the increasing value of the set elements, the children of each set-trie node $n$ can only be the elements larger than the element $n$.
+For a given set $s$ and a set-trie $S$, the set containment operations search solely a sub-tree of $S$ that includes all the sets (paths from a root to a set-trie node) that are the possible subsets or supersets of $s$.
+%The size of such sub-tree depends on the size of $S$ and the shape of $s$.
 
-\subsection{Inverted file}
+The data structure multiset-trie generalizes the set-trie allowing to store a set of multisets.
+When the multiset-trie is restricted to store a set of sets, the underlying data structure becomes a simple binary tree.
+Moreover, all the operations of the set-trie are also supported by the multiset-trie.
+The generalization comes with a small penalty in performance if we compare the multiset-trie with the set-trie in the performance of the set containment operations.
+The downside of such a generalization is that multiset-trie no longer supporting path compression that was obtained in set-trie.
+However, the design of multiset-trie provides storage of multisets with constant worst-case time complexity of the set containment operations.
+
+\subsection{Inverted file\label{rel-invfile}}
 % inverted index:
@@
 
-The performance study~\cite{Helmer2003} shows that the inverted file is the most efficient data structure for set containment queries 
-among data structures: sequential signature files, signature trees, extendible signature hashing and inverted files. 
-In our work we replicate the experiment with inverted file and compare the performance of set containment queries between inverted file and multiset-trie data structures.
-We user the same methodology as in related paper in order to implement containment queries for inverted file.
-The results of this comparison can be found the Experiments section~\ref{s:exp5}.
+The inverted file \cite{zobel1992efficient,zobel1998inverted,zobel2006inverted} is the most common data structure used to represent a collection of (multi)sets.
+It is composed of two parts: a dictionary and the postings.
+Originally, in the area of Information retrieval \cite{manning2008introduction}, the inverted files are used for searching documents that contain a given set of words.
+The dictionary maps each word to a list of document identifiers together with the locations of words in documents.
+The dictionary is most often implemented by a variant of a search tree such as, for example, a B+-tree.
+The postings are implemented as a list of positions that is stored on the disk because of the huge amount of documents usually indexed by the inverted file.
+Since we have a large number of postings for a given word, the postings are compressed.
+Furthermore, several possible optimizations exist in the representation and implementation of postings \cite{zobel2006inverted}, such as sorting of postings, a technique called skipping, and some others.
+
+The empirical analyses~\cite{Helmer2003,zobel1998inverted} show that the inverted file is the most efficient data structure for containment queries among the data structures: the sequential signature file, the signature tree, the extendible signature hashing, and the inverted file.
+In our work we replicate the experiment with the inverted file and compare the performance of the containment queries between the inverted file and the multiset-trie index.
+We designed the implementation of the inverted file and the containment operations as suggested in Helmer et al.
+in \cite{Helmer2003}.
+The results of this comparison can be found in Section~\ref{s:exp5}.
 
@@
 
-\subsection{Set containment joins}
-[WIP]
-% set containment joins:
-% - review the paper: Set containment join revisited\cite{bouros2016set}
-% - multiset containment join
+\subsection{Signature trees\label{rel-signature}}
 
-\subsection{Multisets in relational databases}
-[WIP]
+A dynamically balanced signature tree \cite{Deppisch1986,Pfaltz1980}, or S-tree, is an alternative data structure for the representation of multisets.
+An S-tree stores objects on the basis of their attributes represented in the form of signatures.
+A signature of an object is formed by the discretization of object attributes.
+Each attribute is discretized by mapping the attribute values to a sequence of bits.
+The bit sequences are the abstractions of the values of object attributes.
+They are glued together to form a signature of an object.
+The mappings from attribute values to sequences of bits are defined in such a way that allows superimposing a set of signatures by a single signature.
+Such a signature is often formed by using the operation OR.
+This property of signatures allows for the construction of a hierarchy of signatures that is utilized for the efficient search.
+The mutisets can be effectively represented by means of signatures, and the superposition operation can be implemented by the operation OR.
+The use of the signature tree for the containment operations was studied by Tousidou et al.
+\cite{tousidou2002sigstruc}.
+They show that S-tree that uses linear hash partitioning can be used to implement the containment operations efficiently.
+
+\subsection{Multisets in relational databases\label{rel-dbms}}
 % multisets in RDBMS and ORDBMS:
@@
 
-\subsection{Application of multiset-trie}
-[WIP]
-% application of multiset-trie
-% - indexing of multisets
-% - containment queries
-% - multiset containment join
-% - information retrieval
+The index structures for the efficient implementation of the (multi)set containment queries were studied in the frame of the relational DBMS as well as the object-relational DBMS where we can use multivalued attributes including sets, multisets (bags), and lists.
+Zhang et al.\ \cite{Zhang2001} compared the performance of the containment queries implemented in a standard relational DBMS (abbr.
+RDBMS) to an Information retrieval engine.
+The results show that, in general, the IR engine performs better than a RDBMS on containment queries.
+They have identified the problems that reason for the poor performance of the containment operations in a RDBMS and showed that with some modifications, a RDBMS could perform this class of queries more efficiently.
 
-%
-% Set theory, multiest theory, bag-of-words model
-%\subsection{Multiset}
-%Multiset is a widely used data structure in different areas of mathematics, physics 
-%and computer science \cite{singh2007overview}. The theory of multisets is based 
-%entirely on the theory of sets. However, classical mathematics does not deal with 
-%multisets directly. Instead, one can define a multiset to be a \emph{family} of sets 
-%or the functions on ordered pairs, where the members of a pair are an element and its 
-%multiplicity. This means that mathematically the concepts of set such as cardinality, 
-%set-containment operation, power set, equivalence classes and others are well defined 
-%for multisets in terms of sets \cite{blizard1988multiset}. 
-%
-%The concept of a multiset can also be referred to the \emph{bag-of-words model}. 
-%This model takes its origin from a linguistic context studied by Harris~\cite{harris1954distributional}. 
-%According to the bag-of-words model, text can be represented as a bag (multiset) 
-%of words, where an element is a word and the number of its occurrences in the 
-%text is multiplicity. A bag of words does not keep track of grammar and ordering 
-%of words.
-
-% Information retrieval
-%\subsection{Information retrieval}
-%\emph{Information retrieval (IR)} refers to a problem of finding material of an 
-%unstructured nature that satisfies an information need~\cite{manning2008introduction}. 
-%Usually, one is searching for a specific documents in a significantly large text 
-%documents database. The size of a database makes the search a time consuming 
-%operation. In order to resolve the issue IR systems pre-process data and create 
-%indexes for future use in search operation. 
-%
-%The bag-of-words model is widely used in IR. In particular, such a representation of 
-%text documents is used in database indexes when a full text search of a database 
-%is required. The full-text search problem refers to indexing techniques for full-text 
-%databases. The most efficient index nowadays uses the concept of an inverted index 
-%\cite{zobel1992efficient}. 
-%
-%The proposed data structure multiset-trie can be used as an alternative implementation 
-%of the search structure of an inverted index. It represents words as multisets and 
-%stores them into data structure. The query processing is achieved using boolean 
-%retrieval model \cite{manning2008introduction} and multiset containment operations. 
-%Multiset containment operations of the multiset-trie implement the nearest neighbor 
-%search queries which retrieve not only exact but also the most relevant results 
-%to a user.
-
-%\subsection{Generalized search tree}
-%The properties and operations of the multiset-trie makes it a competitor to the 
-%most efficient implementation of a search tree the \emph{Generalized Search Tree 
-%(GiST)}~\cite{broder2006indexing,hellerstein1995generalized,kornacker1999high} 
-%that is used in inverted index. GiST is a very flexible data structure that can be 
-%customized in order to behave like B+-tree, R-tree or RD-tree. It also provides 
-%support for an extensible set of queries and data types that B+-tree, R-tree or RD-tree 
-%do not support originally. GiST supports all the basic search tree operations such as 
-%insert, delete and search, and in addition provides such extensions as the 
-%nearest-neighbor search and multiset containment operations. The extensions 
-%provided by GiST are native in the multiset-trie. The multiset-trie also has a fixed 
-%height while GiST is a self-balanced tree and has to use additional methods in 
-%order to preserve its balance.
+The joins in an object-relational DBMS can be defined by means of the containment operations.
+A number of \emph{containment join} methods have been proposed \cite{Ramasamy2000,Melnik2003,Jampani2005,Luo2015}.
+Ramasamy et al.
+propose the use of the partitioning set join that relies on the representation of sets by using signatures \cite{Ramasamy2000}.
+The signature-based representation allows efficient implementation of the set comparison operations.
+The partitioning set join was further improved by Melnik at al.
+\cite{Melnik2003} to handle large sets and to speed-up the partitioning phase of the algorithm.
+Further, Jampani et al.
+introduce the PRETTI join algorithm that combines an inverted file with a prefix tree for the efficient implementation of the containment joins.
+The algorithm for $R\bowtie T$ recursively computes record identifiers from $T$ while traversing a prefix tree storing the sets from $R$.
+The algorithm uses a single intersection of two lists to enumerate the matching pairs of rid-s.
+The PRETTI algorithm was improved by Luo at al.
+\cite{Luo2015} by replacing the prefix tree with the Patricia tree.

### 2022-07-25 tex/tex/abstract.tex
first draft (of my corrections).

@@
-This paper proposes a new data structure \emph{multiset-trie} that is designed for storing and efficiently processing of a set of multisets.
+This paper proposes a new data structure \emph{multiset-trie} that is designed for storing and efficiently processing a set of multisets.
 Moreover, multiset-trie can operate on a set of sets without efficiency loss.
@@
 The experiments illuminate the time complexity space of the multiset containment operations.
-Finally, the comparison of the multiset-trie with the inverted index shows that the multiset-trie outperforms the inverted index by up to 4 orders of magnitude.
+%Finally, the comparison of the multiset-trie with the inverted index shows that the multiset-trie outperforms the inverted index by up to 4 orders of magnitude.

### 2022-07-25 tex/tex/s_intro.tex
first draft (of my corrections).

@@
 
-%\subsubsection{Problem definition.}
+%\medskip\noindent\textbf{Problem definition.}\;
 In this paper, we address the problems of storing, indexing, and querying the sets of multisets.
@@
 
-% description of the proposed data structure
-%\subsubsection{Proposed index data structure.}
+%\medskip\noindent\textbf{Proposed data structure.}\;
 To improve the efficiency of containment operations, we propose a data structure \emph{multiset-trie} that is designed for storing and processing a finite bounded set of multisets.
-It generalizes the \emph{set-trie} data structure proposed by Savnik~\cite{savnik2013index,savnik2021plos} that was designed for storage and processing of a finite bounded set of sets.
+A multiset-trie generalizes the \emph{set-trie} data structure proposed by Savnik~\cite{savnik2013index,savnik2021plos} that was designed for storage and processing of a finite bounded set of sets.
 The set-trie is an extension of a trie data structure to provide fast search and retrieval of sets and to implement the set containment operations efficiently.
@@
 
-Multiset-trie generalizes the set-trie data structure.
-It provides a space-efficient representation of a set of multisets and efficient multiset containment operations.
+Multiset-trie provides a space-efficient representation of a set of multisets and efficient multiset containment operations.
 As in the case of set-trie, the ordering of the elements from the alphabet is not important for the representation of multisets.
@@
 
-%\subsubsection{Contributions.}
-The main contribution of this paper is a novel data structure \emph{multiset-trie} for storing and querying a set of multisets that provides efficient multiset containment operations.
+\medskip\noindent\textbf{Contributions.}\;
+The main contributions of this paper are as follows.
+First, a  multiset-trie is a novel data structure \emph{multiset-trie} for storing and querying a set of multisets that provides efficient multiset containment operations.
 Furthermore, the contributions of this work are thorough mathematical and empirical analyses of the proposed data structure multiset-trie.
 
-A mathematical model is developed to analyze the complexity of multiset containment operations.
+Second, a mathematical model is developed to analyze the complexity of multiset containment operations.
 In particular, we estimate the size of a sub-tree traversed by a containment query and give an insight into the time complexity of containment queries.
@@
 
-%\subsubsection{Paper organization.}
+\medskip\noindent\textbf{Paper organization.}\; 
 The paper is organized as follows. In the following Section~\ref{c:description}, we present the multiset-trie data structure in detail.

### 2022-07-25 tex/tex/s_relwork.tex
first draft (of my corrections).

@@
 The empirical analyses~\cite{Helmer2003,zobel1998inverted} show that the inverted file is the most efficient data structure for containment queries among the data structures: the sequential signature file, the signature tree, the extendible signature hashing, and the inverted file.
-In our work we replicate the experiment with the inverted file and compare the performance of the containment queries between the inverted file and the multiset-trie index.
-We designed the implementation of the inverted file and the containment operations as suggested in Helmer et al.
-in \cite{Helmer2003}.
-The results of this comparison can be found in Section~\ref{s:exp5}.
 

### 2022-07-26 tex/tex/s_intro.tex
s_intro updated.

@@
 In the area of expert systems, multisets are used for the representation and querying of the preconditions of rules \cite{forgy1982}.
-Finally, in recent internet applications, efficient representation, and search of multisets (such as user requests and object features) have become essential \cite{bayardo2007simlar,xiao2011tods,wang2017vldb} for data cleaning, information integration, community mining, and entity resolution.
+Finally, in recent internet applications, efficient representation and search of multisets (such as user requests and object features) have become essential \cite{bayardo2007simlar,xiao2011tods,wang2017vldb} for data cleaning, information integration, community mining, and entity resolution.
 
@@
 
-The existing solutions to the implementation of the containment queries include the inverted index \cite{zobel1992efficient,zobel1998inverted,broder2006indexing}, the signature tree \cite{tousidou2002sigstruc,yangjun2005stree,zobel1998inverted} and, the B+ tree~\cite{Helmer2003}.
+The existing solutions to the implementation of the containment queries use the inverted indexes \cite{zobel1992efficient,zobel1998inverted,zobel2006csur,broder2006indexing,terrovitis2006cikm,terrovitis2011icdt}, the signature trees \cite{deppisch1986sigir,tousidou2002sigstruc,yangjun2005stree,zobel1998inverted} and, the B+ trees~\cite{Helmer2003,terrovitis2011icdt}.
 These solutions provide a key-value look-up, where a key is an element of a multiset and a value is a corresponding multiset.
-The containment operations require additional processing of partial results such as intersection or union of multisets.
+The containment operations require multiple key-value index accesses and also additional processing of partial results such as intersection or union of multisets.
 
@@
 A multiset-trie generalizes the \emph{set-trie} data structure proposed by Savnik~\cite{savnik2013index,savnik2021plos} that was designed for storage and processing of a finite bounded set of sets.
-The set-trie is an extension of a trie data structure to provide fast search and retrieval of sets and to implement the set containment operations efficiently.
+The set-trie is an extension of a trie data structure to provide, besides fast search and retrieval of sets, also the efficient implementation of the set containment operations.
 The set-trie is also a form of the binomial tree \cite{corman2001}.
@@
 Multiset-trie provides a space-efficient representation of a set of multisets and efficient multiset containment operations.
-As in the case of set-trie, the ordering of the elements from the alphabet is not important for the representation of multisets.
-The efficiency of the multiset containment operations is the consequence of selecting the specific ordering of multiset elements.
-The ordering can be exploited for the efficient search in a multiset-trie.
+As in the case of set-trie, the ordering of the elements in multiset-trie is not relevant for the representation of multisets.
+As a consequence, the efficiency of the multiset containment operations is obtained by selecting the \emph{specific} ordering of multiset elements.
+This ordering can be exploited for the efficient search in a multiset-trie.
 
@@
 The main contributions of this paper are as follows.
-First, a  multiset-trie is a novel data structure \emph{multiset-trie} for storing and querying a set of multisets that provides efficient multiset containment operations.
+First, a  multiset-trie is a novel data structure for storing and querying a set of multisets that provides efficient multiset containment operations.
 Furthermore, the contributions of this work are thorough mathematical and empirical analyses of the proposed data structure multiset-trie.
@@
 We carefully designed the experiments to unravel the main features of the time complexity space.
+%%% relats btwn math and emprcl anlyss
+%%% concrete results? bounds, ...
+%%% experiments confirm findings of math analysis?
 
 Finally, the mathematical as well as empirical analyses show an influence of ordering the elements of multisets on the efficiency of storing and processing multisets.
-We show that the ordering which is based on the frequencies of the multiset elements can speed-up the multiset containment queries by orders of magnitude.
+We show that the ordering, which is based on the frequencies of the multiset elements, can speed up the multiset containment queries by orders of magnitude.
 
@@
 These include the basic operations of search trees and multiset containment operations.
-The algorithms are presented in detail using the pseudo code.
+The algorithms are presented in detail using pseudo code.
 
@@
 The main assumption is that multisets are constructed uniformly at random with bounded cardinality.
-By using probabilistic tools we describe time complexity of the algorithms and space complexity of the structure.
+By using probabilistic tools, we describe the time complexity of the algorithms and the space complexity of the structure.
 
 In Section~\ref{c:experiments} we present an empirical study of the multiset-trie.
-Synthetic data sets are used in experiments that are designed to study the performance of the multiset containment operations.
+Synthetic and real-world data sets are used in experiments that are designed to study the performance of multiset containment operations.
 The experiments highlight the methods for optimizing a multiset-trie.
@@
 The related work is reviewed in Section~\ref{c:relwork}.
-We present the related approaches including a description of a set-trie data structure, the inverted indexes, the signature indexes, and the multisets in relational database systems.
+We present the related approaches, including a description of a set-trie data structure, the inverted indexes, the signature indexes, and the multisets in relational database systems.
 Finally, the concluding remarks and the future work are presented in Section~\ref{c:conclusions}.

### 2022-07-27 tex/tex/s_relwork.tex
English checked in s_relwork.

@@
 We mainly focus on the related data structures and indexes that efficiently support the set and multiset containment queries.
-Firstly, we summarize our previous work done on the storage and retrieval of sets in Section \ref{rel-strie}.
-Next, we present in Section \ref{rel-invfile} the related work on the inverted files, i.e., the index structure that serves as a central data structure in the area of Information retrieval but also for storing sets and multisets in database management systems.
+Firstly, we summarize our previous work on the data structure for managing sets in Section \ref{rel-strie}.
+Next, we present in Section \ref{rel-invfile} the related work on the inverted files, i.e., the index structure that serves as a central data structure in the area of Information Retrieval (abbr.
+IR) but also for storing sets and multisets in database management systems.
 The alternative to the inverted file is the signature tree that is presented in Section \ref{rel-signature}.
-Finally, we describe the related work in the area of the database management systems in Section \ref{rel-dbms}.
+Finally, we describe the related work in the area of database management systems in Section \ref{rel-dbms}.
 We review the novel index structures used for the containment queries and the proposed containment join algorithms.
@@
 A set-trie is a trie data structure that is adapted for the efficient storage and retrieval of sets instead of the sequences of symbols.
-The set-trie provides the set containment operations such as retrieval of the nearest sub- and super-sets as well as retrieval of all sub- and super-sets from the sets of sets.
+The set-trie provides the set containment operations such as retrieval of the \emph{nearest} subset or supersets as well as retrieval of \emph{all} subsets and supersets from the sets of sets.
 
@@
 Since all sets from a set-trie are ordered by the increasing value of the set elements, the children of each set-trie node $n$ can only be the elements larger than the element $n$.
-For a given set $s$ and a set-trie $S$, the set containment operations search solely a sub-tree of $S$ that includes all the sets (paths from a root to a set-trie node) that are the possible subsets or supersets of $s$.
+For a given set $s$ and a set-trie $S$, the set containment operations search solely the sub-tree of $S$ that includes all the sets (paths from a root to a set-trie node) that are the possible subsets or supersets of $s$.
 %The size of such sub-tree depends on the size of $S$ and the shape of $s$.
 
-The data structure multiset-trie generalizes the set-trie allowing to store a set of multisets.
+The data structure multiset-trie generalizes the set-trie by providing storage for the set of multisets.
 When the multiset-trie is restricted to store a set of sets, the underlying data structure becomes a simple binary tree.
@@
 The generalization comes with a small penalty in performance if we compare the multiset-trie with the set-trie in the performance of the set containment operations.
-The downside of such a generalization is that multiset-trie no longer supporting path compression that was obtained in set-trie.
+The downside of such a generalization is that multiset-trie no longer supports path compression that was obtained in set-trie.
 However, the design of multiset-trie provides storage of multisets with constant worst-case time complexity of the set containment operations.
@@
 The inverted file \cite{zobel1992efficient,zobel1998inverted,zobel2006inverted} is the most common data structure used to represent a collection of (multi)sets.
+In the area of IR \cite{manning2008introduction} the inverted files are used for searching documents that contain a given set of words.
 It is composed of two parts: a dictionary and the postings.
-Originally, in the area of Information retrieval \cite{manning2008introduction}, the inverted files are used for searching documents that contain a given set of words.
 The dictionary maps each word to a list of document identifiers together with the locations of words in documents.
-The dictionary is most often implemented by a variant of a search tree such as, for example, a B+-tree.
-The postings are implemented as a list of positions that is stored on the disk because of the huge amount of documents usually indexed by the inverted file.
-Since we have a large number of postings for a given word, the postings are compressed.
-Furthermore, several possible optimizations exist in the representation and implementation of postings \cite{zobel2006inverted}, such as sorting of postings, a technique called skipping, and some others.
+The dictionary is most often implemented by a variant of a search tree, such as a B+ tree.
+The postings are implemented as a list of positions that are stored on the disk because of the huge amount of documents usually indexed by the inverted file.
+Since we can have a large number of postings for one word, the postings are compressed.
+Furthermore, several possible optimizations exist in the representation and implementation of postings \cite{zobel2006inverted}, such as sorting of postings, a technique called skipping, and others.
 
@@
 Such a signature is often formed by using the operation OR.
-This property of signatures allows for the construction of a hierarchy of signatures that is utilized for the efficient search.
-The mutisets can be effectively represented by means of signatures, and the superposition operation can be implemented by the operation OR.
+This property of signatures provides the means for the construction of the hierarchy of signatures that is utilized for efficient search.
+The multisets can be effectively represented using signatures, and the superposition operation can be implemented by the operation OR.
 The use of the signature tree for the containment operations was studied by Tousidou et al.
@@
 
-The index structures for the efficient implementation of the (multi)set containment queries were studied in the frame of the relational DBMS as well as the object-relational DBMS where we can use multivalued attributes including sets, multisets (bags), and lists.
+The index structures for the efficient implementation of the (multi)set containment queries were studied in the frame of the relational DBMS as well as the object-relational DBMS, where we can use multivalued attributes, including sets, multisets (bags), and lists.
 Zhang et al.\ \cite{Zhang2001} compared the performance of the containment queries implemented in a standard relational DBMS (abbr.
 RDBMS) to an Information retrieval engine.
-The results show that, in general, the IR engine performs better than a RDBMS on containment queries.
-They have identified the problems that reason for the poor performance of the containment operations in a RDBMS and showed that with some modifications, a RDBMS could perform this class of queries more efficiently.
+The results show that, in general, the IR engine performs better than an RDBMS on containment queries.
+They have identified the problems that reason for the poor performance of the containment operations in an RDBMS and showed that with some modifications, an RDBMS could perform this class of queries more efficiently.
 
@@
 The partitioning set join was further improved by Melnik at al.
-\cite{Melnik2003} to handle large sets and to speed-up the partitioning phase of the algorithm.
+\cite{Melnik2003} to handle large sets and to speed up the partitioning phase of the algorithm.
 Further, Jampani et al.

### 2022-07-29 tex/tex/s_data_structure.tex
Riste remarks (from the bus)

@@
 
-Let us have an example of a multiset-trie data structure. Let $\sigma = 2$ and 
-$\Sigma = I = \{ 1,2 \}$ respectively, so the mapping $\phi$ is an identity mapping. 
-Fix the degree of a node $n=3,$ so the maximal multiplicity of an element in 
-a multiset is $n-1=2.$ The Figure~\ref{fig:sketch} presents the multiset-trie that 
-contains multisets $\emptyset, \{ 1,1,2 \},$ $\{ 1,2,2 \},$ $\{ 2 \},$ $\{ 1,2 \},$ $\{ 2,2 \}.$ 
-The \emph{Null} children are omitted on the figure.
+An example of a multiset-trie data structure with $\sigma = 2$ and 
+$\Sigma = I = \{ 1,2 \}$ (i.e. the mapping $\phi$ is an identity mapping) is shown on Figure~\ref{fig:sketch}. 
+On the figure, which stores elements of $\{\emptyset, \{ 1,1,2 \},$ $\{ 1,2,2 \},$ $\{ 2 \},$ $\{ 1,2 \},$ $\{ 2,2 \}\}$, the degree of a node is set to be $n=3,$ so the maximal multiplicity of an element in
+a multiset is $n-1=2.$
 
@@
 \end{tikzpicture}
-\caption{Example of multiset-trie structure.}
+\caption{Example of multiset-trie structure containing multisets multisets $\emptyset, \{ 1,1,2 \},$ $\{ 1,2,2 \},$ $\{ 2 \},$ $\{ 1,2 \},$ $\{ 2,2 \}$. The \emph{Null} children are omitted.}
 \label{fig:sketch}

### 2022-07-29 tex/tex/s_math-noqed.tex
before plos submission
Riste remarks (from the bus)

--- 2021-04-25 tex/tex/s_math.tex
@@
 cardinality $\sigma,$ such that $\Sigma = \{ 1,2, \ldots, \sigma \}.$ Define $N$
-to be the set of all possible multisets that can be inserted in
-multiset-trie. Let $n$ be the maximal degree of a node in multiset-trie.
+to be the set of all possible multisets that can be inserted in a
+multiset-trie. Let $n$ be the maximal degree of a node in a multiset-trie.
 Then the maximal multiplicity of an element in a multiset is equal to $n-1.$
@@
 The claim follows by taking the complement probability of the above result.
-\hspace*{\fill}\qed
+
 %
@@
 which completes the proof.
-\hspace*{\fill}\qed
+
 \end{proof}
@@
 property~(\cref{eq:prop_exp-pgf}) we complete the proof.
-\hspace*{\fill}\qed
+
 \end{proof}
@@
 \end{eqnarray*}
-\hspace*{\fill}\qed
+
 \end{proof}
@@
 supermultiset).
-\hspace*{\fill}\qed
+
 \end{proof}

### 2022-08-01 tex/tex/s_experiments.tex
before plos submission
English checked in s_experiments.tex.

--- 2019-03-14 tex/tex/s_experiments.tex
@@
 %
-This section contains results of experiments that were performed on the multiset-trie 
+This section contains the results of experiments that were performed on the multiset-trie 
 data structure. In particular, we will test the functions: \textsc{submsetExistence}, 
@@
 
-The implementation of multiset-trie is done in the \CC { programming} language. 
+The multiset-trie is implemented in the \CC { programming} language. 
 The current implementation uses only the standard library of \CC14 version of the 
 standard and has a command line interface~\cite{akulich2019mstrie}. The implementation of the program was 
-optimized for testing and therefore, the program operates with files, in order to 
-process queries. After processing all the queries the results are stored in files for further analysis.
+optimized for testing, and therefore, the program operates with files to 
+process queries. After processing all the queries, the results are stored in files for further analysis.
 
@@
 
-Before we start, we will give a few definitions about the parameters 
+Before we start, we will give a few definitions of the parameters 
 that will be varied throughout the experiments and discuss the experimental data 
@@
 
-Let $M$ be a set of multisets that are inserted to multiset-trie and let $n$ be 
+Let $M$ be a set of multisets inserted to multiset-trie and let $n$ be 
 the maximal node degree. Let $N$ be the power multiset of $\Sigma,$ where 
@@
 
-The selected parameters of the data structure that will be varied in experiments 
+The selected parameters of the data structure that will be varied in the experiments 
 are as follows:
@@
 mentioned in Section~\ref{c:description}, the mapping $\phi$ determines the 
-correspondence of letters to levels in multiset-trie, i.e. it defines the ordering of 
-levels in multiset-trie. It is also true, that $\phi$ defines the ordering in multisets.
+correspondence of letters to levels in multiset-trie, i.e., it defines the ordering of 
+levels in multiset-trie. It is also true that $\phi$ defines the ordering in multisets.
 
 % anouncement of the experiments
-In the next sections we will present the behavior of the multiset-trie data 
-structure depending on the selected parameters as well as the comparative 
-benchmark of the multiset-trie against B+ tree implementation of inverted index. 
-% outline of the experiment section
-We start with experiments that are performed on an artificially generated data in 
-order to give a general picture of the multiset-trie performance. In the 
+In the following sections, we will present the behavior of the multiset-trie data 
+structure in four experiments.
+The first three experiments use artificially generated data, and the fourth experiment
+uses real world data. In the 
 Experiment~\hyperref[s:exp1]{1} a special case of the multiset-trie is considered. 
-Only sets are allowed to be stored in the data structure, i.e. the maximal allowed 
+Only sets are allowed to be stored in the data structure, i.e., the maximally allowed 
 multiplicity is set to 1. The performance is measured with respect to the density 
 of the multiset-trie.
-%
+
 The Experiment~\hyperref[s:exp2]{2} is an extension of the previous one. Here, 
@@
 the data structure is populated with multisets.
-%
-Summarizing the tests of performance depending on the density we present the 
-Experiment~\hyperref[s:exp3]{3}. It shows a non linearity of the performance 
+
+Summarizing the tests of performance depending on the density, we present the 
+Experiment~\hyperref[s:exp3]{3}. It shows a nonlinearity of the performance 
 with respect to the density of the multiset-trie.
-%
-The next experiment on the multiset-trie uses the real world data. In 
+
+Finally, the fourth experiment on the multiset-trie uses real-world data. In 
 Experiment~\hyperref[s:exp4]{4} the influence of the mapping $\phi$ is studied. 
-The input data is obtained by mapping of the real words from English dictionary 
+The input data is obtained by mapping the real words from the English dictionary 
 to the set of consecutive integers using the function $\phi.$ The experiment 
@@
 of real data demonstrating the high performance of search queries.
+
 % 
-After all the experiments we present an empirical comparison of multiset-trie 
-data structure with B+ tree based inverted index. We use inverted index 
-to store and retrieve multisets in the same way as it is described in the paper by 
-Helmer and Moerkotte~\cite{Helmer2003} for sets. In the comparison we use 
-three types of queries exact, submultiset and supermultiset retrieval.
+%After all the experiments we present an empirical comparison of multiset-trie 
+%data structure with B-tree based inverted index. We use inverted index 
+%to store and retrieve multisets in the same way as it is described in the paper by 
+%Helmer and Moerkotte~\cite{Helmer2003} for sets. In the comparison we use 
+%three types of queries exact, submultiset and supermultiset retrieval.
 
@@
 We denote by \emph{input data} the data that is used to fill the structure prior 
-to testing and by \emph{test data} the set of queries that is used to test the 
+to testing and by \emph{test data} the set of queries that are used to test the 
 performance of the functions.
@@
 probability $p$ gives a collection $M$ of multisets that are sampled from $N$ 
-with uniform distribution. Uniform distribution is chosen in order to simulate a 
+with uniform distribution. Uniform distribution is chosen in order to simulate
 random user input.
@@
 value of multiset size is 1500. In other words, we perform 1500 experiments 
-in order to measure the number of visited nodes for the queries with test multiset 
-of a distinct size. The final value of visited nodes is calculated by taking an 
+in order to measure the number of visited nodes for the queries with a test multiset 
+of distinct sizes. The final value of visited nodes is calculated by taking an 
 arithmetic mean among all 1500 measurements.
@@
 to make a closer comparison with the \emph{set-trie} data structure~\cite{savnik2013index}.
-In this case we set the maximal node degree $n$ to be $2$ and $\sigma$ to be 25. 
-The mapping $\phi$ does not have an influence in this particular experiment, 
+In this case, we set the maximal node degree $n$ to be $2$ and $\sigma$ to be 25. 
+The mapping $\phi$ does not have an influence in this particular experiment
 because the input data is generated artificially with uniform distribution. On 
-average the results will be the same for any $\phi,$ since all the multisets are 
+average, the results will be the same for any $\phi,$ since all the multisets are 
 equally likely to appear in $M.$ The parameter $|M|$ varies from 10000 sets up 
@@
 	\center
-\includegraphics[width=.4\textwidth]{exp1-m1.pdf}
-\caption{Experiment 1, submsetExistence function.}
-\label{fig:e1m1}
+	\subcaptionbox{submsetExistence \label{fig:e1m1}}{\includegraphics[width=.45\textwidth]{exp1-m1.pdf}}
+	\subcaptionbox{supermsetExistence\label{fig:e1m2}}{\includegraphics[width=.45\textwidth]{exp1-m2.pdf}}
+	\caption{Existence functions of Experiment 1.}
 \end{figure}
 
-\begin{figure}
+\begin{figure}[ht]
 \center
-\includegraphics[width=.4\textwidth]{exp1-m2.pdf}
-\caption{Experiment 1, supermsetExistence function.}
-\label{fig:e1m2}
-\end{figure}
-
-\begin{figure}
-\center
-\includegraphics[width=.4\textwidth]{exp1-m3.pdf}
-\caption{Experiment 1, getAllSubmsets function.}
-\label{fig:e1m3}
-\end{figure}
-
-\begin{figure}
-\center
-\includegraphics[width=.4\textwidth]{exp1-m4.pdf}
-\caption{Experiment 1, getAllSupermsets function.}
-\label{fig:e1m4}
+\subcaptionbox{getAllSubmsets
+\label{fig:e1m3}}{\includegraphics[width=.45\textwidth]{exp1-m3.pdf}
+}
+\subcaptionbox{getAllSupermsets
+\label{fig:e1m4}}{\includegraphics[width=.45\textwidth]{exp1-m4.pdf}
+}
+\caption{Exhaustive functions of Experiment 1.}
 \end{figure}
@@
 \textsc{supermsetExistence} increases as the density increases (see figures~\ref{fig:e1m1}
-and~\ref{fig:e1m2}). The results are as expected, because the increase of the 
-density increases the probability of finding submultiset or supermultiset in 
-multiset-trie, which leads to the lower number of visited nodes.
+and~\ref{fig:e1m2}). The results are as expected because the increase of the 
+density increases the probability of finding sub-multiset or super-multiset in 
+multiset-trie, which leads to a lower number of visited nodes.
 
@@
 we can deduce that at least 7-15 multisets were checked in order to find 
-submultiset or supermultiset, which is from $\num{0.02e-3}$ to $\num{1.5e-3}$ of the 
+sub-multiset or super-multiset, which is from $\num{0.02e-3}$ to $\num{1.5e-3}$ of the 
 multiset-trie and from $\num{1.9e-7}$ to $\num{4.5e-7}$ of the complete 
@@
 
-As the density increases the peaks shift from the center to the left, or to the right, 
+As the density increases, the peaks shift from the center to the left or to the right, 
 for \textsc{submsetExistence} and \textsc{supermsetExistence} respectively. 
 The shifts are the consequence of the uniform distribution of sets in $M.$ 
-Since every set has the same probability to appear in $M,$ the distribution of set 
-sizes in $M$ is normal. Consequently, with increase of the density of the 
+Since every set has the same probability of appearing in $M,$ the distribution of set 
+sizes in $M$ is normal. Consequently, with the increase in the density of the 
 multiset-trie the number of sets in $M$ with cardinality $\frac{1}{2}\sigma$ will be 
@@
 \textsc{supermsetExistence} increases the multiplicity of some elements 
-(in this case adding new elements) in order to find the closest superset. 
+(in this case, adding new elements) in order to find the closest superset. 
 Thus, the peak shifts to the right.
@@
 Note that despite the peak shifts both functions \textsc{submsetExistence} and 
-\textsc{supermsetExistence} have approximately the same worst case performance.
+\textsc{supermsetExistence} have approximately the same worst-case performance.
 
@@
 This happens because the number of multisets in multiset-trie increases, which means 
-that any multiset in the data structure will have more sub- and supermultisets. 
+that any multiset in the data structure will have more sub- and super-multisets. 
 The maxima for both functions varies from $\num{8.0e4}$ to $\num{1.5e6}$ visited nodes. 
@@
 explanation is very simple. In order to find all submultisets of a small set the 
-function has to traverse a small part of multist-trie. As the size of a set 
-increases the part of a multiset-trie where all the submultisets of a given set 
+function has to traverse a small part of the multiset-trie. As the size of a set 
+increases, the part of a multiset-trie where all the submultisets of a given set 
 are stored also increases. The opposite holds for the function 
@@
 In the Experiment~\hyperref[s:exp2]{2} we demonstrate the performance of 
-the unrestricted multiset-trie allowing \emph{multisets} to be inserted into data structure. 
+the unrestricted multiset-trie allowing \emph{multisets} to be inserted into the data structure. 
 We set $n$ to be 6 and retain $\sigma = 25$ as it was in Experiment~\hyperref[s:exp1]{1}. 
-The mapping $\phi$ does not have an influence on results, since the input 
+The mapping $\phi$ does not have an influence on the results, since the input 
 data is generated artificially with uniform distribution. The cardinality of $M$ 
@@
 
-\begin{figure}
+\begin{figure}[ht]
 \center
-\includegraphics[width=.4\textwidth]{exp2-m1.pdf}
-\caption{Experiment 2, submsetExistence function.}
-\label{fig:e2m1}
+
+\subcaptionbox{submsetExistence
+\label{fig:e2m1}}{\includegraphics[width=.45\textwidth]{exp2-m1.pdf}}
+\subcaptionbox{supermsetExistence
+\label{fig:e2m2}}{\includegraphics[width=.45\textwidth]{exp2-m2.pdf}}
+\caption{Existence functions in Experiment 2.}
 \end{figure}
 
-\begin{figure}
+\begin{figure}[ht]
 \center
-\includegraphics[width=.4\textwidth]{exp2-m2.pdf}
-\caption{Experiment 2, supermsetExistence function.}
-\label{fig:e2m2}
-\end{figure}
-
-\begin{figure}
-\center
-\includegraphics[width=.4\textwidth]{exp2-m3.pdf}
-\caption{Experiment 2, getAllSubmsets function.}
-\label{fig:e2m3}
-\end{figure}
-
-\begin{figure}
-\center
-\includegraphics[width=.4\textwidth]{exp2-m4.pdf}
-\caption{Experiment 2, getAllSupermsets function.}
-\label{fig:e2m4}
+\subcaptionbox{Experiment 2, getAllSubmsets function.
+\label{fig:e2m3}}{\includegraphics[width=.45\textwidth]{exp2-m3.pdf}}
+\subcaptionbox{Experiment 2, getAllSupermsets function.
+\label{fig:e2m4}}{\includegraphics[width=.45\textwidth]{exp2-m4.pdf}}
+\caption{Exhaustive functions in Experiment 2.}
 \end{figure}
@@
 \textsc{supermsetExistence} becomes worse as the density increases. 
-In this case the number $|M|$ is slightly larger than in the 
-Experiment~\hyperref[s:exp1]{1}, but the density is very small. Consequently 
-multiset-trie become more sparse. Multisets in a sparse multiset-trie differs more, 
-which leads to the larger number of visited nodes.
+In this case, the number $|M|$ is slightly larger than in the 
+Experiment~\hyperref[s:exp1]{1}, but the density is very small. Consequently, 
+multiset-trie becomes more sparse. Multisets in a sparse multiset-trie differ more, 
+which leads to a larger number of visited nodes.
 
-The maxima for both functions varies from 7500 to 25000 visited nodes. According 
-to those maxima at least 300-1000 multisets were checked in order to find 
-submultiset or supermultiset, which is from $\num{1.5e-3}$ to $\num{7.5e-3}$ of the entire 
+The maxima for both functions vary from 7500 to 25000 visited nodes. According 
+to those maxima, at least 300-1000 multisets were checked in order to find 
+sub-multiset or super-multiset, which is from $\num{1.5e-3}$ to $\num{7.5e-3}$ of the entire 
 multiset-trie and from $\num{1.1e-17}$ to $\num{3.4e-17}$ of the complete 
@@
 the percentage of visited multiset with respect to complete multiset-trie, then 
-in case of Experiment~\hyperref[s:exp2]{2} it is less by 10 orders than in the 
+in the case of Experiment~\hyperref[s:exp2]{2} it is less by 10 orders than in the 
 Experiment~\hyperref[s:exp1]{1}.
@@
 The peaks are shifted from the center to the left and right for 
-\textsc{submsetExistence} and \textsc{supermsetExistence} respectively. Such a 
+\textsc{submsetExistence} and \textsc{supermsetExistence} respectively. Such 
 behavior was previously observed in the Experiment~\hyperref[s:exp1]{1}. The 
-explanation is the same: the input data has uniform distribution, implying that 
+explanation is the same: the input data has a uniform distribution, implying that 
 the size of multisets in $M$ is normally distributed. Because of the normal 
-distribution of size of multisets the shift of the peak occurs as the density increases.
+distribution of the size of multisets, the shift of the peak occurs as the density increases.
 
-It can be also observed that as in previous Experiment~\hyperref[s:exp1]{1} both 
+It can also be observed that, as in previous Experiment~\hyperref[s:exp1]{1}, both 
 functions \textsc{submsetExistence} and \textsc{supermsetExistence} have similar 
@@
 their performance as the density increases (see figures~\ref{fig:e2m3} 
-and~\ref{fig:e2m4}). It happens, because the number of multisets increases as 
-the density increases. So there are more nodes have to be visited in order to 
-retrieve all sub- or supermultisets of some multiset. The maximum for both functions 
+and~\ref{fig:e2m4}). This happens because the number of multisets increases as 
+the density increases. So there are more nodes that have to be visited in order to 
+retrieve all sub- or super-multisets of some multiset. The maximum for both functions 
 varies from $\num{0.9e5}$ to $\num{1.5e7}$ visited nodes. As it was observed in 
-Experiment~\hyperref[s:exp1]{1} the maxima occur at the opposite points. For the 
-function \textsc{getAllSubmsets} it will always be at the largest size of multiset, 
+Experiment~\hyperref[s:exp1]{1}, the maxima occur at the opposite points. For the 
+function \textsc{getAllSubmsets} it will always be at the largest size of the multiset, 
 which is 125 in our case. Conversely the maximum for the \textsc{getAllSupermsets} 
@@
 
-%\subsection{Experiment 3}
-%In this experiment we vary $\sigma$ from 10 up to 70. The maximal degree node 
-%$n$ is set to 6 and the input data $|M|$ is set to 640000 multisets. This way 
-%we extremly decrease density $d$ as we increase $\sigma.$ 
-%The density varies from $1.0\%$ to $\num{2.1e-47}\%.$
-%
-%\begin{figure}
-%\center
-%\includegraphics[width=.9\textwidth]{exp3-m1.pdf}
-%\caption{Experiment 3, submsetExistence function.}
-%\end{figure}
-%
-%\begin{figure}
-%\center
-%\includegraphics[width=.9\textwidth]{exp3-m2.pdf}
-%\caption{Experiment 3, supermsetExistence function.}
-%\end{figure}
-%
-%\begin{figure}
-%\center
-%\includegraphics[width=.9\textwidth]{exp3-m3.pdf}
-%\caption{Experiment 3, getAllSubmsets function.}
-%\end{figure}
-%
-%\begin{figure}
-%\center
-%\includegraphics[width=.9\textwidth]{exp3-m4.pdf}
-%\caption{Experiment 3, getAllSupermsets function.}
-%\end{figure}
-
 \subsection{Experiment 3} \label{s:exp3}
@@
 functions \textsc{submsetExistence} and \textsc{supermsetExistence} can both get 
-better and worse. The reason of such a behavior is that the dependence of the 
-number of visited nodes on density is not a linear function. It is obvious that the 
-performance of the mentioned above functions is maximal when multiset-trie is 
-complete. As multiset-trie becomes more sparse (the density is small) multisets 
-differ more and the number of visited nodes increases. However, when the density 
-is high, multisets differ less, so the number of visited nodes decreases. Since 
+better and worse. The reason for such a behavior is that the dependence of the 
+number of visited nodes on density is not a linear function. 
+The performance of the abovementioned functions is maximal when multiset-trie is 
+complete. As multiset-trie becomes more sparse (the density is small), multisets
+differ more, and the number of visited nodes increases. However, multisets differ
+less when the density is high, so the number of visited nodes decreases. Since 
 the dependence of the number of visited nodes on the density of multiset-trie 
 is a continuous function on the interval $[0,1],$ there exists a global maximum. 
-In other words there exists such a value of density where the number of visited 
+In other words, there exists such a value of density where the number of visited 
 nodes is maximal.
@@
 \textsc{submsetExistence} and \textsc{supermsetExistence}. The parameters 
-$\sigma$ and $n$ are set to 12 and 5 respectively. The density varies from 
+$\sigma$ and $n$ are set to 12 and 5, respectively. The density varies from 
 $\num{1.0e-6}$ to $\num{1.0e-2}.$ The number of visited nodes was chosen to be 
-maximal for each value of particular density.
+maximal for each value of a particular density.
 
@@
 \center
-\includegraphics[width=.4\textwidth]{exp4-m1.pdf}
-\caption{Experiment 3, submsetExistence function.}
-\label{fig:e3m1}
-\end{figure}
-
-\begin{figure}
-\center
-\includegraphics[width=.4\textwidth]{exp4-m2.pdf}
-\caption{Experiment 3, supermsetExistence function.}
-\label{fig:e3m2}
+\subcaptionbox{submsetExistence
+\label{fig:e3m1}}{\includegraphics[width=.45\textwidth]{exp4-m1.pdf}}
+\subcaptionbox{supermsetExistence
+\label{fig:e3m2}}{\includegraphics[width=.45\textwidth]{exp4-m2.pdf}}
+\caption{Exsitence functions in Experiment 3.}
 \end{figure}
@@
 \subsection{Experiment 4} \label{s:exp4}
-In previous experiments the input was generated artificially with uniform 
+In previous experiments, the input was generated artificially with uniform 
 distribution, so there was no influence of the mapping function $\phi$ on 
-performance of tested functions. This experiment shows the influence of the 
+the performance of tested functions. This experiment shows the influence of the 
 mapping $\phi$ from alphabet $\Sigma$ to a set of consecutive integers. 
-We obtain the influence by taking the real world data as an input data.
+We obtain the influence by taking the real-world data as input data.
 
-The data is taken from English dictionary which contains 235883 different words. 
+The data is taken from the English dictionary, which contains 235883 different words. 
 Those words are mapped to multisets of integers according to the $\phi.$ In 
 particular, we are interested in cases when $\phi(\Sigma)$ enumerates 
-letters by their relative frequency in English language. We say that $\phi(\Sigma)$ 
+letters by their relative frequency in the English language. We say that $\phi(\Sigma)$ 
 maps letters in \emph{ascending order} if the most frequent letter is mapped to 
 number $\sigma.$ Conversely, in \emph{descending order} this letter is mapped to 
-number $1.$ The size of the alphabet $\sigma$ is set to the size of the English 
-alphabet 26. The degree of a node $n$ is set to 10. On average the multiplicity 
-of letters is of course less than 10. We choose such a large node degree allowing 
-the multiplicity to be up to 10, because the dictionary contains such words.
+the number $1.$ The size of the alphabet $\sigma$ is set to the size of the English 
+alphabet 26. The degree of a node $n$ is set to 10. On average, the multiplicity 
+of letters is, of course, less than 10. We choose such a large node degree allowing 
+the multiplicity to be up to 10 because the dictionary contains such words.
 
@@
 \center
-\includegraphics[width=.4\textwidth]{exp3-m1.pdf}
-\caption{Experiment 4, submsetExistence function.}
-\label{fig:e4m1}
-\end{figure}
-
-\begin{figure}
-\center
-\includegraphics[width=.4\textwidth]{exp3-m2.pdf}
-\caption{Experiment 4, supermsetExistence function.}
-\label{fig:e4m2}
+\subcaptionbox{submsetExistence
+\label{fig:e4m1}}{\includegraphics[width=.45\textwidth]{exp3-m1.pdf}}
+\subcaptionbox{supermsetExistence
+\label{fig:e4m2}}{\includegraphics[width=.45\textwidth]{exp3-m2.pdf}}
+\caption{Existence functions in Experiment 4.}
 \end{figure}
@@
 According to the design of the data structure multiset-trie, we can say 
-something about multiset only if we try to reach it, i.e. to find the complete 
-path that corresponds to a particular multiset. It means that in order to give 
-an answer whether some multiset exists or not one have to check the leaf level in 
+something about multiset only if we try to reach it, i.e., to find the complete 
+path that corresponds to a particular multiset. This means that in order to give 
+an answer about the existance of some multiset, one has to check the leaf level in 
 multiset-trie.
@@
 multiset-trie according to ascending order of letters by frequency. This means 
-that the search becomes narrower, because a lot of invalid paths will be 
+that the search becomes narrower because a lot of invalid paths will be 
 discarded on top most levels. Thus, multiset-trie can be traversed faster.
@@
 \textsc{getAllSupermsets} were not tested in this experiment. Those functions 
-are not affected by variations of the mapping $\phi,$ because for any multiset 
+are not affected by variations of the mapping $\phi,$ because for any multiset, 
 they retrieve all sub/supermultisets. This means that the number of visited 
 nodes will not be changed as $\phi$ varies.
-
-\subsection{Experiment 5} \label{s:exp5}
-In this experiment, we demonstrate the performance of multiset-trie data structure compared to the inverted index based on the B+ tree.
-Both data structures are implemented in the programming language \CC, providing in this way an experimental setup for a fair comparison \cite{akulich2019mstrie}.
-
-((Brief description of the implementation of the inverted files?))
-
-The experiment uses the input data for the construction of the given data structure and the test data for the execution of the operations on the given data structure.
-The input data comprises a set of randomly generated multisets used for the construction of a data structure.
-The test data includes the set of multisets together with the operations that are evaluated.
-The input and test data were generated with respect to parameters $\sigma$ and $n$ as presented in Table~\ref{t:benchmark}.
-
-\begin{table}[h]
-\center
-\begin{tabular}{|c|c|}
-\hline
-$\sigma$ & $n$ \\
-\hline
-5		& 1\\
-\hline
-30	& 1 \\
-\hline
-5		& 3 \\
-\hline
-15	& 3 \\
-\hline
-30	& 3 \\
-\hline
-10	& 10 \\
-\hline
-\end{tabular}
-\caption{Configuration of $\sigma$ and $n$ in benchmark.}
-\label{t:benchmark}
-\end{table}
-
-We tested all three types of a query on all of the configurations from Table 1 which results in 18 experiments total, i.e., 6 experiments per query type.
-In each experiment, we measured an average time consumed by the data structure to process the query.
-The results of the exact search, sub-multiset and super-multiset search experiments are presented in tables Table~\ref{t:res_ex}, Table~\ref{t:res_sub} and Table~\ref{t:res_sup}, respectively.
-
-\begin{table}[h]
-\center
-\begin{tabular}{|r|r|r|r|}
-\hline
-\multicolumn{1}{|c|}{$\sigma$} & 
-\multicolumn{1}{c|}{$n$} & 
-\multicolumn{1}{c|}{Multiset-trie ($\mu s$)} & 
-\multicolumn{1}{c|}{Inverted index ($\mu s$)} \\
-\hline
-5		& 1 & 3.45 & 17782.35\\
-\hline
-30	& 1 & 4.18 & 24865.93\\
-\hline
-5		& 3 & 2.20 & 1508.81\\
-\hline
-15	& 3 & 4.39 & 2146.36\\
-\hline
-30	& 3 & 10.67 & 3639.97\\
-\hline
-10	& 10 & 6.93 & 384.05\\
-\hline
-\end{tabular}
-\caption{Exact search.}
-\label{t:res_ex}
-\end{table}
-
-\begin{table}[h]
-\center
-\begin{tabular}{|r|r|r|r|}
-\hline
-\multicolumn{1}{|c|}{$\sigma$} & 
-\multicolumn{1}{c|}{$n$} & 
-\multicolumn{1}{c|}{Multiset-trie ($\mu s$)} & 
-\multicolumn{1}{c|}{Inverted index ($\mu s$)} \\
-\hline
-5		& 1			& 8.96 & 73500.84\\
-\hline
-30	& 1			& 17.33 & 547572.74\\
-\hline
-5		& 3			& 117.95 & 162360.43\\
-\hline
-15	& 3			& 20.74 & 443321.39\\
-\hline
-30	& 3 		& 23.75 & 947706.14\\
-\hline
-10	& 10		& 55.59 & 466022.68\\
-\hline
-\end{tabular}
-\caption{Sub-multiset search.}
-\label{t:res_sub}
-\end{table}
-
-\begin{table}[h]
-\center
-\begin{tabular}{|r|r|r|r|}
-\hline
-\multicolumn{1}{|c|}{$\sigma$} & 
-\multicolumn{1}{c|}{$n$} & 
-\multicolumn{1}{c|}{Multiset-trie ($\mu s$)} & 
-\multicolumn{1}{c|}{Inverted index ($\mu s$)} \\
-\hline
-5		& 1 		& 10.63 & 63073.86\\
-\hline
-30	& 1 		& 14.65 & 449251.68\\
-\hline
-5		& 3 		& 171.04 & 163256.77\\
-\hline
-15	& 3 		& 43.42 & 425733.80\\
-\hline
-30	& 3 		& 22.06 & 729831.34\\
-\hline
-10	& 10 	& 58.32 & 373784.81\\
-\hline
-\end{tabular}
-\caption{Super-multiset search.}
-\label{t:res_sup}
-\end{table}
-
-We can see that multiset-trie outperforms inverted index in all of the experiments by up to 4 orders of magnitude.
-In an exact search, multiset-trie has to traverse only up to $\sigma+1$ nodes to get query result.
-It can be seen from results that with the increase of $\sigma$ the processing time for multiset-tire also increases.
-Multiplicity also affects the processing time; however, this happens passively.
-Multiplicity, or degree of a node $n$, defines the shape of multisets that are stored in multiset-trie.
-Thus, it affects the structure and density of the multiset-trie.
-
-As for the inverted index, all three operations must first fetch all postings for each particular element of a test multi-set.
-Afterward, the intersection of postings is computed to answer the query.
-The operations use more processing time than a simple tree traversal, which we can see from results.
-Postings are filtered on-the-fly to reduce the cost of the intersection.
-
-Implementations of the sub-multiset and super-multiset are similar in the case of the inverted file.
-The algorithm consists of the same steps.
-First, the postings are fetched for each element of the test multiset.
-Depending on the particular operation, postings are filtered on-the-fly.
-Finally, the union or the intersection of the filtered set of postings is computed.
-Note that the processing time increases with the size of the inverted index because of the increased sizes of postings.
-
-In the case of multiset-trie, only a traversal of the tree is required which is much faster than the processing of postings as we can see from results.
-In the worst case the whole tree is traversed, but so is for an inverted file.

### 2022-08-01 tex/tex/s_intro.tex
Riste remarks (from the bus)
English checked and updated in s_data_structure and s_operations.

@@
 First, a  multiset-trie is a novel data structure for storing and querying a set of multisets that provides efficient multiset containment operations.
-Furthermore, the contributions of this work are thorough mathematical and empirical analyses of the proposed data structure multiset-trie.
 
@@
 We carefully designed the experiments to unravel the main features of the time complexity space.
-%%% relats btwn math and emprcl anlyss
-%%% concrete results? bounds, ...
-%%% experiments confirm findings of math analysis?
 
-Finally, the mathematical as well as empirical analyses show an influence of ordering the elements of multisets on the efficiency of storing and processing multisets.
+Finally, the mathematical, as well as empirical analyses, show an influence of ordering the elements of multisets on the efficiency of storing and processing multisets.
 We show that the ordering, which is based on the frequencies of the multiset elements, can speed up the multiset containment queries by orders of magnitude.

### 2022-08-03 tex/tex/s_conclusions.tex
latexdiff related, soma updates to gitignore, and a bit of changes in conclusions

@@
 % Experiments conclusions
-%
 One of the conclusions of studying the multiset-trie both theoretically and empirically is that our data structure is input sensitive.
 Input sensitivity implies a non-consistent performance on different input data.
-However, our argument that the performance can be optimized by pre-processing the input data is confirmed in the Experiment~\hyperref[ss:exp3]{4}.
+However, our argument that the performance can be optimized by pre-processing the input data is confirmed in  Experiment~\hyperref[ss:exp3]{4}.
 Pre-processing determines the optimal encoding for input data and ensures the best performance of the multiset-trie on particular input data.
 For example, in the case of storing words in the multiset-trie, the search queries can always be optimized based on the frequencies of letters in a specific language.
-We also see from Experiments~\hyperref[s:exp1]{1} and~\hyperref[s:exp2]{2} that dependence of the multiset-trie performance on the density is not a linear function.
-Yet the function is continuous, and the point of inflection is unique on the whole domain as it is shown in Experiment~\hyperref[s:exp3]{3}.
+We also see from Experiments~\hyperref[s:exp1]{1} and~\hyperref[s:exp2]{2} that the dependence of the multiset-trie performance on the density is not a linear function.
+Yet the function is continuous, and the point of inflection is unique on the whole domain, as shown in Experiment~\hyperref[s:exp3]{3}.
 This allows us to predict whether multiset-trie can be used for some particular application, serving a high performance.
@@
 % Mathematical analysis conclusions
-%
 The mathematical analysis of the space complexity shows that multiset-trie requires only $O(|M|)$ space, which is the minimal possible space required by any data structure for storage of $|M|$ objects.
@@
 
-The multiset-trie is an input sensitive data structure because the size of multiset-trie $|\mathcal{M}|$ depends on the distribution of multisets in $M.$ Our mathematical model assumes that multisets $m$ in $M$ are distributed uniformly.
-However, in real-world models, such an assumption is not true in a lot of cases.
-Specifically, the probability $P(m\in M)$ may vary dramatically and can be even equal to $0.$ For example, if words are mapped to multisets, then the sample space contains very large multisets.
-Nonetheless, most of the words will have zero probability to appear in $M$, because a word that would correspond to such a multiset does not exist.%
+The multiset-trie is an input-sensitive data structure because the size of multiset-trie $|\mathcal{M}|$ depends on the distribution of multisets in $M.$
+Our mathematical model assumes that multisets $m$ in $M$ are distributed uniformly;  however, in real-world models, such an assumption is not generally true. 
+%Specifically, the probability $P(m\in M)$ may vary dramatically and can even be equal to $0.$ For example, if words are mapped to multisets, then the sample space contains very large multisets. 
+For example, many multisets will have zero probability of appearing in $M$ because a word that would correspond to such a multiset does not exist in the dictionary.%
 

### 2022-08-04 tex/tex/s_experiments.tex
English checked and corrected in s_math-noqed.tex.

@@
 The first three experiments use artificially generated data, and the fourth experiment
-uses real world data. In the 
+uses real-world data. In the 
 Experiment~\hyperref[s:exp1]{1} a special case of the multiset-trie is considered. 
@@
 functions \textsc{submsetExistence} and \textsc{supermsetExistence} have similar 
-worst case performance.
+worst-case performance.
 

### 2022-08-06 tex/response.txt
responses to reviewer (iztok).

Comments to the Author

1. Is the manuscript technically sound, and do the data support the conclusions?

Reviewer #1: Partly

2. Has the statistical analysis been performed appropriately and rigorously?
Reviewer #1: N/A

3. Have the authors made all data underlying the findings in their manuscript fully available?

Reviewer #1: No

NAŠ ODZIV NA ZGORNJE:

? (to ask Mikita)

4. Is the manuscript presented in an intelligible fashion and written in standard English?

Reviewer #1: Yes

5. Review Comments to the Author

Reviewer #1: 
The motivation is not clear, the authors claim that this is an important problem in object-relational databases, information retrieval, data mining and expert systems. However, for example, when authors claim to be a relevant problem in information retrieval, they provide two citations to complete books of information retrieval, without any indication of section or chapter. Please, please be more specific. A good practice is to show the specific problems (examples) of the application.

NAŠ ODZIV NA ZGORNJE:

In the second pharagraph of the introduction (Section 1) we overview
the CS areas where the use of a data structure for fast retrieval of
sets is needed. In all of the presented areas the set containment
operational are essential.

It would be nice if the authors provide running examples of their algorithms, they would be easier to understand.

NAŠ ODZIV NA ZGORNJE:

?

All figures are of very poor quality and can hardly be read.

NAŠ ODZIV NA ZGORNJE:

?

An important problem for me is that in the paper it is always assumed that the multisets are always ordered according to a certain order, more precisely, that specified in the alphabet \Sigma, or as stated on page 10, line 269 "for simplicity that all elements are ordered in an ascending order". Therefore, the authors claim that the functions search, delete and insert have complexity \sigma (the size of \Sigma). I do not agree with this, in practical implementations, for example in an object-relational database, we cannot assure that the multiset queries or insertions will have the elements in a specific order. For your method to work, a previous sort and sum are required, in order to build an instance of the type Multiset. This requires O(n log n) cost, being n the cardinality of the multiset in the query, be aware, not the distinct symbols in the multiset, all (including repeated) symbols.

NAŠ ODZIV NA ZGORNJE:

In multisets ordering is not important. We use a specific ordering of
the multiset elements to speed up the set containment queries. This
specific ordering is used when we construct the multiset-trie, so the
ordering is a part of the data structure. Please have a look at the
description of the set containment operations (Section 3) to see the
details.

In the second pharagraph of the introduction (Section 1) we overview
the CS areas where the use of a data structure for fast retrieval of
sets is needed. In all of the presented areas the set containment
operational are essential.

We have updated the introduction (Section 1) to:
1) emphasise that the ordering is not important in multisets, and
2) to depict the utilization of the selected ordering for multiset
containment operations.

The experiments are not convincing for the same reason, as the whole paper obviates the ordering, whereas the inverted index uses a B-tree to search for the inverted list of each element in the multiset.
The experiments must consider that the multisets are not ordered nor added, that is, an input query could be {1,2,2, 1, 3, 3, 1, 2, 2}

The values of n, the maximum multiplicity of each element of the multiset, are small, most experiments (as shown in Table 1) are 3, and one with 10. Experiments with larger values of n are needed, and, as I will argue next, authors must include the size of the resulting data structures. The scalability must be studied, and experiments showing how the size of n affects the query times and the size of the data structure must be included.

NAŠ ODZIV NA ZGORNJE:

The multiset-trie data structure is not optimized but it is given in
the original and complete form. The absence of the compression makes
the multiset-trie a perfectly height-balanced tree.

The multiplicity of an element in a multiset is in a multiset-trie
implemented by means of an array. There exist a number of possible
optimizations. For example, the multiplicity of an element can be
implemented using a small, customly designed, hash table.

The idea of minimality of the data structure size stated in line 808 page 27, for me, is wrong. A simple analysis reveals that the space complexity O(|M|) is a very coarse bound, and it is not minimal. First, the size of the data structure is better bounded by the number of nodes, which is analyzed in Section 4.1.1. Second, the space efficiency is not optimal because you are using a complex method to determine the label of a node (the symbol a node contains). That is, to know the label of a node, it is necessary to know the position of the pointer that points to the node of interest, within the \sigma pointers of the parent node. Then each node needs an array of \sigma entries, even though only some of them are filled, so you are wasting many empty entries. In addition, since the label of a node is given by a pointer, unless you use a modern tree (that exists, but you have not mentioned that) I assume that you require 32 bits to represent the pointer (label) of a node. Instead, if you explicitly represent the symbol of each node, you only need log \sigma bits.

NAŠ ODZIV NA ZGORNJE:

We dropped the comment about the space minimality of the data
structure in the conclusions (Section 7). Minimality of the data
structure is not mentioned in any other part of the paper.

(Matjaž, so preostale ocene v istem odstavku o minimalnosti korektne?)

I can continue giving arguments, for example, you even recognize that your structure is based on a trie, and not on a patricia trie or compact prefix tree, that is, your structure does not support path compression, and this added to the aforementioned array of pointers at each node, yields a poor space efficiency.
Therefore, the space consumption of your data structure must be reported in the experimental section and compared with that of the inverted index.

NAŠ ODZIV NA ZGORNJE:

See the previous reply.

Finally, the intersection algorithm used in your experiments is not described. There are different intersection algorithms, please refer to, for example :
J. Shane Culpepper and Alistair Moffat. 2011. Efficient set intersection for inverted indexing. ACM Trans. Inf. Syst. 29, 1, Article 1 (December 2010), 25 pages. https://doi.org/10.1145/1877766.1877767

That with the better behaviour in your case must be selected.

NAŠ ODZIV NA ZGORNJE:

We have omitted the empirical comparison of the multiset-trie with the
inverted file from the paper.

Related work is poor, especially the explanation of the set-trie. Although Section 6.1 sketches the differences between multiset-trie and set-trie, more explanations are needed. First, the differences in the data structures and then, the differences in operations must be clarified.

NAŠ ODZIV NA ZGORNJE:

We have added a more detailed presentation of the set-trie in the
related work.

Page 3 line 37 "at a certan level" -> "at a certain
level"

NAŠ ODZIV NA ZGORNJE:

Corrected.

### 2022-08-13 tex/tex/s_conclusions.tex
paragraph added to s_conclusions.tex.

@@
 % Mathematical analysis conclusions
-The mathematical analysis of the space complexity shows that multiset-trie requires only $O(|M|)$ space, which is the minimal possible space required by any data structure for storage of $|M|$ objects.
-As for the running time complexity of algorithms, the basic tree functions such as \textsc{insert}, \textsc{search}, and \textsc{delete} all have a constant complexity once the multiset-trie is defined.
+% The mathematical analysis of the space complexity shows that multiset-trie requires only $O(|M|)$ space, which is the minimal possible space required by any data structure for storage of $|M|$ objects.
+As for the running time complexity of algorithms,
+The basic tree functions such as \textsc{insert}, \textsc{search}, and \textsc{delete} all have a constant complexity once the multiset-trie is defined.
 The "getAll" multiset containment functions have worst-case running time complexity of $O(|\mathcal{M}|),$ where $|\mathcal{M}|$ is the cardinality of the multiset-trie data structure.
@@
 For example, many multisets will have zero probability of appearing in $M$ because a word that would correspond to such a multiset does not exist in the dictionary.%
+
+The implementation of the multiset-trie is not optimized.
+The main reason for space-inefficiency is in implementing the links from a node to its children.
+A node at the $i$-th level of a tree has $n-i$ possible children where $n$ is the size of the alphabet.
+Our implementation uses an array data structure to link a node to its children where each element of an array includes a link to a sub-tree.
+A custom-implemented small and extendable hash table would significantly decrease the amount the space needed, especially since the number of children tends to decrease with the increasing levels of a tree.
+There are more additional optimizations possible.
+For example, the ...
 

### 2022-08-17 tex/tex/s_conclusions.tex
s_conclusions updated.

@@
 
-The multiset-trie is an input-sensitive data structure because the size of multiset-trie $|\mathcal{M}|$ depends on the distribution of multisets in $M.$
-Our mathematical model assumes that multisets $m$ in $M$ are distributed uniformly;  however, in real-world models, such an assumption is not generally true. 
-%Specifically, the probability $P(m\in M)$ may vary dramatically and can even be equal to $0.$ For example, if words are mapped to multisets, then the sample space contains very large multisets. 
-For example, many multisets will have zero probability of appearing in $M$ because a word that would correspond to such a multiset does not exist in the dictionary.%
+%The multiset-trie is an input-sensitive data structure because the size of multiset-trie $|\mathcal{M}|$ depends on the distribution of multisets in $M.$
+%Our mathematical model assumes that multisets $m$ in $M$ are distributed uniformly;  however, in real-world models, such an assumption is not generally true. 
+%%Specifically, the probability $P(m\in M)$ may vary dramatically and can even be equal to $0.$ For example, if words are mapped to multisets, then the sample space contains very large multisets. 
+%For example, many multisets will have zero probability of appearing in $M$ because a word that would correspond to such a multiset does not exist in the dictionary.%
 
 The implementation of the multiset-trie is not optimized.
-The main reason for space-inefficiency is in implementing the links from a node to its children.
-A node at the $i$-th level of a tree has $n-i$ possible children where $n$ is the size of the alphabet.
-Our implementation uses an array data structure to link a node to its children where each element of an array includes a link to a sub-tree.
-A custom-implemented small and extendable hash table would significantly decrease the amount the space needed, especially since the number of children tends to decrease with the increasing levels of a tree.
-There are more additional optimizations possible.
-For example, the ...
+For example, the main reason for space-inefficiency is in implementing the links from a node to its children.
+Our implementation uses an array data structure to link a node to its children where each element of an array, indexed by the multiplicity of the element of the next level, includes a link to a sub-tree.
+A custom-implemented small and extendable hash table would significantly decrease the amount the space needed to represent a multiset-trie.
+% There are more additional optimizations possible. For example, the ...
 

### 2022-08-17 answers-to-reviewers.tex
added dataset to bibliography, added answers to reviewers

\documentclass[11pt,a4paper]{article}
\usepackage{amssymb,amsmath}
\usepackage[hmargin=2.5cm,vmargin=2.5cm]{geometry}
\usepackage[color=green!40]{todonotes}
\usepackage[basic]{complexity}
\usepackage{enumitem}
\setlist{labelsep=5pt, labelindent=0pt, leftmargin=*, itemindent=0pt}
\usepackage{amssymb}
\usepackage{microtype}
\usepackage{mathrsfs}
%%Quote environment
\newlength{\quotemargin}
\setlength{\quotemargin}{1.7cm}
\renewenvironment{quote}
  {\list{}{\rightmargin=\quotemargin \leftmargin=\quotemargin}%
   \item\relax}
  {\endlist}

%%Justification for teletype (tt) font
\newcommand*\justify{%
  \fontdimen2\font=0.4em%interword space
  \fontdimen3\font=0.2em%interword stretch
  \fontdimen4\font=0.1em%interword shrink
  \fontdimen7\font=0.1em%\extra space
  \hyphenchar\font=`\-%allowing hyphenation
}

%%'\comment' and '\answer' commands
\newcommand{\comment}[1]{\begin{itemize}
    \item {\tt\justify #1}
\end{itemize}}
\newcommand{\answer}[1]{{\sc Answer:} #1}

\begin{document}

	\rightline{\today}

\parindent0cm	
\noindent

\vskip0.6truecm

Dear editor,

\vskip0.3truecm

\bigskip

We would like to thank the reviewers for carefully reading the paper and for their insightful comments and suggestions.

Please find below the details about the changes we made to the paper, organised with respect to the reviewers comments.

In addition to addressing the reviewers' comments, we made the following improvements:

\medskip
\parindent 0cm

\vskip0.6truecm

%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Reviewer 1}
\comment{
1. Is the manuscript technically sound, and do the data support the conclusions?

Reviewer \#1: Partly

2. Has the statistical analysis been performed appropriately and rigorously?
Reviewer \#1: N/A

	3. Have the authors made all data underlying the findings in their manuscript fully available?

	Reviewer \#1: No}
\answer{We published the data that was used to preform experiments from the paper, so that the whole dataset is now accessable by using the following digital object identifier
	 \begin{quote}
	 	\textbf{DOI:10.5281/zenodo.7004689}
	 \end{quote}
 We refer to the above repository within the paper as appropriate.}

\comment{
4. Is the manuscript presented in an intelligible fashion and written in standard English?

Reviewer \#1: Yes
}

\comment{
5. Review Comments to the Author

Reviewer \#1:
The motivation is not clear, the authors claim that this is an important problem in object-relational databases, information retrieval, data mining and expert systems. However, for example, when authors claim to be a relevant problem in information retrieval, they provide two citations to complete books of information retrieval, without any indication of section or chapter. Please, please be more specific. A good practice is to show the specific problems (examples) of the application.}

\answer{
We followed the advice of the referee, and replaced references with the more specific ones, as requested.
This can be found in the second pharagraph of the introduction (Section 1), where we overview the related areas of computer science.
Notice that in those presented areas the set containment operations play an essential role.}

\comment{It would be nice if the authors provide running examples of their algorithms, they would be easier to understand.}

\answer{
We acknowledge the remark of the reviewer, however we feel that this paper is already substentilly long, and providing an additional figure for each operation might not give the desirable outcome.
We feel that an example of the underlying multiset-trie data structure should suffice, see Figure 1.}

\comment{All figures are of very poor quality and can hardly be read.}

\answer{All figures presenting various plots are generated by matplotlib python library, and exported in vector format.
We believe this to be industry-standard and kindly ask for detailed explanation on how to improve the quality of our figures.
%Forthermore we believe that those figures were carefully designed
}

\comment{An important problem for me is that in the paper it is always assumed that the multisets are always ordered according to a certain order, more precisely, that specified in the alphabet $\Sigma$, or as stated on page 10, line 269 "for simplicity that all elements are ordered in an ascending order". Therefore, the authors claim that the functions search, delete and insert have complexity $\sigma$ (the size of $\Sigma$). I do not agree with this, in practical implementations, for example in an object-relational database, we cannot assure that the multiset queries or insertions will have the elements in a specific order. For your method to work, a previous sort and sum are required, in order to build an instance of the type Multiset. This requires $O(n \log n)$ cost, being $n$ the cardinality of the multiset in the query, be aware, not the distinct symbols in the multiset, all (including repeated) symbols.}

\answer{In multisets ordering is not important.
That said, in our implementation for storing multisets,  we make use of (arbitrary) specific ordering of the multiset elements which results in speeding up the set containment queries.
The corresponding ordering of the elements from input query is  hence indeed required, as is clearly specified throughout the paper.

We have updated the introduction (Section 1) to:
\begin{enumerate}[nosep]
	\item emphasise that the ordering is not important in multisets, and
 	\item  to depict the utilization of the selected ordering for multiset containment operations.
\end{enumerate}
}

\comment{The experiments are not convincing for the same reason, as the whole paper obviates the ordering, whereas the inverted index uses a B-tree to search for the inverted list of each element in the multiset.
The experiments must consider that the multisets are not ordered nor added, that is, an input query could be $\{1, 2, 2, 1, 3, 3, 1, 2, 2\}$}

\answer{We kindly point out that the specification of all our algorithms is designed so that the input queries are already in the so-called Multiset type (see Section 3 for more details).
This type assumes an order to the input elemens, as well as precomputed multiplicity of each of them.
We may have two distinct orderings of multisets.
The first is the application ordering where particular element can be at many places in the multiset.
The second ordering is the ordering of the symbols that is used for implementation of multiset-trie.
In particular this implies that the preprocessing step from the above comment is not covered by our complexity analysis.
We acknowledge that different applications may require the choice of our definitions to be modified.}

\comment{The values of n, the maximum multiplicity of each element of the multiset, are small, most experiments (as shown in Table 1) are 3, and one with 10. Experiments with larger values of n are needed, and, as I will argue next, authors must include the size of the resulting data structures. The scalability must be studied, and experiments showing how the size of n affects the query times and the size of the data structure must be included.
}

\answer{We acknowledge that our description of the multiset-trie data structure is not optimized, as we instead strive for a clean represeantation of a new idea.
We also point out that the absence of the compression makes the multiset-trie a perfectly height-balanced tree, which simplifies the analysis.

That said, there are various ways on how to optimize this data structure further.
For example, the multiplicity of an element can be implemented using a small, customly designed, hash table (instead of a simple array as in the paper).

In fact, we added this and some other possible optimiziations to the conclusion of our paper.
}

\comment{The idea of minimality of the data structure size stated in line 808 page 27, for me, is wrong. 
A simple analysis reveals that the space complexity $O(|M|)$ is a very coarse bound, and it is not minimal. 
First, the size of the data structure is better bounded by the number of nodes, which is analyzed in Section 4.1.1. Second, the space efficiency is not optimal because you are using a complex method to determine the label of a node (the symbol a node contains). That is, to know the label of a node, it is necessary to know the position of the pointer that points to the node of interest, within the $\sigma$ pointers of the parent node. Then each node needs an array of $\sigma$ entries, even though only some of them are filled, so you are wasting many empty entries. In addition, since the label of a node is given by a pointer, unless you use a modern tree (that exists, but you have not mentioned that) I assume that you require 32 bits to represent the pointer (label) of a node. Instead, if you explicitly represent the symbol of each node, you only need log $\sigma$ bits.
}

\answer{We agree with the reviewer, and we dropped the comment about the space minimality of the data
structure in the conclusions (Section 7), accordingly.
The minimality of the data structure is not mentioned in any other part of the paper.
}
(Matjaž, so preostale ocene v istem odstavku o minimalnosti korektne?)

\comment{I can continue giving arguments, for example, you even recognize that your structure is based on a trie, and not on a patricia trie or compact prefix tree, that is, your structure does not support path compression, and this added to the aforementioned array of pointers at each node, yields a poor space efficiency.
Therefore, the space consumption of your data structure must be reported in the experimental section and compared with that of the inverted index.
}

\answer{Please see our previous reply.}

\comment{Finally, the intersection algorithm used in your experiments is not described. There are different intersection algorithms, please refer to, for example :
J. Shane Culpepper and Alistair Moffat. 2011. Efficient set intersection for inverted indexing. ACM Trans. Inf. Syst. 29, 1, Article 1 (December 2010), 25 pages. https://doi.org/10.1145/1877766.1877767

That with the better behaviour in your case must be selected.
}

\answer{We understand the comment of reviewer. However, one of the authors of this paper recently went to work in industry.
As he was in charge of the development of inverted index implementation, it seems that a new implementation is hardly possible at the moment.
That said, we agree that in the current form the Experiment 5 does not fairly compare both data structures.
Accordingly, we have omitted the empirical comparison of the multiset-trie with the inverted file from the paper.
}

\comment{Related work is poor, especially the explanation of the set-trie. Although Section 6.1 sketches the differences between multiset-trie and set-trie, more explanations are needed. First, the differences in the data structures and then, the differences in operations must be clarified.
}

\answer{We have added a more detailed presentation of the set-trie in the related work.
Most notable changes can be found in the second paragraph of Section 6.1.
In the current form, also differences in operations between the two data structures are, in our opinion, exposed better (see third paragraph of Section 6.1).
}
\comment{Page 3 line 37 "at a certan level" -> "at a certain level"
}

\answer{We corrected the typo.}
\end{document}

### 2022-08-23 answers-to-editors.tex
submit R2

\documentclass[11pt,a4paper]{article}
\usepackage{amssymb,amsmath}
\usepackage[hmargin=2.5cm,vmargin=2.5cm]{geometry}
\usepackage[color=green!40]{todonotes}
\usepackage[basic]{complexity}
\usepackage{enumitem}
\setlist{labelsep=5pt, labelindent=0pt, leftmargin=*, itemindent=0pt}
\usepackage{amssymb}
\usepackage{microtype}
\usepackage{mathrsfs}
%%Quote environment
\newlength{\quotemargin}
\setlength{\quotemargin}{1.7cm}
\renewenvironment{quote}
  {\list{}{\rightmargin=\quotemargin \leftmargin=\quotemargin}%
   \item\relax}
  {\endlist}

%%Justification for teletype (tt) font
\newcommand*\justify{%
  \fontdimen2\font=0.4em%interword space
  \fontdimen3\font=0.2em%interword stretch
  \fontdimen4\font=0.1em%interword shrink
  \fontdimen7\font=0.1em%\extra space
  \hyphenchar\font=`\-%allowing hyphenation
}

%%'\comment' and '\answer' commands
\newcommand{\comment}[1]{\begin{itemize}
    \item {\tt\justify #1}
\end{itemize}}
\newcommand{\answer}[1]{{\sc Answer:} #1}

\begin{document}

	\rightline{August 20, 2022}

\parindent0cm	
\noindent

\vskip0.6truecm

Dear editor,

\vskip0.3truecm

\bigskip

We would like to thank the editors for overseeing the paper and for their insightful comments and suggestions.
In addition to addressing the reviewers' comments, we provide a document which emphasizes all additional changes that were made in this revision. 
Those minor improvements can be found in file diff.pdf, which we attach to this submission.
Please find below the responses to both comments.

\medskip
\parindent 0cm

\vskip0.6truecm

%%%%%%%%%%%%%%%%%%%%%%%%

\comment{Please ensure that your manuscript meets PLOS ONE's style requirements, including those for file naming. The PLOS ONE style templates can be found at

{\small \textrm{https://journals.plos.org/plosone/s/file?id=wjVg/PLOSOne\_formatting\_sample\_main\_body.pdf}} and

{\small
	\textrm{https://journals.plos.org/plosone/s/file?id=ba62/PLOSOne\_formatting\_sample\_title\_authors\_affiliations.pdf}}
}
\answer{Thank you for those links. We used the suggested tool (PACE) to convert as well as rename all our figures according to the standards of PLOS One. Please find those attached in a single archive (PLOS\_20220821\_83046AM.zip)
We verified that style used in our LaTeX project is the official PLOS template. 
We acknowledge that at this stage the sources are not yet supposed to be attached to this revision.  }

\comment{We note that the grant information you provided in the ‘Funding Information’ and ‘Financial Disclosure’ sections do not match. When you resubmit, please ensure that you provide the correct grant numbers for the awards you received for your study in the ‘Funding Information’ section.
}
\answer{We added the precise grant number (P1-0383) in the “Funding Information” section, in the Editorial manager.
We also mentioned the same grant at the very end of the paper, so that both sections now match. } 
\medskip
\parindent 0cm

\vskip0.6truecm
Please let us know if any further changes are needed.
\end{document}

### 2022-08-23 answers-to-reviewers.tex
submit R2

@@
 
-	\rightline{\today}
+	\rightline{August 20, 2022}
 
@@
 We would like to thank the reviewers for carefully reading the paper and for their insightful comments and suggestions.
-
+In addition to addressing the reviewers' comments, we provide a document which emphasizes all additional changes that were made in this revision. 
+Those minor improvements can be found in file diff.pdf, which we attach to this submission.
 Please find below the details about the changes we made to the paper, organised with respect to the reviewers comments.
-
-In addition to addressing the reviewers' comments, we made the following improvements:
 

### 2022-08-23 tex/tex/s_conclusions.tex
submit R2

@@
 % Mathematical analysis conclusions
-% The mathematical analysis of the space complexity shows that multiset-trie requires only $O(|M|)$ space, which is the minimal possible space required by any data structure for storage of $|M|$ objects.
-As for the running time complexity of algorithms,
-The basic tree functions such as \textsc{insert}, \textsc{search}, and \textsc{delete} all have a constant complexity once the multiset-trie is defined.
+The mathematical analysis section provides a non-trivial insight regarding the behavior of multiset-trie datastructure when used in randomized data.
+It is estimated that the space complexity of multiset-trie is of order $O(|M|)$, which is the minimal possible space required by any data structure for storage of $|M|$ objects.
+As for the running time complexity of algorithms, the basic tree functions such as \textsc{insert}, \textsc{search}, and \textsc{delete} all have a constant complexity once the multiset-trie is defined.
 The "getAll" multiset containment functions have worst-case running time complexity of $O(|\mathcal{M}|),$ where $|\mathcal{M}|$ is the cardinality of the multiset-trie data structure.

### 2022-12-09 tex/referee-data/response.txt
MDPI template
threw everything into MDPI template.
created parsed file for math section (uses {Theorem} instead of {theorem})

--- 2022-08-06 tex/response.txt
@@
 NAŠ ODZIV NA ZGORNJE:
+We published the data that was used to preform experimets from the paper, and refered to those accordingly.
 
-? (to ask Mikita)
+? (to ask Mikita, both iztok and matjaž
+Matjaž takes care of publishing data)
 
@@
 NAŠ ODZIV NA ZGORNJE:
+We followed the advice of referee, and replaced references with the more specific ones, as requested.
+This can be found in the second pharagraph of the introduction (Section 1), where we overview the related areas of computer science.
+Notice that in those presented areas the set containment operations play an essential role.
 
-In the second pharagraph of the introduction (Section 1) we overview
-the CS areas where the use of a data structure for fast retrieval of
-sets is needed. In all of the presented areas the set containment
-operational are essential.
+---
 
@@
 NAŠ ODZIV NA ZGORNJE:
+We acknowledge the remark of the reviewer, however we feel that this paper is already substentilly long, and providing an additional figure for each operation might not give the desirable outcome.
 
-?
+Please notice that we did include an example of the underlying multiset-trie data structure.
+
+---
 
@@
 NAŠ ODZIV NA ZGORNJE:
-
-?
+All figures presenting various plots are generated by matplotlib python library, and exported in vector format.
+We believe this to be industry-standard and kindly ask for detailed explanation on how to improve their quality.
+%Forthermore we believe that those figures were carefully designed
 
@@
 
-In multisets ordering is not important. We use a specific ordering of
-the multiset elements to speed up the set containment queries. This
-specific ordering is used when we construct the multiset-trie, so the
-ordering is a part of the data structure. Please have a look at the
-description of the set containment operations (Section 3) to see the
-details.
-
-In the second pharagraph of the introduction (Section 1) we overview
-the CS areas where the use of a data structure for fast retrieval of
-sets is needed. In all of the presented areas the set containment
-operational are essential.
+In multisets ordering is not important.
+We use a specific ordering of the multiset elements to speed up the set containment queries.
+This specific ordering is used when we construct the multiset-trie, so the
+ordering is a part of the data structure.
 
@@
 1) emphasise that the ordering is not important in multisets, and
-2) to depict the utilization of the selected ordering for multiset
-containment operations.
+2) to depict the utilization of the selected ordering for multiset containment operations.
+
+---
 
 The experiments are not convincing for the same reason, as the whole paper obviates the ordering, whereas the inverted index uses a B-tree to search for the inverted list of each element in the multiset.
-The experiments must consider that the multisets are not ordered nor added, that is, an input query could be {1,2,2, 1, 3, 3, 1, 2, 2}
+The experiments must consider that the multisets are not ordered nor added, that is, an input query could be {1, 2, 2, 1, 3, 3, 1, 2, 2}
 
+NAŠ ODZIV NA ZGORNJE:
+
+We kindly point out that the specification of all our algorithms is designed so that the input queries are already in the so-called Multiset type (see Section 3 for more details).
+This type assumes an order to the input elemens, as well as precomputed multiplicity of each of them.
+We may have two distinct orderings of multisets.
+The first is the application ordering where particular element can be at many places in the multiset.
+The second ordering is the ordering of the symbols that is used for implementation of multiset-trie.
+In particular this implies that the preprocessing step from the above comment is not covered by our complexity analysis.
+We acknowledge that different applications may require the choice of our definitions to be modified.
+
+----
 The values of n, the maximum multiplicity of each element of the multiset, are small, most experiments (as shown in Table 1) are 3, and one with 10.
@@
 
-NAŠ ODZIV NA ZGORNJE:
+NAŠ ODZIV:
+We acknowledge that our description of the multiset-trie data structure is not optimized, as we instead strive for a clean represeantation of a new idea.
+We also point out that the absence of the compression makes the multiset-trie a perfectly height-balanced tree, which simplifies the analysis.
 
-The multiset-trie data structure is not optimized but it is given in
-the original and complete form. The absence of the compression makes
-the multiset-trie a perfectly height-balanced tree.
+That said, there are various ways on how to optimize this data structure further.
+For example, the multiplicity of an element can be implemented using a small, customly designed, hash table (instead of a simple array as in the paper).
 
-The multiplicity of an element in a multiset is in a multiset-trie
-implemented by means of an array. There exist a number of possible
-optimizations. For example, the multiplicity of an element can be
-implemented using a small, customly designed, hash table.
+In fact, we added this and some other possible optimiziations to the conclusion of our paper.
+
+-----
 
@@
 
-We dropped the comment about the space minimality of the data
-structure in the conclusions (Section 7). Minimality of the data
-structure is not mentioned in any other part of the paper.
+We agree with the reviewer, and we dropped the comment about the space minimality of the data
+structure in the conclusions (Section 7), accordingly.
+The minimality of the data structure is not mentioned in any other part of the paper.
 
@@
 See the previous reply.
+
+----
 
@@
 
-We have omitted the empirical comparison of the multiset-trie with the
-inverted file from the paper.
+We understand the comment of reviewer. However, one of the authors of this paper recently went to work in industry.
+As he was in charge of the development of inverted index implementation, it seems that a new implementation is hardly possible at the moment.
+That said, we agree that in the current form the Experiment 5 does not fairly compare both data structures.
+Accordingly, we have omitted the empirical comparison of the multiset-trie with the inverted file from the paper.
+
+-----
 
@@
 
-We have added a more detailed presentation of the set-trie in the
-related work.
+We have added a more detailed presentation of the set-trie in the related work.
+Most notable changes can be found in the second paragraph of Section 6.1.
+In the current form, also differences in operations between the two data structures are, in our opinion, exposed better (see third paragraph of Section 6.1).
 
-Page 3 line 37 "at a certan level" -> "at a certain
-level"
+Page 3 line 37 "at a certan level" -> "at a certain level"
 

### 2022-12-09 tex/tex/s_math-Theorems.tex
MDPI template
threw everything into MDPI template.
created parsed file for math section (uses {Theorem} instead of {theorem})

--- 2022-07-29 tex/tex/s_math-noqed.tex
@@
 %
-In this chapter we present theoretical results of time and space complexity of
+In this chapter, we present theoretical results of time and space complexity of
 the multiset-trie data structure. In the following Section~\ref{s:timecomplexity} 
 we discuss the running time complexity of the presented algorithms. First, in 
-Section~\ref{ss:mathmodel}, we present our mathematical model that we use 
+Section~\ref{ss:mathmodel}, we present the mathematical model that we use 
 to describe the distribution of multisets in the multiset-trie and input data. 
-Using probabilistic approach and tools from a Galton-Watson process we 
+Using a probabilistic approach and tools from a Galton-Watson process, we 
 measure the expected cardinality of the multiset-trie in Theorem~\ref{thm:exp_nodes}. 
@@
 \textsc{getAllSubmsets} and \textsc{getAllSupermsets}. We observe that the 
-complexity of functions is exponential. Moreover, the worst case running time 
-complexity is the same for both functions and its upper bound is the cardinality of 
+complexity of functions is exponential. Moreover, the worst-case running time 
+complexity is the same for both functions, and its upper bound is the cardinality of 
 the multiset-trie.
@@
 The remaining "existence" functions are discussed in the Section~\ref{ss:exists}. 
-We observe that out of scope of our mathematical model unlike in functions 
+We observe that out of the scope of our mathematical model, unlike in functions 
 \textsc{getAllSubmsets} and \textsc{getAllSupermsets} the mapping $\phi$ has an 
@@
 We find that the performance of the functions \textsc{submsetExistence} and 
-\textsc{supermsetExistence} in the worst case scenario is also exponential and 
+\textsc{supermsetExistence} in the worst-case scenario is also exponential and 
 does not depend on the outcome of the functions. We give a quite precise upper 
-bound for the worst case running time complexity, which appears to be the same 
-for both functions. However, it must be stressed that for the positive outcome an 
-exponential behavior holds only on specific cases, such as presence of the emptyset 
+bound for the worst-case running time complexity, which appears to be the same 
+for both functions. However, it must be stressed that for the positive outcome, an 
+exponential behavior holds only on specific cases, such as the presence of the empty set 
 in the multiset-trie.
@@
 Finalizing the mathematical analysis, we present the study of space complexity 
-of the multiset-trie in the Section~\ref{s:spacecomplexity}. We show that the space 
+of the multiset-trie in Section~\ref{s:spacecomplexity}. We show that the space 
 used for the storage is asymptotically equal to the size of the input data.
@@
 The performance of the functions will be measured by the number of 
-visited nodes in a multiset-trie during execution of a particular query by the 
+visited nodes in a multiset-trie during the execution of a particular query by the 
 functions \textsc{search}, \textsc{delete}, \textsc{submsetExistence}, 
@@
 \textsc{delete} and the procedure \textsc{insert} is $O(1).$ Nonetheless, in the general 
-case the complexity is $O(\sigma).$
+case, the complexity is $O(\sigma).$
 
-In what follows, we focus on analysis of the more involved functions:
+In what follows, we focus on the analysis of the more involved functions:
 \textsc{submsetExistence}, \textsc{supermsetExistence}, \textsc{getAllSubmsets}
@@
 Because a multiset does not distinguish different orderings, it is assumed, for
-simplicity that all elements are ordered in an ascending order. A multiset $m$ is
+simplicity, that all elements are ordered in ascending order. A multiset $m$ is
 represented as $\{1^{k_1},2^{k_2},\ldots, \sigma^{k_\sigma}\},$ where $e^{k_e}$
@@
 Observe that every internal non-root node has a degree at least~1. Indeed an
-insertion of a multiset requires a construction of a path of length~$\sigma + 1,$
-meaning that if an internal node exists in a multiset-trie it must have a degree
+insertion of a multiset requires construction of a path of length~$\sigma + 1,$
+meaning that if an internal node exists in a multiset-trie, it must have a degree
 at least~1. It also follows that the height of a multiset-trie is always~$\sigma +1$
@@
 Our model assumes that all the inserted multisets are chosen with the same probability,
-meaning that for some $p\in (0,1)$ the following holds:
+meaning that for some $p\in (0,1)$, the following holds:
 \[
@@
 Without conditioning on the existence of any node in multiset-trie, 
-it is easy to describe the probability of existence of any individual node.
+it is easy to describe the probability of the existence of any individual node.
 
-\begin{lemma}\label{l:prob-node-existence}
+\begin{Lemma}\label{l:prob-node-existence}
 Any potential node on a fixed level $i,$ where $i\in \{ 1,2,\ldots, \sigma +1 \}$ exists, %independently of others,
@@
 \end{equation}
-\end{lemma}
+\end{Lemma}
 \begin{proof}
@@
 multiset-trie is $\sigma + 1$ we can calculate the height of the $v$-subtree.
-Taking in account that the root node has height 1, the height of the $v$-subtree is
+Taking into account that the root node has height 1, the height of the $v$-subtree is
 %
@@
 
-\begin{lemma}\label{l:prob-children}
+\begin{Lemma}\label{l:prob-children}
 Suppose that a node $v$ exists at level $1\leq i\leq\sigma$.
@@
 Consider an arbitrary node $v$ on level $1\leq i\leq\sigma.$ According to the
-definition of the multiset-trie a node exists at level $i$ if and only if
+definition of the multiset-trie, a node exists at level $i$ if and only if
 it has at least one child. Note that this is not true for the nodes on the leaf
@@
 Knowing the probability density and probability generating functions of $\xi_{ij}$ 
-from Lemma~\ref{l:prob-children}, we now can estimate the number of nodes in 
+from Lemma~\ref{l:prob-children}, we can now estimate the number of nodes in 
 a randomly generated multiset-trie as follows:
@@
 
-In order to evaluate~(\cref{eq:num_nodes}) we will use some of the tools from 
+In order to evaluate~(\ref{eq:num_nodes}) we will use some of the tools from 
 a Galton-Watson process, see Gardiner~\cite{gardiner1985stochastic} for 
 an introduction.
-Using the equations~(\cref{eq:sum_recursive}) and~(\cref{eq:generating_func}) we can 
+Using the equations~(\ref{eq:sum_recursive}) and~(\ref{eq:generating_func}) we can 
 derive the probability generating function for the random variable $\xi_{i+1}$ as 
@@
 \end{equation}
-Since there is always precisely one node at the root-level, we have $P(\xi_1 = 1) = 1.$ 
+Since there is always precisely one node at the root level, we have $P(\xi_1 = 1) = 1.$ 
 Hence, the probability generating function for the random variable $\xi_1$ is 
@@
 \end{equation}
-which is the initial condition for the recursive equation~(\cref{eq:g_rec}).
+which is the initial condition for the recursive equation~(\ref{eq:g_rec}).
 
-\begin{proposition}\label{prop:exp_rec}
+\begin{Proposition}\label{prop:exp_rec}
 The expectation of the random variable $\xi_{i+1}$ can be expressed 
@@
 for $1\leq i\leq\sigma.$
-\end{proposition}
+\end{Proposition}
 \begin{proof}
@@
 the expectation for the random variable $\xi_{i+1}$ can be derived in terms of 
-the equation~(\cref{eq:g_rec}).
+the equation~(\ref{eq:g_rec}).
 \begin{eqnarray}\label{eq:exp}
@@
 \end{eqnarray}
-According to~(\cref{eq:pdf}) and~(\cref{eq:generating_func}) the value of $G_i(z)$ at 
+According to~(\ref{eq:pdf}) and~(\ref{eq:generating_func}) the value of $G_i(z)$ at 
 $1$ is $1$ and the value of its derivative at $1$ is $\mathbb{E}(\mathcal{B}_0(n,p_{i+1})).$ 
 Substituting the values of $G_i(1^-)$ and $G_i'(1^-),$ and applying the 
-property~(\cref{eq:prop_exp-pgf}) we complete the proof.
+property~(\ref{eq:prop_exp-pgf}) we complete the proof.
 
@@
 
-\begin{theorem}\label{thm:exp_level}
+\begin{Theorem}\label{thm:exp_level}
 Let $\mathcal{M}$ be a multiset-trie defined with parameters $n,$ $\sigma,$ and denote the number 
@@
 \end{equation}
-\end{theorem}
+\end{Theorem}
 \begin{proof}
-According to~(\cref{eq:g_xi_one}) the expected number of nodes on the first level is 1. \\
+According to~(\ref{eq:g_xi_one}) the expected number of nodes on the first level is 1. \\
 Using $\mathbb{E}(\xi_1) = 1$ and the result from Proposition~\ref{prop:exp_rec}
 we get
-\begin{eqnarray*}
-\mathbb{E}(\xi_{i}) &=& \prod_{j=2}^{i} \frac{n p_j}{1-(1-p_j)^n}
+\begin{align*}
+\mathbb{E}(\xi_{i}) &= \prod_{j=2}^{i} \frac{n p_j}{1-(1-p_j)^n}
 = \prod_{j=2}^{i} n \frac{1-(1-p)^{n^{\sigma +1-j}}}{1-(1-p)^{n^{\sigma + 2 -j}}} \\
-&=& n^{i-1} \frac{1-(1-p)^{n^{\sigma +1 -i}}}{1-(1-p)^{n^{\sigma}}}
-\end{eqnarray*}
-
+&= n^{i-1} \frac{1-(1-p)^{n^{\sigma +1 -i}}}{1-(1-p)^{n^{\sigma}}}
+\end{align*}
 \end{proof}
@@
 
-\begin{theorem}\label{thm:exp_nodes}
+\begin{Theorem}\label{thm:exp_nodes}
 The expected cardinality of a multiset-trie defined on parameters $n,$ $\sigma$ 
@@
 where $r = (1-p)^n,$ so $r\in(0,1).$
-\end{theorem}
+\end{Theorem}
 \begin{proof}
 Using the results obtained from Theorem~\ref{thm:exp_level} we compute 
-\begin{eqnarray*}
-\mathbb{E}( | \mathcal{M} | ) &=& \mathbb{E}\left[ \sum_{i=1}^{\sigma+1} \xi_i \right] \\
-&=& \sum_{i=1}^{\sigma + 1} n^{i-1} \frac{1-(1-p)^{n^{\sigma + 1 -i}}}{1-(1-p)^{n^{\sigma}}} \\
-\end{eqnarray*}
-
+\begin{align*}
+\mathbb{E}( | \mathcal{M} | ) &= \mathbb{E}\left[ \sum_{i=1}^{\sigma+1} \xi_i \right] \\
+&= \sum_{i=1}^{\sigma + 1} n^{i-1} \frac{1-(1-p)^{n^{\sigma + 1 -i}}}{1-(1-p)^{n^{\sigma}}}
+\end{align*}
 \end{proof}
@@
 in $\mathcal{M}$ parametrized by an input multiset $m.$ The subtrees that we 
-are interested in are the ones that contain all the submultisets or all the 
-supermultisets of $m.$ In order to calculate the expected cardinality of such 
-subtrees we need the following definition.
+are interested in are the ones that contain all the sub-multisets or all the 
+super-multisets of $m.$ In order to calculate the expected cardinality of such 
+subtrees, we need the following definition.
 
-\begin{definition}\label{def:params}
+\begin{Definition}\label{def:params}
 Let $m = \{ 1^{k_1}, 2^{k_2}, \ldots, \sigma^{k_\sigma} \},$ where 
@@
 \end{equation*}
-\end{definition}
+\end{Definition}
 
@@
 
-\begin{corollary}\label{cor:exp_nodes_param}
+\begin{Corollary}\label{cor:exp_nodes_param}
 Let $M_1, M_2, \alpha_i$ and $\beta_i$ be defined as in previous Definition~\ref{def:params}, 
@@
 \end{equation}
-\end{corollary}
+\end{Corollary}
 %
@@
 Using the results from Theorem~\ref{thm:exp_level} and Theorem~\ref{thm:exp_nodes} 
-we derive the formulas~(\cref{eq:nodes_sub_param}) and~(\cref{eq:nodes_super_param}) 
+we derive the formulas~(\ref{eq:nodes_sub_param}) and~(\ref{eq:nodes_super_param}) 
 by specifying the possible number of nodes on every level in the multiset-trie according 
-to the multiset $m.$ Note that the formula~(\cref{eq:nodes_level}) assumes that on every 
-level but the first one there are $n$ possible nodes. Given submultiset or supermultiset 
+to the multiset $m.$ Note that the formula~(\ref{eq:nodes_level}) assumes that on every 
+level but the first one, there are $n$ possible nodes. Given sub-multiset or super-multiset 
 query and an input multiset $m$ the number of nodes that will be traversed on level $i$ is 
@@
 there is only one root node in any multiset-trie $\mathcal{M},$ which always exists 
-if $M\neq\emptyset$ and is traversed for any type of query (submultiset and 
-supermultiset).
+if $M\neq\emptyset$ and is traversed for any type of query (sub-multiset and 
+super-multiset).
 
@@
 \textsc{getAllSubmsets} and \textsc{getAllSupermsets}. It is obvious that any 
-other algorithm for retrieving all the submultisets or supermultisets has worst 
-case running time complexity at least $O(|M|).$ Hence, the functions 
-\textsc{getAllSubmsets} and \textsc{getAllSupermsets} have the worst case 
+other algorithm for retrieving all the sub-multisets or super-multisets has worst-case
+running time complexity of at least $O(|M|).$ Hence, the functions 
+\textsc{getAllSubmsets} and \textsc{getAllSupermsets} have the worst-case 
 running time complexity $O(|\mathcal{M}|).$ Indeed, the case when the algorithms 
@@
 Now let us consider the function \textsc{getAllSupermsets}. Similarly, the function 
-takes a multiset $m$ as an input argument. However, in this case it returns the 
+takes a multiset $m$ as an input argument. However, in this case, it returns the 
 set of multisets $\{ x\in M : x\supseteq m \}$ from the multiset-trie 
-$\mathcal{M}.$ In order to obtain a traversing of all the multiset-trie one must set 
-$m$ to the smallest possible multiset, i.e. an empty multiset
+$\mathcal{M}.$ In order to obtain a traversing of all the multiset-trie, one must set 
+$m$ to the smallest possible multiset, i.e., an empty multiset
 \begin{equation*}
@@
 
-Thus, we can conclude that the worst case running time complexity of the 
+Thus, we can conclude that the worst-case running time complexity of the 
 functions \textsc{getAllSubmsets} and \textsc{getAllSupermsets} is $O(\mathbb{E}( |\mathcal{M}| )).$ 
@@
 \end{equation*}
-According to the Theorem~\ref{cor:exp_nodes_param} the worst case running 
+According to the Theorem~\ref{cor:exp_nodes_param} the worst-case running 
 time complexity given an input multiset $m$ for the function \textsc{getAllSubmsets} is 
@@
 that the probability distribution function $P(m\in M)$ has an impact on the size of 
-mulitset-trie $\mathcal{M}.$ Moreover, this distribution influences on the performance 
+multiset-trie $\mathcal{M}.$ Moreover, this distribution influences on the performance 
 of the functions \textsc{submsetExistence} and \textsc{supermsetExistence} even more.
 
-For a real world model, such that $P(m\in M)\neq const$ the performance of the search 
+For a real-world model, such that $P(m\in M)\neq const$ the performance of the search 
 algorithms directly depends on the number of nodes on every level $\xi_i.$ When 
-the search functions check if a multiset is in multiset-trie the complete path that 
-corresponds to that multiset is checked. Knowing that fact the search can be optimized 
+the search functions check if a multiset is in multiset-trie, the complete path that 
+corresponds to that multiset is checked. Knowing that fact, the search can be optimized 
 during the construction of a multiset-trie.
@@
 and $\phi.$ Let the frequency of an element $e$ in a multiset $m$ be the multiplicity of 
-$e$ in $m,$ denoted by $\emph{mult}_m(e).$ Then the frequency of an element $e$ 
+$e$ in $m,$ denoted by $\emph{mult}_m(e).$ Then, the frequency of an element $e$ 
 can be defined as a sum $\sum_{m\in M} \emph{mult}_m(e).$ According to the frequencies 
 of elements in $\Sigma,$ the performance of the multiset-trie can be optimized by 
-the mapping $\phi : \Sigma \rightarrow I.$ Indeed the ordering of elements by their 
+the mapping $\phi : \Sigma \rightarrow I.$ Indeed, the ordering of elements by their 
 frequencies has an influence on the performance.
@@
 The frequency of an element $e\in\Sigma$ affects the distribution of $\xi_{\phi(e)}$ 
-as follows. The larger the frequency of $e$ the larger the number of nodes on 
+as follows. The larger the frequency of $e$, the larger the number of nodes on 
 $\phi(e)$ level. 
@@
 %
-%\begin{lemma}\label{l:frequency}
+%\begin{Lemma}\label{l:frequency}
 %Let $M$ be a collection of multisets. Let the frequency of an element $e$ in a 
@@
 %an element the smaller its index.
-%\end{lemma}
+%\end{Lemma}
 %%
@@
 Let us now switch back to our mathematical model and note that the influence 
-of the mapping function $\phi$ in our model has inessential impact on performance, 
-because all the multisets are equally likely and the whole domain $N$ is used for 
+of the mapping function $\phi$ in our model has an inessential impact on performance, 
+because all the multisets are equally likely, and the whole domain $N$ is used for 
 sampling multisets.
@@
 Whenever the result is \emph{false}, i.e. no multiset in $M$ is 
-a submultiset or supermultiset of an input multiset $m,$ both functions in the 
+a sub-multiset or super-multiset of an input multiset $m,$ both functions in the 
 worst case visit all the nodes in $\mathcal{M}$ but the nodes on leaf level. 
-Of course such a case would be very rare assuming a random input model, but 
+Of course, such a case would be very rare, assuming a random input model, but 
 it can be constructed as follows.
@@
 
-Thus, the worst case running time complexity of the functions 
+Thus, the worst-case running time complexity of the functions 
 \textsc{submsetExistence} and \textsc{supermsetExistence} is $O(|\mathcal{M}| - |M|).$ 
@@
 \end{equation*}
-According to Theorem~\ref{cor:exp_nodes_param} the worst case running 
+According to Theorem~\ref{cor:exp_nodes_param} the worst-case running 
 time given an input multiset $m$ for the function \textsc{submsetExistence} is 
@@
 \textsc{supermsetExistence} is \emph{true} one has to guarantee the termination 
-of the algorithm at some node on the leaf level. The worst case scenario can be 
+of the algorithm at some node on the leaf level. The worst-case scenario can be 
 constructed in the same way as for the \emph{false} outcome but with two more 
@@
 will also visit the same amount of nodes as for the \emph{false} case plus one more 
-for the maximal multiset. Hence, the worst case running time complexity for both 
+for the maximal multiset. Hence, the worst-case running time complexity for both 
 outcomes (\emph{true} and \emph{false}) is the same.
@@
 \subsection{Space complexity}\label{s:spacecomplexity}
-As in any efficient algorithm there is always some trade-off between space and 
-time complexity. While offering efficient sub- and supermultiset queries an 
+As in any efficient algorithm, there is always some trade-off between space and 
+time complexity. While offering efficient sub- and super-multiset queries, an 
 additional space must be provided for multisets storage. Clearly, the cardinality 
@@
 Since the ratio $\sigma + 1$ can be obtained for a very specific case only and 
-with a small increase of probability the ratio drops rapidly it can be concluded 
+with a small increase in probability, the ratio drops rapidly it can be concluded 
 that the space complexity of the multiset-trie is $O(|M|).$

### 2023-02-20 tex/tex/s_math-Theorems.tex
all mathematical improvements

@@
 The claim follows by taking the complement probability of the above result. 
-
-%
 \end{proof}
 
-However, in order to determine the distribution of $\xi_{ij}$, one needs a lemma of a different kind.
+Recall that for any given discrete random variable $X$ taking values over positive non-negative integral values, one can define 
+a so-called \emph{probability generating function} (PGF, for short) $G_X$ to be a formal power series defined as 
+\begin{align*}
+G_X(z)=\sum_{i\ge 0} Pr(X=i)z^i.
+\end{align*}
+While the PGFs are usually not meant to be evaluated for concrete values of $z$, certain values have special interpretation when used in PGF, or in a derivation(s) of PGF. 
+For instance, $Pr(X=0)=G_X(0)$, and also $\mathbb{E}(X)=G'_X(1)$.
+For many values of $z$, the function $G_X$ may not converge to any finite value. For this reason it is a common notation to write, in particular
+\[
+G_X(1^-)=\lim_{z\nearrow 1}G_X(z).
+\]
+
+In what follows we will denote a Bernoulli-distributed random variable with parameter $p$ as $Bernoulli(p)$.
+Furthermore we denote
+zero-truncated binomially
+distributed random variable on parameters $n$ and $p_{i+1}$ by $\mathcal{B}_0(n,p_{i+1})$. Finally, $\overset{d}{=}$ indicates so-called equality in distribution.
+
+\begin{Example}
+The probability generating function of a binomial random variable $\mathcal B(n,p)$, the number of successes in $n$ trials, with probability $p$ of success in each trial, is
+
+\begin{align*}
+ G_{\mathcal B(n,p)}(z)=G(z)
+ &=\sum_{k = 0}^n \binom n k p^k  (1 - p)^{n - k} z^k\\
+ &= \sum_{k = 0}^n \binom n k  (p z)^k  (1 - p)^{n - k}\\
+ &=\left((1-p)+pz\right)^{n}.
+\end{align*}
+
+It is thus easy to compute expectation
+\[
+\mathbb{E}(\mathcal{B}(n,p))=G'(1)=np.
+\]
+
+\end{Example}
+
+We now focus our attention to the distribution of $\xi_{ij}$.
 
@@
 \begin{equation}\label{eq:generating_func}
+\mathcal{B}_0(n,p_{i+1})\overset{d}{=}
 G_i(z) = \frac{(1+p_{i+1}(z-1))^n - (1-p_{i+1})^n}{1-(1-p_{i+1})^n}.
 \end{equation}
+
 \end{Lemma}
@@
 \[
-\sum_{k=1}^n X_k \sim\mathcal{B}_0(n,p_{i+1})
+\sum_{k=1}^n X_k \overset{d}{=} \mathcal{B}_0(n,p_{i+1}),
 \] 
@@
 \begin{Corollary}\label{cor:exp_nodes_param}
-Let $M_1, M_2, \alpha_i$ and $\beta_i$ be defined as in previous Definition~\ref{def:params}, 
+Let $M_1, M_2, \alpha_i$ and $\beta_i$ be defined as in  Definition~\ref{def:params}, 
 then the expected cardinality of a multiset-trie subtree $\mathcal{M}_{M_1}$ that contains all the multisets 
@@
 that the probability distribution function $P(m\in M)$ has an impact on the size of 
-multiset-trie $\mathcal{M}.$ Moreover, this distribution influences on the performance 
+multiset-trie $\mathcal{M}.$ 
+Moreover, this distribution influences the performance 
 of the functions \textsc{submsetExistence} and \textsc{supermsetExistence} even more.
@@
 time complexity. While offering efficient sub- and super-multiset queries, an 
-additional space must be provided for multisets storage. Clearly, the cardinality 
-of the set $M$ is smaller than the size of $\mathcal{M},$ because the number 
-of multisets in $\mathcal{M}$ is equal to the number of nodes only on the leaf 
-level. The figure~\ref{f:exp-nodes} demonstrates the relation between the number 
-of multisets stored and the number of nodes needed for storage, where parameters 
-$\sigma$ and $n$ are $26$ and $10,$ respectively.
+additional space must be provided for multisets storage. 
+%Clearly, the cardinality 
+%of the set $M$ is smaller than the size of $\mathcal{M},$ because the number 
+%of multisets in $\mathcal{M}$ is equal to the number of nodes only on the leaf 
+%level. 
+As we increase the parameter $p$, the number of sets we need to store (i.e.
+number of leaves) increases linearly.
+The figure~\ref{f:exp-nodes} shows that the additional overhead space that our data-structure uses does not increase as the tree is becoming more and more dense.
 
@@
 \includegraphics[width=.4\textwidth, keepaspectratio]{exp-nodes-on-probab.pdf}
-\caption{$\mathbb{E}(|\mathcal{M}|)$ and $\mathbb{E}(|M|)$ on probability.}
+\caption{Size of a tree ($\mathbb{E}(|\mathcal{M}|)$) compared to the number of stored sets ($\mathbb{E}(|M|)$), as the probability parameter $p$ increases.
+Parameters 
+$\sigma$ and $n$ are $26$ and $10,$ respectively.
+The additional overhead space that our data-structure uses does not increase as the tree is becoming more and more dense.
+}
 \label{f:exp-nodes}
@@
 \center
-\includegraphics[width=.4\textwidth, keepaspectratio]{ratio-exp-nodes-and-msets-on-prob.pdf}
-\caption{Ratio $\mathbb{E}(\frac{|\mathcal{M}|}{|M|})$ on $p.$}
+\includegraphics[width=.4\textwidth, keepaspectratio]{ratio-exp-nodes-and-msets-on-prob1.pdf}
+\caption{The ratio of "data overhead" derived from Figure \ref{f:exp-nodes}, as the density $p$ of our data-structure increases.
+Unless we are in a very sparse setting, the trade-off in terms of space is small.}
 \label{f:ratio-exp-msets}
@@
 
-Note that analyzing the graph on figure~\ref{f:ratio-exp-msets} we can safely 
+Analyzing the graph on figure~\ref{f:ratio-exp-msets} we can safely 
 say that the upper bound for the ratio is $\sigma + 1.$ The argument holds, 
@@
 Since the ratio $\sigma + 1$ can be obtained for a very specific case only and 
-with a small increase in probability, the ratio drops rapidly it can be concluded 
+with a small increase in probability, the ratio drops rapidly and it can be concluded 
 that the space complexity of the multiset-trie is $O(|M|).$
+
+\subsection{An informal comment on the mathematical analysis}
+In the next section we show how our data-structure preforms in certain environments.
+We will see, in our view, great results throughout various experiments, even when compared with state-of-the-art approaches like Inverted Index.
+
+It was very frustrating to us that, despite those indications, it seems that this structure is very difficult to analyze. 
+Even after restricting the analysis to the (usually simple) case when the input data is uniformly distributed, and after utilising reliable probabilistic tools, our approach unfortunately did not give as clear results as we hoped for.
+We would be very keen to learn of the approach or technique which would shed more insight into the mathematical analysis of our data-structure and its algorithms.

### 2023-02-20 tex/tex/s_math-Theorems.tex
responses to reviewers, and a bit more math

@@
 In order to prove the lemma, we have to show that
-$\xi_{iv}\sim\mathcal{B}_0(n, p_{i+1}).$
+$\xi_{iv}\overset{d}{=}\mathcal{B}_0(n, p_{i+1}).$
 Consider an arbitrary node $v$ on level $1\leq i\leq\sigma.$ According to the
@@
 As it was shown in previous Lemma~\ref{l:prob-children}, the distribution of 
-$X_k$ is $X_k\sim Bernoulli(p_{i+1}).$ Since our model assumes that all the 
+$X_k$ is $X_k\overset{d}{=} Bernoulli(p_{i+1}).$ Since our model assumes that all the 
 multisets in $M$ are chosen uniformly at random, the variables 

### 2023-02-20 tex/tex/s_experiments.tex
Reviews.docx prepared. some responses to comments updated and added. associated tex files updated.

@@
 % anouncement of the experiments
-In the following sections, we will present the behavior of the multiset-trie data 
-structure in four experiments.
-The first three experiments use artificially generated data, and the fourth experiment
-uses real-world data. In the 
-Experiment~\hyperref[s:exp1]{1} a special case of the multiset-trie is considered. 
+In the following Sections 1-5 we will present the behavior of the multiset-trie data 
+structure depending on the selected parameters as well as the comparative 
+benchmark of the multiset-trie against B-tree implementation of inverted index. 
+We start with experiments that are performed on an artificially generated data in 
+order to give a general picture of the multiset-trie performance.
+
+In the Experiment~\hyperref[s:exp1]{1} a special case of the multiset-trie is considered. 
 Only sets are allowed to be stored in the data structure, i.e., the maximally allowed 
@@
 
-Finally, the fourth experiment on the multiset-trie uses real-world data. In 
+The fourth experiment on the multiset-trie uses real-world data. In 
 Experiment~\hyperref[s:exp4]{4} the influence of the mapping $\phi$ is studied. 
@@
 
-% 
-%After all the experiments we present an empirical comparison of multiset-trie 
-%data structure with B-tree based inverted index. We use inverted index 
-%to store and retrieve multisets in the same way as it is described in the paper by 
-%Helmer and Moerkotte~\cite{Helmer2003} for sets. In the comparison we use 
-%three types of queries exact, submultiset and supermultiset retrieval.
+Finally, the empirical comparison of multiset-trie data structure with
+B-tree based inverted index is presemted in Experiment~\hyperref[s:exp5]{5}. We use
+inverted index to store and retrieve multisets in the same way as it
+is described in the paper by Helmer and Moerkotte~\cite{Helmer2003}
+for sets. In the comparison we use three types of queries: exact,
+submultiset and supermultiset retrieval.
 
@@
 \label{fig:e3m2}}{\includegraphics[width=.45\textwidth]{exp4-m2.pdf}}
-\caption{Exsitence functions in Experiment 3.}
+\caption{Existence functions in Experiment 3.}
 \end{figure}
@@
 nodes will not be changed as $\phi$ varies.
+
+\subsection{Experiment 5} \label{s:exp5}
+
+In this experiment, we demonstrate the performance of multiset-trie
+data structure compared to the inverted index based on the
+B-tree. Both data structures are implemented in the programming
+language \CC, providing in this way an experimental setup for a fair
+comparison \cite{akulich2019mstrie}.
+
+%((Brief description of the implementation of the inverted files?))
+The inverted index is implemented using an idea
+from~\cite{Helmer2003}. An inverted index structure consists of two
+parts: a dictionary and postings. In our case, a dictionary is
+implemented as an in-memory B-tree where keys are all distinct values
+from a domain represented by a set $\Sigma.$ The postings are
+represented by lists of multisets that contain a particular element
+from $\Sigma.$ Each list item in postings contains a cardinality of a
+multiset, which speeds up the containment queries.
+
+The experiment uses the input data for the construction of the given
+data structure and the test data for the execution of the operations
+on the given data structure. The input data comprises a set of
+randomly generated multisets used for the construction of a data
+structure. The test data includes the set of multisets together with
+the operations that are evaluated. The input and test data were
+generated with respect to parameters $\sigma$ and $n$ as presented in
+xTable~\ref{t:benchmark}.
+
+\begin{table}[h]
+\center
+\begin{tabular}{|c|c|}
+\hline
+$\sigma$ & $n$ \\
+\hline
+5		& 1\\
+\hline
+30	& 1 \\
+\hline
+5		& 3 \\
+\hline
+15	& 3 \\
+\hline
+30	& 3 \\
+\hline
+10	& 10 \\
+\hline
+\end{tabular}
+\caption{Configuration of $\sigma$ and $n$ in benchmark.}
+\label{t:benchmark}
+\end{table}
+
+We tested all three types of a query on all of the configurations from
+Table 1 results in 18 experiments total, i.e., 6 experiments per
+query type. In each experiment, we measured the average time consumed
+by the data structure to process the query. The results of the exact
+search, sub-multiset, and super-multiset search experiments are
+presented in tables Table~\ref{t:res_ex}, Table~\ref{t:res_sub} and
+Table~\ref{t:res_sup}, respectively.
+
+\begin{table}[h]
+\center
+\begin{tabular}{|r|r|r|r|}
+\hline
+\multicolumn{1}{|c|}{$\sigma$} & 
+\multicolumn{1}{c|}{$n$} & 
+\multicolumn{1}{c|}{Multiset-trie ($\mu s$)} & 
+\multicolumn{1}{c|}{Inverted index ($\mu s$)} \\
+\hline
+5		& 1 & 3.45 & 17782.35\\
+\hline
+30	& 1 & 4.18 & 24865.93\\
+\hline
+5		& 3 & 2.20 & 1508.81\\
+\hline
+15	& 3 & 4.39 & 2146.36\\
+\hline
+30	& 3 & 10.67 & 3639.97\\
+\hline
+10	& 10 & 6.93 & 384.05\\
+\hline
+\end{tabular}
+\caption{Exact search.}
+\label{t:res_ex}
+\end{table}
+
+\begin{table}[h]
+\center
+\begin{tabular}{|r|r|r|r|}
+\hline
+\multicolumn{1}{|c|}{$\sigma$} & 
+\multicolumn{1}{c|}{$n$} & 
+\multicolumn{1}{c|}{Multiset-trie ($\mu s$)} & 
+\multicolumn{1}{c|}{Inverted index ($\mu s$)} \\
+\hline
+5		& 1			& 8.96 & 73500.84\\
+\hline
+30	& 1			& 17.33 & 547572.74\\
+\hline
+5		& 3			& 117.95 & 162360.43\\
+\hline
+15	& 3			& 20.74 & 443321.39\\
+\hline
+30	& 3 		& 23.75 & 947706.14\\
+\hline
+10	& 10		& 55.59 & 466022.68\\
+\hline
+\end{tabular}
+\caption{Sub-multiset search.}
+\label{t:res_sub}
+\end{table}
+
+\begin{table}[h]
+\center
+\begin{tabular}{|r|r|r|r|}
+\hline
+\multicolumn{1}{|c|}{$\sigma$} & 
+\multicolumn{1}{c|}{$n$} & 
+\multicolumn{1}{c|}{Multiset-trie ($\mu s$)} & 
+\multicolumn{1}{c|}{Inverted index ($\mu s$)} \\
+\hline
+5		& 1 		& 10.63 & 63073.86\\
+\hline
+30	& 1 		& 14.65 & 449251.68\\
+\hline
+5		& 3 		& 171.04 & 163256.77\\
+\hline
+15	& 3 		& 43.42 & 425733.80\\
+\hline
+30	& 3 		& 22.06 & 729831.34\\
+\hline
+10	& 10 	& 58.32 & 373784.81\\
+\hline
+\end{tabular}
+\caption{Super-multiset search.}
+\label{t:res_sup}
+\end{table}
+
+We can see that multiset-trie outperforms the inverted index in all of the
+experiments by up to 4 orders of magnitude. In an exact search,
+multiset-trie has to traverse only up to $\sigma+1$ nodes to get a query
+result. It can be seen from the results that with the increase of $\sigma$
+the processing time for multiset-tire also increases. Multiplicity
+also affects the processing time; however, this happens
+passively. Multiplicity, or degree of a node $n$, defines the shape of
+multisets that are stored in multiset-trie. Thus, it affects the
+structure and density of the multiset-trie.
+
+As for the inverted index, all three operations must first fetch all
+postings for each particular element of a test multiset. Afterward,
+the intersection of postings is computed to answer the query. The
+operations use more processing time than a simple tree traversal,
+which we can see from the results. Postings are filtered on-the-fly to
+reduce the cost of the intersection.
+
+Implementations of the sub-multiset and super-multiset are similar in
+the case of the inverted file. The algorithm consists of the same
+steps. First, the postings are fetched for each element of the test
+multiset. Depending on the particular operation, postings are filtered
+on-the-fly. Finally, the union or the intersection of the filtered set
+of postings is computed. Note that the processing time increases with
+the size of the inverted index because of the increased sizes of
+postings.
+
+In the case of multiset-trie, only a traversal of the tree is
+required, which is much faster than the processing of postings, as we
+can see from the results. In the worst case, the whole tree is
+traversed, but the same is for an inverted file.

### 2023-02-20 tex/tex/s_operations.tex
Reviews.docx prepared. some responses to comments updated and added. associated tex files updated.

@@
 an array $m$ of constant length $\sigma,$ where $i$-th cell represents the element 
-$\phi^{-1}(i)$ from $\Sigma$ with multiplicity $m[i].$ From now on we 
+$\phi^{-1}(i)$ from $\Sigma$ with multiplicity $m[i].$ From now on, we 
 agree that the first cell of an array has index 1. Let us have an example of a 
@@
 
-In the following subsections we will present each operation of the multiset-trie 
+In the following subsections, we will present each operation of the multiset-trie 
 data structure separately.
@@
 type Multiset into multiset-trie $\mathcal{M}$. If the complete path already 
-exists, then procedure leaves the structure unchanged. Otherwise it extends 
+exists, then the procedure leaves the structure unchanged. Otherwise, it extends 
 partially existing or creates a new complete path. The procedure does not return 
@@
 \subsection{Delete} \label{s:delete}
-The function \textsc{delete}($\mathcal{M},$ $m$) searches for the complete path 
+Function \textsc{delete}($\mathcal{M},$ $m$) searches for the complete path 
 that corresponds to $m$ in order to remove it. If the path can not be found, the 
-function immediately returns false. During search, the function keeps track of the 
+function immediately returns false. During the search, the function keeps track of the 
 number of children for every node. It marks the nodes that have more than one child 
-as \emph{parent nodes} and remembers the label of the child which is a potential node 
+as \emph{parent nodes} and remembers the label of the child, which is a potential node 
 where the sub-tree will be cut to remove the multiset. The parent node is needed to 
-perform a removal, because the multiset-tire is an explicit data structure. When search 
-is completed, the function removes the sub-tree of the last found parent node, and 
-returns true. In such a way after deletion all the prefixes for other multisets are 
+perform a removal because the multiset-trie is an explicit data structure. When the search 
+is completed, the function removes the sub-tree of the last found parent node and 
+returns true. In such a way, after deletion, all the prefixes for other multisets are 
 preserved in $\mathcal{M}$ and $m$ is removed. The function \textsc{delete} is 
@@
 not found in $\mathcal{M},$ the function uses multiset-trie to find the closest 
-(the largest) sub-multiset of $m$ in $\mathcal{M}$ by decreasing multiplicity of 
+(the largest) sub-multiset of $m$ in $\mathcal{M}$ by decreasing the multiplicity of 
 elements in $m.$ The parameter $dev$ is used to limit a maximal deviation of 
 multiplicity for a particular element in $x$ with respect to $m.$ 
-At every level the function tries to proceed with the largest possible multiplicity of 
+At every level, the function tries to proceed with the largest possible multiplicity of 
 an element that is provided by $m.$ However, when the function reaches some level 
-where it meets a \emph{Null} node and can not go further using path provided by 
+where it meets a \emph{Null} node and can not go further using the path provided by 
 $m,$ it decreases the multiplicity of an element 
 that corresponds to a current level with respect to the specified maximal deviation. 
-Thus, the function can decrease multiplicity of an element or eventually skip it in 
+Thus, the function can decrease the multiplicity of an element or eventually skip it in 
 order to find the closest $x\subseteq m.$ The function \textsc{submsetExistence} 
@@
 smallest) super-multiset $x$ of $m$ in $\mathcal{M}$ by increasing multiplicity 
-of elements in $m.$ At every level the function tries to proceed with the smallest 
-possible multiplicity of an element that is provided by $m.$ However, when function 
+of elements in $m.$ At every level, the function tries to proceed with the smallest 
+possible multiplicity of an element that is provided by $m.$ However, when the function 
 reaches some level where it meets a \emph{Null} node and can not go further using 
-path provided by $m,$ it increases the multiplicity of an element that corresponds to a 
+the path provided by $m,$ it increases the multiplicity of an element that corresponds to a 
 current level according to the deviation parameter $dev.$ Thus, the function \textsc{supermsetExistence} 
@@
 terminate on the first existing sub/super-multiset, but store the results and 
-continue procedure until all existing sub/super-multisets in $\mathcal{M}$ are 
+continue the procedure until all existing sub/super-multisets in $\mathcal{M}$ are 
 found and stored. The functions \textsc{getAllSubmsets} and \textsc{getAllSupermsets} 
@@
 
-In order to record a multiset during multiset-trie traversal we use the variable $x$ in the algorithms.
+In order to record a multiset during multiset-trie traversal, we use the variable $x$ in the algorithms.
 It is an empty array of size $\sigma$ where we store multiplicities of elements at 
@@
 sub-multisets of $m$ found during traversal. Both variables $x$ and $result$ are presented 
-as global, however they could be passed to the recursive function as parameters.
+as global, however, they could be passed to the recursive function as parameters.
 

### 2023-02-21 tex/tex/s_data_structure.tex
reponses to some comments and updates of corresponding tex files.

@@
 %
-Let $\Sigma$ be a set of distinct symbols that define an alphabet and let 
+Let $\Sigma$ be a set of distinct symbols that define an alphabet, and let 
 $\sigma$ be the cardinality of $\Sigma.$ The \emph{multiset-trie} data structure 
 stores multisets that are composed of symbols from the alphabet $\Sigma.$ It 
-provides the basic tree data structure operations such as insert, delete and 
-search together with multiset containment and membership operations such as 
-submultiset and supermultiset that will be discussed in the next section 
-in greater details.
+provides the basic tree data structure operations such as insert, delete, and 
+search, together with multiset containment operations for searching sub-multisets
+and super-multisets that will be discussed in the next section in greater detail.
 
-Multiset ignores the ordering of its elements by definition, which allows us to 
+A multiset ignores the ordering of its elements by definition, which allows us to 
 define a bijective mapping $\phi:\Sigma \rightarrow I,$ where $I$ is the set of 
-integers $\{ 1,2,3, \ldots, \sigma \}.$ In this way, we obtain an indexing of 
+integers $\{ 1,2,3, \ldots, \sigma \}.$ In this way, we obtain indexing of 
 elements from the alphabet $\Sigma,$ so we can work directly with integers 
-rather then with specific symbols from $\Sigma.$
+rather than with specific symbols from $\Sigma.$
 
-The multiset-trie is an $n$-ary tree based data structure with the properties of 
-trie. A node in multiset-trie always has degree $n,$ i.e. $n$ children. Some of 
+The multiset-trie is an $n$-ary tree-based data structure with the properties of 
+the trie. A node in multiset-trie always has degree $n,$ i.e., $n$ children. Some of 
 the children may be \emph{Null} (non-existing), but the number of \emph{Null} 
@@
 children, are labeled from left to right with labels $c_j,$ where $j\in \{ 0, 1, \ldots, n-1 \}.$ 
-Every two child nodes $u$ and $v$ that share the same parent node have different labels.
+Every pair of child nodes $u$ and $v$ that share the same parent node have different labels.
 
@@
 %
-Levels in multiset-trie are enumerated by their height, i.e. a level $L_i$ has 
+Levels in multiset-trie are enumerated by their height, i.e., a level $L_i$ has 
 height $i.$ The connection between level height in a multiset-trie and symbols 
@@
 We define a \emph{complete path} to be the path of length $\sigma$ in a 
-multiset-trie with the end points at root node (the 1st level) and $LL$. Thus, 
+multiset-trie with the endpoints at the root node (the 1st level) and $LL$. Thus, 
 a multiset $m$ is inserted into a multiset-trie if and only if there exists a 

### 2023-02-21 tex/tex/s_intro.tex
reponses to some comments and updates of corresponding tex files.

@@
 % multisets def; not ordered
-A multiset is a collection of elements where a particular element can have more than one instance.
+A multiset is a collection of elements that can have more than one instance.
 As in the case of ordinary sets, the ordering of the elements in multisets is not relevant.
@@
 However, it is not as efficient for  containment queries.
-The search based on the containment relation requires access to the collections $C\subseteq\/S$ of multisets that are related to a multiset $m$ either by a sub-multiset or a super-multisets relationship.
+The search based on the containment relation requires access to the collections $C\subseteq\/S$ of multisets that are related to a multiset $m$ either by a sub-multiset or a super-multiset relationship.
 
@@
 In particular, we estimate the size of a sub-tree traversed by a containment query and give an insight into the time complexity of containment queries.
+
 In addition to the mathematical analysis, the size of a sub-tree visited by a multiset containment query is also the central focus of the empirical analysis.
-We carefully designed the experiments to unravel the main features of the time complexity space.
+We carefully designed the experiments to unravel the main features of the search space.
+We observe how the number of visited nodes depends on various parameters, such as the density of the tree, and the ordering of multiset elements.
 
-Finally, the mathematical, as well as empirical analyses, show an influence of ordering the elements of multisets on the efficiency of storing and processing multisets.
+Finally, the mathematical, as well as empirical analyses show the influence of ordering the elements of multisets on the efficiency of storing and processing multisets.
 We show that the ordering, which is based on the frequencies of the multiset elements, can speed up the multiset containment queries by orders of magnitude.
@@
 Synthetic and real-world data sets are used in experiments that are designed to study the performance of multiset containment operations.
+Further, the multiset-trie is empirically compared to the inverted index in a separate experiment.
 The experiments highlight the methods for optimizing a multiset-trie.
@@
 The related work is reviewed in Section~\ref{c:relwork}.
-We present the related approaches, including a description of a set-trie data structure, the inverted indexes, the signature indexes, and the multisets in relational database systems.
+This section describes a set-trie data structure, the inverted indexes, the signature indexes, and the multisets in relational database systems.
 Finally, the concluding remarks and the future work are presented in Section~\ref{c:conclusions}.

### 2023-02-21 tex/tex/s_math-Theorems.tex
a lot of changes and added text to responses.

@@
 We find that the performance of the functions \textsc{submsetExistence} and 
-\textsc{supermsetExistence} in the worst-case scenario is also exponential and 
-does not depend on the outcome of the functions. We give a quite precise upper 
+\textsc{supermsetExistence} in the worst-case scenario is also proportional to the size of the constructed trie. We give a rather technical upper 
 bound for the worst-case running time complexity, which appears to be the same 
-for both functions. However, it must be stressed that for the positive outcome, an 
-exponential behavior holds only on specific cases, such as the presence of the empty set 
+for both functions. 
+However, it must be stressed that for the positive outcome, this 
+worst-case behavior holds only on specific cases, such as the presence of the empty set 
 in the multiset-trie.

### 2023-02-22 tex/tex/s_operations.tex
responses added; two XXX left; s_operations.tex updated.

@@
 %
-The operations supported by the multiset-trie data structure are as follows.
+The operations supported by the multiset-trie data structure are as
+follows. 
 %
@@
 \end{enumerate}
+
+The parameter $dev$ is used to specify the maximal deviation in the multiplicity
+of multiset containment operations. It is utilized to limit the search in multiset
+containment queries to the sub-multisets and super-multisets that are
+the closest to the input multiset $m$. In addition, we use $dev$ for
+the implementation of multiset similarity search that, given $m$, retrieves
+from $\mathcal{M}$ all sub-multisets or super-multisets that are similar to $m$
+with respect to the deviation $dev$.
 
@@
 
-\subsection{Sub-multiset existence} \label{s:subexists}
+\subsection{Sub-multiset and super-multiset existence}
+\label{s:subexists} \label{s:superexists}
+The functions \textsc{submsetExistence} and \textsc{supermsetExistence} are
+symmetrical in the following sense. Let a multiset $m$ represent the borderline
+in $\mathcal{M}$ defined by a path from the root to a leaf by following
+the elements from $m$. The operation \textsc{submsetExistence} searches
+the left part of $\mathcal{M}$ and the operation \textsc{supermsetExistence}
+the right part of $\mathcal{M}$.
+
 The function \textsc{submsetExistence}($\mathcal{M},m,dev$) checks if there exists 
@@
 
-\subsection{Super-multiset existence} \label{s:superexists}
+%\subsection{Super-multiset existence} \label{s:superexists}
 The function \textsc{supermsetExistence}($\mathcal{M},m,dev$) checks if there 
 exists super-multiset $x$ of a given multiset $m$ in $\mathcal{M},$ such that 
-condition $| x[i] - m[i] |\leq dev$ is satisfied, where $1\leq i \leq \sigma.$ By analogy 
-to the function \textsc{submsetExistence}, the function \textsc{supermsetExistence} 
-starts by searching for an exact match $x=m$ in $\mathcal{M}.$ If an exact 
-match is not found in $\mathcal{M},$ the function searches for the closest (the 
-smallest) super-multiset $x$ of $m$ in $\mathcal{M}$ by increasing multiplicity 
-of elements in $m.$ At every level, the function tries to proceed with the smallest 
-possible multiplicity of an element that is provided by $m.$ However, when the function 
-reaches some level where it meets a \emph{Null} node and can not go further using 
-the path provided by $m,$ it increases the multiplicity of an element that corresponds to a 
-current level according to the deviation parameter $dev.$ Thus, the function \textsc{supermsetExistence} 
-can increase multiplicity of an element up to $n-1,$ where $n$ is the degree of 
-a node in $\mathcal{M},$ to find the closest super-multiset $x\supseteq m$ in 
-$\mathcal{M}.$ The function \textsc{supermsetExistence} is presented in 
+condition $| x[i] - m[i] |\leq dev$ is satisfied, where $1\leq i \leq \sigma.$
+Symmetrically to the function \textsc{submsetExistence}, the function
+\textsc{supermsetExistence} searches first for an exact match $x=m$ in
+$\mathcal{M}.$ If such $x$ does not exist in $\mathcal{M},$ then
+the function searches on the right side of the borderline (an exact match)
+defined by $m$. Since we would like to find the closest (the smallest)
+super-multiset we increase the multiplicity of elements in $m$ at every
+level of $\mathcal(M)$ starting with the multiplicities of $m.$
+The function \textsc{supermsetExistence} is presented in
 Algorithm~\ref{alg:superxists}.

### 2023-02-22 tex/tex/s_experiments.tex
all reviewers comments answered.

@@
 
-%Performance of the functions will be measured by 
+The current implementation of multiset-trie is not optimized for space consumption.
+A node in the multiset-trie is implemented as a record indicating the level of the tree
+that corresponds to a symbol from $\Sigma$, and including an array of pointers to
+its children indexed by the multiplicities of the child symbols. The arrays have
+fixed length $n$ from the interval $[0,n].$
+Only some elements of an array may point to a child node, while others include
+$Null$ value.
+
+% Performance of the functions will be measured by 
 %the number of visited nodes in multiset-trie by the particular function. In 
@@
 
-Implementations of the sub-multiset and super-multiset are similar in
+Implementations of the multiset containment operations
+for searching sub-multisets and super-multisets are similar in
 the case of the inverted file. The algorithm consists of the same

### 2023-02-22 tex/tex/s_operations.tex
all reviewers comments answered.

@@
 symmetrical in the following sense. Let a multiset $m$ represent the borderline
-in $\mathcal{M}$ defined by a path from the root to a leaf by following
+in $\mathcal{M}$ defined by a path from the root to a leaf following
 the elements from $m$. The operation \textsc{submsetExistence} searches
@@
 $\mathcal{M}.$ If such $x$ does not exist in $\mathcal{M},$ then
-the function searches on the right side of the borderline (an exact match)
-defined by $m$. Since we would like to find the closest (the smallest)
-super-multiset we increase the multiplicity of elements in $m$ at every
-level of $\mathcal(M)$ starting with the multiplicities of $m.$
-The function \textsc{supermsetExistence} is presented in
-Algorithm~\ref{alg:superxists}.
+the function searches on the right side of the borderline representing
+an exact match of $m$ in $\mathcal{M}.$ Since we would like to find
+the closest (the smallest) super-multiset we increase the multiplicity of
+elements in $m$ at every level of $\mathcal(M)$ starting with
+the multiplicities of $m.$ The function \textsc{supermsetExistence}
+is presented in Algorithm~\ref{alg:superxists}.
 

### 2023-02-24 tex/tex/s_relwork.tex
a lot of changes and added text to responses.
additional comments to reviewers added, and s_relwork.tex changed!

@@
 The downside of such a generalization is that multiset-trie no longer supports path compression that was obtained in set-trie.
-However, the design of multiset-trie provides storage of multisets with constant worst-case time complexity of the set containment operations.
+%However, the design of multiset-trie provides storage of multisets with constant worst-case time complexity of the set containment operations.
 
@@
 The results show that, in general, the IR engine performs better than an RDBMS on containment queries.
-They have identified the problems that reason for the poor performance of the containment operations in an RDBMS and showed that with some modifications, an RDBMS could perform this class of queries more efficiently.
+The significant causes that differentiate the performance of the IR and RDBMS implementations are the join algorithms employed and the hardware cache utilization.
+The differences in join algorithms and the cache access method that makes IR queries faster were identified precisely.
+First, the multi-predicate merge join of the IR engine is different from the standard merge join, and the index nested-loop join algorithms.
+Second, the study of the utilization of cache in the multi-predicate merge join, standard merge join, and index-nested loop joins has been done, identifying more precisely the differences in the algorithms.
+They have presented that with some modifications, an RDBMS could perform this class of queries more efficiently.
 

### 2023-03-09 tex/tex/s_experiments.tex
update for marked points in the paper (in s_operations.tex and s_experimanets.tex)

@@
 generated with respect to parameters $\sigma$ and $n$ as presented in
-xTable~\ref{t:benchmark}.
+Table~\ref{t:benchmark}.
 
@@
 \end{tabular}
-\caption{Configuration of $\sigma$ and $n$ in benchmark.}
+\caption{Selected values of parameters $\sigma$ and $n$ in Experiment 5.}
 \label{t:benchmark}
@@
 
-We tested all three types of a query on all of the configurations from
+The three types of query tested on all of the configurations from
 Table 1 results in 18 experiments total, i.e., 6 experiments per

### 2023-03-10 tex/algorithms-2122309-proof.tex
mdpi editing

%  LaTeX support: latex@mdpi.com 
%  For support, please attach all files needed for compiling as well as the log file, and specify your operating system, LaTeX version, and LaTeX editor.

%=================================================================
\documentclass[algorithms,article,accept,pdftex,moreauthors]{Definitions/mdpi} 
% For posting an early version of this manuscript as a preprint, you may use "preprints" as the journal and change "submit" to "accept". The document class line would be, e.g., \documentclass[preprints,article,accept,moreauthors,pdftex]{mdpi}. This is especially recommended for submission to arXiv, where line numbers should be removed before posting. For preprints.org, the editorial staff will make this change immediately prior to posting.
%\usepackage{showframe}
%--------------------
% Class Options:
%--------------------
%----------
% journal
%----------
% Choose between the following MDPI journals:
% acoustics, actuators, addictions, admsci, adolescents, aerospace, agriculture, agriengineering, agronomy, ai, algorithms, allergies, alloys, analytica, animals, antibiotics, antibodies, antioxidants, applbiosci, appliedchem, appliedmath, applmech, applmicrobiol, applnano, applsci, aquacj, architecture, arts, asc, asi, astronomy, atmosphere, atoms, audiolres, automation, axioms, bacteria, batteries, bdcc, behavsci, beverages, biochem, bioengineering, biologics, biology, biomass, biomechanics, biomed, biomedicines, biomedinformatics, biomimetics, biomolecules, biophysica, biosensors, biotech, birds, bloods, blsf, brainsci, breath, buildings, businesses, cancers, carbon, cardiogenetics, catalysts, cells, ceramics, challenges, chemengineering, chemistry, chemosensors, chemproc, children, chips, cimb, civileng, cleantechnol, climate, clinpract, clockssleep, cmd, coasts, coatings, colloids, colorants, commodities, compounds, computation, computers, condensedmatter, conservation, constrmater, cosmetics, covid, crops, cryptography, crystals, csmf, ctn, curroncol, currophthalmol, cyber, dairy, data, dentistry, dermato, dermatopathology, designs, diabetology, diagnostics, dietetics, digital, disabilities, diseases, diversity, dna, drones, dynamics, earth, ebj, ecologies, econometrics, economies, education, ejihpe, electricity, electrochem, electronicmat, electronics, encyclopedia, endocrines, energies, eng, engproc, ent, entomology, entropy, environments, environsciproc, epidemiologia, epigenomes, est, fermentation, fibers, fintech, fire, fishes, fluids, foods, forecasting, forensicsci, forests, foundations, fractalfract, fuels, futureinternet, futureparasites, futurepharmacol, futurephys, futuretransp, galaxies, games, gases, gastroent, gastrointestdisord, gels, genealogy, genes, geographies, geohazards, geomatics, geosciences, geotechnics, geriatrics, hazardousmatters, healthcare, hearts, hemato, heritage, highthroughput, histories, horticulturae, humanities, humans, hydrobiology, hydrogen, hydrology, hygiene, idr, ijerph, ijfs, ijgi, ijms, ijns, ijtm, ijtpp, immuno, informatics, information, infrastructures, inorganics, insects, instruments, inventions, iot, j, jal, jcdd, jcm, jcp, jcs, jdb, jeta, jfb, jfmk, jimaging, jintelligence, jlpea, jmmp, jmp, jmse, jne, jnt, jof, joitmc, jor, journalmedia, jox, jpm, jrfm, jsan, jtaer, jzbg, kidney, kidneydial, knowledge, land, languages, laws, life, liquids, literature, livers, logics, logistics, lubricants, lymphatics, machines, macromol, magnetism, magnetochemistry, make, marinedrugs, materials, materproc, mathematics, mca, measurements, medicina, medicines, medsci, membranes, merits, metabolites, metals, meteorology, methane, metrology, micro, microarrays, microbiolres, micromachines, microorganisms, microplastics, minerals, mining, modelling, molbank, molecules, mps, msf, mti, muscles, nanoenergyadv, nanomanufacturing, nanomaterials, ncrna, network, neuroglia, neurolint, neurosci, nitrogen, notspecified, nri, nursrep, nutraceuticals, nutrients, obesities, oceans, ohbm, onco, oncopathology, optics, oral, organics, organoids, osteology, oxygen, parasites, parasitologia, particles, pathogens, pathophysiology, pediatrrep, pharmaceuticals, pharmaceutics, pharmacoepidemiology, pharmacy, philosophies, photochem, photonics, phycology, physchem, physics, physiologia, plants, plasma, pollutants, polymers, polysaccharides, poultry, powders, preprints, proceedings, processes, prosthesis, proteomes, psf, psych, psychiatryint, psychoactives, publications, quantumrep, quaternary, qubs, radiation, reactions, recycling, regeneration, religions, remotesensing, reports, reprodmed, resources, rheumato, risks, robotics, ruminants, safety, sci, scipharm, seeds, sensors, separations, sexes, signals, sinusitis, skins, smartcities, sna, societies, socsci, software, soilsystems, solar, solids, sports, standards, stats, stresses, surfaces, surgeries, suschem, sustainability, symmetry, synbio, systems, taxonomy, technologies, telecom, test, textiles, thalassrep, thermo, tomography, tourismhosp, toxics, toxins, transplantology, transportation, traumacare, traumas, tropicalmed, universe, urbansci, uro, vaccines, vehicles, venereology, vetsci, vibration, viruses, vision, waste, water, wem, wevj, wind, women, world, youth, zoonoticdis

%---------
% article
%---------
% The default type of manuscript is "article", but can be replaced by: 
% abstract, addendum, article, book, bookreview, briefreport, casereport, comment, commentary, communication, conferenceproceedings, correction, conferencereport, entry, expressionofconcern, extendedabstract, datadescriptor, editorial, essay, erratum, hypothesis, interestingimage, obituary, opinion, projectreport, reply, retraction, review, perspective, protocol, shortnote, studyprotocol, systematicreview, supfile, technicalnote, viewpoint, guidelines, registeredreport, tutorial
% supfile = supplementary materials

%----------
% submit
%----------
% The class option "submit" will be changed to "accept" by the Editorial Office when the paper is accepted. This will only make changes to the frontpage (e.g., the logo of the journal will get visible), the headings, and the copyright information. Also, line numbering will be removed. Journal info and pagination for accepted papers will also be assigned by the Editorial Office.

%------------------
% moreauthors
%------------------
% If there is only one author the class option oneauthor should be used. Otherwise use the class option moreauthors.

%---------
% pdftex
%---------
% The option pdftex is for use with pdfLaTeX. If eps figures are used, remove the option pdftex and use LaTeX and dvi2pdf.

%=================================================================
% MDPI internal commands
\firstpage{1} 
\makeatletter 
\setcounter{page}{\@firstpage} 
\makeatother
\pubvolume{1}
\issuenum{1}
\articlenumber{0}
\pubyear{2023}
\copyrightyear{2023}
\externaleditor{Academic Editor:Frank Werner}%MDPI: Please add Academic Editor.
\datereceived{10 December 2022 } 
\daterevised{22 February 2023  } % Comment out if no revised date
\dateaccepted{8 March 2023 } 
\datepublished{ } 
%\datecorrected{} % For corrected papers: "Corrected: XXX" date in the original paper.
%\dateretracted{} % For corrected papers: "Retracted: XXX" date in the original paper.
\hreflink{https://doi.org/} % If needed use \linebreak
%\doinum{}
%\pdfoutput=1 % Uncommented for upload to arXiv.org
%------------------------------------------------------------------
% The following line should be uncommented if the LaTeX file is uploaded to arXiv.org
%\pdfoutput=1

%=================================================================
% Add packages and commands here. The following packages are loaded in our class file: fontenc, inputenc, calc, indentfirst, fancyhdr, graphicx, epstopdf, lastpage, ifthen, lineno, float, amsmath, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, tabto, xcolor, soul, multirow, microtype, tikz, totcount, changepage, attrib, upgreek, cleveref, amsthm, hyphenat, natbib, hyperref, footmisc, url, geometry, newfloat, caption

%tikz stuff
%\usepackage{tikz}
\usetikzlibrary{arrows}
\tikzset{
	treenode/.style = {align=center, inner sep=0pt, font=\sffamily\bfseries, 
		text=black, draw=black, thick},
	node1/.style = {treenode, rectangle, rounded corners=5pt, fill=green!10, 
		minimum width=2em, minimum height=1.5em, text width=1.5em},
	dummy/.style = {treenode, rectangle, rounded corners=2.5pt, fill=green!10, 
		minimum width=1em, minimum height=0.75em},
	emp/.style = {treenode, circle, fill=white, minimum width=1.5em},
	none/.style = {treenode, rectangle, fill=black, minimum width=0.5em, 
		minimum height=0.5em}
}

% for our algorithms
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}

%cref
%\usepackage[capitalise]{cleveref}

\usepackage{siunitx}

%\usepackage{subcaption}
\usepackage[labelformat=simple]{subcaption}
\renewcommand\thesubfigure{\alph{subfigure}}
\DeclareCaptionLabelFormat{subcaptionlabel}{\normalfont(\textbf{#2}\normalfont)}
\captionsetup[subfigure]{labelformat=subcaptionlabel}
% for math enviroments

% graphics paths
%\usepackage{graphicx}
\graphicspath{{img/ex1/}{img/ex2/}{img/ex3/}{img/ex4/}{img/ex5/}{img/other/}{img/benchmark/}}

% pretty C++
\newcommand{\CC}{C\nolinebreak\hspace{-.05em}\raisebox{.4ex}{\tiny\bf ++}}

%=================================================================
%% Please use the following mathematics environments: Theorem, Lemma, Corollary, Proposition, Characterization, Property, Problem, Example, ExamplesandDefinitions, Hypothesis, Remark, Definition, Notation, Assumption
%% For proofs, please use the proof environment (the amsthm package is loaded by the MDPI class).

%=================================================================
% Full title of the paper (Capitalized)
\Title{Multiset-Trie Data Structure}

% MDPI internal command: Title for citation in the left column
\TitleCitation{Multiset-Trie Data Structure}

% Author Orchid ID: enter ID or remove command
%\newcommand{\orcidauthorA}{0000-0000-0000-000X} % Add \orcidA{} behind the author's name
\newcommand{\orcidauthorB}{0000-0002-3994-4805} % Add \orcidB{} behind the author's name
\newcommand{\orcidauthorC}{0000-0002-4960-8901} % Add \orcidC{} behind the author's name
\newcommand{\orcidauthorD}{0000-0001-6851-3214} % Add \orcidD{} behind the author's name

% Authors, for the paper (add full first names)
\Author{Mikita Akulich $^{1,\dagger}$, Iztok Savnik $^{1,2,\dagger}$\orcidB{}, Matjaž Krnc $^{1,2,3,\dagger}$\orcidC{} and Riste Škrekovski $^{1,2,3,\dagger,}$*\orcidD{}}

%\longauthorlist{yes}

% MDPI internal command: Authors, for metadata in PDF
\AuthorNames{Mikita Akulich, Iztok Savnik, Matjaž Krnc and Riste Škrekovski}

% MDPI internal command: Authors, for citation in the left column
\AuthorCitation{Akulich, M.; Savnik, I.; Krnc, K.; Škrekovski, R.}
% If this is a Chicago style journal: Lastname, Firstname, Firstname Lastname, and Firstname Lastname.

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Faculty of Mathematics, Natural Sciences and Information Technologies, University of Primorska,  6000 Koper, Slovenia; mikita.akulich@gmail.com (M.A.); iztok.savnik@upr.si (I.S.); matjaz.krnc@upr.si (M.K.)\\
$^{2}$ \quad Faculty of Information Studies,  8000 Novo Mesto,  Slovenia\\
$^{3}$ \quad Department of Mathematics, Faculty of Mathematics and Physics,  University of Ljubljana, \mbox{1000 Ljubljana, Slovenia}
}

% Contact information of the corresponding author
\corres{Correspondence: skrekovski@gmail.com %Tel.: (optional; include country code; if there are multiple corresponding authors, add author initials) +xx-xxxx-xxx-xxxx (F.L.)
%\corres{Correspondence: iztok.savnik@upr.si %Tel.: (optional; include country code; if there are multiple corresponding authors, add author initials) +xx-xxxx-xxx-xxxx (F.L.)
}

% Current address and/or shared authorship
%\firstnote{Current address: Affiliation 3.} 
\firstnote{These authors contributed equally to this work.}
% The commands \thirdnote{} till \eighthnote{} are available for further notes

%\simplesumm{} % Simple summary

%\conference{} % An extended version of a conference paper

% Abstract (Do not insert blank lines, i.e., \\) 
\abstract{This paper proposes a new data structure, \highlighting{\emph{multiset-trie}}, %MDPI: Please confirm if the italics is unnecessary and can be removed. Please check all italic.
 that is designed for storing and efficiently processing a set of multisets. Moreover, multiset-trie can operate on a set of sets without efficiency loss. 
The multiset-trie structure is a search tree with properties similar to those of a trie. It implements all standard search tree operations together with the multiset containment operations, such as sub-multiset and super-multiset. Suppose that we have a set of multisets $S$ and a multiset $X$. The multiset containment operations retrieve multisets from $S$ that are either sub-multisets or super-multisets of $X$.  
We present the mathematical analysis of a multiset-trie that gives the time complexity of the algorithms and the space complexity of the data structure. Further, the empirical analysis of the data structure is implemented in a series of experiments. The experiments illuminate the time complexity space of the multiset containment operations.
%Finally, the comparison of the multiset-trie with the inverted index shows that the multiset-trie outperforms the inverted index by up to 4 orders of magnitude.
%%% more rounded...!
}

% Keywords
\keyword{trie data structure; multiset; containment queries}

% The fields PACS, MSC, and JEL may be left empty or commented out if not applicable
%\PACS{J0101}
%\MSC{}
%\JEL{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Diversity
%\LSID{\url{http://}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Applied Sciences
%\featuredapplication{Authors are encouraged to provide a concise description of the specific application or a potential application of the work. This section is not mandatory.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Data
%\dataset{DOI number or link to the deposited data set if the data set is published separately. If the data set shall be published as a supplement to this paper, this field will be filled by the journal editors. In this case, please submit the data set as a supplement.}
%\datasetlicense{License under which the data set is made available (CC0, CC-BY, CC-BY-SA, CC-BY-NC, etc.)}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Toxins
%\keycontribution{The breakthroughs or highlights of the manuscript. Authors can write one or two sentences to describe the most important part of the paper.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Encyclopedia
%\encyclopediadef{For entry manuscripts only: please provide a brief overview of the entry title instead of an abstract.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Advances in Respiratory Medicine
%\addhighlights{yes}
%\renewcommand{\addhighlights}{%

%\noindent This is an obligatory section in “Advances in Respiratory Medicine”, whose goal is to increase the discoverability and readability of the article via search engines and other scholars. Highlights should not be a copy of the abstract, but a simple text allowing the reader to quickly and simplified find out what the article is about and what can be cited from it. Each of these parts should be devoted up to 2~bullet points.\vspace{3pt}\\
%\textbf{What are the main findings?}
% \begin{itemize}[labelsep=2.5mm,topsep=-3pt]
% \item First bullet.
% \item Second bullet.
% \end{itemize}\vspace{3pt}
%\textbf{What is the implication of the main finding?}
% \begin{itemize}[labelsep=2.5mm,topsep=-3pt]
% \item First bullet.
% \item Second bullet.
% \end{itemize}
%}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction} \label{c:introduction}

% multisets def; not ordered
A multiset is a collection of elements that can have more than one instance. As in the case of ordinary sets, the ordering of the elements in multisets is not relevant. For example, multisets $\{1,1,2\}$ and $\{1,2,1\}$ represent the same multiset.

% motivation for the design
Multisets appear in a wide variety of domains and applications \cite{savnik2021plos}. The index structures for storing sets of multisets were studied in the area of object-relational database systems to store, compress and query multiset-valued attributes efficiently \cite{bouros2016set,gripon2012compressing,ross2004symmetric,steinruecken2015compressing}. The need to efficiently manage multisets also appears in information retrieval \cite{zobel1992efficient,zobel2006inverted,manning2008introduction}, where texts are represented as multisets. In data mining, sets are often used to represent and efficiently search hypotheses in the knowledge discovery process \cite{mannila1997,flach1999aicom}. In the area of expert systems, multisets are used for the representation and querying of the preconditions of rules \cite{forgy1982}. Finally, in recent internet applications, the efficient representations and searches of multisets (such as user requests and object features) have become essential \cite{bayardo2007simlar,xiao2011tods,wang2017vldb} for data cleaning, information integration, community mining, and entity resolution.

%\medskip\noindent\textbf{Problem definition.}\;
In this paper, we address the problems of storing, indexing, and querying the sets of multisets. In particular, we deal with the design of an index data structure that provides an efficient implementation of the multiset containment queries. Let $S$ be an index storing a set of multisets. For a given input multiset $m$, a \emph{containment query} searches for either sub-multisets or super-multisets of $m$ in $S$.

Existent indexes for storing a set of multisets are rooted in search trees \cite{corman2001}. The elements of a search tree can be accessed through keys. This approach is efficient for checking the membership of individual multisets $m$ in $S$. However, it is not as efficient for  containment queries. The search based on the containment relation requires access to the collections $C\subseteq\/S$ of multisets that are related for a multiset $m$ either by a sub-multiset or a super-multiset relationship.

The existing solutions to the implementation of the containment queries use the inverted indexes \cite{zobel1992efficient,zobel1998inverted,zobel2006csur,broder2006indexing,terrovitis2006cikm,terrovitis2011icdt}, the signature trees \cite{deppisch1986sigir,tousidou2002sigstruc,yangjun2005stree,zobel1998inverted}, and the B+ trees~\cite{Helmer2003,terrovitis2011icdt}. These solutions provide a key--value look-up, where a key is an element of a multiset and a value is a corresponding multiset. The containment operations require multiple key--value index accesses and also the additional processing of partial results such as the intersection or union of~multisets.

%\medskip\noindent\textbf{Proposed data structure.}\;
To improve the efficiency of containment operations, we propose a data structure, \emph{multiset-trie}, that is designed for storing and processing a finite bounded set of multisets. A multiset-trie generalizes the \emph{set-trie} data structure proposed by Savnik~\cite{savnik2013index,savnik2021plos}, which was designed for the storage and processing of a finite bounded set of sets. 
The set-trie is an extension of a trie data structure to provide, besides the fast search and retrieval of sets, also the efficient implementation of the set containment operation. The set-trie is also a form of the binomial tree \cite{corman2001}.

Multiset-trie provides a space-efficient representation of a set of multisets and efficient multiset containment operations. As in the case of set-trie, the ordering of the elements in multiset-trie is not relevant for the representation of multisets. As a consequence, the efficiency of the multiset containment operations is obtained by selecting the \emph{specific} ordering of multiset elements. This ordering can be exploited for the efficient search in a multiset-trie.

% intro into mstrie
The multiset-trie is an $n$-ary tree data structure. Each multiset in a multiset-trie is represented by a path from the root to a leaf node. Multiset elements are symbols from an alphabet. Each symbol from the alphabet is represented by a node at a certain level of the multiset-trie. The node stores the multiplicity of an element in a multiset.

A multiset-trie is also a kind of search tree. Similar to a trie, it uses common prefixes for a shared data representation. Unlike the compact prefix tree \cite{Sedgewick:2011:ALG:2011916}, the multiset-trie does not support path compression. However, the absence of path compression makes the multiset-trie a perfectly height-balanced tree. Moreover, when multiset-trie is full, it forms a complete $n$-ary tree.

\medskip\noindent\textbf{Contributions.}\;
The main contributions of this paper are as follows. First, a  multiset-trie is a novel data structure for storing and querying a set of multisets that provides efficient multiset containment operations.

Second, a mathematical model is developed to analyze the complexity of multiset containment operations. In particular, we estimate the size of a subtree traversed by a containment query and give an insight into the time complexity of containment queries.

In addition to the mathematical analysis, the size of a subtree visited by a multiset containment query is also the central focus of the empirical analysis. We carefully designed the experiments to unravel the main features of the search space. We observe how the number of visited nodes depends on various parameters, such as the density of the tree and the ordering of multiset elements.

Finally, the mathematical as well as empirical analyses show the influence of ordering the elements of multisets on the efficiency of storing and processing multisets. We show that the ordering, which is based on the frequencies of the multiset elements, can speed up the multiset containment queries by orders of magnitude.

\medskip\noindent\textbf{Paper organization.}\; 
The paper is organized as follows. In the following Section~\ref{c:description}, we present the multiset-trie data structure in detail.
Section~\ref{c:operations} presents the operations of the multiset-trie. These include the basic operations of search trees and multiset containment operations. The algorithms are presented in detail using pseudocode.

The description of multiset-trie operations is followed by the mathematical analysis of their complexity in Section~\ref{c:analysis}. The main assumption is that multisets are constructed uniformly at random with bounded cardinality. By using probabilistic tools, we describe the time complexity of the algorithms and the space complexity of the structure.

In Section~\ref{c:experiments}, we present an empirical study of the multiset-trie. Synthetic and real-world data sets are used in experiments that are designed to study the performance of multiset containment operations. Further, the multiset-trie is empirically compared to the inverted index in a separate experiment. The experiments highlight the methods for optimizing a multiset-trie.

The related work is reviewed in Section~\ref{c:relwork}. This section describes a set-trie data structure, the inverted indexes, the signature indexes, and the multisets in relational database systems. Finally, the concluding remarks and the future work are presented in Section~\ref{c:conclusions}.

%\input{tex/s_intro}
% section 2
%\input{tex/s_data_structure}
\section{Multiset-Trie Data Structure} \label{c:description}
%
Let $\Sigma$ be a set of distinct symbols that define an alphabet, and let 
$\sigma$ be the cardinality of $\Sigma.$ The \emph{multiset-trie} data structure 
stores multisets that are composed of symbols from the alphabet $\Sigma.$ It 
provides the basic tree data structure operations, such as insert, delete, and 
search, together with multiset containment operations for searching sub-multisets
and super-multisets, which will be discussed in the next section in greater detail.

A multiset ignores the ordering of its elements by definition, which allows us to 
define a bijective mapping $\phi:\Sigma \rightarrow I,$ where $I$ is the set of 
integers $\{ 1,2,3, \ldots, \sigma \}.$ In this way, we obtain the indexing of 
elements from the alphabet $\Sigma,$ so we can work directly with integers 
rather than with specific symbols from $\Sigma.$

The multiset-trie is an $n$-ary tree-based data structure with the properties of 
the trie. A node in multiset-trie always has degree $n,$ i.e., $n$ children. Some of 
the children may be \emph{Null} (non-existing), but the number of \emph{Null} 
children can be at most $n-1.$ All the children of a node, including the \emph{Null} 
children, are labeled from left to right with labels $c_j,$ where $j\in \{ 0, 1, \ldots, n-1 \}.$ 
Every pair of child nodes $u$ and $v$ that share the same parent node have different labels.

%The multiset-trie is an $n$-ary tree based structure. A node can have a degree 
%from $1$ to $n,$ depending on the number of existing children. The edges from 
%parent node to children have labels $c_j,$ where $j\in \{ 0, 1, \ldots, n-1 \}.$ 
%The edges are labeled from left to right with the condition that there are no 
%two edges with the same label coming from a parent node.

Nodes that have equal height in a multiset-trie form a level. The height of a 
multiset-trie is always $\sigma+1$ if at least one multiset is in the structure. 
The height of the root node (the first level) is defined to be 1.
%
Levels in multiset-trie are enumerated by their height, i.e., a level $L_i$ has 
height $i.$ The connection between level height in a multiset-trie and symbols 
from alphabet $\Sigma$ is defined as follows. A level $L_i,$ where 
$i\in\{ 1,2,\ldots, \sigma \}$ represents a symbol $s\in\Sigma,$ such that 
$\phi^{-1}(i) = s.$ The last level $L_{\sigma+1}$ does not represent any symbol 
and is named the \emph{leaf level} ($LL$ for short).

Since every level, except $LL$, represents a symbol from $\Sigma,$ we can define 
a transition between nodes that are located at different levels in a multiset-trie. 
%
Consider two nodes $u,v$ in a multiset-trie at levels $L_i, L_{i+1}$, respectively, 
where $i\in\{1,2,\ldots,\sigma\}.$ Let a node $u$ be a parent node of a node $v$ 
and consequently a node $v$ be a child node of a node $u.$ Suppose that a child 
node $v$ is not \emph{Null} and has a label $c_j,$ where $j\in\{ 0,1,\ldots, n-1 \}.$ 
%
Then, the \emph{path} $u\rightarrow v$ represents a symbol $s\in\Sigma$ with 
multiplicity $j,$ such that $\phi^{-1}(i) = s.$ 
%
Such a transition $u\rightarrow v$ is called a \emph{path of length} $1$ and is 
allowed if and only if a node $v$ is not \emph{Null} and $u$ is a parent node of 
a node $v.$ If a node $v$ has label $c_0,$ then the path $u\rightarrow v$ 
represents a symbol with the multiplicity $0$ respectively, i.e., an empty symbol.

We define a \emph{complete path} to be the path of length $\sigma$ in a 
multiset-trie with the endpoints at the root node (the first level) and $LL$. Thus, 
a multiset $m$ is inserted into a multiset-trie if and only if there exists a 
complete path in a multiset-trie that corresponds to $m.$
%
Note that every complete path in a multiset-trie is unique. Therefore, the multisets 
that share a common prefix in a multiset-trie can have a common path of length at 
most $\sigma-1.$ The complete path that passes through nodes labeled by $c_0$ 
on all levels represents an empty multiset or an empty set.
%
Thus, any multiset $m$ that is composed of symbols from $\Sigma$ with maximum 
multiplicity not greater than $n-1$ can be represented by a complete path in a multiset-trie.

An example of a multiset-trie data structure with $\sigma = 2$ and 
$\Sigma = I = \{ 1,2 \}$ (i.e., the mapping $\phi$ is an identity mapping) is shown in Figure~\ref{fig:sketch}. 
In the figure, which stores elements of $\{\emptyset, \{ 1,1,2 \},$ $\{ 1,2,2 \},$ $\{ 2 \},$ $\{ 1,2 \},$ $\{ 2,2 \}\}$, the degree of a node is set to be $n=3,$ so the maximal multiplicity of an element in 
a multiset is $n-1=2.$

\begin{figure}[H]
\begin{tikzpicture}[>=stealth',
level 1/.style={sibling distance = 2.5cm},
level 2/.style={sibling distance = 1cm},
level distance = 1.5cm]
\node (R) [node1] {\tiny Root}
	child { node [node1] {} 
		child { node [node1] {} edge from parent node [near end, left] {$c_0$}}
		child { node [node1] {} edge from parent node [near end, left] {$c_1$}}
		child { node [node1] {} edge from parent node [near end, right] {$c_2$}}
		edge from parent node [near end, left] {$c_0$}	
	}
	child { node [node1] {}
		child { node [node1] {} edge from parent node [near end, left] {$c_1$}}
		child { node [node1] {} edge from parent node [near end, right] {$c_2$}}
		edge from parent node [near end, left] {$c_1$}
	}
	child { node [node1] {} 
		child { node [node1] {} edge from parent node [near end, left] {$c_1$}}
		edge from parent node [near end, right] {$c_2$} 
};	
%----------------- Level labels --------------	
	\begin{scope}[every node/.style={right}]
		\path (R 				-| R-3-1) ++(5mm,0) node {$L_1$};
		\path (R-3 			-| R-3-1) ++(5mm,0) node {$L_2$};
		\path (R-3-1		-| R-3-1) ++(5mm,0) node {$LL$};
	\end{scope}			
\end{tikzpicture}
\caption{Example of multiset-trie structure containing multisets $\emptyset, \{ 1,1,2 \},$ $\{ 1,2,2 \},$ $\{ 2 \},$ $\{ 1,2 \},$ $\{ 2,2 \}$. The \emph{Null} children are omitted.}
\label{fig:sketch}
\end{figure}

Let a pair $(L_i, c_j)$ represent a node with label $c_j$ at a level $L_i.$ 
The pair $(L_1,c_j)$ is equivalent to $(L_1,root),$ since the first level has 
the root node only. According to Figure~\ref{fig:sketch}, we can extract the 
inserted multisets as follows:
\begin{eqnarray*}
(L_1,root) \rightarrow (L_2,c_0) \rightarrow (LL,c_0) & = & \{ 1^0, 2^0 \} = \emptyset \\
(L_1,root) \rightarrow (L_2,c_0) \rightarrow (LL,c_1) & = & \{ 1^0, 2^1 \} = \{ 2 \} \\
(L_1,root) \rightarrow (L_2,c_0) \rightarrow (LL,c_2) & = & \{ 1^0, 2^2 \} = \{ 2,2 \} \\
(L_1,root) \rightarrow (L_2,c_1) \rightarrow (LL,c_1) & = & \{ 1^1, 2^1 \} = \{ 1,2 \} \\
(L_1,root) \rightarrow (L_2,c_1) \rightarrow (LL,c_2) & = & \{ 1^1, 2^2 \} = \{ 1,2,2 \} \\
(L_1,root) \rightarrow (L_2,c_2) \rightarrow (LL,c_1) & = & \{ 1^2, 2^1 \} = \{ 1,1,2 \} 
\end{eqnarray*}
where $e^k$ represents an element $e$ with multiplicity $k.$
% section 3
%\input{tex/s_operations}
\section{Multiset-Trie Operations} \label{c:operations}

%
Let $\mathcal{M}$ be a multiset-trie and let $M$ be a set of multisets that are 
inserted into the multiset-trie $\mathcal{M}.$ We define a type \emph{Multiset} in 
order to use it as a representation of a multiset. The type \emph{Multiset} is 
an array $m$ of constant length $\sigma,$ where the $i$-th cell represents the element 
$\phi^{-1}(i)$ from $\Sigma$ with multiplicity $m[i].$ From now on, we 
agree that the first cell of an array has index 1. Let us give an example of a 
\emph{Multiset} instance with $\sigma = 2:$
%
\begin{center}
\begin{tabular}{ccc}
Multiset & & Instance of type Multiset \\
$\{ 1,1,2 \}$ & $\cong $ & 
\begin{tabular}{|c|c|}
\hline 
2 & 1 \\
\hline 
\multicolumn{1}{c}{\tiny 1} & \multicolumn{1}{c}{\tiny 2} \\
\end{tabular}
\end{tabular}
\end{center}
%
\hl{The} %MDPI: Whether the indentation of the paragraph after the formula is necessary, please confirm.
 operations supported by the multiset-trie data structure are as
follows. 
%
\begin{enumerate}
\item \textsc{insert}($\mathcal{M}$, $m$): inserts a multiset $m$ into 
$\mathcal{M}$ if $m\not\in M;$
%
\item \textsc{search}($\mathcal{M}$, $m$): returns true if a multiset $m\in M$ 
for a given $\mathcal{M},$ and returns false otherwise;
%
\item \textsc{delete}($\mathcal{M}$, $m$): returns true if a multiset $m$ was 
successfully deleted from $\mathcal{M},$ and returns false otherwise (in case 
$m\not\in M$);
%
\item \textsc{submsetExistence}($\mathcal{M}$, $m$, $dev$): returns true if 
there exists a $x\in M$ for a given $\mathcal{M}$ such that $x\subseteq m$ 
and $| x[i] - m[i] |\leq dev$ for $1\leq i\leq \sigma$, and returns false otherwise; 
%
\item \textsc{supermsetExistence}($\mathcal{M}$, $m$, $dev$): returns true if 
there exists a $x\in M$ for a given $\mathcal{M}$ such that $x\supseteq m$ 
and $| x[i] - m[i] |\leq dev$ for $1\leq i\leq \sigma$, and returns false otherwise; 
%
\item \textsc{getAllSubmsets}($\mathcal{M}$, $m$, $dev$): returns the set of multisets 
$\{ x \in M : x\subseteq m \wedge |x[i]-m[i]|\leq dev \}$ for a given 
$\mathcal{M},$ where $1\leq i\leq \sigma;$
%
\item \textsc{getAllSupermsets}($\mathcal{M}$, $m$, $dev$): returns the set of multisets 
$\{ x\in M : x\supseteq m \wedge |x[i]-m[i]|\leq dev \}$ for a given $\mathcal{M},$
where $1\leq i\leq \sigma.$
%
\end{enumerate}

The parameter $dev$ is used to specify the maximal deviation in the multiplicity
of multiset containment operations. It is utilized to limit the search in multiset
containment queries to the sub-multisets and super-multisets that are
the closest to the input multiset $m$. In addition, we use $dev$ for
the implementation of a multiset similarity search that, given $m$, retrieves
from $\mathcal{M}$ all sub-multisets or super-multisets that are similar to $m$
with respect to the deviation $dev$.

In the following subsections, we will present each operation of the multiset-trie 
data structure separately.

Firstly, we would like to describe some notations that will be used. The 
multiset-trie data structure is a recursive data structure. Hence, any subtree 
of a multiset-trie $\mathcal{M}$ is again a multiset-trie. This fact allows 
us to use the root node of a multiset-trie as its representative. Thus, the notation 
$\mathcal{M}$ will be used instead of $\mathcal{M}.root$ to refer to the root 
node of $\mathcal{M}.$ Non-existing or \emph{Null} nodes in multiset-trie will 
be marked as \emph{Null} and existing nodes at the level $LL$ will be marked 
as \emph{accepting} nodes. The array slicing operation will be used as follows. 
For a given array $a,$ $a[i:]$ represents the array obtained from $a$ by taking 
only the cells from index $i$ until the last cell.

\subsection{Insert} \label{s:insert}
The procedure \textsc{insert}($\mathcal{M}$, $m$) inserts a new instance $m$ of 
type Multiset into multiset-trie $\mathcal{M}$. If the complete path already 
exists, then the procedure leaves the structure unchanged. Otherwise, it extends 
partially existing paths or creates a new complete path. The procedure does not return 
any result. The pseudocode for procedure \textsc{insert} is presented in 
Algorithm~\ref{alg:insert}.

\begin{algorithm}[H]
\caption{Procedure \textsc{insert}.}
\label{alg:insert}
\begin{algorithmic}[1]
\Procedure{\textsc{insert}}{$\mathcal{M}$, $m$}
\State $currentNode \gets \mathcal{M}$
\For{$i=1$ to $\sigma$}
\If {child $c_{m[i]}$ of $currentNode$ is \emph{Null}}
\State create new child $c_{m[i]}$ of $currentNode$
\EndIf
\State $currentNode \gets c_{m[i]}$
\EndFor
\State mark $currentNode$ as \emph{accepting}
\EndProcedure
\end{algorithmic}
\end{algorithm}

\subsection{Search}\label{s:search}
The function \textsc{search}($\mathcal{M}$, $m$) checks if the complete path corresponding to 
a given multiset $m$ exists in the structure $\mathcal{M}.$ The function returns 
true if the multiset $m$ exists in $\mathcal{M}$, and returns false otherwise. The 
function \textsc{search} is presented in Algorithm~\ref{alg:search}.

\begin{algorithm}[H]
\caption{Function \textsc{search}.}
\label{alg:search}
\begin{algorithmic}[1]
\Function{\textsc{search}}{$\mathcal{M}$, $m$}
\State $currentNode \gets \mathcal{M}$
\For{$i=1$ to $\sigma$}
\If {child $c_{m[i]}$ of $currentNode$ is \emph{Null}}
\State \Return False
\EndIf
\State $currentNode \gets c_{m[i]}$
\EndFor
\State \Return True
\EndFunction
\end{algorithmic}
\end{algorithm}

\subsection{Delete} \label{s:delete}
Function \textsc{delete}($\mathcal{M},$ $m$) searches for the complete path 
that corresponds to $m$ in order to remove it. If the path cannot be found, the 
function immediately returns false. During the search, the function keeps track of the 
number of children for every node. It marks the nodes that have more than one child 
as \emph{parent nodes} and remembers the label of the child, which is a potential node 
where the subtree will be cut to remove the multiset. The parent node is needed to 
perform a removal because the multiset-trie is an explicit data structure. When the search 
is completed, the function removes the subtree of the last found parent node and 
returns true. In such a way, after deletion, all the prefixes for other multisets are 
preserved in $\mathcal{M}$ and $m$ is removed. The function \textsc{delete} is 
presented in Algorithm~\ref{alg:delete}.

\begin{algorithm}[H]
\caption{Function \textsc{delete}.}
\label{alg:delete}
\begin{algorithmic}[1]
\Function{\textsc{delete}}{$\mathcal{M},$ $m$}
\State $currentNode \gets \mathcal{M}$
\State $parent \gets currentNode$ 
\State $position \gets 1$
\For {$i=1$ to $\sigma$}
\If {child $c_{m[i]}$ of $currentNode$ is \emph{Null}}
\State \Return False
\EndIf
\State $numChildren \gets 0$
\For {$j=0$ to $n-1$}
\If {child $c_j$ of $currentNode$ is not \emph{Null}}
\State $numChildren\gets numChildren+1$
\EndIf
\EndFor
\If {$numChildren$ is not $1$}
\State $parent\gets currentNode$
\State $position \gets i$
\EndIf
\State $currentNode \gets c_{m[i]}$
\EndFor
\State child $c_{m[position]}$ of $parent\gets$ \emph{Null}
\State \Return True
\EndFunction
\end{algorithmic}
\end{algorithm}

\subsection{Sub-Multiset and Super-Multiset Existence}
\label{s:subexists} \label{s:superexists}
The functions \textsc{submsetExistence} and \textsc{supermsetExistence} are
symmetrical in the following sense. Let a multiset $m$ represent the borderline
in $\mathcal{M}$ defined by a path from the root to a leaf following
the elements from $m$. The operation \textsc{submsetExistence} searches
the left part of $\mathcal{M}$ and the operation \textsc{supermsetExistence}
the right part of $\mathcal{M}$.

The function \textsc{submsetExistence}($\mathcal{M},m,dev$) checks if there exists 
a multiset $x$ in $\mathcal{M},$ that satisfies the condition $x\subseteq m$ and 
$| x[i] - m[i] | \leq dev,$ where $1\leq i \leq \sigma.$ 
The function starts with searching for an exact match $x=m$ in $\mathcal{M},$ 
since $m\subseteq m$ by definition of sub-multiset inclusion. If an exact match is 
not found in $\mathcal{M},$ the function uses multiset-trie to find the closest 
(the largest) sub-multiset of $m$ in $\mathcal{M}$ by decreasing the multiplicity of 
elements in $m.$ The parameter $dev$ is used to limit a maximal deviation of 
multiplicity for a particular element in $x$ with respect to $m.$ 
At every level, the function tries to proceed with the largest possible multiplicity of 
an element that is provided by $m.$ However, when the function reaches some level 
where it meets a \emph{Null} node and cannot go further using the path provided by 
$m,$ it decreases the multiplicity of an element 
that corresponds to a current level with respect to the specified maximal deviation. 
Thus, the function can decrease the multiplicity of an element or eventually skip it in 
order to find the closest $x\subseteq m.$ The function \textsc{submsetExistence} 
is presented in Algorithm~\ref{alg:subexists}.

\begin{algorithm}[H]
\caption{Function \textsc{submsetExistence}.}
\label{alg:subexists}
\begin{algorithmic}[1]
\Function{\textsc{submsetExistence}}{$\mathcal{M}, m, dev$}
\State $currentNode \gets \mathcal{M}$
\If {$currentNode$ is \emph{accepting}}
\State \Return True
\EndIf
\For {$i=m[1]$ down to max($0, m[1]-dev$)}
\If {child $c_i$ of $currentNode$ is not \emph{Null}}
\If {\textsc{submsetExistence}($c_i,m[2:], dev$)}
\State \Return True
\EndIf
\EndIf
\EndFor
\State \Return False
\EndFunction
\end{algorithmic}
\end{algorithm}

%\subsection{Super-multiset existence} \label{s:superexists}
The function \textsc{supermsetExistence}($\mathcal{M},m,dev$) checks if there 
exists super-multiset $x$ of a given multiset $m$ in $\mathcal{M},$ such that 
condition $| x[i] - m[i] |\leq dev$ is satisfied, where \mbox{$1\leq i \leq \sigma.$}
Symmetrically to the function \textsc{submsetExistence}, the function
\textsc{supermsetExistence} searches first for an exact match $x=m$ in
$\mathcal{M}.$ If such $x$ does not exist in $\mathcal{M},$ then
the function searches on the right side of the borderline representing
an exact match of $m$ in $\mathcal{M}.$ Since we would like to find
the closest (the smallest) super-multiset, we increase the multiplicity of
elements in $m$ at every level of $\mathcal(M)$ starting with
the multiplicities of $m.$ The function \textsc{supermsetExistence}
is presented in Algorithm~\ref{alg:superxists}.

\begin{algorithm}[H]
\caption{Function \textsc{supermsetExistence}.}
\label{alg:superxists}
\begin{algorithmic}[1]
\Function{\textsc{supermsetExistence}}{$\mathcal{M},m,dev$}
\State $currentNode \gets \mathcal{M}$
\If {$currentNode$ is \emph{accepting} }
\State \Return True
\EndIf
\For {$i=m[1]$ to min($n-1, m[1] + dev$)}
\If {child $c_i$ of $currentNode$ is not \emph{Null}}
\If {\textsc{supermsetExistence}($c_i,$ $m[2:], dev$)}
\State \Return True
\EndIf
\EndIf
\EndFor
\State \Return False
\EndFunction
\end{algorithmic}
\end{algorithm}

\subsection{Get All Sub-Multisets and Get All Super-Multisets} \label{s:getall}
The algorithms for functions \textsc{getAllSubmsets} and 
\textsc{getAllSupermsets} are based entirely on algorithms for 
\textsc{submsetExistence} and \textsc{supermsetExistence} functions that do not 
terminate on the first existing sub/super-multiset, but store the results and 
continue the procedure until all existing sub/super-multisets in $\mathcal{M}$ are 
found and stored. The functions \textsc{getAllSubmsets} and \textsc{getAllSupermsets} 
are presented in Algorithm~\ref{alg:getallsub} and Algorithm~\ref{alg:getallsup}, respectively.

In order to record a multiset during multiset-trie traversal, we use the variable $x$ in the algorithms.
It is an empty array of size $\sigma$ where we store multiplicities of elements at 
each level as we traverse the tree. The variable $result$ is used as a container for storing 
sub-multisets of $m$ found during traversal. Both variables $x$ and $result$ are presented 
as global; however, they could be passed to the recursive function as parameters.

\begin{algorithm}[H]
\caption{Function \textsc{getAllSubmsets}.}
\label{alg:getallsub}
\begin{algorithmic}[1]
\State $result \gets$ empty container
\State $x \gets$ empty array of size $\sigma$
\Function{\textsc{getAllSubmsets}}{$\mathcal{M}, m, dev$}
\State $currentNode \gets \mathcal{M}$
\If {$currentNode$ is \emph{accepting}}
\State add copy of $x$ to $result$ 
\EndIf
\For {$i=m[1]$ down to max($0, m[1]-dev$)}
\If {child $c_i$ of $currentNode$ is not \emph{Null}}
\State $x[1]\gets i$
\State \textsc{getAllSubmsets}($c_i,m[2:], dev$)
\EndIf
\EndFor
\EndFunction
\end{algorithmic}
\end{algorithm}

\begin{algorithm}[H]
\caption{Function \textsc{getAllSupermsets}.}
\label{alg:getallsup}
\begin{algorithmic}[1]
\State $result \gets$ empty container
\State $x \gets$ empty array of size $\sigma$
\Function{\textsc{getAllSupermsets}}{$\mathcal{M},m,dev$}
\State $currentNode \gets \mathcal{M}$
\If {$currentNode$ is \emph{accepting} }
\State add copy of $x$ to $result$ 
\EndIf
\For {$i=m[1]$ to min($n-1, m[1] + dev$)}
\If {child $c_i$ of $currentNode$ is not \emph{Null}}
\State $x[1]\gets i$
\textsc{getAllSupermsets}($c_i,$ $m[2:], dev$)
\EndIf
\EndFor
\EndFunction
\end{algorithmic}
\end{algorithm}
% section 4
%\input{tex/s_math-Theorems}
\section{Mathematical Analysis of the Structure} \label{c:analysis}
%
In this section, we present theoretical results of the time and space complexity of
the multiset-trie data structure. In the following Section~\ref{s:timecomplexity}, 
we discuss the running time complexity of the presented algorithms. First, in 
Section~\ref{ss:mathmodel}, we present the mathematical model that we use 
to describe the distribution of multisets in the multiset-trie and input data. 
Using a probabilistic approach and tools from a Galton--Watson process, we 
measure the expected cardinality of the multiset-trie in Theorem~\ref{thm:exp_nodes}. 
Further, we derive the expected cardinality of the searched subtree of the 
multiset-trie parametrized by an input multiset in Corollary~\ref{cor:exp_nodes_param}.

In Section~\ref{ss:getall}, we discuss the running time complexity of the functions 
\textsc{getAllSubmsets} and \textsc{getAllSupermsets}. We observe that the 
complexity of the functions is exponential. Moreover, the worst-case running time 
complexity is the same for both functions, and its upper bound is the cardinality of 
the multiset-trie.

The remaining ``existence'' functions are discussed in Section~\ref{ss:exists}. 
We observe that, beyond the scope of our mathematical model, unlike in functions 
\textsc{getAllSubmsets} and \textsc{getAllSupermsets}, the mapping $\phi$ has an 
impact on the performance of the functions \textsc{submsetExistence} and 
\textsc{supermsetExistence}. In particular, the frequency analysis of the symbols 
from $\Sigma$ in input data determines such a $\phi$ that gives a boost in performance.

We find that the performance of the functions \textsc{submsetExistence} and 
\textsc{supermsetExistence} in the worst-case scenario is also proportional to the size of the constructed trie. We give a rather technical upper 
bound for the worst-case running time complexity, which appears to be the same 
for both functions. 
However, it must be stressed that for the positive outcome, this 
worst-case behavior holds only on specific cases, such as the presence of the empty set 
in the multiset-trie.

Finalizing the mathematical analysis, we present the study of the space complexity 
of the multiset-trie in Section~\ref{s:spacecomplexity}. We show that the space 
used for the storage is asymptotically equal to the size of the input data.

\subsection{Time Complexity of the Algorithms}\label{s:timecomplexity}
The performance of the functions will be measured by the number of 
visited nodes in a multiset-trie during the execution of a particular query by the 
functions \textsc{search}, \textsc{delete}, \textsc{submsetExistence}, 
\textsc{supermsetExistence}, \textsc{getAllSubmsets}, \textsc{getAllSupermsets} 
and the procedure \textsc{insert}.
%

By the design of the multiset-trie, it is easy to see that the functions \textsc{search},
\textsc{delete} and the procedure \textsc{insert} have complexity of $O(\sigma).$
Because $\sigma$ is defined when the structure is initialized and does not depend
on the user input afterwards, the asymptotic complexity of the functions \textsc{search}, 
\textsc{delete} and the procedure \textsc{insert} is $O(1).$ Nonetheless, in the general 
case, the complexity is $O(\sigma).$

In what follows, we focus on the analysis of the more involved functions:
\textsc{submsetExistence}, \textsc{supermsetExistence}, \textsc{getAllSubmsets},
and \textsc{getAllSupermsets}.

\subsubsection{Mathematical Model} \label{ss:mathmodel}
We start with the basics of our mathematical model. Let $\Sigma$ be an alphabet of
cardinality $\sigma,$ such that $\Sigma = \{ 1,2, \ldots, \sigma \}.$ Define $N$
to be the set of all possible multisets that can be inserted in a
multiset-trie. Let $n$ be the maximal degree of a node in a multiset-trie.
Then, the maximal multiplicity of an element in a multiset is equal to $n-1.$
Thus, the number of multisets in a complete multiset-trie is $ |N| = n^{\sigma}.$
%
Let $M$ be a collection of multisets inserted into multiset-trie $\mathcal{M}.$
All the multisets in $M$ are constructed from the alphabet $\Sigma$ according
to the parameters $\sigma$ and $n$. Hence, any multiset $m\in M,$ 
has at most $\sigma$ distinct elements that are members of $\Sigma$, and 
every distinct element in $m$ has multiplicity strictly less than $n.$
%
Because a multiset does not distinguish different orderings, it is assumed, for
simplicity, that all elements are ordered in ascending order. A multiset $m$ is
represented as $\{1^{k_1},2^{k_2},\ldots, \sigma^{k_\sigma}\},$ where $e^{k_e}$
represents an element $e\in\Sigma$ with multiplicity $k_e.$
%

Denote the nodes of multiset-trie on all levels but on $\sigma + 1$ as \emph{internal}
and nodes on the leaf level as \emph{leaf} nodes.
%
Observe that every internal non-root node has a degree of at least~1. Indeed, the
insertion of a multiset requires the construction of a path of length~$\sigma + 1,$
meaning that if an internal node exists in a multiset-trie, it must have a degree of
at least~1. It also follows that the height of a multiset-trie is always~$\sigma +1$
as soon as at least one multiset is inserted into the data structure.

Our model assumes that all the inserted multisets are chosen with the same probability,
meaning that for some $p\in (0,1)$, the following holds:
\[
P(m\in M) = p, \quad \forall~m\in N.
\]
%
\hl{Let} %MDPI: Whether the indentation of the paragraph after the formula is necessary, please confirm. Please check all.
 $\xi_1, \xi_2, \ldots, \xi_{\sigma+1}$ be random variables such that $\xi_i$
represents the number of nodes in a multiset-trie on the $i$-th level. For every node $j$ 
on $i$-th level, we assign a random variable $\xi_{ij}$ to be the number of its children, 
such that $j\in[1,\xi_i].$ Then, for every $i\in[1,\sigma]$, the following~holds:%
\begin{equation}\label{eq:sum_recursive}
\xi_{i+1} = \sum_{j=1}^{\xi_i} \xi_{ij},
\end{equation}
%
where $\xi_1 = 1.$
%
It is easy to see that the variable $\xi_{i+1}$ can have values in the interval
$[\xi_i,n^{i}]$ and the value of the variable $\xi_{ij}$ is within the interval $[1,n].$
Without conditioning on the existence of any node in multiset-trie, 
it is easy to describe the probability of the existence of any individual node.

\begin{Lemma}\label{l:prob-node-existence}
Any potential node on a fixed level $i,$ where $i\in \{ 1,2,\ldots, \sigma +1 \}$ exists, %independently of others,
with probability
\begin{equation}
p_i=1-(1-p)^{n^{\sigma + 1 -i}}.
\end{equation}
\end{Lemma}
\begin{proof}
Let $v$ be an arbitrary node in a multiset-trie on an arbitrary level $i.$ Consider
the subtree with the root $v$ and call it the $v$-subtree. Since the height of the
multiset-trie is $\sigma + 1$, we can calculate the height of the $v$-subtree.
Taking into account that the root node has height 1, the height of the $v$-subtree is
%
\[
h_v = \sigma + 1 - i.
\]
%
A node in a multiset-trie exists if at least one node exists on the leaf level of its
subtree, i.e., a node on the level $\sigma + 1$ that belongs to $v$-subtree. The possible
number of nodes on the leaf level of $v$-subtree can be easily calculated knowing its height.
It is equal to
%
\[
n^{\sigma + 1 - i}.
\]
%
A node at level $\sigma+1$ exists with probability $p,$ where $p = P(m\in M).$
Thus, the probability that there are no nodes on the leaf level in $v$-subtree is
%
\[
(1-p)^{n^{\sigma +1 - i}}.
\]
%
The claim follows by taking the complement probability of the above result. 
\end{proof}

Recall that for any given discrete random variable $X$ taking values over positive non-negative integral values, one can define 
a so-called \emph{probability generating function} (PGF, for short) $G_X$ to be a formal power series defined as 
\begin{align*}
G_X(z)=\sum_{i\ge 0} Pr(X=i)z^i.
\end{align*}
While the PGFs are usually not meant to be evaluated for concrete values of $z$, certain values have special interpretation when used in PGF, or in a derivation(s) of PGF. 
For instance, $Pr(X=0)=G_X(0)$, and also $\mathbb{E}(X)=G'_X(1)$.
For many values of $z$, the function $G_X$ may not converge to any finite value. For this reason, it is a common notation to write, in particular,
\[
G_X(1^-)=\lim_{z\nearrow 1}G_X(z).
\]

In what follows, we will denote a Bernoulli-distributed random variable with parameter $p$ as $Bernoulli(p)$.
Furthermore, we denote
a zero-truncated binomially
distributed random variable on parameters $n$ and $p_{i+1}$ by $\mathcal{B}_0(n,p_{i+1})$. Finally, $\overset{d}{=}$ indicates so-called equality in distribution.

\begin{Example}
The probability generating function of a binomial random variable $\mathcal B(n,p)$, the number of successes in $n$ trials, with probability $p$ of success in each trial, is

\begin{align*}
 G_{\mathcal B(n,p)}(z)=G(z)
 &=\sum_{k = 0}^n \binom n k p^k  (1 - p)^{n - k} z^k\\
 &= \sum_{k = 0}^n \binom n k  (p z)^k  (1 - p)^{n - k}\\
 &=\left((1-p)+pz\right)^{n}.
\end{align*}

It is thus easy to compute expectation
\[
\mathbb{E}(\mathcal{B}(n,p))=G'(1)=np.
\]

\end{Example}

We now focus our attention on the distribution of $\xi_{ij}$.

\begin{Lemma}\label{l:prob-children}
Suppose that a node $v$ exists at level $1\leq i\leq\sigma$.
Then, the number of its children $\xi_{iv}$ is modeled by a zero-truncated binomially
distributed random variable on parameters $n$ and $p_{i+1}$. In particular,
the probability of node $v$ having $k$ children equals
\begin{equation}\label{eq:pdf}
P(\xi_{iv} = k) = \frac{\binom{n}{k} (1-p_{i+1})^{n-k}}{1-(1-p_{i+1})^n}
\end{equation}
and the corresponding probability generating function equals
\begin{equation}\label{eq:generating_func}
\mathcal{B}_0(n,p_{i+1})\overset{d}{=}
G_i(z) = \frac{(1+p_{i+1}(z-1))^n - (1-p_{i+1})^n}{1-(1-p_{i+1})^n}.
\end{equation}

\end{Lemma}
\begin{proof}
In order to prove the lemma, we have to show that
$\xi_{iv}\overset{d}{=}\mathcal{B}_0(n, p_{i+1}).$
Consider an arbitrary node $v$ on level $1\leq i\leq\sigma.$ According to the
definition of the multiset-trie, a node exists at level $i$ if and only if
it has at least one child. Note that this is not true for the nodes on the leaf
level $\sigma + 1.$ This implies that a node on level $i$ can have
$k\in\{ 1,2,\ldots, n\}$ children. Let $X_0, X_1, \ldots, X_{n-1}$ be random
variables; they are defined as follows:
\[
X_k = \begin{cases}
0 & \textrm{child $k$ of node $v$ does not exist} \\
1 & \textrm{child $k$ of node $v$ exists} \\
\end{cases}
\]
As was shown in the previous Lemma~\ref{l:prob-children}, the distribution of 
$X_k$ is $X_k\overset{d}{=} Bernoulli(p_{i+1}).$ Since our model assumes that all the 
multisets in $M$ are chosen uniformly at random, the variables 
$X_k,X_l$ are independent for $k\neq l.$ However, in our case, the node $v$ cannot 
have 0 children, so the sum $\sum_{k=1}^n X_k$ has a zero-truncated binomial 
distribution,
%
\[
\sum_{k=1}^n X_k \overset{d}{=} \mathcal{B}_0(n,p_{i+1}),
\] 
%
which completes the proof.
\end{proof}
%
Knowing the probability density and probability generating functions of $\xi_{ij}$ 
from Lemma~\ref{l:prob-children}, we can now estimate the number of nodes in 
a randomly generated multiset-trie as follows:
%
\begin{equation}\label{eq:num_nodes}
\mathbb{E}( | \mathcal{M} | ) = \mathbb{E}\left[ \sum_{i=1}^{\sigma+1} \xi_i \right].
\end{equation}
%

%Note that in a real world models multisets in $M$ are not chosen with the same 
%probability. Specifically, in some cases, the probability $P(m\in M)$ varies 
%dramatically and can be even equal to $0.$ For example, if words are mapped 
%to multisets, then the sample space $N$ contains very large multisets. However, 
%most of them will have probability $P(m\in M)=0,$ because a word that would correspond 
%to such a multiset simply does not exist. So, we can safely conclude that multiset-trie 
%is an input sensitive data structure, because the size of multiset-trie $|\mathcal{M}|$ 
%depends on the probability distribution function $P(m\in M).$
%

In order to evaluate~(\ref{eq:num_nodes}), we will use some of the tools from 
a Galton--Watson process; see Gardiner~\cite{gardiner1985stochastic} for 
an introduction.
Using Equations~(\ref{eq:sum_recursive}) and~(\ref{eq:generating_func}), we can 
derive the probability generating function for the random variable $\xi_{i+1}$ as 
\begin{equation}\label{eq:g_rec}
G_{\xi_{i+1}}(z) = G_{\xi_i}(G_i(z)).
\end{equation}
Since there is always precisely one node at the root level, we have $P(\xi_1 = 1) = 1.$ 
Hence, the probability generating function for the random variable $\xi_1$ is 
\begin{equation}\label{eq:g_xi_one}
G_{\xi_1}(z) = z^1 = z
\end{equation}
which is the initial condition for the recursive Equation~(\ref{eq:g_rec}).

\begin{Proposition}\label{prop:exp_rec}
The expectation of the random variable $\xi_{i+1}$ can be expressed 
as follows.
\[
\mathbb{E}(\xi_{i+1}) = \mathbb{E}(\xi_i)\mathbb{E}(\mathcal{B}_0(n,p_{i+1}))
\]
for $1\leq i\leq\sigma.$
\end{Proposition}
\begin{proof}
Using the following property of probability generating function
\begin{equation}\label{eq:prop_exp-pgf}
G_X'(1^-) = \mathbb{E}(X)
\end{equation}
the expectation for the random variable $\xi_{i+1}$ can be derived in terms of Equation~(\ref{eq:g_rec}).
\begin{eqnarray}\label{eq:exp}
\mathbb{E}(\xi_{i+1}) &=& G_{\xi_{i+1}}'(1^-) \nonumber \\
&=& G_{\xi_i}'(G_i(1^-))G_i'(1^-).
\end{eqnarray}
According to~(\ref{eq:pdf}) and~(\ref{eq:generating_func}), the value of $G_i(z)$ at 
$1$ is $1$ and the value of its derivative at $1$ is $\mathbb{E}(\mathcal{B}_0(n,p_{i+1})).$ 
Substituting the values of $G_i(1^-)$ and $G_i'(1^-),$ and applying the 
property~(\ref{eq:prop_exp-pgf}), we complete the proof.
\end{proof}
%
From Proposition~\ref{prop:exp_rec} above and Lemma~\ref{l:prob-children}, we 
can conclude that 
\begin{eqnarray}
\mathbb{E}(\xi_{i}) &=& \mathbb{E}(\xi_{i-1})\mathbb{E}\left( \mathcal{B}_0(n,p_{i}) \right) \nonumber \\
& = & \mathbb{E}(\xi_{i-1})\frac{np_{i}}{1-(1-p_{i})^n}.
\end{eqnarray}

\begin{Theorem}\label{thm:exp_level}
Let $\mathcal{M}$ be a multiset-trie defined with parameters $n,$ $\sigma,$ and denote the number 
of nodes on every level $i$ by a random variable $\xi_i.$ Furthermore, let all multisets appear in 
$\mathcal{M}$ with equal probability $p\in (0,1).$ Then, the expected number 
of nodes on every level of $\mathcal{M},$ i.e., $\mathbb{E}(\xi_i)$, is defined as 
\begin{equation}\label{eq:nodes_level}
\mathbb{E}(\xi_{i}) = n^{i-1} \frac{1-(1-p)^{n^{\sigma +1 -i}}}{1-(1-p)^{n^{\sigma}}}.
\end{equation}
\end{Theorem}
\begin{proof}
According to~(\ref{eq:g_xi_one}), the expected number of nodes on the first level is 1. \\
Using $\mathbb{E}(\xi_1) = 1$ and the result from Proposition~\ref{prop:exp_rec},
we obtain
\begin{align*}
\mathbb{E}(\xi_{i}) &= \prod_{j=2}^{i} \frac{n p_j}{1-(1-p_j)^n}
= \prod_{j=2}^{i} n \frac{1-(1-p)^{n^{\sigma +1-j}}}{1-(1-p)^{n^{\sigma + 2 -j}}} \\
&= n^{i-1} \frac{1-(1-p)^{n^{\sigma +1 -i}}}{1-(1-p)^{n^{\sigma}}}
\end{align*}
\end{proof}

Having derived the expected number of nodes on every level of multiset-trie, the
expected value of the total number of nodes in a multiset-trie can be calculated 
with respect to the parameters $n,$ $\sigma$ and $p.$ This result is obtained in 
the next theorem.

\begin{Theorem}\label{thm:exp_nodes}
The expected cardinality of a multiset-trie defined on parameters $n,$ $\sigma$, 
and $p$ can be computed as 
\begin{equation}
\mathbb{E}( | \mathcal{M} | ) = \sum_{i=1}^{\sigma+1} n^{i-1} \frac{1-(1-p)^{n^{\sigma +1 -i}}}{1-(1-p)^{n^{\sigma}}},
\end{equation}
where $r = (1-p)^n,$ so $r\in(0,1).$
\end{Theorem}
\begin{proof}
Using the results obtained from Theorem~\ref{thm:exp_level}, we compute 
\begin{align*}
\mathbb{E}( | \mathcal{M} | ) &= \mathbb{E}\left[ \sum_{i=1}^{\sigma+1} \xi_i \right] \\
&= \sum_{i=1}^{\sigma + 1} n^{i-1} \frac{1-(1-p)^{n^{\sigma + 1 -i}}}{1-(1-p)^{n^{\sigma}}}
\end{align*}
\end{proof}

With the expected number of nodes in a multiset-trie $\mathcal{M}$ obtained from 
Theorem~\ref{thm:exp_nodes}, we can now generalize the result for a subtree 
in $\mathcal{M}$ parametrized by an input multiset $m.$ The subtrees that we 
are interested in are the ones that contain all the sub-multisets or all the 
super-multisets of $m.$ In order to calculate the expected cardinality of such 
subtrees, we need the following definition.

\begin{Definition}\label{def:params}
Let $m = \{ 1^{k_1}, 2^{k_2}, \ldots, \sigma^{k_\sigma} \},$ where 
$e^{k_e}$ is an element $e$ with multiplicity $k_e.$ Let $M_1, M_2$ be 
the subsets of the set $M,$ such that $M_1 = \{ x\in M : x\subseteq m \}$ and 
$M_2 = \{ x\in M : x\supseteq m \}.$
Define $\alpha_i$ and $\beta_i$ as follows: 
\begin{equation*}
\alpha_i = \begin{cases}
1, & i=0 \\
\prod_{j=1}^i (k_j+1), & 1\leq i\leq\sigma \\
\end{cases}
\end{equation*} 
and 
\begin{equation*}
\beta_i = \begin{cases}
1, & i=0 \\
\prod_{j=1}^i (n-k_j-1), & 1\leq i\leq\sigma \\
\end{cases}.
\end{equation*}
\end{Definition}

The expected cardinality of the subtrees containing the multisets from $M_1$ or $M_2$ 
is defined in the following corollary.

\begin{Corollary}\label{cor:exp_nodes_param}
Let $M_1, M_2, \alpha_i$ and $\beta_i$ be defined as in  Definition~\ref{def:params}; 
then, the expected cardinality of a multiset-trie subtree $\mathcal{M}_{M_1}$ that contains all the multisets 
from the set $M_1$ is equal to 
\vspace{6pt} 
\begin{equation}\label{eq:nodes_sub_param}
\mathbb{E}( |\mathcal{M}_{M_1}| ) = \sum_{i=1}^{\sigma + 1} \alpha_{i-1} \frac{1-(1-p)^{\alpha_{i-1}}}{1-(1-p)^{\alpha_{\sigma}}}.
\end{equation}
The expected cardinality of a multiset-trie subtree $\mathcal{M}_{M_2}$ that contains all the multisets 
from the set $M_2$ is equal to 
\begin{equation}\label{eq:nodes_super_param}
\mathbb{E}( |\mathcal{M}_{M_2}| ) = \sum_{i=1}^{\sigma + 1} \beta_{i-1} \frac{1-(1-p)^{\beta_{i-1}}}{1-(1-p)^{\beta_{\sigma}}}.
\end{equation}
\end{Corollary}
%
\begin{proof}
Using the results from Theorems~\ref{thm:exp_level} and~\ref{thm:exp_nodes}, 
we derive the formulas~(\ref{eq:nodes_sub_param}) and~(\ref{eq:nodes_super_param}) 
by specifying the possible number of nodes on every level in the multiset-trie according 
to the multiset $m.$ Note that the formula~(\ref{eq:nodes_level}) assumes that, on every 
level but the first one, there are $n$ possible nodes. Given a sub-multiset or super-multiset 
query and an input multiset $m$, the number of nodes that will be traversed on level $i$ is 
defined by the number $k_{i-1}+1$ or $n-k_{i-1}-1$ for $i\geq 2.$ On level $i=1$, 
there is only one root node in any multiset-trie $\mathcal{M},$ which always exists 
if $M\neq\emptyset$ and is traversed for any type of query (sub-multiset and 
super-multiset).
\end{proof}

\subsubsection{GetAllSubmsets and GetAllSupermsets}\label{ss:getall}
In this subsection, we discuss the running time complexity of the functions 
\textsc{getAllSubmsets} and \textsc{getAllSupermsets}. It is obvious that any 
other algorithm for retrieving all the sub-multisets or super-multisets has a worst-case
running time complexity of at least $O(|M|).$ Hence, the functions 
\textsc{getAllSubmsets} and \textsc{getAllSupermsets} have the worst-case 
running time complexity $O(|\mathcal{M}|).$ Indeed, the case when the algorithms 
retrieve all the multisets stored in a multiset-trie by traversing the whole structure 
can be easily~constructed.

Consider the function \textsc{getAllSubmsets}. The function takes some multiset 
$m$ as an input argument. Then, it returns a set of multisets $\{x\in M : x\subseteq m \}$ 
from the multiset-trie $\mathcal{M}$. Having a multiset $m$ set to the largest 
possible multiset in $N$ (it can also be larger), 
\[
m = \{ 1^{n-1},2^{n-1},\ldots, \sigma^{n-1} \}
\]
the whole multiset-trie is traversed during the \textsc{getAllSubmsets} query.

Now, let us consider the function \textsc{getAllSupermsets}. Similarly, the function 
takes a multiset $m$ as an input argument. However, in this case, it returns the 
set of multisets $\{ x\in M : x\supseteq m \}$ from the multiset-trie 
$\mathcal{M}.$ In order to obtain the traversing of all the multiset-trie, one must set 
$m$ to the smallest possible multiset, i.e., an empty multiset
\begin{equation*}
m = \{ \emptyset \} = \{ 1^0,2^0, \ldots, \sigma^0 \}.
\end{equation*}

Thus, we can conclude that the worst-case running time complexity of the 
functions \textsc{getAllSubmsets} and \textsc{getAllSupermsets} is $O(\mathbb{E}( |\mathcal{M}| )).$ 
According to Theorem~\ref{thm:exp_nodes}, the expected number of visited 
nodes in the worst case is 
\begin{equation*}
O(\sum_{i=1}^{\sigma+1} n^{i-1} \frac{1-(1-p)^{n^{\sigma +1 -i}}}{1-(1-p)^{n^{\sigma}}}).
\end{equation*}
According to Theorem~\ref{cor:exp_nodes_param}, the worst-case running 
time complexity given an input multiset $m$ for the function \textsc{getAllSubmsets} is 
\begin{equation*}
O(\sum_{i=1}^{\sigma + 1} \alpha_{i-1} \frac{1-(1-p)^{\alpha_{i-1}}}{1-(1-p)^{\alpha_{\sigma}}})
\end{equation*}
and for the function \textsc{getAllSupermsets} is 
\vspace{6pt} 
\begin{equation*}
O(\sum_{i=1}^{\sigma + 1} \beta_{i-1} \frac{1-(1-p)^{\beta_{i-1}}}{1-(1-p)^{\beta_{\sigma}}}).
\end{equation*}

%Taking into account that $\sigma |M| \geq |\mathcal{M}|,$ we can say that the 
%running time complexity is $O(\sigma |M|).$ However, this is a very rough upper 
%bound, which overestimates the complexity a lot.

\subsubsection{SubmsetExistence and SupermsetExistence}\label{ss:exists}
We start the analysis of the functions \textsc{submsetExistence} and 
\textsc{supermsetExistence} with an observation. Our theoretical model assumes 
that all the multisets are inserted into multiset-trie at random. It was already concluded 
that the probability distribution function $P(m\in M)$ has an impact on the size of 
multiset-trie $\mathcal{M}.$ 
Moreover, this distribution influences the performance 
of the functions \textsc{submsetExistence} and \textsc{supermsetExistence} even more.

For a real-world model, such that $P(m\in M)\neq const$, the performance of the search 
algorithms directly depends on the number of nodes on every level $\xi_i.$ When 
the search functions check if a multiset is in multiset-trie, the complete path that 
corresponds to this multiset is checked. Knowing this fact, the search can be optimized 
during the construction of a multiset-trie.

Recall that a multiset-trie is defined on parameters $n,$ $\Sigma,$ $\sigma = |\Sigma|$, 
and $\phi.$ Let the frequency of an element $e$ in a multiset $m$ be the multiplicity of 
$e$ in $m,$ denoted by $\emph{mult}_m(e).$ Then, the frequency of an element $e$ 
can be defined as a sum $\sum_{m\in M} \emph{mult}_m(e).$ According to the frequencies 
of elements in $\Sigma,$ the performance of the multiset-trie can be optimized by 
the mapping $\phi : \Sigma \rightarrow I.$ Indeed, the ordering of elements by their 
frequencies has an influence on the performance.
%
The frequency of an element $e\in\Sigma$ affects the distribution of $\xi_{\phi(e)}$ 
as follows. The larger the frequency of $e$, the larger the number of nodes on the
$\phi(e)$ level. 
Thus, if the number of nodes on lower levels is greater than on higher levels, then 
the search functions will discard complete paths that do not satisfy the query 
faster. Hence, the closest match will be found faster.

%
%\begin{Lemma}\label{l:frequency}
%Let $M$ be a collection of multisets. Let the frequency of an element $e$ in a 
%multiset $m$ be the multiplicity of $e$ in $m,$ denoted by $\emph{mult}_m(e).$ 
%Then the frequency of an element $e$ is the sum 
%\[
%\sum_{m\in M} \emph{mult}_m(e).
%\] 
%According to the frequencies of elements in $\Sigma,$ the performance of the 
%multiset-trie can be optimized. In particular, the mapping $\phi : \Sigma \rightarrow I$ 
%must order elements by their frequencies as follows. The smaller the frequency of 
%an element the smaller its index.
%\end{Lemma}
%%
%\begin{proof}
%The frequency of an element $e\in\Sigma$ affects the distribution of $\xi_{\phi(e)}$ 
%as follows. The larger the frequency of $e$ the larger the number of nodes on 
%$\phi(e)$ level. 
%So, if the number of nodes on lower levels is greater than on higher levels, then 
%the search functions will discard complete paths that do not satisfy the query 
%faster. Hence, the closest match will be found faster.
%\end{proof}
%%

Let us now switch back to our mathematical model and note that the influence 
of the mapping function $\phi$ in our model has an inessential impact on the performance, 
because all the multisets are equally likely, and the whole domain $N$ is used for 
sampling multisets.

Consider both functions \textsc{submsetExistence} and \textsc{supermsetExistence}. 
Whenever the result is \emph{false}, i.e., no multiset in $M$ is 
a sub-multiset or super-multiset of an input multiset $m,$ both functions in the 
worst case visit all the nodes in $\mathcal{M}$ but the nodes on the leaf level. 
Of course, such a case would be very rare, assuming a random input model, but 
it can be constructed as follows.

Consider the function \textsc{submsetExistence}. Then, given an input multiset 
$m = \{ 1^{k_1}, 2^{k_2}, \ldots, \sigma^{k_\sigma} \},$ the collection of inserted 
multisets $M$ must be equal to $M = \{ x\in M : k_{x,\sigma} > k_{m,\sigma} \}.$ 
Analogically, for the function \textsc{supermsetExistence} with an input multiset 
$m = \{ 1^{k_1}, 2^{k_2}, \ldots, \sigma^{k_\sigma} \},$ the collection of inserted 
multisets $M$ must be equal to $M = \{ x\in M : k_{x,\sigma} < k_{m,\sigma} \}.$

Thus, the worst-case running time complexity of the functions 
\textsc{submsetExistence} and \textsc{supermsetExistence} is $O(|\mathcal{M}| - |M|).$ 
According to Theorem~\ref{thm:exp_nodes}, this value is 
\begin{equation*}
O(\sum_{i=1}^{\sigma} n^{i-1} \frac{1-(1-p)^{n^{\sigma +1 -i}}}{1-(1-p)^{n^{\sigma}}}).
\end{equation*}
According to Theorem~\ref{cor:exp_nodes_param}, the worst-case running 
time given an input multiset $m$ for the function \textsc{submsetExistence} is 
\begin{equation*}
O(\sum_{i=1}^{\sigma} \alpha_{i-1} \frac{1-(1-p)^{\alpha_{i-1}}}{1-(1-p)^{\alpha_{\sigma}}})
\end{equation*}
and for the function \textsc{supermsetExistence} is 
\begin{equation*}
O(\sum_{i=1}^{\sigma} \beta_{i-1} \frac{1-(1-p)^{\beta_{i-1}}}{1-(1-p)^{\beta_{\sigma}}}).
\end{equation*}
Note that the summation goes only up to $\sigma$ and not up to $\sigma + 1$ as 
in Theorem~\ref{thm:exp_nodes} or in Theorem~\ref{cor:exp_nodes_param}.

As for the case wherein the outcome of the functions \textsc{submsetExistence} and 
\textsc{supermsetExistence} is \emph{true}, one has to guarantee the termination 
of the algorithm at some node on the leaf level. The worst-case scenario can be 
constructed in the same way as for the \emph{false} outcome but with two more 
multisets in $M.$ The first multiset is the empty multiset. With the empty multiset, 
the function \textsc{submsetExistence} will visit the same amount of nodes as for 
the \emph{false} case, plus one more for the empty multiset. The second multiset 
is the maximal possible multiset from $N.$ In this case, the function \textsc{supermsetExistence} 
will also visit the same amount of nodes as for the \emph{false} case, plus one more 
for the maximal multiset. Hence, the worst-case running time complexity for both 
outcomes (\emph{true} and \emph{false}) is the same.

\subsection{Space Complexity}\label{s:spacecomplexity}
As in any efficient algorithm, there is always some trade-off between space and 
time complexity. While offering efficient sub- and super-multiset queries, an 
additional space must be provided for multiset storage. 
%Clearly, the cardinality 
%of the set $M$ is smaller than the size of $\mathcal{M},$ because the number 
%of multisets in $\mathcal{M}$ is equal to the number of nodes only on the leaf 
%level. 
As we increase the parameter $p$, the number of sets we need to store (i.e., number of leaves) increases linearly. Figure~\ref{f:exp-nodes} shows that the additional overhead space that our data structure uses does not increase as the tree is becoming more and more dense.

\begin{figure}[H]

\includegraphics[width=.64\textwidth, keepaspectratio]{exp-nodes-on-probab.pdf}
\caption{\hl{Size} %MDPI: Please change the terms into scientific notations in the figure,  e.g., "$8 \times 10^{3}$", not "8E3".
 of a tree ($\mathbb{E}(|\mathcal{M}|)$) compared to the number of stored sets ($\mathbb{E}(|M|)$), as the probability parameter $p$ increases.
Parameters 
$\sigma$ and $n$ are $26$ and $10,$ respectively.
The additional overhead space that our data structure uses does not increase as the tree is becoming more and more dense.
}
\label{f:exp-nodes}
\end{figure}

As we see in Figure~\ref{f:exp-nodes}, the value of $|\mathcal{M}|$ is slightly 
shifted with respect to the value of $|M|.$

Now, we demonstrate a more descriptive comparison between $|\mathcal{M}|$ and 
$|M|.$ Figure~\ref{f:ratio-exp-msets} shows the ratio between the expected cardinality 
of a multiset-trie $|\mathcal{M}|$ and the actual number of multisets stored $|M|$ for 
parameters $n$ and $\sigma$ being $10$ and $26$, respectively.

Analyzing the graph in Figure~\ref{f:ratio-exp-msets}, we can safely 
say that the upper bound for the ratio is $\sigma + 1.$ The argument holds 
because of the limit 
\begin{equation}
\lim_{p\rightarrow 0^+} \mathbb{E}(\xi_i) = 1,
\end{equation}
where $\xi_i$ is the number of nodes on the $i$-th level and $1\leq i \leq \sigma + 1.$

However, the ratio $\sigma + 1$ can be obtained only with a very small cardinality 
of the set $M,$ in particular $|M| = 1.$ In order to obtain such a case, the 
probability $p$ must be at most~$\frac{1}{n^\sigma}.$

The lower bound for the ratio is obviously at $p=1$ and is equal to 1:
\begin{equation}
\lim_{n,\sigma \rightarrow\infty} \frac{n^{\sigma + 1} - 1}{n^\sigma (n-1)} = 1.
\end{equation}

Since the ratio $\sigma + 1$ can be obtained for a very specific case only and 
with a small increase in probability, the ratio drops rapidly and it can be concluded 
that the space complexity of the multiset-trie is $O(|M|).$

\begin{figure}[H]

\includegraphics[width=.64\textwidth, keepaspectratio]{ratio-exp-nodes-and-msets-on-prob1.pdf}
\caption{The ratio of ``data overhead'' derived from Figure \ref{f:exp-nodes}, as the density $p$ of our data structure increases.
Unless we are in a very sparse setting, the trade-off in terms of space is small.}
\label{f:ratio-exp-msets}
\end{figure}

\subsection{An Informal Comment on the Mathematical Analysis}
In the next section, we show how our data structure preforms in certain environments.
We will see, in our view, great results throughout various experiments, even when compared with state-of-the-art approaches such as Inverted Index.

It was very frustrating to us that, despite these indications, it seems that this structure is very difficult to analyze. 
Even after restricting the analysis to the (usually simple) case wherein the input data are uniformly distributed, and after utilizing reliable probabilistic tools, our approach unfortunately did not provide as clear results as we hoped for. 
We would be very keen to learn of an approach or technique that would offer more insight into the mathematical analysis of our data structure and its algorithms.

% section 5
%\input{tex/s_experiments}
\section{Experiments} \label{c:experiments}

%
This section contains the results of experiments that were performed on the multiset-trie 
data structure. In particular, we tested the functions \textsc{submsetExistence}, 
\textsc{supermsetExistence}, \textsc{getAllSubmsets}, and \textsc{getAllSupermsets}.

The multiset-trie is implemented in the \CC { programming} language. 
The current implementation uses only the standard library of the \CC14 version of the 
standard and has a command line interface~\cite{akulich2019mstrie}. The implementation of the program was 
optimized for testing, and, therefore, the program operates with files to 
process queries. After processing all the queries, the results are stored in files for further analysis.

The current implementation of multiset-trie is not optimized for space consumption.
A node in the multiset-trie is implemented as a record indicating the level of the tree
that corresponds to a symbol from $\Sigma$, and including an array of pointers to
its children indexed by the multiplicities of the child symbols. The arrays have
fixed length $n$ from the interval $[0,n].$
Only some elements of an array may point to a child node, while others include a
$Null$~value.

% Performance of the functions will be measured by 
%the number of visited nodes in multiset-trie by the particular function. In 
%particular the performance is inversely proportional to the number of visited 
%nodes.

Before we start, we will give a few definitions of the parameters 
that were varied throughout the experiments and discuss the experimental data 
that were used.

Let $M$ be a set of multisets inserted into multiset-trie and let $n$ be 
the maximal node degree. Let $N$ be the power multiset of $\Sigma,$ where 
the multiplicity of each element is bounded from above by $n-1.$ We define the 
\emph{density} of a multiset-trie to be the ratio $\frac{|M|}{|N|},$ where 
$|\cdot|$ denotes cardinality.

The selected parameters of the data structure that were varied in the experiments 
were as follows:
%
\begin{itemize}
\item $\sigma$---the cardinality of the alphabet $\Sigma;$
%
\item $n$---the maximal degree of a node, which explicitly defines the maximal 
multiplicity of elements in a multiset;
%
\item $\phi$---mapping of letters from $\Sigma$ into a set of consecutive 
integers;
%
\item $d$---density of a multiset-trie.
%
\end{itemize}
The cardinality of a power multiset $N$ is equal to $n^\sigma,$ which means that 
density $d$ of a multiset-trie depends on parameters $|M|,$ $\sigma$ and $n.$ 
Because parameters $\sigma$ and $n$ are set when a multiset-trie is initialized, 
the parameter $|M|$ was varied to change the density in the experiments. As we 
mentioned in Section~\ref{c:description}, the mapping $\phi$ determines the 
correspondence of letters to levels in multiset-trie, i.e., it defines the ordering of 
levels in multiset-trie. It is also true that $\phi$ defines the ordering in multisets.

% anouncement of the experiments
In the following Sections \ref{c:introduction}--\ref{c:experiments}, we will present the behavior of the multiset-trie data 
structure depending on the selected parameters, as well as the comparative 
benchmark of the multiset-trie against the B-tree implementation of Inverted Index. 
We start with experiments that were performed on artificially generated data in 
order to give a general picture of the multiset-trie's performance.

In Experiment~\hyperref[s:exp1]{1}, a special case of the multiset-trie is considered. 
Only sets are allowed to be stored in the data structure, i.e., the maximally allowed 
multiplicity is set to 1. The performance is measured with respect to the density 
of the multiset-trie.

Experiment~\hyperref[s:exp2]{2} was an extension of the previous one. Here, 
we also measured the performance of the multiset-trie depending on its density. 
The difference was that the allowed multiplicity of an element was raised, i.e., 
the data structure was populated with multisets.

Summarizing the tests of the performance depending on the density, we present 
Experiment~\hyperref[s:exp3]{3}. It shows the nonlinearity of the performance 
with respect to the density of the multiset-trie.

The fourth experiment on the multiset-trie uses real-world data. In 
Experiment~\hyperref[s:exp4]{4}, the influence of the mapping $\phi$ is studied. 
The input data are obtained by mapping real words from the English dictionary 
to the set of consecutive integers using the function $\phi.$ The experiment 
shows that the performance of the multiset-trie is noticeably influenced by 
different mappings $\phi.$ It also shows the usability of the multiset-trie in terms 
of real data, demonstrating the high performance of search queries.

Finally, the empirical comparison of the multiset-trie data structure with
the B-tree-based Inverted Index is presented in Experiment~\hyperref[s:exp5]{5}. We use
Inverted Index to store and retrieve multisets in the same way as
is described in the paper by Helmer and Moerkotte~\cite{Helmer2003}
for sets. In the comparison, we use three types of queries: exact,
sub-multiset, and super-multiset~retrieval.

\subsection{\hl{Data} %MDPI: Section headings should be numbered sequentially, e.g., Section 2.1, Section 2.2.1. Please confirm this revision.
 Generation}
We denote by \emph{input data} the data that are used to fill the structure prior 
to testing and by \emph{test data} the set of queries that are used to test the 
performance of the functions.

The artificially generated input data are obtained by sampling $|M|$ multisets 
from $N.$ All the multisets in $N$ are constructed according to parameters 
$\sigma$ and $n$ and represent the power multiset of the alphabet $\Sigma.$ 
Every multiset in $M$ is chosen from $N$ with equal probability $p.$ Thus, the 
probability $p$ gives a collection $M$ of multisets that are sampled from $N$ 
with uniform distribution. Uniform distribution is chosen in order to simulate
random user~input.

% test data explanation
The test data are generated artificially and constructed as follows. Given the 
parameters $\sigma$ and $n,$ the possible size of a multiset varies from~1 
to~$\sigma n.$ The number of randomly generated test multisets for every 
value of multiset size is 1500. In other words, we perform 1500 experiments 
in order to measure the number of visited nodes for the queries with a test multiset 
of distinct sizes. The final value of visited nodes is calculated by taking an 
arithmetic mean among all 1500 measurements.

\subsection{Experiment 1} \label{s:exp1}
This experiment shows the performance of multiset-trie being used for storing 
and retrieving \emph{sets} instead of \emph{multisets}. We restrict multiset-trie in order 
to obtain a closer comparison with the \emph{set-trie} data structure~\cite{savnik2013index}.
In this case, we set the maximal node degree $n$ to be $2$ and $\sigma$ to be 25. 
The mapping $\phi$ does not have an influence in this particular experiment
because the input data are generated artificially with uniform distribution. On 
average, the results will be the same for any $\phi,$ since all the multisets are 
equally likely to appear in $M.$ The parameter $|M|$ varies from \hl{10,000} %MDPI: We added commas to separate thousands for numbers with five or more digits. Please confirm.
 sets up 
to \hl{320,000} sets. According to the parameters $n$ and $\sigma,$ the cardinality of 
$N$ is $33,554,432\approx \num{3.36e+7}.$ Thus, the calculated density of the 
multiset-trie with respect to $|M|$ varies from~$\num{0.3e-3}$ to~$\num{9.5e-3}.$

The performance of the functions \textsc{submsetExistence} and 
\textsc{supermsetExistence} increases as the density increases (see Figure~\ref{fig:4}a,b). The results are as expected because the increase in the 
density increases the probability of finding a sub-multiset or super-multiset in 
multiset-trie, which leads to a lower number of visited nodes.

\begin{figure}[H]
	\subcaptionbox{\centering  \label{fig:e1m1}}{\includegraphics[width=.485\textwidth]{exp1-m1.pdf}}
	\subcaptionbox{\centering \label{fig:e1m2}}{\includegraphics[width=.485\textwidth]{exp1-m2.pdf}}
	\caption{\hl{Existence} %MDPI: Please use commas to separate thousands for numbers with five or more digits (not four digits) in the picture, e.g., "10000" should be "10,000".
 functions of Experiment 1. (\textbf{a}) \hl{submsetExistence;} %MDPI: We moved the subfigure explanations into the figure caption. Please confirm. The following are same.
 (\textbf{b}) \hl{supermsetExistence}.  \label{fig:4}}
\end{figure}

The maxima are located between 175 and 375 for \textsc{submsetExistence} and 
between 175 and 350 for \textsc{supermsetExistence}. According to these maxima,
we can deduce that at least 7--15 multisets were checked in order to find 
a sub-multiset or super-multiset, which is from $\num{0.02e-3}$ to $\num{1.5e-3}$ of the 
multiset-trie and from $\num{1.9e-7}$ to $\num{4.5e-7}$ of the complete 
multiset-trie.

As the density increases, the peaks shift from the center to the left or to the right, 
for \textsc{submsetExistence} and \textsc{supermsetExistence}, respectively. 
The shifts are the consequence of the uniform distribution of sets in $M.$ 
Since every set has the same probability of appearing in $M,$ the distribution of set 
sizes in $M$ is normal. Consequently, with the increase in the density of the 
multiset-trie, the number of sets in $M$ with cardinality $\frac{1}{2}\sigma$ will be 
larger than the number of sets with cardinality $\frac{1}{2}\sigma\pm\epsilon,$ 
for $\frac{1}{2}\sigma > \epsilon > 0.$ Thus, the function \textsc{submsetExistence} 
needs to visit less nodes for test sets of size $\frac{1}{2}\sigma$ than for test 
sets of size $\frac{1}{2}\sigma\pm\epsilon.$ The function decreases the 
multiplicity of some elements (in some cases, it skips them) in order to find the 
closest subset. Hence, the peak shifts to the left. Oppositely, the function 
\textsc{supermsetExistence} increases the multiplicity of some elements 
(in this case, adding new elements) in order to find the closest superset. 
Thus, the peak shifts to the right.

Note that despite the peak shifts, both functions \textsc{submsetExistence} and 
\textsc{supermsetExistence} have approximately the same worst-case performance.

The performance of the functions \textsc{getAllSubmsets} and \textsc{getAllSupermsets} 
decreases as the density increases (see Figure~\ref{fig:5}a,b). 
This happens because the number of multisets in multiset-trie increases, which means 
that any multiset in the data structure will have more sub- and super-multisets. 
The maxima for both functions vary from $\num{8.0e4}$ to $\num{1.5e6}$ visited nodes. 
We can notice that the local maxima for the functions \textsc{getAllSubmsets} and 
\textsc{getAllSupermsets} differ with respect to the length of input. The 
explanation is very simple. In order to find all sub-multisets of a small set, the 
function has to traverse a small part of the multiset-trie. As the size of a set 
increases, the part of a multiset-trie where all the sub-multisets of a given set 
are stored also increases. The opposite holds for the function 
\textsc{getAllSupermsets}.

\begin{figure}[H]

\subcaptionbox{\centering 
\label{fig:e1m3}}{\includegraphics[width=.485\textwidth]{exp1-m3.pdf}
}
\subcaptionbox{\centering 
\label{fig:e1m4}}{\includegraphics[width=.485\textwidth]{exp1-m4.pdf}
}
\caption{\hl{Exhaustive} %MDPI: Please use commas to separate thousands for numbers with five or more digits (not four digits) in the picture, e.g., "10000" should be "10,000". please check all figures.
 functions of Experiment 1. (\textbf{a}) \hl{getAllSubmsets}; (\textbf{b}) \hl{getAllSupermsets}.\label{fig:5}}
\end{figure}

Despite the fact, that for a lookup of any set/multiset, $\sigma$ nodes must be visited 
in multiset-trie on average, the data structure has very similar performance 
results in comparison to the \emph{set-trie} data structure.

\subsection{Experiment 2} \label{s:exp2}
In Experiment~\hyperref[s:exp2]{2}, we demonstrate the performance of 
the unrestricted multiset-trie allowing \emph{multisets} to be inserted into the data structure. 
We set $n$ to be 6 and retain $\sigma = 25$, as in Experiment~\hyperref[s:exp1]{1}. 
The mapping $\phi$ does not have an influence on the results, since the input 
data are generated artificially with uniform distribution. The cardinality of $M$ 
varies from \hl{40,000} to \hl{640,000} multisets. Thus, the calculated density $d$ varies 
from $\num{1.4e-15}$ to $\num{2.25e-14}.$ The density is much smaller than 
in Experiment~\hyperref[s:exp1]{1}, because now we allow multisets to be stored 
in the data structure, and according to the parameters $n$ and $\sigma$, the 
cardinality of $N$ is $6^{25} = \num{2.84e19}.$

As we can see from the graphs in Figure~\ref{fig:e2m1}a,b, 
the performance of the functions \textsc{submsetExistence} and 
\textsc{supermsetExistence} becomes worse as the density increases. 
In this case, the number $|M|$ is slightly larger than in 
Experiment~\hyperref[s:exp1]{1}, but the density is very small. Consequently, 
multiset-trie becomes more sparse. Multisets in a sparse multiset-trie differ more, 
which leads to a larger number of visited nodes.

The maxima for both functions vary from 7500 to 25,000 visited nodes. According 
to these maxima, at least 300--1000 multisets were checked in order to find a 
sub-multiset or super-multiset, which is from $\num{1.5e-3}$ to $\num{7.5e-3}$ of the entire 
multiset-trie and from $\num{1.1e-17}$ to $\num{3.4e-17}$ of the complete 
multiset-trie. The percentage of visited multisets with respect to $|M|$ is 
larger than in Experiment~\hyperref[s:exp1]{1}. However, if one would compare 
the percentage of visited multisets with respect to the complete multiset-trie, then, 
in the case of Experiment~\hyperref[s:exp2]{2}, it is less by 10 orders than in 
Experiment~\hyperref[s:exp1]{1}.

\begin{figure}[H]
\subcaptionbox{\centering
}{\includegraphics[width=.485\textwidth]{exp2-m1.pdf}}
\subcaptionbox{\centering
\label{fig:e2m2}}{\includegraphics[width=.485\textwidth]{exp2-m2.pdf}}
\caption{Existence functions in Experiment 2. (\textbf{a}) \hl{submsetExistence}; (\textbf{b}) \hl{supermsetExistence}.\label{fig:e2m1}}
\end{figure}

The peaks are shifted from the center to the left and right for 
\textsc{submsetExistence} and \textsc{supermsetExistence}, respectively. Such 
behavior was previously observed in Experiment~\hyperref[s:exp1]{1}. The 
explanation is the same: the input data have a uniform distribution, implying that 
the size of multisets in $M$ is normally distributed. Because of the normal 
distribution of the size of multisets, the shift in the peak occurs as the density increases.

It can also be observed that, as in the previous Experiment~\hyperref[s:exp1]{1}, both 
functions \textsc{submsetExistence} and \textsc{supermsetExistence} have similar 
worst-case performance.

The functions \textsc{getAllSubmsets} and \textsc{getAllSupermsets} decrease 
their performance as the density increases (see Figure~\ref{fig:e2m3}a,b). This happens because the number of multisets increases as 
the density increases. Thus, there are more nodes that have to be visited in order to 
retrieve all sub- or super-multisets of some multiset. The maximum for both functions 
varies from $\num{0.9e5}$ to $\num{1.5e7}$ visited nodes. As was observed in 
Experiment~\hyperref[s:exp1]{1}, the maxima occur at opposite points. For the 
function \textsc{getAllSubmsets}, it will always be at the largest size of the multiset, 
which is 125 in our case. Conversely, the maximum for \textsc{getAllSupermsets} 
is at the smallest size of multiset, which is 0 (an empty set).

\begin{figure}[H]

\subcaptionbox{\centering \hl{Experiment} %Please move sub-caption to Caption.
2, getAllSubmsets function.
}{\includegraphics[width=.485\textwidth]{exp2-m3.pdf}}
\subcaptionbox{\centering \hl{Experiment} 2, getAllSupermsets function.
\label{fig:e2m4}}{\includegraphics[width=.485\textwidth]{exp2-m4.pdf}}
\caption{Exhaustive functions in Experiment 2.\label{fig:e2m3}}
\end{figure}

The results of Experiment~\hyperref[s:exp1]{1} show that the performance 
of functions \textsc{submsetExistence} and \textsc{supermsetExistence} increases 
as the density increases. However, we observe the opposite behavior in 
Experiment~\hyperref[s:exp2]{2}. We explain the reason for such a contradiction 
in the next Experiment~\hyperref[s:exp3]{3}

\subsection{Experiment 3} \label{s:exp3}
The results of Experiment~\hyperref[s:exp1]{1} and Experiment~\hyperref[s:exp2]{2} 
have shown that as the density of a multiset-trie increases, the performance of 
functions \textsc{submsetExistence} and \textsc{supermsetExistence} can both become 
better and worse. The reason for such behavior is that the dependence of the 
number of visited nodes on the density is not a linear function. 
The performance of the abovementioned functions is maximal when the multiset-trie is 
complete. As the multiset-trie becomes more sparse (the density is small), the multisets
differ more, and the number of visited nodes increases. However, the multisets differ
less when the density is high, so the number of visited nodes decreases. Since 
the dependence of the number of visited nodes on the density of multiset-trie 
is a continuous function in the interval $[0,1],$ there exists a global maximum. 
In other words, there exists such a value of density where the number of visited 
nodes is maximal.

In this experiment, we empirically find the extremum of density for functions 
\textsc{submsetExistence} and \textsc{supermsetExistence}. The parameters 
$\sigma$ and $n$ are set to 12 and 5, respectively. The density varies from 
$\num{1.0e-6}$ to $\num{1.0e-2}.$ The number of visited nodes was chosen to be 
maximal for each value of a particular density.

As we see in Figure~\ref{fig:e3m1}a,b both functions 
\textsc{submsetExistence} and \textsc{supermsetExistence} have the maximum 
around $d\approx \num{7.0e-5}.$ The maximum is less than $\num{0.3e-3}$ and 
greater than $\num{1.4e-15},$ which explains the behavior of multiset-trie in 
Experiment~\hyperref[s:exp1]{1} and Experiment~\hyperref[s:exp2]{2}. It is safe 
to say that the maximum may vary depending on parameters $n$ and $\sigma,$ but 
such a maximum always exists. Therefore, we omit the experiments with different 
parameters $n$ and $\sigma.$

\begin{figure}[H]

\subcaptionbox{\centering}{\includegraphics[width=.485\textwidth]{exp4-m1.pdf}}
\subcaptionbox{\centering 
\label{fig:e3m2}}{\includegraphics[width=.485\textwidth]{exp4-m2.pdf}}
\caption{Existence functions in Experiment 3. (\textbf{a}) \hl{submsetExistence}; (\textbf{b}) supermsetExistence.\label{fig:e3m1}}
\end{figure}

\subsection{Experiment 4} \label{s:exp4}
In previous experiments, the input was generated artificially with a uniform 
distribution, so there was no influence of the mapping function $\phi$ on 
the performance of the tested functions. This experiment shows the influence of the 
mapping $\phi$ from alphabet $\Sigma$ to a set of consecutive integers. 
We obtain the influence by taking the real-world data as input data.

The data are taken from the English dictionary, which contains \hl{235,883} different words. 
These words are mapped to multisets of integers according to the $\phi.$ In 
particular, we are interested in cases where $\phi(\Sigma)$ enumerates 
letters by their relative frequency in the English language. We say that $\phi(\Sigma)$ 
maps letters in \emph{ascending order} if the most frequent letter is mapped to 
number $\sigma.$ Conversely, in \emph{descending order}, this letter is mapped to 
the number $1.$ The size of the alphabet $\sigma$ is set to the size of the English 
alphabet: 26. The degree of a node $n$ is set to 10. On average, the multiplicity 
of letters is, of course, less than 10. We choose such a large node degree allowing 
the multiplicity to be up to 10 because the dictionary contains such words.

%\begin{figure}
%\includegraphics[width=\textwidth]{exp4-m3.pdf}
%\caption{Experiment 4, getAllSubmsets function.}
%\end{figure}
%
%\begin{figure}
%\includegraphics[width=\textwidth]{exp4-m4.pdf}
%\caption{Experiment 4, getAllSupermsets function.}
%\end{figure}

The results in Figure~\ref{fig:e4m1}a,b are more balanced when 
letters are ordered by frequency in ascending order. The maxima for the functions 
\textsc{submsetExistence} and \textsc{supermsetExistence} are at 250 visited nodes.

\begin{figure}[H]

\subcaptionbox{\centering}{\includegraphics[width=.475\textwidth]{exp3-m1.pdf}}
\subcaptionbox{\centering 
\label{fig:e4m2}}{\includegraphics[width=.475\textwidth]{exp3-m2.pdf}}
\caption{Existence functions in Experiment 4. (\textbf{a}) submsetExistence; (\textbf{b}) supermsetExistence. \label{fig:e4m1}}
\end{figure}

According to the design of the data structure multiset-trie, we can say 
something about multiset only if we try to reach it, i.e., to find the complete 
path that corresponds to a particular multiset. This means that in order to give 
an answer about the existance of some multiset, one has to check the leaf level in 
multiset-trie.

Letters that have the least frequencies are now located at the top of 
multiset-trie according to the ascending order of letters by frequency. This means 
that the search becomes narrower because a great deal of invalid paths will be 
discarded at the top-most levels. Thus, multiset-trie can be traversed faster.

As may have been noticed, the functions \textsc{getAllSubmsets} and 
\textsc{getAllSupermsets} were not tested in this experiment. These functions 
are not affected by variations in the mapping $\phi,$ because, for any multiset, 
they retrieve all sub-/super-multisets. This means that the number of visited 
nodes will not be changed as $\phi$ varies.

\subsection{Experiment 5} \label{s:exp5}

In this experiment, we demonstrate the performance of the multiset-trie
data structure compared to the Inverted Index based on the
B-tree. Both data structures are implemented in the programming
language \CC, providing in this way an experimental setup for a fair
comparison \cite{akulich2019mstrie}.

%((Brief description of the implementation of the inverted files?))
The Inverted Index is implemented using an idea
from~\cite{Helmer2003}. An inverted index structure consists of two
parts: a dictionary and postings. In our case, a dictionary is
implemented as an in-memory B-tree where keys are all distinct values
from a domain represented by a set $\Sigma.$ The postings are
represented by lists of multisets that contain a particular element
from $\Sigma.$ Each list item in postings contains a cardinality of a
multiset, which speeds up the containment queries.

The experiment uses the input data for the construction of the given
data structure and the test data for the execution of the operations
on the given data structure. The input data comprise a set of
randomly generated multisets used for the construction of a data
structure. The test data include the set of multisets together with
the operations that are evaluated. The input and test data were
generated with respect to parameters $\sigma$ and $n$, as presented in
\hl{x} %MDPI: Please confirm if this can be removed.
 Table~\ref{t:benchmark}.

\begin{table}[H]
\caption{\hl{Configuration} %MDPI: We removed the vertical lines. Please confirm.
 of $\sigma$ and $n$ in benchmark.}
\label{t:benchmark}
\newcolumntype{C}{>{\centering\arraybackslash}X}
\begin{tabularx}{\textwidth}{CC}
\toprule
\boldmath$\sigma$ & \boldmath$n$ \\
\midrule
5		& 1\\
%\hline
30	& 1 \\
%\hline
5		& 3 \\
%\hline
15	& 3 \\
%\hline
30	& 3 \\
%\hline
10	& 10 \\
\bottomrule
\end{tabularx}
\end{table}

We tested all three types of query on all of the configurations from
Table \ref{t:benchmark}, resulting in 18 experiments in total, i.e., 6 experiments per
query type. In each experiment, we measured the average time consumed
by the data structure to process the query. The results of the exact
search, sub-multiset, and super-multiset search experiments are
presented in Tables~\ref{t:res_ex},~\ref{t:res_sub}, and~\ref{t:res_sup}, respectively.

\begin{table}[H]
\caption{Exact search.}
\label{t:res_ex}
\newcolumntype{r}{>{\raggedleft\arraybackslash}X}
\begin{tabularx}{\textwidth}{rrrr}
\toprule
\multicolumn{1}{c}{\boldmath$\sigma$} & 
\multicolumn{1}{c}{\boldmath$n$} & 
\multicolumn{1}{c}{\textbf{Multiset-Trie (\boldmath$\mu s$)}} & 
\multicolumn{1}{c}{\textbf{Inverted Index (\boldmath$\mu s$)}} \\
\midrule
5		& 1 & 3.45 & 17,782.35\\
%\hline
30	& 1 & 4.18 & 24,865.93\\
%\hline
5		& 3 & 2.20 & 1508.81\\
%\hline
15	& 3 & 4.39 & 2146.36\\
%\hline
30	& 3 & 10.67 & 3639.97\\
%\hline
10	& 10 & 6.93 & 384.05\\
\bottomrule
\end{tabularx}
\end{table}

\begin{table}[H]
\caption{Sub-multiset search.}
\label{t:res_sub}
\newcolumntype{r}{>{\raggedleft\arraybackslash}X}
\begin{tabularx}{\textwidth}{rrrr}
\toprule
\multicolumn{1}{c}{\boldmath$\sigma$} & 
\multicolumn{1}{c}{\boldmath$n$} & 
\multicolumn{1}{c}{\textbf{Multiset-Trie (\boldmath$\mu s$)}} & 
\multicolumn{1}{c}{\textbf{Inverted Index (\boldmath$\mu s$)}} \\
\midrule
5		& 1			& 8.96 & 73,500.84\\
%\hline
30	& 1			& 17.33 & 547,572.74\\
%\hline
5		& 3			& 117.95 & 162,360.43\\
%\hline
15	& 3			& 20.74 & 443,321.39\\
%\hline
30	& 3 		& 23.75 & 947,706.14\\
%\hline
10	& 10		& 55.59 & 466,022.68\\
\bottomrule
\end{tabularx}
\end{table}

\begin{table}[H]
\caption{Super-multiset search.}
\label{t:res_sup}
\newcolumntype{r}{>{\raggedleft\arraybackslash}X}
\begin{tabularx}{\textwidth}{rrrr}
\toprule
\multicolumn{1}{c}{\boldmath$\sigma$} & 
\multicolumn{1}{c}{\boldmath$n$} & 
\multicolumn{1}{c}{\textbf{Multiset-Trie (\boldmath$\mu s$)}} & 
\multicolumn{1}{c}{\textbf{Inverted Index (\boldmath$\mu s$)}} \\
\midrule
5		& 1 		& 10.63 & 63,073.86\\
%\hline
30	& 1 		& 14.65 & 449,251.68\\
%\hline
5		& 3 		& 171.04 & 163,256.77\\
%\hline
15	& 3 		& 43.42 & 425,733.80\\
%\hline
30	& 3 		& 22.06 & 729,831.34\\
%\hline
10	& 10 	& 58.32 & 373,784.81\\
\bottomrule
\end{tabularx}
\end{table}

We can see that multiset-trie outperforms the Inverted Index in all of the
experiments by up to four orders of magnitude. In an exact search,
multiset-trie has to traverse only up to $\sigma+1$ nodes to obtain a query
result. It can be seen from the results that with the increase in $\sigma$,
the processing time for multiset-tire also increases. Multiplicity
also affects the processing time; however, this happens
passively. Multiplicity, or the degree of a node $n$, defines the shape of
multisets that are stored in multiset-trie. Thus, it affects the
structure and density of the multiset-trie.

As for the Inverted Index, all three operations must first fetch all
postings for each particular element of a test multiset. Afterward,
the intersection of postings is computed to answer the query. The
operations use more processing time than a simple tree traversal,
which we can see from the results. Postings are filtered on-the-fly to
reduce the cost of the intersection.

Implementations of the multiset containment operations
for searching sub-multisets and super-multisets are similar in
the case of the inverted file. The algorithm consists of the same
steps. First, the postings are fetched for each element of the test
multiset. Depending on the particular operation, postings are filtered
on-the-fly. Finally, the union or the intersection of the filtered set
of postings is computed. Note that the processing time increases with
the size of the inverted index because of the increased sizes of
postings.

In the case of multiset-trie, only a traversal of the tree is
required, which is much faster than the processing of postings, as we
can see from the results. In the worst case, the whole tree is
traversed, but the same is for an inverted file.

% section 6
%\input{tex/s_relwork}
\section{Related Work} \label{c:relwork}

% intro to related work
The data structure multiset-trie is related to the data structures and indexes designed to store and manage sets and multisets. We mainly focus on the related data structures and indexes that efficiently support the set and multiset containment queries. Firstly, we summarize our previous work on the data structure for managing sets in Section \ref{rel-strie}. Next, we present in Section \ref{rel-invfile} the related work on the inverted files, i.e., the index structure that serves as a central data structure in the area of Information Retrieval (IR), but also for storing sets and multisets in database management systems. The alternative to the inverted file is the signature tree that is presented in Section \ref{rel-signature}. Finally, we describe the related work in the area of database management systems in Section \ref{rel-dbms}. We review the novel index structures used for the containment queries and the proposed containment join algorithms.

% Construction of multiset-trie

\subsection{Set-Trie\label{rel-strie}}
% set-trie: 
% - data structure that multiset-trie generalizes
% - stores a set of sets
% - supports set containment operations

The multiset-trie is closely related to the set-trie data structure introduced by Savnik in~\cite{savnik2013index,savnik2021plos}. A set-trie is a trie data structure that is adapted for the efficient storage and retrieval of sets instead of the sequences of symbols. The set-trie provides the set containment operations, such as the retrieval of the \emph{nearest} subset or supersets, as well as the retrieval of \emph{all} subsets and supersets from the sets of sets.

Since we are storing sets where each element of the set can appear only once, and the ordering of elements is not important, the ordering of the elements from the alphabet can be used for guiding the search in set containment operations. Each set is represented in a set-trie by a path including the increasing elements of a set represented by set-trie nodes. Since all sets from a set-trie are ordered by the increasing value of the set elements, the children of each set-trie node $n$ can only be elements larger than the element $n$. For a given set $s$ and a set-trie $S$, the set containment operations search solely the subtree of $S$ that includes all the sets (paths from a root to a set-trie node) that are the possible subsets or supersets of $s$.
%The size of such subtree depends on the size of $S$ and the shape of $s$.

The data structure multiset-trie generalizes the set-trie by providing storage for the set of multisets. When the multiset-trie is restricted to store a set of sets, the underlying data structure becomes a simple binary tree. Moreover, all the operations of the set-trie are also supported by the multiset-trie. The generalization comes with a small penalty in performance if we compare the multiset-trie with the set-trie in the performance of the set containment operations. The downside of such a generalization is that multiset-trie no longer supports the path compression that was obtained in set-trie. 
%However, the design of multiset-trie provides storage of multisets with constant worst-case time complexity of the set containment operations.

\subsection{Inverted File\label{rel-invfile}}
% inverted index:
% - performance study of set containment queries (sequential signature files, signature trees, extendible signature hashing, inverted files)
% - performance comparison of multiset-trie to inverted index
% - containment query issues of inverted index

The inverted file \cite{zobel1992efficient,zobel1998inverted,zobel2006inverted} is the most common data structure used to represent a collection of (multi)sets. In the area of IR \cite{manning2008introduction}, the inverted files are used for searching documents that contain a given set of words. It is composed of two parts: a dictionary and the postings. The dictionary maps each word to a list of document identifiers together with the locations of words in documents. The dictionary is most often implemented by a variant of a search tree, such as a B+ tree. The postings are implemented as a list of positions that are stored on the disk because of the huge amount of documents usually indexed by the inverted file. Since we can have a large number of postings for one word, the postings are compressed. Furthermore, several possible optimizations exist in the representation and implementation of postings \cite{zobel2006inverted}, such as the sorting of postings, a technique called skipping, and others.

The empirical analyses~\cite{Helmer2003,zobel1998inverted} show that the inverted file is the most efficient data structure for containment queries among the data structures: the sequential signature file, the signature tree, the extendible signature hashing, and the inverted file.

% Application areas of multiset-trie

\subsection{Signature Trees\label{rel-signature}}

A dynamically balanced signature tree \cite{Deppisch1986,Pfaltz1980}, or S-tree, is an alternative data structure for the representation of multisets. An S-tree stores objects on the basis of their attributes represented in the form of signatures. A signature of an object is formed by the discretization of object attributes. Each attribute is discretized by mapping the attribute values to a sequence of bits. The bit sequences are the abstractions of the values of object attributes. They are glued together to form the signature of an object. The mappings from attribute values to sequences of bits are defined in such a way that allows superimposing a set of signatures by a single signature. Such a signature is often formed by using the operation OR. This property of signatures provides the means for the construction of the hierarchy of signatures that is utilized for efficient search. The multisets can be effectively represented using signatures, and the superposition operation can be implemented by the operation OR. The use of the signature tree for the containment operations was studied by Tousidou et~al.~\cite{tousidou2002sigstruc}. They show that an S-tree that uses linear hash partitioning can be used to implement the containment operations efficiently.

\subsection{Multisets in Relational Databases\label{rel-dbms}}
% multisets in RDBMS and ORDBMS:
% - set-valued attributes
% - multiset-valued attributes
% - underlying datastructures for storage of set- and multiset-valued attributes

The index structures for the efficient implementation of the (multi)set containment queries were studied in the framework of the relational DBMS as well as the object-relational DBMS, where we can use multivalued attributes, including sets, multisets (bags), and lists. Zhang et al.\ \cite{Zhang2001} compared the performance of the containment queries implemented in a standard relational DBMS (RDBMS) to an information retrieval engine. The results show that, in general, the IR engine performs better than an RDBMS on containment queries. They identified the problems that lead to the poor performance of the containment operations in an RDBMS and showed that, with some modifications, an RDBMS could perform this class of queries more efficiently.

The joins in an object-relational DBMS can be defined by means of the containment operations. A number of \emph{containment join} methods have been proposed \cite{Ramasamy2000,Melnik2003,Jampani2005,Luo2015}. Ramasamy et al. propose the use of the partitioning set join that relies on the representation of sets by using signatures \cite{Ramasamy2000}. The signature-based representation allows the efficient implementation of the set comparison operations. The partitioning set join was further improved by Melnik et al. \cite{Melnik2003} to handle large sets and to speed up the partitioning phase of the algorithm. Further, Jampani et al. introduced the PRETTI join algorithm, which combines an inverted file with a prefix tree for the efficient implementation of the containment joins. The algorithm for $R\bowtie T$ recursively computes record identifiers from $T$ while traversing a prefix tree storing the sets from $R$. The algorithm uses a single intersection of two lists to enumerate the matching pairs of rid-s. The PRETTI algorithm was improved by Luo at al. \cite{Luo2015} by replacing the prefix tree with the Patricia tree.

% section 7
%\input{tex/s_conclusions}
\section{Conclusions and Future Work} \label{c:conclusions}

%
% Experiments conclusions
One of the conclusions of studying the multiset-trie both theoretically and empirically is that our data structure is input-sensitive. Input sensitivity implies non-consistent performance on different input data. However, our argument that the performance can be optimized by pre-processing the input data is confirmed in  Experiment~\hyperref[ss:exp3]{4}. Pre-processing determines the optimal encoding for input data and ensures the best performance of the multiset-trie on particular input data. For example, in the case of storing words in the multiset-trie, the search queries can always be optimized based on the frequencies of letters in a specific language. We also see from Experiments~\hyperref[s:exp1]{1} and~\hyperref[s:exp2]{2} that the dependence of the multiset-trie's performance on the density is not a linear function. However, the function is continuous, and the point of inflection is unique on the whole domain, as shown in Experiment~\hyperref[s:exp3]{3}. This allows us to predict whether multiset-trie can be used for some particular application, offering high performance.

%
% Mathematical analysis conclusions
The mathematical analysis section provides a non-trivial insight regarding the behavior of the multiset-trie data structure when used in randomized data. It is estimated that the space complexity of multiset-trie is of order $O(|M|)$, which is the minimal possible space required by any data structure for the storage of $|M|$ objects. 
As for the running time complexity of algorithms, the basic tree functions such as \textsc{insert}, \textsc{search}, and \textsc{delete} all have constant complexity once the multiset-trie is defined. The ``getAll'' multiset containment functions have the worst-case running time complexity of $O(|\mathcal{M}|),$ where $|\mathcal{M}|$ is the cardinality of the multiset-trie data structure. The ``existence'' multiset containment functions have the worst-case running time complexity of $O(|\mathcal{M}| - |M|),$ where $|\mathcal{M}|$ is the cardinality of the multiset-trie and $|M|$ is the number of inserted multisets (nodes on leaf level).

%The multiset-trie is an input-sensitive data structure because the size of multiset-trie $|\mathcal{M}|$ depends on the distribution of multisets in $M.$
%Our mathematical model assumes that multisets $m$ in $M$ are distributed uniformly;  however, in real-world models, such an assumption is not generally true. 
%%Specifically, the probability $P(m\in M)$ may vary dramatically and can even be equal to $0.$ For example, if words are mapped to multisets, then the sample space contains very large multisets. 
%For example, many multisets will have zero probability of appearing in $M$ because a word that would correspond to such a multiset does not exist in the dictionary.%

The implementation of the multiset-trie is not optimized. For example, the main reason for the space inefficiency is in implementing the links from a node to its children. Our implementation uses an array data structure to link a node to its children, where each element of an array, indexed by the multiplicity of the element of the next level, includes a link to a subtree. A custom-implemented small and extendable hash table would significantly decrease the amount of space needed to represent a multiset-trie.
% There are more additional optimizations possible. For example, the ...

% Future work notes
%
%The above results have opened even more interesting questions for future research. 
Further steps in our research will be to extend the functionality of the multiset-trie. We are interested in more flexible multiset containment queries where additional conditions constrain the sub- and super-multisets. For example, the multiplicity of an element in a multiset can be bounded in operations getAllSubmsets and getAllSupermsets. Furthermore, the similarity search on multisets can be implemented by modifying the algorithms for searching the sub- and super-multisets. 
%
The second line of research is to investigate the multiset-trie as a database index data structure. A disk-based index data structure allows for storing and managing a huge amount of multisets. The mapping from a multiset-trie, i.e., a $n$-ary search tree, to a block-based index can be easily defined because of the regularity of multiset-trie. It will be interesting to compare the multiset-trie with other existing disk-based index data structures.

\vspace{6pt}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
%\supplementary{The following supporting information can be downloaded at:  \linksupplementary{s1}, Figure S1: title; Table S1: title; Video S1: title.}

% Only for the journal Methods and Protocols:
% If you wish to submit a video article, please do so with any other supplementary material.
% \supplementary{The following supporting information can be downloaded at: \linksupplementary{s1}, Figure S1: title; Table S1: title; Video S1: title. A supporting video article is available at doi: link.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontributions{\hl{ .}}%For research articles with several authors, a short paragraph specifying their individual contributions must be provided. The following statements should be used ``Conceptualization, X.X. and Y.Y.; methodology, X.X.; software, X.X.; validation, X.X., Y.Y. and Z.Z.; formal analysis, X.X.; investigation, X.X.; resources, X.X.; data curation, X.X.; writing---original draft preparation, X.X.; writing---review and editing, X.X.; visualization, X.X.; supervision, X.X.; project administration, X.X.; funding acquisition, Y.Y. All authors have read and agreed to the published version of the manuscript.'', please turn to the  \href{http://img.mdpi.org/data/contributor-role-instruction.pdf}{CRediT taxonomy} for the term explanation. Authorship must be limited to those who have contributed substantially to the work~reported.}

\funding{\hl{The authors} %MDPI: We moved the last paragraph here, please confirm.
 acknowledge the support in part of the Slovenian Research Agency, research program P1-0383.}

\institutionalreview{\hl{ .}}%In this section, you should add the Institutional Review Board Statement and approval number, if relevant to your study. You might choose to exclude this statement if the study did not require ethical approval. Please note that the Editorial Office might ask you for further information. Please add “The study was conducted in accordance with the Declaration of Helsinki, and approved by the Institutional Review Board (or Ethics Committee) of NAME OF INSTITUTE (protocol code XXX and date of approval).” for studies involving humans. OR “The animal study protocol was approved by the Institutional Review Board (or Ethics Committee) of NAME OF INSTITUTE (protocol code XXX and date of approval).” for studies involving animals. OR “Ethical review and approval were waived for this study due to REASON (please provide a detailed justification).” OR “Not applicable” for studies not involving humans or animals.}

\informedconsent{\hl{ .}}%Any research article describing a study involving humans should contain this statement. Please add ``Informed consent was obtained from all subjects involved in the study.'' OR ``Patient consent was waived due to REASON (please provide a detailed justification).'' OR ``Not applicable'' for studies not involving humans. You might also choose to exclude this statement if the study did not involve humans. Written informed consent for publication must be obtained from participating patients who can be identified (including by the patients themselves). Please state ``Written informed consent has been obtained from the patient(s) to publish this paper'' if applicable.}

\dataavailability{\hl{ .}}%We encourage all authors of articles published in MDPI journals to share their research data. In this section, please provide details regarding where data supporting reported results can be found, including links to publicly archived datasets analyzed or generated during the study. Where no new data were created, or where data is unavailable due to privacy or ethical re-strictions, a statement is still required. Suggested Data Availability Statements are available in section “MDPI Research Data Policies” at \url{https://www.mdpi.com/ethics}.}

%\acknowledgments{}

\conflictsofinterest{\hl{The} %MDPI: Newly added information. Please confirm.
 authors declare no conflicts of interest.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\paragraph{\textbf{Funding information.}} Authors acknowledge the 
%support in part by the Slovenian Research Agency, research program P1-0383.

%\nolinenumbers

%\newpage
\begin{adjustwidth}{-\extralength}{0cm}
\reftitle{References}
\begin{thebibliography}{999}

\bibitem[Savnik \em{et~al.}(2021)Savnik, Akulich, Krnc, and
  Škrekovski]{savnik2021plos}
Savnik, I.; Akulich, M.; Krnc, M.; Škrekovski, R.
\newblock Data structure set-trie for storing and querying sets: Theoretical
  and empirical analysis.
\newblock {\em PLoS ONE} {\bf 2021}, {\em 2}, e0245122. https://doi.org/10.1371/journal.pone.0245122.

\bibitem[Bouros \em{et~al.}(2016)Bouros, Mamoulis, Ge, and
  Terrovitis]{bouros2016set}
Bouros, P.; Mamoulis, N.; Ge, S.; Terrovitis, M.
\newblock Set containment join revisited.
\newblock {\em Knowl. Inf. Syst.} {\bf 2016}, {\em
  49},~375--402.

\bibitem[Gripon \em{et~al.}(2012)Gripon, Rabbat, Skachek, and
  Gross]{gripon2012compressing}
Gripon, V.; Rabbat, M.; Skachek, V.; Gross, W.J.
\newblock Compressing multisets using tries.
\newblock In Proceedings of the 2012 IEEE Information Theory Workshop, \hl{Lausanne, Switzerland, 3--7 September} %MDPI: We added the location and date of the conference. Please confirm.
  2012; pp. 642--646.

\bibitem[Ross and Stoyanovich(2004)]{ross2004symmetric}
Ross, K.A.; Stoyanovich, J.
\newblock Symmetric relations and cardinality-bounded multisets in database
  systems.
\newblock In Proceedings of the Thirtieth International
  Conference on Very Large Data Bases-Volume 30. VLDB \hl{Endowment,} %MDPI: Please add the location and date of the conference. 
  2004; pp.
  912--923.

\bibitem[Steinruecken(2015)]{steinruecken2015compressing}
Steinruecken, C.
\newblock Compressing sets and multisets of sequences.
\newblock {\em IEEE Trans. Inf. Theory} {\bf 2015}, {\em
  61},~1485--1490.

\bibitem[Zobel \em{et~al.}(1992)Zobel, Moffat, and
  Sacks-Davis]{zobel1992efficient}
Zobel, J.; Moffat, A.; Sacks-Davis, R.
\newblock An efficient indexing technique for full-text database systems.
\newblock In Proceedings of the VLDB '92: Proceedings of the 18th International Conference on Very Large Data Bases, \hl{IEEE,} %MDPI: Please add the location and date of the conference. 
 23--27 August 1992; pp. 352--352.

\bibitem[Zobel and Moffat(2006)]{zobel2006inverted}
\hl{Zobel, J.; Moffat, A.}  %MDPI: Refs. 7 and 17 are duplicated. Please remove duplicated references and rearrange all the references to appear in numerical order. Please ensure that there are no duplicated references.
\newblock Inverted files for text search engines.
\newblock {\em ACM Comput. Surv. (CSUR)} {\bf 2006}, {\em 38},~6.

\bibitem[Manning \em{et~al.}(2008)Manning, Raghavan, Sch{\"u}tze,
  et~al.]{manning2008introduction}
Manning, C.D.; Raghavan, P.; Sch{\"u}tze, H.;  \hl{et~al.} %MDPI: Please include the first ten authors' names before using "et al." in the references.
\newblock {\em Introduction to Information Retrieval}; Cambridge
  University Press: Cambridge, UK, 2008; Volume~1.

\bibitem[Mannila and Toivonen(1997)]{mannila1997}
Mannila, H.; Toivonen, H.
\newblock Levelwise Search and Borders of Theories in Knowledge Discovery.
\newblock {\em Data Min. Knowl. Discov.} {\bf 1997}, {\em
  1},~241--258.

\bibitem[Flach and Savnik(1999)]{flach1999aicom}
Flach, P.A.; Savnik, I.
\newblock Database Dependency Discovery: A Machine Learning Approach.
\newblock {\em AI Commun.} {\bf 1999}, {\em 12},~139--160.

\bibitem[Forgy(1982)]{forgy1982}
Forgy, C.
\newblock Rete: A Fast Algorithm for the Many Pattern/Many Object Pattern Match
  Problem.
\newblock {\em Artif. Intell.} {\bf 1982}, {\em 19},~17--37.

\bibitem[Bayardo \em{et~al.}(2007)Bayardo, Ma, and Srikant]{bayardo2007simlar}
Bayardo, R.J.; Ma, Y.; Srikant, R.
\newblock Scaling up All Pairs Similarity Search.
\newblock In WWW'07: 16th International World Wide Web Conference, \hl{Banff, AB, Canada, 8--12 May} %MDPI: We added the location and date of the conference. Please confirm.
 2007;  Association for Computing Machinery: New York,
  NY, USA,  2007; pp. 131–140.

\bibitem[Xiao \em{et~al.}(2011)Xiao, Wang, Lin, Yu, and Wang]{xiao2011tods}
Xiao, C.; Wang, W.; Lin, X.; Yu, J.X.; Wang, G.
\newblock Efficient Similarity Joins for Near-Duplicate Detection.
\newblock {\em ACM Trans. Database Syst.} {\bf 2011}, {\em 36}, 15.

\bibitem[Wang \em{et~al.}(2017)Wang, Qin, Lin, Zhang, and Chang]{wang2017vldb}
Wang, X.; Qin, L.; Lin, X.; Zhang, Y.; Chang, L.
\newblock Leveraging set relations in exact set similarity join.
\newblock {\em Proc. VLDB Endow.} {\bf 2017}, {\em
  10},~925--936.

\bibitem[H.~Cormen \em{et~al.}(2001)H.~Cormen, Leiserson, L.~Rivest, and
  Stein]{corman2001}
H.~Cormen, T.; Leiserson, C.; L.~Rivest, R.; Stein, C.
\newblock {\em Introduction to Algorithms, Second Edition}; \hl{The MIT Press: Cambridge, MA, USA,} %MDPI: We added the name of the publisher and location. Please confirm.
 2001.

\bibitem[Zobel \em{et~al.}(1998)Zobel, Moffat, and
  Ramamohanarao]{zobel1998inverted}
Zobel, J.; Moffat, A.; Ramamohanarao, K.
\newblock Inverted files versus signature files for text indexing.
\newblock {\em ACM Trans. Database Syst. (TODS)} {\bf 1998}, {\em
  23},~453--490.

\bibitem[Zobel and Moffat(2006)]{zobel2006csur}
\hl{Zobel, J.; Moffat, A.}
\newblock Inverted files for text search engines.
\newblock {\em ACM Comput. Surv. (CSUR)} {\bf 2006}, {\em 38},~6.

\bibitem[Broder \em{et~al.}(2006)Broder, Eiron, Fontoura, Herscovici, Lempel,
  McPherson, Qi, and Shekita]{broder2006indexing}
Broder, A.Z.; Eiron, N.; Fontoura, M.; Herscovici, M.; Lempel, R.; McPherson,
  J.; Qi, R.; Shekita, E.
\newblock Indexing shared content in information retrieval systems.
\newblock In \emph{Proceedings of the International Conference on Extending Database
  Technology}; Springer:  \hl{Berlin/Heidelberg, Germany,} %We added the location of publisher. Please confirm
  2006; pp. 313--330.

\bibitem[Terrovitis \em{et~al.}(2006)Terrovitis, Passas, Vassiliadis, and
  Sellis]{terrovitis2006cikm}
Terrovitis, M.; Passas, S.; Vassiliadis, P.; Sellis, T.
\newblock A Combination of Trie-trees and Inverted Files for the Indexing of
  Set-valued Attributes.
\newblock In \emph{Proceedings of the Proceedings of the 15th ACM International
  Conference on Information and Knowledge Management}; ACM: New York, NY, USA,
  2006; pp. 728--737.

\bibitem[Terrovitis \em{et~al.}(2011)Terrovitis, Bouros, Vassiliadis, Sellis,
  and Mamoulis]{terrovitis2011icdt}
Terrovitis, M.; Bouros, P.; Vassiliadis, P.; Sellis, T.; Mamoulis, N.
\newblock Efficient Answering of Set Containment Queries for Skewed Item
  Distributions.
\newblock In \emph{Proceedings of the Proceedings of the 14th International
  Conference on Extending Database Technology}; ACM: New York, NY, USA,  2011;
   pp. 225--236.

\bibitem[Deppisch(1986)]{deppisch1986sigir}
\hl{Deppisch, U.} %MDPI: Refs. 21 and 29 are duplicated. Please remove duplicated references and rearrange all the references to appear in numerical order. Please ensure that there are no duplicated references.
\newblock S-tree: A Dynamic Balanced Signature Index for Office Retrieval.
\newblock In \emph{Proceedings of the 9th Annual International ACM
  SIGIR Conference on Research and Development in Information Retrieval}; ACM:
  New York, NY, USA,  1986;  pp. 77--87.

\bibitem[Tousidou \em{et~al.}(2002)Tousidou, Bozanis, and
  Manolopoulos]{tousidou2002sigstruc}
Tousidou, E.; Bozanis, P.; Manolopoulos, Y.
\newblock Signature-based Structures for Objects with Set-valued Attributes.
\newblock {\em Inf. Syst.} {\bf 2002}, {\em 27},~93--121.
\newblock {{https://doi.org/10.1016/S0306-4379(01)00047-3}}.

\bibitem[Chen and Shi(2005)]{yangjun2005stree}
Chen, Y.; Shi, Y.
\newblock Signature Files and Signature File Construction.
\newblock In {\em Encyclopedia of Database Technologies and Applications}; \hl{IGI Global: } %MDPI: Please add the location of the publisher (City and Country).
 {2005}.
\newblock {{https://doi.org/10.4018/978-1-59140-560-3.ch105}}.

\bibitem[Helmer and Moerkotte(2003)]{Helmer2003}
Helmer, S.; Moerkotte, G.
\newblock A performance study of four index structures for set-valued
  attributes of low cardinality.
\newblock {\em  VLDB J.} {\bf 2003}, {\em 12},~244--261.
\newblock {{https://doi.org/10.1007/s00778-003-0106-0}}.

\bibitem[Savnik(2013)]{savnik2013index}
Savnik, I.
\newblock Index data structure for fast subset and superset queries.
\newblock In \emph{Proceedings of the International Conference on Availability,
  Reliability, and Security}; Springer:  \hl{Berlin/Heidelberg, Germany,} %We added the location of publisher. Please confirm
  2013; pp. 134--148.

\bibitem[Sedgewick and Wayne(2011)]{Sedgewick:2011:ALG:2011916}
Sedgewick, R.; Wayne, K.
\newblock {\em Algorithms}, 4th ed.; \hl{Addison-Wesley} %MDPI: Please add the location of the publisher (City and Country).
 Professional:  2011.

\bibitem[Gardiner(1985)]{gardiner1985stochastic}
Gardiner, C.W.
\newblock {\em Stochastic Methods}; Springer: Berlin/Heidelberg, Germany, New
  York, NY, USA, Tokyo, Japan, 1985.

\bibitem[Akulich(2019)]{akulich2019mstrie}
Akulich, M.
\newblock Mstrie Repository. 2019. Available online:  \url{https://github.com/nick-ak96/mstrie} (\hl{accessed} %MDPI: Please add the access date (format: Date Month Year), e.g., accessed on 1 January 2020.
 on).

\bibitem[Deppisch(1986)]{Deppisch1986}
\hl{Deppisch, U.} %MDPI: Refs. 21 and 29 are duplicated. Please remove duplicated references and rearrange all the references to appear in numerical order. Please ensure that there are no duplicated references.
\newblock S-tree: A Dynamic Balanced Signature Index for Office Retrieval.
\newblock In \emph{Proceedings of the 9th Annual International ACM
  SIGIR Conference on Research and Development in Information Retrieval}; ACM:
  New York, NY, USA,  1986;  pp. 77--87.
\newblock {{https://doi.org/10.1145/253168.253189}}.

\bibitem[Pfaltz \em{et~al.}(1980)Pfaltz, Berman, and Cagley]{Pfaltz1980}
Pfaltz, J.L.; Berman, W.J.; Cagley, E.M.
\newblock Partial-match Retrieval Using Indexed Descriptor Files.
\newblock {\em Commun. ACM} {\bf 1980}, {\em 23},~522--528.
\newblock {{https://doi.org/10.1145/359007.359013}}.

\bibitem[Zhang \em{et~al.}(2001)Zhang, Naughton, DeWitt, Luo, Luo, and
  Lohman]{Zhang2001}
Zhang, C.; Naughton, J.; DeWitt, D.; Luo, Q.; Luo, Q.; Lohman, G.
\newblock On Supporting Containment Queries in Relational Database Management
  Systems.
\newblock In \emph{Proceedings of the  SIGMOD}; ACM: New York, NY, USA,
  2001; pp. 425--436.
\newblock {{https://doi.org/10.1145/375663.375722}}.

\bibitem[Ramasamy \em{et~al.}(2000)Ramasamy, Patel, Naughton, and
  Kaushik]{Ramasamy2000}
Ramasamy, K.; Patel, J.M.; Naughton, J.F.; Kaushik, R.
\newblock Set containment joins: The good, the bad and the ugly.
\newblock In Proceedings of the \hl{Proc. of VLDB,} %MDPI: Please provide more information about the article type, such as book (please provide the name and location of the publisher); online resource (please provide the URL of the website and the date it was accessed (Date Month Year)); or journal article (please provide the name of the journal, the year and volume in which it was published, and the page number). Please refer to https://www.mdpi.com/authors/references for full reference formatting guides.
  2000.

\bibitem[Melnik and Garcia-Molina(2003)]{Melnik2003}
Melnik, S.; Garcia-Molina, H.
\newblock Adaptive Algorithms for Set Containment Joins.
\newblock {\em ACM Trans. Database Syst.} {\bf 2003}, {\em 28},~56--99.
\newblock {{https://doi.org/10.1145/762471.762474}}.

\bibitem[Jampani and Pudi(2005)]{Jampani2005}
Jampani, R.; Pudi, V.
\newblock Using Prefix-Trees for Efficiently Computing Set Joins.
\newblock In Proceedings of the Database Systems for Advanced Applications,
  10th International Conference, {DASFAA} 2005, Beijing, China,  17--20 April
  2005; pp. 761--772.
\newblock {{https://doi.org/10.1007/11408079\_69}}.

\bibitem[Luo \em{et~al.}(2015)Luo, Fletcher, Hidders, and Bra]{Luo2015}
Luo, Y.; Fletcher, G.H.L.; Hidders, J.; Bra, P.D.
\newblock Efficient and scalable trie-based algorithms for computing set
  containment relations.
\newblock In Proceedings of the 31st {IEEE} International Conference on Data
  Engineering, {ICDE} 2015, Seoul, Republic of Korea,  13--17 April 2015; pp.
  303--314.
\newblock {{https://doi.org/10.1109/ICDE.2015.7113293}}.

\end{thebibliography}

\PublishersNote{}
%%\input{bibliography.bbl}
%\bibliographystyle{Definitions/MDPI.bst}
%\bibliography{multiset-trie.bib}
\end{adjustwidth}
\end{document}

### 2023-03-15 tex/algorithms-2122309-proof.tex
a few updates...
updates for the %MDPI comments. numbers remain to be corrected.

@@
 % Abstract (Do not insert blank lines, i.e., \\) 
-\abstract{This paper proposes a new data structure, \highlighting{\emph{multiset-trie}}, %MDPI: Please confirm if the italics is unnecessary and can be removed. Please check all italic.
+\abstract{This paper proposes a new data structure, multiset-trie, %MDPI: Please confirm if the italics is unnecessary and can be removed. Please check all italic. %RES Italics removed.
  that is designed for storing and efficiently processing a set of multisets. Moreover, multiset-trie can operate on a set of sets without efficiency loss. 
@@
 \end{center}
-%
-\hl{The} %MDPI: Whether the indentation of the paragraph after the formula is necessary, please confirm.
+
+The %MDPI: Whether the indentation of the paragraph after the formula is necessary, please confirm. %RES Indentation added.
  operations supported by the multiset-trie data structure are as
@@
 
-\includegraphics[width=.64\textwidth, keepaspectratio]{ratio-exp-nodes-and-msets-on-prob1.pdf}
+\includegraphics[width=.64\textwidth, keepaspectratio]{ratio-exp-nodes-and-msets-on-prob.pdf}
 \caption{The ratio of ``data overhead'' derived from Figure \ref{f:exp-nodes}, as the density $p$ of our data structure increases.
@@
 \textsc{supermsetExistence}, \textsc{getAllSubmsets}, and \textsc{getAllSupermsets}. 
+We refer to the former two operations as the \emph{existence operations}, and the
+later two as the \emph{exhaustive operations}. %RES A sentence added to introduce the terms existence and exhaustive operations used later.
 
@@
 
-\subsection{\hl{Data} %MDPI: Section headings should be numbered sequentially, e.g., Section 2.1, Section 2.2.1. Please confirm this revision.
+\subsection{Data%MDPI: Section headings should be numbered sequentially, e.g., Section 2.1, Section 2.2.1. Please confirm this revision. %RES It is Ok as it is.
  Generation}
@@
 average, the results will be the same for any $\phi,$ since all the multisets are 
-equally likely to appear in $M.$ The parameter $|M|$ varies from \hl{10,000} %MDPI: We added commas to separate thousands for numbers with five or more digits. Please confirm.
+equally likely to appear in $M.$ The parameter $|M|$ varies from 10,000 %MDPI: We added commas to separate thousands for numbers with five or more digits. Please confirm. %RES We confirm.
  sets up 
-to \hl{320,000} sets. According to the parameters $n$ and $\sigma,$ the cardinality of 
+to 320,000 sets. According to the parameters $n$ and $\sigma,$ the cardinality of 
 $N$ is $33,554,432\approx \num{3.36e+7}.$ Thus, the calculated density of the 
@@
 	\caption{\hl{Existence} %MDPI: Please use commas to separate thousands for numbers with five or more digits (not four digits) in the picture, e.g., "10000" should be "10,000".
- functions of Experiment 1. (\textbf{a}) \hl{submsetExistence;} %MDPI: We moved the subfigure explanations into the figure caption. Please confirm. The following are same.
- (\textbf{b}) \hl{supermsetExistence}.  \label{fig:4}}
+ functions of Experiment 1. (\textbf{a}) submsetExistence; %MDPI: We moved the subfigure explanations into the figure caption. Please confirm. The following are same. %RES We confirm.
+ (\textbf{b}) supermsetExistence.  \label{fig:4}}
 \end{figure}
@@
 data are generated artificially with uniform distribution. The cardinality of $M$ 
-varies from \hl{40,000} to \hl{640,000} multisets. Thus, the calculated density $d$ varies 
+varies from 40,000 to 640,000 multisets. Thus, the calculated density $d$ varies 
 from $\num{1.4e-15}$ to $\num{2.25e-14}.$ The density is much smaller than 
@@
 \label{fig:e3m2}}{\includegraphics[width=.485\textwidth]{exp4-m2.pdf}}
-\caption{Existence functions in Experiment 3. (\textbf{a}) \hl{submsetExistence}; (\textbf{b}) supermsetExistence.\label{fig:e3m1}}
+\caption{Existence functions in Experiment 3. (\textbf{a}) submsetExistence; (\textbf{b}) supermsetExistence.\label{fig:e3m1}}
 \end{figure}
@@
 
-The data are taken from the English dictionary, which contains \hl{235,883} different words. 
+The data are taken from the English dictionary, which contains 235,883 different words. 
 These words are mapped to multisets of integers according to the $\phi.$ In 
@@
 generated with respect to parameters $\sigma$ and $n$, as presented in
-\hl{x} %MDPI: Please confirm if this can be removed.
+%MDPI: Please confirm if this can be removed. %RES Yes, we removed it.
  Table~\ref{t:benchmark}.
@@
 \begin{table}[H]
-\caption{\hl{Configuration} %MDPI: We removed the vertical lines. Please confirm.
+\caption{Configuration %MDPI: We removed the vertical lines. Please confirm. %RES We confirm.
  of $\sigma$ and $n$ in benchmark.}

### 2023-03-15 tex/algorithms-2122309-proof.tex
a sentence in abstract updated.
references updated on comments %MDPI.

@@
 The multiset-trie structure is a search tree with properties similar to those of a trie.
-It implements all standard search tree operations together with the multiset containment operations, such as sub-multiset and super-multiset.
+It implements all standard search tree operations together with the multiset containment operations for searching sub-multisets and super-multisets.
 Suppose that we have a set of multisets $S$ and a multiset $X$.
@@
 
-The existing solutions to the implementation of the containment queries use the inverted indexes \cite{zobel1992efficient,zobel1998inverted,zobel2006csur,broder2006indexing,terrovitis2006cikm,terrovitis2011icdt}, the signature trees \cite{deppisch1986sigir,tousidou2002sigstruc,yangjun2005stree,zobel1998inverted}, and the B+ trees~\cite{Helmer2003,terrovitis2011icdt}.
+The existing solutions to the implementation of the containment queries use the inverted indexes \cite{zobel1992efficient,zobel1998inverted,zobel2006inverted,broder2006indexing,terrovitis2006cikm,terrovitis2011icdt}, the signature trees \cite{deppisch1986sigir,tousidou2002sigstruc,yangjun2005stree,zobel1998inverted}, and the B+ trees~\cite{Helmer2003,terrovitis2011icdt}.
 These solutions provide a key--value look-up, where a key is an element of a multiset and a value is a corresponding multiset.
@@
 \begin{figure}[H]
-
-\subcaptionbox{\centering \hl{Experiment} %Please move sub-caption to Caption.
+\subcaptionbox{\centering \hl{Experiment} %MDPI Please move sub-caption to Caption.
 2, getAllSubmsets function.
@@
 
-A dynamically balanced signature tree \cite{Deppisch1986,Pfaltz1980}, or S-tree, is an alternative data structure for the representation of multisets.
+A dynamically balanced signature tree \cite{deppisch1986sigir,Pfaltz1980}, or S-tree, is an alternative data structure for the representation of multisets.
 An S-tree stores objects on the basis of their attributes represented in the form of signatures.
@@
 \newblock Compressing multisets using tries.
-\newblock In Proceedings of the 2012 IEEE Information Theory Workshop, \hl{Lausanne, Switzerland, 3--7 September} %MDPI: We added the location and date of the conference. Please confirm.
+\newblock In Proceedings of the 2012 IEEE Information Theory Workshop, Lausanne, Switzerland, 3--7 September %MDPI: We added the location and date of the conference. Please confirm. %RES We confirm.
   2012; pp. 642--646.
@@
 \newblock In Proceedings of the Thirtieth International
-  Conference on Very Large Data Bases-Volume 30. VLDB \hl{Endowment,} %MDPI: Please add the location and date of the conference. 
-  2004; pp.
-  912--923.
+  Conference on Very Large Data Bases, Volume 30, VLDB Endowment, Toronto, Canada, 29 August--3 September 2004; %MDPI: Please add the location and date of the conference. %RES Location and date added.
+  pp. 912--923.
 
@@
 \newblock An efficient indexing technique for full-text database systems.
-\newblock In Proceedings of the VLDB '92: Proceedings of the 18th International Conference on Very Large Data Bases, \hl{IEEE,} %MDPI: Please add the location and date of the conference. 
- 23--27 August 1992; pp. 352--352.
+\newblock In Proceedings of the 18th Conference on Very Large Data Bases, Morgan Kaufmann, Vancouver, Canada, 23--27 August, 1992; %MDPI: Please add the location and date of the conference.
+%RES Location and date added.
+ pp. 352--352.
 
 \bibitem[Zobel and Moffat(2006)]{zobel2006inverted}
-\hl{Zobel, J.; Moffat, A.}  %MDPI: Refs.
+Zobel, J.; Moffat, A.
+%MDPI: Refs.
 7 and 17 are duplicated.
@@
 Please ensure that there are no duplicated references.
+%RES Corrected.
 \newblock Inverted files for text search engines.
@@
 
-\bibitem[Manning \em{et~al.}(2008)Manning, Raghavan, Sch{\"u}tze,
-  et~al.]{manning2008introduction}
-Manning, C.D.; Raghavan, P.; Sch{\"u}tze, H.;  \hl{et~al.} %MDPI: Please include the first ten authors' names before using "et al." in the references.
+\bibitem[Manning \em{et~al.}(2008)Manning, Raghavan, Sch{\"u}tze]{manning2008introduction}
+Manning, C.D.; Raghavan, P.; Sch{\"u}tze, H.; %MDPI: Please include the first ten authors' names before using "et al." in the references. %RES All authors included.
 \newblock {\em Introduction to Information Retrieval}; Cambridge
@@
 \newblock Scaling up All Pairs Similarity Search.
-\newblock In WWW'07: 16th International World Wide Web Conference, \hl{Banff, AB, Canada, 8--12 May} %MDPI: We added the location and date of the conference. Please confirm.
- 2007;  Association for Computing Machinery: New York,
-  NY, USA,  2007; pp. 131–140.
+\newblock In WWW'07: 16th International World Wide Web Conference, ACM, Banff, Alberta, Canada, 8--12 May %MDPI: We added the location and date of the conference.
+Please confirm.
+%RES Updated and confirmed.
+ 2007; pp. 131–140.
 
@@
 H.~Cormen, T.; Leiserson, C.; L.~Rivest, R.; Stein, C.
-\newblock {\em Introduction to Algorithms, Second Edition}; \hl{The MIT Press: Cambridge, MA, USA,} %MDPI: We added the name of the publisher and location. Please confirm.
+\newblock {\em Introduction to Algorithms, Second Edition}; The MIT Press: Cambridge, MA, USA, %MDPI: We added the name of the publisher and location. Please confirm. %RES Confirmed.
  2001.
@@
 
-\bibitem[Zobel and Moffat(2006)]{zobel2006csur}
-\hl{Zobel, J.; Moffat, A.}
-\newblock Inverted files for text search engines.
-\newblock {\em ACM Comput. Surv. (CSUR)} {\bf 2006}, {\em 38},~6.
-
 \bibitem[Broder \em{et~al.}(2006)Broder, Eiron, Fontoura, Herscovici, Lempel,
@@
 \newblock In \emph{Proceedings of the International Conference on Extending Database
-  Technology}; Springer:  \hl{Berlin/Heidelberg, Germany,} %We added the location of publisher. Please confirm
+  Technology}; Springer: Berlin/Heidelberg, Germany, %We added the location of publisher. Please confirm %RES Confirmed.
   2006; pp. 313--330.
@@
 \bibitem[Deppisch(1986)]{deppisch1986sigir}
-\hl{Deppisch, U.} %MDPI: Refs.
+Deppisch, U.
+%MDPI: Refs.
 21 and 29 are duplicated.
@@
 Please ensure that there are no duplicated references.
+%RES The other reference deleted and updated.
 \newblock S-tree: A Dynamic Balanced Signature Index for Office Retrieval.
@@
 \newblock Signature Files and Signature File Construction.
-\newblock In {\em Encyclopedia of Database Technologies and Applications}; \hl{IGI Global: } %MDPI: Please add the location of the publisher (City and Country).
+\newblock In {\em Encyclopedia of Database Technologies and Applications}; IGI Global: Hershey, Pennsylvania, USA, %MDPI: Please add the location of the publisher (City and Country).
+%RES Location of pub.
+added.
  {2005}.
@@
   attributes of low cardinality.
-\newblock {\em  VLDB J.} {\bf 2003}, {\em 12},~244--261.
+\newblock {\em  VLDB Journal}, {\bf 2003}, {\em 12},~244--261.
 \newblock {{https://doi.org/10.1007/s00778-003-0106-0}}.
@@
 \newblock In \emph{Proceedings of the International Conference on Availability,
-  Reliability, and Security}; Springer:  \hl{Berlin/Heidelberg, Germany,} %We added the location of publisher. Please confirm
+  Reliability, and Security}; Springer: Berlin/Heidelberg, Germany, %We added the location of publisher. Please confirm %RES Confirmed.
   2013; pp. 134--148.
@@
 Sedgewick, R.; Wayne, K.
-\newblock {\em Algorithms}, 4th ed.; \hl{Addison-Wesley} %MDPI: Please add the location of the publisher (City and Country).
- Professional:  2011.
+\newblock {\em Algorithms}, 4th ed.; Addison-Wesley: %MDPI: Please add the location of the publisher (City and Country). %RES Added.
+ Professional: Boston, USA, 2011.
 
@@
 2019.
-Available online:  \url{https://github.com/nick-ak96/mstrie} (\hl{accessed} %MDPI: Please add the access date (format: Date Month Year), e.g., accessed on 1 January 2020.
- on).
-
-\bibitem[Deppisch(1986)]{Deppisch1986}
-\hl{Deppisch, U.} %MDPI: Refs.
-21 and 29 are duplicated.
-Please remove duplicated references and rearrange all the references to appear in numerical order.
-Please ensure that there are no duplicated references.
-\newblock S-tree: A Dynamic Balanced Signature Index for Office Retrieval.
-\newblock In \emph{Proceedings of the 9th Annual International ACM
-  SIGIR Conference on Research and Development in Information Retrieval}; ACM:
-  New York, NY, USA,  1986;  pp. 77--87.
-\newblock {{https://doi.org/10.1145/253168.253189}}.
+Available online:  \url{https://github.com/nick-ak96/mstrie} (accessed %MDPI: Please add the access date (format: Date Month Year), e.g., accessed on 1 January 2020.
+%RES Added
+ on 10 March 2023).
 
@@
 \newblock Set containment joins: The good, the bad and the ugly.
-\newblock In Proceedings of the \hl{Proc.
-of VLDB,} %MDPI: Please provide more information about the article type, such as book (please provide the name and location of the publisher); online resource (please provide the URL of the website and the date it was accessed (Date Month Year)); or journal article (please provide the name of the journal, the year and volume in which it was published, and the page number).
+\newblock In Proceedings of the 26th VLDB Conference, Morgan Kaufmann, Cairo, Egypt, 10--14 September, 2000.
+%MDPI: Please provide more information about the article type, such as book (please provide the name and location of the publisher); online resource (please provide the URL of the website and the date it was accessed (Date Month Year)); or journal article (please provide the name of the journal, the year and volume in which it was published, and the page number).
 Please refer to https://www.mdpi.com/authors/references for full reference formatting guides.
-  2000.
+%RES Details on reference Added.
 

### 2023-03-15 tex/algorithms-2122309-proof.tex
author contributions added.

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\authorcontributions{\hl{ .}}%For research articles with several authors, a short paragraph specifying their individual contributions must be provided.
+\authorcontributions{The individual contributions of the authors are as follows.
+First, Mikita Akulich was the original inventor of the proposed data structure.
+He also worked on software implementation, formal analysis, visualization, and writing---the original draft.
+Second, Iztok Savnik contributed to the work on the investigation, supervision, conceptualization, validation, and writing--- original draft, review, and editing.
+Third, Matjaž Krnc has contributed with the formal analysis, supervision, conceptualization, validation, and writing---review and editing.
+Finally, Riste Skrekovski worked on conceptualization, supervision, formal analysis, validation, project administration, and funding acquisition.}%For research articles with several authors, a short paragraph specifying their individual contributions must be provided.
 The following statements should be used ``Conceptualization, X.X.
@@
 
-\funding{\hl{The authors} %MDPI: We moved the last paragraph here, please confirm.
+\funding{The authors %MDPI: We moved the last paragraph here, please confirm. %RES We confirm.
  acknowledge the support in part of the Slovenian Research Agency, research program P1-0383.}
 
-\institutionalreview{\hl{ .}}%In this section, you should add the Institutional Review Board Statement and approval number, if relevant to your study.
+%\institutionalreview{\hl{ .}}%In this section, you should add the Institutional Review Board Statement and approval number, if relevant to your study.
 You might choose to exclude this statement if the study did not require ethical approval.
@@
 
-\informedconsent{\hl{ .}}%Any research article describing a study involving humans should contain this statement.
+%\informedconsent{\hl{ .}}%Any research article describing a study involving humans should contain this statement.
 Please add ``Informed consent was obtained from all subjects involved in the study.'' OR ``Patient consent was waived due to REASON (please provide a detailed justification).'' OR ``Not applicable'' for studies not involving humans.
@@
 
-\dataavailability{\hl{ .}}%We encourage all authors of articles published in MDPI journals to share their research data.
+%\dataavailability{\hl{ .}}%We encourage all authors of articles published in MDPI journals to share their research data.
 In this section, please provide details regarding where data supporting reported results can be found, including links to publicly archived datasets analyzed or generated during the study.
@@
 
-\conflictsofinterest{\hl{The} %MDPI: Newly added information. Please confirm.
+\conflictsofinterest{The %MDPI: Newly added information. Please confirm. %RES We confirm.
  authors declare no conflicts of interest.}

### 2023-03-15 tex/algorithms-2122309-proof.tex
a paragraph added to related work (omitted in the previous version).

@@
 The results show that, in general, the IR engine performs better than an RDBMS on containment queries.
-They identified the problems that lead to the poor performance of the containment operations in an RDBMS and showed that, with some modifications, an RDBMS could perform this class of queries more efficiently.
+The significant causes that differentiate the performance of the IR and RDBMS implementations are the join algorithms employed and the hardware cache utilization.
+The differences in join algorithms and the cache access method that makes IR queries faster were identified precisely.
+First, the multi-predicate merge join of the IR engine is different from the standard merge join, and the index nested-loop join algorithms.
+Second, the study of the utilization of cache in the multi-predicate merge join, standard merge join, and index-nested loop joins has been done, identifying more precisely the differences in the algorithms.
+They have presented that with some modifications, an RDBMS could perform this class of queries more efficiently.
 

### 2023-03-20 tex/algorithms-2122309-proof.tex
changes in mathematical section

@@
 %
-\hl{Let} %MDPI: Whether the indentation of the paragraph after the formula is necessary, please confirm. Please check all.
+Let % MDPI: Whether the indentation of the paragraph after the formula is necessary, please confirm. Please check all.
+%RES: We do not want to start new paragraph here, hence no indentation!
  $\xi_1, \xi_2, \ldots, \xi_{\sigma+1}$ be random variables such that $\xi_i$
@@
 \begin{figure}[H]
-
-\includegraphics[width=.64\textwidth, keepaspectratio]{exp-nodes-on-probab.pdf}
-\caption{\hl{Size} %MDPI: Please change the terms into scientific notations in the figure,  e.g., "$8 \times 10^{3}$", not "8E3".
+\centering
+\includegraphics[width=.64\textwidth, keepaspectratio]{exp-nodes-on-probab-1.pdf}
+\caption{Size % MDPI: Please change the terms into scientific notations in the figure,  e.g., "$8 \times 10^{3}$", not "8E3". %RES: done.
  of a tree ($\mathbb{E}(|\mathcal{M}|)$) compared to the number of stored sets ($\mathbb{E}(|M|)$), as the probability parameter $p$ increases.
@@
 \begin{figure}[H]
-
+\centering
 \includegraphics[width=.64\textwidth, keepaspectratio]{ratio-exp-nodes-and-msets-on-prob.pdf}
@@
 
-\subsection{Data%MDPI: Section headings should be numbered sequentially, e.g., Section 2.1, Section 2.2.1. Please confirm this revision. %RES It is Ok as it is.
+\subsection{Data %MDPI: Section headings should be numbered sequentially, e.g., Section 2.1, Section 2.2.1. Please confirm this revision. %RES It is Ok as it is.
  Generation}

### 2023-03-20 tex/algorithms-2122309-proof.tex
removed higlighting, ZIP

@@
 	\subcaptionbox{\centering \label{fig:e1m2}}{\includegraphics[width=.485\textwidth]{exp1-m2.pdf}}
-	\caption{\hl{Existence} %MDPI: Please use commas to separate thousands for numbers with five or more digits (not four digits) in the picture, e.g., "10000" should be "10,000".
+	\caption{{Existence} %MDPI: Please use commas to separate thousands for numbers with five or more digits (not four digits) in the picture, e.g., "10000" should be "10,000".
  functions of Experiment 1. (\textbf{a}) submsetExistence; %MDPI: We moved the subfigure explanations into the figure caption. Please confirm. The following are same. %RES We confirm.
@@
 }
-\caption{\hl{Exhaustive} %MDPI: Please use commas to separate thousands for numbers with five or more digits (not four digits) in the picture, e.g., "10000" should be "10,000".
-please check all figures.
- functions of Experiment 1. (\textbf{a}) \hl{getAllSubmsets}; (\textbf{b}) \hl{getAllSupermsets}.\label{fig:5}}
+\caption{{Exhaustive} %MDPI: Please use commas to separate thousands for numbers with five or more digits (not four digits) in the picture, e.g., "10000" should be "10,000". please check all figures.
+ functions of Experiment 1. (\textbf{a}) {getAllSubmsets}; (\textbf{b}) {getAllSupermsets}.\label{fig:5}}
 \end{figure}
@@
 \label{fig:e2m2}}{\includegraphics[width=.485\textwidth]{exp2-m2.pdf}}
-\caption{Existence functions in Experiment 2. (\textbf{a}) \hl{submsetExistence}; (\textbf{b}) \hl{supermsetExistence}.\label{fig:e2m1}}
+\caption{Existence functions in Experiment 2. (\textbf{a}) {submsetExistence}; (\textbf{b}) {supermsetExistence}.\label{fig:e2m1}}
 \end{figure}
@@
 \begin{figure}[H]
-\subcaptionbox{\centering \hl{Experiment} %MDPI Please move sub-caption to Caption.
+\subcaptionbox{\centering {Experiment} %MDPI Please move sub-caption to Caption.
 2, getAllSubmsets function.
 }{\includegraphics[width=.485\textwidth]{exp2-m3.pdf}}
-\subcaptionbox{\centering \hl{Experiment} 2, getAllSupermsets function.
+\subcaptionbox{\centering {Experiment} 2, getAllSupermsets function.
 \label{fig:e2m4}}{\includegraphics[width=.485\textwidth]{exp2-m4.pdf}}
@@
 5		& 1\\
-%\hline
 30	& 1 \\
-%\hline
 5		& 3 \\
-%\hline
 15	& 3 \\
-%\hline
 30	& 3 \\
-%\hline
 10	& 10 \\
